---
title: "Area Quantile EDA"
author: "Codex"
date: "2026-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook explores the area-augmented `stan_ready_data_with_area` object. The goals are:

1. visualize size-distribution trajectories alongside counts,
2. quantify within-line ploidy contrasts over time,
3. test whether early morphology shifts relate to later failure or survival,
4. relate morphology to glucose depletion,
5. port the size-to-cost scaling idea from `workflow/volume_aware_anal.Rmd` onto the row-aligned area dataset.

## Setup

```{r libraries}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(purrr)
  library(ggplot2)
  library(patchwork)
  library(scales)
  library(forcats)
})

`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}

resolve_area_stan_data_path <- function() {
  candidates <- c(
    file.path("data", "stan_ready_data_with_area.Rds"),
    file.path("dev_data", "stan_ready_data_with_area.Rds"),
    file.path("data", "stan_ready_data.Rds"),
    file.path("ecology", "stan_ready_data.Rds")
  )

  for (path in candidates) {
    if (file.exists(path)) {
      return(path)
    }
  }

  stop(
    "Could not find an area-aware stan data object. Tried: ",
    paste(candidates, collapse = ", ")
  )
}

safe_median <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) return(NA_real_)
  median(x)
}

safe_quantile <- function(x, prob = 0.9) {
  x <- x[is.finite(x)]
  if (!length(x)) return(NA_real_)
  as.numeric(stats::quantile(x, probs = prob, names = FALSE, type = 8))
}

safe_trapz <- function(x, y) {
  ok <- is.finite(x) & is.finite(y)
  x <- x[ok]
  y <- y[ok]
  if (length(x) < 2L) return(NA_real_)
  ord <- order(x)
  x <- x[ord]
  y <- y[ord]
  sum(diff(x) * (head(y, -1L) + tail(y, -1L)) / 2)
}

smooth_derivative <- function(hours, value, spar = 0.55) {
  ok <- is.finite(hours) & is.finite(value)
  hours <- hours[ok]
  value <- value[ok]

  if (length(hours) < 4L || length(unique(hours)) < 4L) {
    return(tibble(
      hours = hours,
      value_smooth = value,
      deriv = c(NA_real_, diff(value) / diff(hours))
    ))
  }

  ord <- order(hours)
  hours <- hours[ord]
  value <- value[ord]

  fit <- try(stats::smooth.spline(hours, value, spar = spar), silent = TRUE)
  if (inherits(fit, "try-error")) {
    return(tibble(
      hours = hours,
      value_smooth = value,
      deriv = c(NA_real_, diff(value) / diff(hours))
    ))
  }

  tibble(
    hours = hours,
    value_smooth = as.numeric(stats::predict(fit, x = hours, deriv = 0)$y),
    deriv = as.numeric(stats::predict(fit, x = hours, deriv = 1)$y)
  )
}

lagged_correlations <- function(hours, x, y, lag_hours = seq(-24, 24, by = 8)) {
  out <- lapply(lag_hours, function(lag_h) {
    x_shift <- approx(
      x = hours + lag_h,
      y = x,
      xout = hours,
      rule = 1
    )$y

    tibble(
      lag_hours = lag_h,
      corr = suppressWarnings(cor(x_shift, y, use = "complete.obs")),
      n_overlap = sum(is.finite(x_shift) & is.finite(y))
    )
  })

  bind_rows(out)
}

fit_cost_model <- function(y, x_build, x_maint) {
  ok <- is.finite(y) & is.finite(x_build) & is.finite(x_maint)
  y <- y[ok]
  x_build <- x_build[ok]
  x_maint <- x_maint[ok]

  if (length(y) < 4L) {
    return(c(a = NA_real_, m = NA_real_, rmse = NA_real_, n = length(y)))
  }

  obj <- function(par) {
    pred <- par[1] * x_build + par[2] * x_maint
    sum((y - pred)^2)
  }

  fit0 <- try(lm(y ~ 0 + x_build + x_maint), silent = TRUE)
  par0 <- if (inherits(fit0, "try-error")) {
    c(0.1, 0.001)
  } else {
    pmax(as.numeric(coef(fit0)), 0)
  }
  if (length(par0) != 2L || any(!is.finite(par0))) {
    par0 <- c(0.1, 0.001)
  }

  opt <- optim(
    par = par0,
    fn = obj,
    method = "L-BFGS-B",
    lower = c(0, 0)
  )

  pred <- opt$par[1] * x_build + opt$par[2] * x_maint
  rmse <- sqrt(mean((y - pred)^2))

  c(a = unname(opt$par[1]), m = unname(opt$par[2]), rmse = rmse, n = length(y))
}

stan_path <- resolve_area_stan_data_path()
stan <- readRDS(stan_path)

if (is.null(stan$area_alive_quantiles)) {
  stop("Loaded stan data does not contain area_alive_quantiles.")
}

line_names <- names(stan$line_map)[match(stan$line_id, unname(stan$line_map))]
well_meta <- tibble(
  well_idx = seq_len(stan$N_wells),
  cellLine = line_names,
  line_id = as.integer(stan$line_id),
  ploidy_metric = as.numeric(stan$ploidy_metric),
  ploidy_abs = as.numeric(stan$ploidy_abs),
  G0 = as.numeric(stan$G0_per_well),
  exp_id = as.integer(stan$exp_id),
  has_starvation = as.integer(stan$has_starvation)
) %>%
  mutate(
    ploidy_label = if_else(ploidy_metric == ave(ploidy_metric, cellLine, FUN = min), "baseline", "elevated"),
    G0_label = factor(as.character(G0), levels = as.character(sort(unique(G0))))
  )

area_q <- as.data.frame(stan$area_alive_quantiles, check.names = FALSE)

count_obs <- tibble(
  well_idx = as.integer(stan$well_idx_count),
  rep_id = as.character(stan$rep_id_count),
  hours = as.numeric(stan$t_grid[stan$grid_idx_count]),
  live_cells = as.numeric(stan$N_obs),
  dead_cells = as.numeric(stan$D_obs),
  n_area_alive = as.numeric(stan$n_area_alive %||% rep(NA_real_, length(stan$N_obs))),
  q0 = area_q[["q0"]],
  q01 = area_q[["q0.01"]],
  q05 = area_q[["q0.05"]],
  q10 = area_q[["q0.1"]],
  q25 = area_q[["q0.25"]],
  q50 = area_q[["q0.5"]],
  q75 = area_q[["q0.75"]],
  q90 = area_q[["q0.9"]],
  q95 = area_q[["q0.95"]],
  q99 = area_q[["q0.99"]],
  q100 = area_q[["q1"]]
) %>%
  mutate(
    total_cells = live_cells + dead_cells,
    dead_fraction = if_else(total_cells > 0, dead_cells / total_cells, NA_real_),
    tail_width = q99 - q50,
    tail_ratio = q99 / pmax(q50, 1e-8)
  ) %>%
  left_join(well_meta, by = "well_idx")

glucose_obs <- tibble(
  well_idx = as.integer(stan$well_idx_gluc),
  hours = as.numeric(stan$t_grid[stan$grid_idx_gluc]),
  lum = as.numeric(stan$lum_obs),
  dilution = as.numeric(stan$dilution),
  is_censored = as.integer(stan$is_censored)
) %>%
  left_join(well_meta %>% select(well_idx, exp_id), by = "well_idx") %>%
  mutate(
    calib_a = as.numeric(stan$calib_a_fixed[exp_id]),
    calib_b = as.numeric(stan$calib_b_fixed[exp_id]),
    glucose_hat = pmax(0, (lum - calib_b) / (calib_a * pmax(dilution, 1e-12)))
) %>%
  group_by(well_idx, hours) %>%
  summarise(
    glucose_hat = {
      uncens <- glucose_hat[is_censored == 0L]
      if (length(uncens)) median(uncens, na.rm = TRUE) else NA_real_
    },
    any_glucose_censored = any(is_censored == 1L, na.rm = TRUE),
    n_glucose_uncensored = sum(is_censored == 0L, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(well_meta, by = "well_idx")

condition_counts <- count_obs %>%
  group_by(cellLine, ploidy_metric, ploidy_label, G0, G0_label, hours) %>%
  summarise(
    n_reps = n(),
    live_cells = median(live_cells, na.rm = TRUE),
    dead_cells = median(dead_cells, na.rm = TRUE),
    total_cells = median(total_cells, na.rm = TRUE),
    dead_fraction = median(dead_fraction, na.rm = TRUE),
    q50 = median(q50, na.rm = TRUE),
    q90 = median(q90, na.rm = TRUE),
    q99 = median(q99, na.rm = TRUE),
    tail_width = median(tail_width, na.rm = TRUE),
    tail_ratio = median(tail_ratio, na.rm = TRUE),
    .groups = "drop"
  )

condition_glucose <- glucose_obs %>%
  group_by(cellLine, ploidy_metric, ploidy_label, G0, G0_label, hours) %>%
  summarise(
    glucose_hat = median(glucose_hat, na.rm = TRUE),
    .groups = "drop"
  )

count_glucose_joint <- condition_counts %>%
  left_join(
    condition_glucose,
    by = c("cellLine", "ploidy_metric", "ploidy_label", "G0", "G0_label", "hours")
  )

paired_ploidy <- condition_counts %>%
  group_by(cellLine, G0, G0_label, hours) %>%
  filter(n_distinct(ploidy_metric) == 2L) %>%
  arrange(ploidy_metric, .by_group = TRUE) %>%
  summarise(
    delta_live = live_cells[2] - live_cells[1],
    delta_dead_fraction = dead_fraction[2] - dead_fraction[1],
    delta_q50 = q50[2] - q50[1],
    delta_q90 = q90[2] - q90[1],
    delta_q99 = q99[2] - q99[1],
    delta_tail_ratio = tail_ratio[2] - tail_ratio[1],
    .groups = "drop"
  )

phase_obs <- count_obs %>%
  group_by(well_idx, rep_id, cellLine, ploidy_metric, ploidy_label, G0, G0_label) %>%
  group_modify(function(df, keys) {
    live_fit <- smooth_derivative(df$hours, df$live_cells)
    death_fit <- smooth_derivative(df$hours, df$dead_cells)
    total_ref <- df %>%
      select(hours, total_cells) %>%
      distinct()

    full_join(
      live_fit %>% rename(live_cells_smooth = value_smooth, live_deriv = deriv),
      death_fit %>% rename(dead_cells_smooth = value_smooth, death_deriv = deriv),
      by = "hours"
    ) %>%
      left_join(total_ref, by = "hours") %>%
      transmute(
        hours = hours,
        live_cells_smooth = live_cells_smooth,
        dead_cells_smooth = dead_cells_smooth,
        live_rate_per_cell = pmax(live_deriv, 0) / pmax(total_cells, 1e-8),
        death_rate_per_cell = pmax(death_deriv, 0) / pmax(total_cells, 1e-8)
      )
  }) %>%
  ungroup()

well_phase_ref <- phase_obs %>%
  group_by(well_idx, rep_id, cellLine, ploidy_metric, ploidy_label, G0, G0_label) %>%
  summarise(
    peak_live_rate_per_cell = safe_quantile(live_rate_per_cell[live_rate_per_cell > 0], prob = 0.75),
    .groups = "drop"
  )

rate_obs <- count_obs %>%
  group_by(well_idx, rep_id, cellLine, ploidy_metric, ploidy_label, G0, G0_label) %>%
  group_modify(function(df, keys) {
    death_fit <- smooth_derivative(df$hours, df$dead_cells)
    area_fit <- smooth_derivative(df$hours, df$q50)
    total_ref <- df %>%
      select(hours, total_cells) %>%
      distinct()

    full_join(
      death_fit %>% rename(death_cells_smooth = value_smooth, death_deriv = deriv),
      area_fit %>% rename(area_q50_smooth = value_smooth, area_rate = deriv),
      by = "hours"
    ) %>%
      left_join(total_ref, by = "hours") %>%
      transmute(
        hours = hours,
        death_cells_smooth = death_cells_smooth,
        death_rate_per_cell = pmax(death_deriv, 0) / pmax(total_cells, 1e-8),
        area_q50_smooth = area_q50_smooth,
        area_rate = area_rate
      )
  }) %>%
  ungroup()

lag_profiles <- rate_obs %>%
  group_by(well_idx, rep_id, cellLine, ploidy_metric, ploidy_label, G0, G0_label) %>%
  group_modify(function(df, keys) {
    lagged_correlations(df$hours, df$area_rate, df$death_rate_per_cell)
  }) %>%
  ungroup()

lag_summary <- lag_profiles %>%
  filter(n_overlap >= 4) %>%
  group_by(cellLine, ploidy_label, G0, G0_label, lag_hours) %>%
  summarise(
    median_corr = median(corr, na.rm = TRUE),
    mean_corr = mean(corr, na.rm = TRUE),
    n_profiles = sum(is.finite(corr)),
    .groups = "drop"
  )

best_lag_tbl <- lag_profiles %>%
  filter(n_overlap >= 4, is.finite(corr)) %>%
  group_by(well_idx, rep_id, cellLine, ploidy_metric, ploidy_label, G0, G0_label) %>%
  summarise(
    best_abs_corr = corr[which.max(abs(corr))],
    best_lag_hours = lag_hours[which.max(abs(corr))],
    .groups = "drop"
  )

glucose_last <- glucose_obs %>%
  group_by(well_idx) %>%
  summarise(
    t_first_glucose = if (any(is.finite(glucose_hat))) min(hours[is.finite(glucose_hat)], na.rm = TRUE) else NA_real_,
    t_last_glucose = if (any(is.finite(glucose_hat))) max(hours[is.finite(glucose_hat)], na.rm = TRUE) else NA_real_,
    glucose_initial = if (is.finite(t_first_glucose)) glucose_hat[match(t_first_glucose, hours)] else NA_real_,
    glucose_last = if (is.finite(t_last_glucose)) glucose_hat[match(t_last_glucose, hours)] else NA_real_,
    glucose_consumed = pmax(0, glucose_initial - glucose_last),
    n_uncensored_timepoints = sum(is.finite(glucose_hat)),
    .groups = "drop"
  )

replicate_cost_base <- count_obs %>%
  left_join(glucose_last, by = "well_idx") %>%
  filter(is.finite(t_last_glucose), hours <= t_last_glucose) %>%
  group_by(well_idx, rep_id, cellLine, ploidy_metric, ploidy_abs, ploidy_label, G0, G0_label, glucose_consumed) %>%
  summarise(
    B0 = total_cells[which.min(hours)],
    B_end = total_cells[which.max(hours)],
    dB = pmax(0, B_end - B0),
    AUC_live = safe_trapz(hours, live_cells),
    area_24h = safe_median(q50[hours >= 20 & hours <= 28]),
    .groups = "drop"
  )

glucose_init_check <- glucose_last %>%
  left_join(well_meta, by = "well_idx")

rate_summary <- rate_obs %>%
  group_by(cellLine, ploidy_label, G0, G0_label, hours) %>%
  summarise(
    death_rate_per_cell = median(death_rate_per_cell, na.rm = TRUE),
    area_rate = median(area_rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(death_rate_per_cell, area_rate),
    names_to = "signal",
    values_to = "value"
  ) %>%
  mutate(
    signal = recode(
      signal,
      death_rate_per_cell = "1/N dD/dt",
      area_rate = "dA/dt"
    )
  )

cellline_ploidy_capacity <- count_obs %>%
  group_by(cellLine, ploidy_metric, ploidy_label) %>%
  summarise(max_total_cells = max(total_cells, na.rm = TRUE), .groups = "drop")

volume_proxy_rows <- count_obs %>%
  left_join(
    phase_obs %>%
      select(well_idx, rep_id, hours, live_rate_per_cell, death_rate_per_cell),
    by = c("well_idx", "rep_id", "hours")
  ) %>%
  left_join(
    well_phase_ref %>%
      select(well_idx, rep_id, peak_live_rate_per_cell),
    by = c("well_idx", "rep_id")
  ) %>%
  left_join(
    cellline_ploidy_capacity,
    by = c("cellLine", "ploidy_metric", "ploidy_label")
  ) %>%
  mutate(
    crowding_frac = total_cells / pmax(max_total_cells, 1e-8),
    core_area = sqrt(pmax(q25, 1e-8) * pmax(q50, 1e-8)),
    likely_attached = hours >= 16,
    growth_phase = live_rate_per_cell > 0.25 * pmax(peak_live_rate_per_cell, 1e-8),
    no_death_phase = death_rate_per_cell <= pmax(0.002, 0.2 * pmax(peak_live_rate_per_cell, 1e-8)),
    usable_for_volume = is.finite(core_area) &
      n_area_alive >= 30 &
      likely_attached &
      crowding_frac <= 0.7 &
      growth_phase &
      no_death_phase
  )

crowding_slopes <- volume_proxy_rows %>%
  filter(usable_for_volume, is.finite(crowding_frac)) %>%
  group_by(cellLine) %>%
  group_modify(function(df, keys) {
    if (nrow(df) < 4L || dplyr::n_distinct(df$crowding_frac) < 2L) {
      return(tibble(beta_crowding = 0, n_fit = nrow(df)))
    }

    fit <- try(stats::lm(log(core_area) ~ crowding_frac, data = df), silent = TRUE)
    if (inherits(fit, "try-error")) {
      tibble(beta_crowding = 0, n_fit = nrow(df))
    } else {
      tibble(
        beta_crowding = unname(stats::coef(fit)[["crowding_frac"]] %||% 0),
        n_fit = nrow(df)
      )
    }
  }) %>%
  ungroup()

volume_proxy_rows <- volume_proxy_rows %>%
  left_join(
    crowding_slopes,
    by = "cellLine"
  ) %>%
  mutate(
    beta_crowding = if_else(is.finite(beta_crowding), beta_crowding, 0),
    log_area_adj = if_else(
      usable_for_volume,
      log(core_area) - beta_crowding * (crowding_frac - 0.2),
      NA_real_
    ),
    adj_area_ref = exp(log_area_adj)
  )

well_volume_proxy <- volume_proxy_rows %>%
  group_by(well_idx, rep_id, cellLine, ploidy_metric, ploidy_abs, ploidy_label, G0, G0_label) %>%
  summarise(
    n_time = sum(is.finite(adj_area_ref)),
    crowding_span = max(crowding_frac, na.rm = TRUE) - min(crowding_frac, na.rm = TRUE),
    n_usable = sum(usable_for_volume, na.rm = TRUE),
    attached_area_hat = safe_quantile(adj_area_ref[usable_for_volume], prob = 0.9),
    latent_volume_index = attached_area_hat^1.5,
    naive_q50_24h = safe_median(q50[hours >= 20 & hours <= 28]),
    naive_volume_24h = naive_q50_24h^1.5,
    .groups = "drop"
  ) %>%
  mutate(
    latent_vs_naive_ratio = latent_volume_index / naive_volume_24h
  )

condition_volume_proxy <- well_volume_proxy %>%
  group_by(cellLine, ploidy_metric, ploidy_abs, ploidy_label, G0, G0_label) %>%
  summarise(
    latent_volume_index = median(latent_volume_index, na.rm = TRUE),
    attached_area_hat = median(attached_area_hat, na.rm = TRUE),
    naive_volume_24h = median(naive_volume_24h, na.rm = TRUE),
    latent_vs_naive_ratio = median(latent_vs_naive_ratio, na.rm = TRUE),
    n_reps = sum(is.finite(attached_area_hat)),
    median_usable_timepoints = median(n_usable, na.rm = TRUE),
    .groups = "drop"
  )

replicate_cost_df <- replicate_cost_base %>%
  left_join(
    well_volume_proxy %>%
      select(well_idx, rep_id, attached_area_hat, latent_volume_index, naive_volume_24h, n_usable),
    by = c("well_idx", "rep_id")
  ) %>%
  filter(
    is.finite(glucose_consumed),
    is.finite(dB),
    is.finite(AUC_live),
    is.finite(attached_area_hat),
    is.finite(latent_volume_index)
  ) %>%
  mutate(
    vol_flat = attached_area_hat,
    vol_sphere = latent_volume_index
  )

cost_fits <- replicate_cost_df %>%
  filter(G0 != 25) %>%
  group_by(cellLine, ploidy_metric, ploidy_abs, ploidy_label) %>%
  summarise(
    fit = list(fit_cost_model(glucose_consumed, dB, AUC_live)),
    .groups = "drop"
  ) %>%
  mutate(
    a = map_dbl(fit, ~ .x[["a"]]),
    m = map_dbl(fit, ~ .x[["m"]]),
    rmse = map_dbl(fit, ~ .x[["rmse"]]),
    n = map_dbl(fit, ~ .x[["n"]])
  ) %>%
  select(-fit) %>%
  left_join(
    replicate_cost_df %>%
      group_by(cellLine, ploidy_metric, ploidy_abs, ploidy_label) %>%
      summarise(
        attached_area_hat = median(attached_area_hat, na.rm = TRUE),
        area_24h = median(area_24h, na.rm = TRUE),
        n_usable = median(n_usable, na.rm = TRUE),
        vol_flat = median(vol_flat, na.rm = TRUE),
        vol_sphere = median(vol_sphere, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("cellLine", "ploidy_metric", "ploidy_abs", "ploidy_label")
  ) %>%
  group_by(cellLine) %>%
  mutate(
    a_base = a[which.min(ploidy_metric)],
    vol_flat_base = vol_flat[which.min(ploidy_metric)],
    vol_sphere_base = vol_sphere[which.min(ploidy_metric)],
    ratio_cost = a / a_base,
    ratio_vol_flat = vol_flat / vol_flat_base,
    ratio_vol_sphere = vol_sphere / vol_sphere_base,
    cost_per_flat_vol = a / vol_flat,
    cost_per_sphere_vol = a / vol_sphere
  ) %>%
  ungroup()

cost_fit_diagnostics <- replicate_cost_df %>%
  left_join(
    cost_fits %>%
      select(cellLine, ploidy_metric, a, m),
    by = c("cellLine", "ploidy_metric")
  ) %>%
  mutate(
    glucose_pred = a * dB + m * AUC_live,
    resid = glucose_consumed - glucose_pred
  )

count_glucose_aligned <- count_obs %>%
  group_by(well_idx, rep_id) %>%
  group_modify(function(df, keys) {
    gluc_df <- glucose_obs %>%
      filter(well_idx == keys$well_idx[[1]]) %>%
      arrange(hours)

    if (!nrow(gluc_df)) {
      return(df %>%
        mutate(
          glucose_hat = NA_real_,
          glucose_lag24 = NA_real_,
          glucose_lag48 = NA_real_
        ))
    }

    approx_glucose <- function(target_hours) {
      ok <- is.finite(gluc_df$hours) & is.finite(gluc_df$glucose_hat)
      if (sum(ok) < 2L) {
        return(rep(NA_real_, length(target_hours)))
      }
      approx(
        x = gluc_df$hours[ok],
        y = gluc_df$glucose_hat[ok],
        xout = target_hours,
        rule = 1
      )$y
    }

    df %>%
      arrange(hours) %>%
      mutate(
        glucose_hat = approx_glucose(hours),
        glucose_lag24 = approx_glucose(hours - 24),
        glucose_lag48 = approx_glucose(hours - 48)
      )
  }) %>%
  ungroup() %>%
  left_join(
    cellline_ploidy_capacity,
    by = c("cellLine", "ploidy_metric", "ploidy_label")
  ) %>%
  filter(total_cells <= 0.7 * max_total_cells)

morph_glucose_summary <- count_glucose_aligned %>%
  mutate(
    glucose_hat_plot = pmax(glucose_hat, 0.02),
    glucose_lag24_plot = pmax(glucose_lag24, 0.02),
    glucose_lag48_plot = pmax(glucose_lag48, 0.02)
  ) %>%
  group_by(cellLine, ploidy_metric, ploidy_label, G0, G0_label, hours) %>%
  summarise(
    q50 = median(q50, na.rm = TRUE),
    glucose_hat_plot = median(glucose_hat_plot, na.rm = TRUE),
    glucose_lag24_plot = median(glucose_lag24_plot, na.rm = TRUE),
    glucose_lag48_plot = median(glucose_lag48_plot, na.rm = TRUE),
    .groups = "drop"
  )

cost_fit_timecourse <- count_glucose_aligned %>%
  left_join(
    cost_fits %>%
      select(cellLine, ploidy_metric, a, m),
    by = c("cellLine", "ploidy_metric")
  ) %>%
  filter(is.finite(glucose_hat), is.finite(a), is.finite(m)) %>%
  group_by(well_idx, rep_id, cellLine, ploidy_metric, ploidy_label, G0, G0_label, a, m) %>%
  arrange(hours, .by_group = TRUE) %>%
  mutate(
    glucose_initial_timecourse = glucose_hat[which.min(hours)],
    glucose_consumed_to_t = pmax(0, glucose_initial_timecourse - glucose_hat),
    AUC_live_to_t = cumsum(c(
      0,
      diff(hours) * (head(live_cells, -1) + tail(live_cells, -1)) / 2
    )),
    B0 = total_cells[which.min(hours)],
    dB_pred = pmax((glucose_consumed_to_t - m * AUC_live_to_t) / pmax(a, 1e-8), 0),
    total_cells_pred = B0 + dB_pred
  ) %>%
  ungroup()

cost_fit_timecourse_summary <- cost_fit_timecourse %>%
  group_by(cellLine, ploidy_metric, ploidy_label, G0, G0_label, hours) %>%
  summarise(
    total_cells_obs = median(total_cells, na.rm = TRUE),
    total_cells_pred = median(total_cells_pred, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(total_cells_obs, total_cells_pred),
    names_to = "series",
    values_to = "value"
  ) %>%
  mutate(
    series = recode(
      series,
      total_cells_obs = "Observed total cells",
      total_cells_pred = "Predicted total cells"
    )
  )
```

Loaded area-aware Stan data from `r stan_path`.

The object contains `r stan$N_wells` conditions, `r length(stan$well_idx_count)` count observations, and `r ncol(stan$area_alive_quantiles)` aligned area quantiles per count row.

## Coverage Check

```{r coverage-check}
coverage_tbl <- tibble(
  metric = c("count rows", "rows with any area summary", "rows missing all area quantiles"),
  value = c(
    nrow(count_obs),
    sum(rowSums(is.na(stan$area_alive_quantiles)) < ncol(stan$area_alive_quantiles)),
    sum(rowSums(is.na(stan$area_alive_quantiles)) == ncol(stan$area_alive_quantiles))
  )
)

knitr::kable(coverage_tbl)
```

The area join is effectively complete: only one of the 5,258 count observations lacks an aligned area summary, so downstream patterns should not be driven by missingness.

## Size Trajectories

These plots treat area as an additional phenotype channel parallel to live/dead counts. `q50` captures the middle of the alive-cell area distribution, while `q99` and `tail_ratio = q99 / q50` track the upper tail.

```{r size-trajectories, fig.height=18}
plot_traj_counts <- condition_counts %>%
  pivot_longer(
    cols = c(live_cells, dead_cells, total_cells),
    names_to = "state",
    values_to = "value"
  ) %>%
  mutate(
    state = recode(
      state,
      live_cells = "Live",
      dead_cells = "Dead",
      total_cells = "Total"
    )
  ) %>%
  ggplot(
    aes(hours, value, color = ploidy_label, group = interaction(ploidy_metric, G0))
  ) +
  geom_line(linewidth = 0.7) +
  facet_grid(state + G0_label ~ cellLine, scales = "free_y") +
  scale_color_manual(values = c("baseline" = "#0072B2", "elevated" = "#D55E00")) +
  theme_bw() +
  labs(
    title = "Count Trajectories",
    x = "Hours",
    y = "Cells",
    color = "Ploidy"
  )

plot_traj_q50 <- ggplot(
  condition_counts,
  aes(hours, q50, color = ploidy_label, group = interaction(ploidy_metric, G0))
) +
  geom_line(linewidth = 0.7) +
  geom_hline(
    data = condition_volume_proxy,
    aes(yintercept = attached_area_hat, color = ploidy_label),
    linetype = 2,
    linewidth = 0.7,
    alpha = 0.8
  ) +
  facet_grid(G0_label ~ cellLine, scales = "free_y") +
  scale_color_manual(values = c("baseline" = "#0072B2", "elevated" = "#D55E00")) +
  theme_bw() +
  labs(
    title = "Median Alive-Cell Area Trajectories",
    subtitle = "Dashed lines show the latent attached-area estimate used to derive the volume proxy",
    x = "Hours",
    y = "Alive-cell area q50",
    color = "Ploidy"
  )

plot_traj_tail <- ggplot(
  condition_counts,
  aes(hours, tail_ratio, color = ploidy_label, group = interaction(ploidy_metric, G0))
) +
  geom_line(linewidth = 0.7) +
  facet_grid(G0_label ~ cellLine, scales = "free_y") +
  scale_color_manual(values = c("baseline" = "#0072B2", "elevated" = "#D55E00")) +
  theme_bw() +
  labs(
    title = "Upper-Tail Inflation Over Time",
    subtitle = "tail_ratio = q99 / q50",
    x = "Hours",
    y = "q99 / q50",
    color = "Ploidy"
  )

(plot_traj_counts / plot_traj_q50 / plot_traj_tail) + plot_layout(heights = c(1.6, 1.2, 1.2))
```

```{r size-trajectory-summary}
size_summary_tbl <- condition_counts %>%
  group_by(cellLine, ploidy_label, G0) %>%
  summarise(
    peak_live = max(live_cells, na.rm = TRUE),
    peak_q50 = max(q50, na.rm = TRUE),
    peak_tail_ratio = max(tail_ratio, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(cellLine, G0, ploidy_label)

knitr::kable(head(size_summary_tbl, 18), digits = 2)
```

The count and area trajectories already suggest that morphology is not just a proxy for cell number. In several conditions, `q50` and `tail_ratio` keep moving after the main expansion phase has slowed, which supports looking at morphology as a state variable rather than just a confluence readout.

## Latent Volume Proxy

If the goal is a single number for cell "volume", the measured 2D area cannot be used literally. A spread, well-attached cell can have a large projected area without having much more biomass than a rounder cell, and crowding pushes projected area downward even when biomass has not shrunk proportionally.

The compromise here is to estimate a **typical attached pre-confluent footprint** and then convert that footprint to a volume-like index:

- use `core_area = sqrt(q25 * q50)` rather than `q99` or even raw `q50`, so the estimate targets the center of the alive-cell distribution and is less sensitive to upper-tail inflation from cycling/mitotic cells,
- identify candidate footprint rows using phase rather than cumulative death burden: `hours >= 16`, positive live-cell growth, negligible ongoing death accumulation, and `crowding_frac <= 0.7`,
- estimate a cell-line-level crowding slope from those candidate rows,
- adjust `log(core_area)` back to a light-crowding reference (`crowding_frac = 0.2`),
- within each replicate, take the 90th percentile of that adjusted area over those candidate rows as the best available approximation to the fully attached footprint,
- convert footprint to a volume-like quantity with `latent_volume_index = attached_area_hat^(3/2)`.

This is still an index, not an absolute micron-cubed volume. But it now handles the main confounders in a more phase-aware way: confluence is adjusted explicitly, attachment is handled by excluding the earliest post-plating rows, death is handled by filtering on ongoing death rate rather than total dead-cell burden, and cell-cycle inflation is downweighted by using the lower-middle part of the area distribution.

```{r latent-volume-proxy, fig.height=12}
p_volume_proxy <- ggplot(
  condition_volume_proxy,
  aes(x = G0, y = latent_volume_index, color = ploidy_label, group = ploidy_metric)
) +
  geom_point(size = 2.4, alpha = 0.9, position = position_dodge(width = 0.3)) +
  geom_line(alpha = 0.7, position = position_dodge(width = 0.3)) +
  facet_wrap(~cellLine, scales = "free_y") +
  scale_color_manual(values = c("baseline" = "#0072B2", "elevated" = "#D55E00")) +
  theme_bw() +
  labs(
    title = "Latent Volume Proxy by Condition",
    subtitle = "Replicate-median attached, pre-confluent footprint converted as area^(3/2)",
    x = "Initial glucose (G0)",
    y = "Latent volume index",
    color = "Ploidy"
  )

p_naive_compare <- ggplot(
  condition_volume_proxy,
  aes(naive_volume_24h, latent_volume_index, color = ploidy_label)
) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey40") +
  geom_point(size = 2.3, alpha = 0.85) +
  facet_wrap(~cellLine, scales = "free") +
  scale_color_manual(values = c("baseline" = "#0072B2", "elevated" = "#D55E00")) +
  theme_bw() +
  labs(
    title = "Latent Proxy Versus Naive 24h Volume",
    subtitle = "Points above the line are cases where detachment/crowding likely depressed the naive area snapshot",
    x = "Naive 24h volume proxy = q50_24h^(3/2)",
    y = "Latent volume index",
    color = "Ploidy"
  )

p_volume_proxy / p_naive_compare
```

```{r latent-volume-table}
latent_volume_tbl <- condition_volume_proxy %>%
  group_by(cellLine, ploidy_metric, ploidy_abs, ploidy_label) %>%
  summarise(
    latent_volume_index = median(latent_volume_index, na.rm = TRUE),
    attached_area_hat = median(attached_area_hat, na.rm = TRUE),
    latent_vs_naive_ratio = median(latent_vs_naive_ratio, na.rm = TRUE),
    median_usable_timepoints = median(median_usable_timepoints, na.rm = TRUE),
    n_conditions = sum(is.finite(latent_volume_index)),
    .groups = "drop"
  ) %>%
  group_by(cellLine) %>%
  mutate(
    baseline_volume = latent_volume_index[which.min(ploidy_metric)],
    volume_ratio_vs_baseline = latent_volume_index / baseline_volume
  ) %>%
  ungroup() %>%
  arrange(cellLine, ploidy_metric)

knitr::kable(latent_volume_tbl, digits = 3)
```

The table above is the cleanest answer to the original question. The most defensible single-number estimate in this dataset is not the observed footprint at one arbitrary timepoint, but the crowding-adjusted upper-envelope footprint of the central alive-cell distribution, restricted to rows that look like true growth phase rather than plating artifact or late collapse. I would treat `volume_ratio_vs_baseline` as the most interpretable downstream quantity because it is less sensitive to the arbitrary units of the segmentation area, while `median_usable_timepoints` is the quick reliability check.

## Ploidy Contrasts Through Time

These contrasts show the elevated-ploidy condition minus the baseline-ploidy condition within each cell line, glucose condition, and time point.

```{r ploidy-contrast-trajectories}
contrast_long <- paired_ploidy %>%
  pivot_longer(
    cols = starts_with("delta_"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      delta_live = "Live Cells",
      delta_dead_fraction = "Dead Fraction",
      delta_q50 = "Area q50",
      delta_q90 = "Area q90",
      delta_q99 = "Area q99",
      delta_tail_ratio = "Tail Ratio"
    )
  )

ggplot(contrast_long, aes(hours, value, color = metric)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey50") +
  geom_line(linewidth = 0.7) +
  facet_grid(G0_label ~ cellLine, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Within-Line Ploidy Contrasts Over Time",
    subtitle = "elevated ploidy minus baseline ploidy",
    x = "Hours",
    y = "Contrast"
  ) +
  theme(legend.position = "bottom")
```

```{r ploidy-contrast-summary}
contrast_summary_tbl <- paired_ploidy %>%
  group_by(cellLine, G0) %>%
  summarise(
    max_abs_delta_live = max(abs(delta_live), na.rm = TRUE),
    max_abs_delta_q50 = max(abs(delta_q50), na.rm = TRUE),
    max_abs_delta_tail_ratio = max(abs(delta_tail_ratio), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(max_abs_delta_q50))

knitr::kable(head(contrast_summary_tbl, 15), digits = 2)
```

The contrast view is useful because it makes it obvious where ploidy effects are primarily morphological versus primarily count-based. The larger `|delta_q50|` and `|delta_tail_ratio|` cases are the ones most worth following up in a mechanistic branch.

## Area-Rate and Death-Rate Coupling

This replaces the earlier early-versus-late scatter with a rate-based view. For each replicate, a smoothing spline is fit to dead-cell counts and to area `q50`. We then examine:

- `death_rate_per_cell = max(dD/dt, 0) / N`
- `area_rate = d(q50)/dt`

and compute lagged correlations between the two signals.

```{r lag-coupling-plots, fig.height=12}
p_lag_profile <- ggplot(
  lag_summary,
  aes(lag_hours, median_corr, color = ploidy_label)
) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey50") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  facet_grid(G0_label ~ cellLine, scales = "free_y") +
  scale_color_manual(values = c("baseline" = "#0072B2", "elevated" = "#D55E00")) +
  theme_bw() +
  labs(
    title = "Lagged Coupling Between Area Change and Death Rate",
    subtitle = "Positive lag means area-rate signal leads death-rate signal",
    x = "Lag (hours)",
    y = "Median replicate-level correlation",
    color = "Ploidy"
  )

rate_summary_scaled <- rate_summary %>%
  group_by(cellLine, G0, G0_label, signal) %>%
  mutate(value_scaled = value / max(abs(value), na.rm = TRUE)) %>%
  ungroup()

p_rate_signals <- ggplot(
  rate_summary_scaled,
  aes(hours, value_scaled, color = ploidy_label, linetype = signal, group = interaction(ploidy_label, signal, G0))
) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey50") +
  geom_line(linewidth = 0.8) +
  facet_grid(G0_label ~ cellLine, scales = "free_y") +
  scale_color_manual(values = c("baseline" = "#0072B2", "elevated" = "#D55E00")) +
  theme_bw() +
  labs(
    title = "Smoothed Signals Used for Lag Analysis",
    subtitle = "Signals are scaled within each cellLine x G0 x signal block to aid visual comparison",
    x = "Hours",
    y = "Scaled signal",
    color = "Ploidy",
    linetype = "Signal"
  )

(p_lag_profile / p_rate_signals) + plot_layout(heights = c(1.1, 1.4))
```

```{r lag-coupling-table}
lag_corr_tbl <- best_lag_tbl %>%
  group_by(cellLine) %>%
  summarise(
    median_best_corr = median(best_abs_corr, na.rm = TRUE),
    median_best_lag_hours = median(best_lag_hours, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(lag_corr_tbl, digits = 3)
```

The lag summaries suggest that for most lines the area-change and death-rate signals are coupled with negative correlation and modest positive lag, meaning shrinking median area often coincides with or slightly precedes rising death pressure. `MCF10A` is the main exception here.

## Morphology and Glucose Depletion

These plots address confluence by dropping count observations above 70% of the maximum total-cell count for each `cellLine x ploidy` combination. They then compare median alive-cell area to current glucose, glucose lagged by 24h, and glucose lagged by 48h. Glucose values are shown on a log scale and clipped to `0.02` only for plotting.

Within each panel, points connected by a line are the median time trajectory for one `cellLine x ploidy x G0` condition after the confluence filter.

```{r morphology-glucose, fig.height=16}
p_gluc_q50 <- ggplot(
  morph_glucose_summary,
  aes(glucose_hat_plot, q50, color = hours, shape = ploidy_label, group = interaction(ploidy_metric, G0))
) +
  geom_point(alpha = 0.8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~cellLine, scales = "free") +
  scale_x_log10() +
  scale_color_viridis_c(option = "C") +
  theme_bw() +
  labs(
    title = "Median Area Versus Remaining Glucose",
    x = "Estimated glucose",
    y = "Area q50",
    color = "Hours",
    shape = "Ploidy"
  )

p_gluc_lag24 <- ggplot(
  morph_glucose_summary,
  aes(glucose_lag24_plot, q50, color = hours, shape = ploidy_label, group = interaction(ploidy_metric, G0))
) +
  geom_point(alpha = 0.8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~cellLine, scales = "free") +
  scale_x_log10() +
  scale_color_viridis_c(option = "C") +
  theme_bw() +
  labs(
    title = "Median Area Versus Glucose Lagged 24h",
    x = "Estimated glucose 24h earlier",
    y = "Area q50",
    color = "Hours",
    shape = "Ploidy"
  )

p_gluc_lag48 <- ggplot(
  morph_glucose_summary,
  aes(glucose_lag48_plot, q50, color = hours, shape = ploidy_label, group = interaction(ploidy_metric, G0))
) +
  geom_point(alpha = 0.8) +
  geom_line(alpha = 0.7) +
  facet_wrap(~cellLine, scales = "free") +
  scale_x_log10() +
  scale_color_viridis_c(option = "C") +
  theme_bw() +
  labs(
    title = "Median Area Versus Glucose Lagged 48h",
    x = "Estimated glucose 48h earlier",
    y = "Area q50",
    color = "Hours",
    shape = "Ploidy"
  )

(p_gluc_q50 / p_gluc_lag24 / p_gluc_lag48) + plot_layout(heights = c(1.2, 1.2, 1.2))
```

```{r morphology-glucose-summary}
morph_glucose_assoc <- morph_glucose_summary %>%
  group_by(cellLine) %>%
  summarise(
    cor_q50_vs_glucose = suppressWarnings(cor(log10(glucose_hat_plot), q50, use = "complete.obs")),
    cor_q50_vs_glucose_lag24 = suppressWarnings(cor(log10(glucose_lag24_plot), q50, use = "complete.obs")),
    cor_q50_vs_glucose_lag48 = suppressWarnings(cor(log10(glucose_lag48_plot), q50, use = "complete.obs")),
    .groups = "drop"
  )

knitr::kable(morph_glucose_assoc, digits = 3)
```

The useful summary here is whether area tracks current glucose or past glucose more strongly. If the 24h- or 48h-lag correlation is stronger, that supports a delayed morphology response to nutrient history rather than an instantaneous glucose effect.

In the current summaries that delayed-history pattern is most obvious for `MCF10A` and `MDA-MB-231`, where lagged glucose correlates with median area more strongly than contemporaneous glucose. The other lines look weaker or less consistent, so this should be treated as line-specific rather than universal.

## Size-To-Cost Scaling

This ports the most interesting idea from `workflow/volume_aware_anal.Rmd`: relate glucose cost per unit cell production to a size proxy. Here the calculation is rebuilt from the row-aligned area dataset and the cost model is explicitly fit as:

`glucose consumed ~ a * biomass built + m * live-cell AUC`

with `a >= 0` and `m >= 0`.

Here `glucose consumed` is not the nominal `G0`. It is the estimated glucose at the first glucose measurement minus the estimated glucose at the last glucose measurement for each well.

The size proxy used below is now the same latent footprint estimate introduced above, not the older `24 h` snapshot. So the cost-scaling analysis uses the crowding-adjusted growth-phase area/volume proxy rather than a single raw-timepoint area.

```{r cost-fit-table}
cost_fit_tbl <- cost_fits %>%
  select(
    cellLine,
    ploidy_metric,
    ploidy_abs,
    ploidy_label,
    a,
    m,
    rmse,
    n,
    attached_area_hat,
    n_usable,
    area_24h,
    vol_flat,
    vol_sphere,
    ratio_cost,
    ratio_vol_flat,
    ratio_vol_sphere
  )

knitr::kable(cost_fit_tbl, digits = 3)
```

The very small `m` estimates and fairly tight `p_build` relationships imply the current accounting fit is behaving almost like a one-parameter biomass-yield model, especially outside `MCF10A`. That is exactly why an implied biomass-trajectory view is worth plotting.

```{r cost-input-diagnostics, fig.height=14}
p_gluc_init <- ggplot(
  glucose_init_check,
  aes(G0, glucose_initial, color = ploidy_label)
) +
  geom_point(size = 2.2, alpha = 0.85) +
  facet_wrap(~cellLine, scales = "free") +
  theme_bw() +
  labs(
    title = "Estimated Initial Glucose Versus Nominal G0",
    x = "Nominal initial glucose (G0)",
    y = "Estimated glucose at first measurement",
    color = "Ploidy"
  )

p_build <- ggplot(
  cost_fit_diagnostics,
  aes(dB, glucose_consumed, color = ploidy_label)
) +
  geom_point(alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~cellLine, scales = "free") +
  theme_bw() +
  labs(
    title = "Glucose Consumed Versus Biomass Built",
    x = "dB = net total-cell gain up to last glucose measurement",
    y = "Estimated glucose consumed",
    color = "Ploidy"
  )

p_auc <- ggplot(
  cost_fit_diagnostics,
  aes(AUC_live, glucose_consumed, color = ploidy_label)
) +
  geom_point(alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~cellLine, scales = "free") +
  theme_bw() +
  labs(
    title = "Glucose Consumed Versus Live-Cell AUC",
    x = "Live-cell AUC up to last glucose measurement",
    y = "Estimated glucose consumed",
    color = "Ploidy"
  )

p_gluc_init / p_build / p_auc
```

```{r cost-scaling-plots, fig.height=12}
p_cost <- ggplot(cost_fits, aes(factor(ploidy_metric), a, fill = cellLine)) +
  geom_col() +
  facet_wrap(~cellLine, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Estimated Construction Cost by Ploidy",
    subtitle = "Parameter a from glucose_consumed ~ a*dB + m*AUC_live",
    x = "Ploidy metric",
    y = "Construction cost"
  )

scaling_data <- bind_rows(
  cost_fits %>%
    transmute(cellLine, ploidy_metric, assumption = "Flat", volume_ratio = ratio_vol_flat, cost_ratio = ratio_cost),
  cost_fits %>%
    transmute(cellLine, ploidy_metric, assumption = "Sphere", volume_ratio = ratio_vol_sphere, cost_ratio = ratio_cost)
) %>%
  filter(is.finite(volume_ratio), is.finite(cost_ratio), volume_ratio > 0, cost_ratio > 0)

p_scaling <- ggplot(scaling_data, aes(volume_ratio, cost_ratio, color = assumption)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey40") +
  geom_point(size = 2.5) +
  geom_line(aes(group = interaction(cellLine, assumption)), alpha = 0.5) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~cellLine, scales = "free") +
  theme_bw() +
  labs(
    title = "Does Glucose Cost Scale With Size?",
    subtitle = "Reference line is isometric scaling",
    x = "Volume ratio relative to baseline ploidy",
    y = "Construction-cost ratio relative to baseline ploidy",
    color = "Volume proxy"
  )

p_cost / p_scaling
```

```{r cost-fit-quality, fig.height=10}
p_fit_quality <- ggplot(
  cost_fit_diagnostics,
  aes(glucose_consumed, glucose_pred, color = ploidy_label)
) +
  geom_point(alpha = 0.85) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey40") +
  facet_wrap(~cellLine, scales = "free") +
  theme_bw() +
  labs(
    title = "Observed Versus Predicted Glucose Consumption",
    x = "Observed glucose consumed",
    y = "Predicted glucose consumed",
    color = "Ploidy"
  )

p_resid <- ggplot(
  cost_fit_diagnostics,
  aes(dB, resid, color = ploidy_label)
) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey40") +
  geom_point(alpha = 0.85) +
  facet_wrap(~cellLine, scales = "free") +
  theme_bw() +
  labs(
    title = "Cost-Model Residuals Versus Biomass Built",
    x = "dB",
    y = "Observed - predicted glucose consumed",
    color = "Ploidy"
  )

p_fit_quality / p_resid
```

```{r cost-implied-timecourse, fig.height=20}
cellline_timecourse_plots <- cost_fit_timecourse_summary %>%
  split(.$cellLine) %>%
  lapply(function(df) {
    ggplot(
      df,
      aes(hours, value, color = series, linetype = series)
    ) +
      geom_line(linewidth = 0.8) +
      facet_grid(G0_label ~ ploidy_label, scales = "free_y") +
      theme_bw() +
      labs(
        title = unique(df$cellLine),
        x = "Hours",
        y = "Total cells",
        color = NULL,
        linetype = NULL
      )
  })

for (p in cellline_timecourse_plots) {
  print(p)
}
```

```{r cost-fit-summary}
cost_fit_summary_tbl <- cost_fit_timecourse_summary %>%
  pivot_wider(names_from = series, values_from = value) %>%
  group_by(cellLine, ploidy_label, G0) %>%
  summarise(
    rmse_timecourse = sqrt(mean((`Observed total cells` - `Predicted total cells`)^2, na.rm = TRUE)),
    cor_timecourse = suppressWarnings(cor(`Observed total cells`, `Predicted total cells`, use = "complete.obs")),
    .groups = "drop"
  ) %>%
  arrange(cellLine, G0, ploidy_label)

knitr::kable(head(cost_fit_summary_tbl, 20), digits = 3)
```

The timecourse summary shows this accounting model is not uniformly good. It looks promising in several non-MCF10A settings, but some conditions, especially parts of `MCF10A`, still have large trajectory error or even pathological correlation. So the result supports a basic biomass ODE as a plausible next branch, but not yet as a globally adequate description.

```{r cost-efficiency}
efficiency_long <- cost_fits %>%
  select(cellLine, ploidy_metric, cost_per_flat_vol, cost_per_sphere_vol) %>%
  pivot_longer(
    cols = starts_with("cost_per_"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      cost_per_flat_vol = "Cost per flat-volume proxy",
      cost_per_sphere_vol = "Cost per sphere-volume proxy"
    )
  )

ggplot(efficiency_long, aes(factor(ploidy_metric), value, fill = factor(ploidy_metric))) +
  geom_col(show.legend = FALSE) +
  facet_grid(metric ~ cellLine, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Glucose Cost Per Unit Size Proxy",
    x = "Ploidy metric",
    y = "Cost / size"
  )
```

## Potential Next Steps

Two next branches look especially well motivated by this notebook.

First, the current latent-volume quantity should be experimentally calibrated rather than left as a purely image-derived index. The most obvious options are:

- acquire matched `z`-stack imaging so projected area can be tied to actual 3D cell volume,
- collect an orthogonal biomass readout such as total protein per cell across the same low/high-ploidy states,
- use those measurements to learn a mapping from alive-cell area quantiles to true volume or biomass, rather than relying on `area^(3/2)` as a geometry prior.

Second, if that calibration looks credible, volume could replace ploidy as the global cross-line effect in the ODE pipeline. Concretely, that would mean:

- carrying a per-line or per-state volume estimate into `stan_ready_data` alongside or instead of the current ploidy metric,
- rewriting the hierarchical global-effect term so the shared shift is parameterized by volume rather than by ploidy label,
- checking whether volume-based transfer outperforms ploidy-based transfer in the current directional holdout workflows.

That second branch is scientifically attractive because it would test whether the apparent ploidy effect is actually a size/biomass effect in disguise, which is a much stronger and more portable claim.

## Notes

- `q50`, `q99`, and `tail_ratio = q99 / q50` are used as the main morphology summaries to avoid redundant quantile-derived features.
- The rate-coupling section is exploratory: it uses smoothed derivatives to estimate lead-lag structure between area change and death burden.
- The cost model still works with replicate-level count rows but well-level glucose summaries, matching the spirit of the original notebook while keeping the area data aligned to the actual count observations.
- `glucose_consumed` is estimated from the first and last glucose measurements within each well, not from nominal `G0`.
- Any strong effect seen here should still be treated as exploratory until promoted into a scripted analysis branch.
