---
title: "Competition Analysis"
author: "Codex"
date: "2026-03-23"
output: html_document
params:
  competition_path: "data/counts/competition/SUM-159-fuse.Rds"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook is a brief empirical report for the SUM-159 mixed-ploidy competition dataset in:

`data/counts/competition/SUM-159-fuse.Rds`

The immediate role of this dataset is not to enter the main Stan training path. Instead, it provides a direct mixed-culture check on the project's selection-strategy branch, which already simulates low- and high-ploidy cells in shared media.

The main questions here are:

1. What was measured in this object?
2. How does mixed-culture composition change across glucose?
3. Which summaries should be carried forward as validation targets for the simulation branch?

## Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(forcats)

competition_path <- params$competition_path

if (!file.exists(competition_path)) {
  stop(sprintf("Competition dataset not found: %s", competition_path))
}

competition_raw <- readRDS(competition_path)

required_cols <- c("frame", "label", "rep", "condition", "ncells", "glucose")
missing_cols <- setdiff(required_cols, names(competition_raw))
if (length(missing_cols)) {
  stop(sprintf(
    "Competition dataset is missing required columns: %s",
    paste(missing_cols, collapse = ", ")
  ))
}

competition_raw <- competition_raw %>%
  mutate(
    frame = as.integer(frame),
    rep = as.character(rep),
    condition = factor(as.character(condition), levels = sort(unique(as.numeric(as.character(condition))))),
    label = factor(as.character(label), levels = c("2N", "4N", "dead")),
    glucose = as.numeric(glucose),
    ncells = as.numeric(ncells)
  )

start_frame <- min(competition_raw$frame, na.rm = TRUE)
end_frame <- max(competition_raw$frame, na.rm = TRUE)

live_wide <- competition_raw %>%
  filter(label %in% c("2N", "4N")) %>%
  mutate(label = as.character(label)) %>%
  tidyr::pivot_wider(
    names_from = label,
    values_from = ncells,
    values_fill = 0
  ) %>%
  rename(n_2N = `2N`, n_4N = `4N`) %>%
  mutate(
    total_live = n_2N + n_4N,
    frac_4N = if_else(total_live > 0, n_4N / total_live, NA_real_),
    frac_2N = if_else(total_live > 0, n_2N / total_live, NA_real_),
    log_ratio_4N_2N = log((n_4N + 1e-8) / (n_2N + 1e-8))
  )

dead_df <- competition_raw %>%
  filter(as.character(label) == "dead") %>%
  transmute(frame, rep, condition, glucose, n_dead = ncells)

competition_df <- live_wide %>%
  left_join(dead_df, by = c("frame", "rep", "condition", "glucose")) %>%
  mutate(
    n_dead = coalesce(n_dead, 0),
    total_cells = total_live + n_dead,
    dead_fraction = if_else(total_cells > 0, n_dead / total_cells, NA_real_)
  )

object_audit <- tibble::tibble(
  metric = c("rows", "conditions", "replicates", "frames", "glucose_levels", "labels"),
  value = c(
    nrow(competition_raw),
    dplyr::n_distinct(competition_raw$condition),
    dplyr::n_distinct(competition_raw$rep),
    dplyr::n_distinct(competition_raw$frame),
    paste(sort(unique(competition_raw$glucose)), collapse = ", "),
    paste(levels(droplevels(competition_raw$label)), collapse = ", ")
  )
)

initial_state <- competition_df %>%
  filter(frame == start_frame) %>%
  select(rep, condition, glucose, n_2N, n_4N, n_dead, total_live, frac_4N, dead_fraction) %>%
  arrange(glucose, condition, rep)

condition_change <- competition_df %>%
  filter(frame %in% c(start_frame, end_frame)) %>%
  select(frame, rep, condition, glucose, frac_4N, log_ratio_4N_2N, total_live, n_dead, dead_fraction) %>%
  pivot_wider(
    names_from = frame,
    values_from = c(frac_4N, log_ratio_4N_2N, total_live, n_dead, dead_fraction),
    names_prefix = "frame_"
  )

resolve_frame_col <- function(base_name, frame_value, available_names) {
  candidates <- c(
    sprintf("%s_frame_%d", base_name, frame_value),
    sprintf("frame_%d_%s", frame_value, base_name)
  )
  hit <- candidates[candidates %in% available_names]
  if (!length(hit)) {
    stop(sprintf(
      "Could not find a column for base_name='%s' and frame=%d. Available columns include: %s",
      base_name,
      as.integer(frame_value),
      paste(utils::head(available_names, 20), collapse = ", ")
    ))
  }
  hit[[1]]
}

cc_names <- names(condition_change)
start_frac_col <- resolve_frame_col("frac_4N", start_frame, cc_names)
end_frac_col <- resolve_frame_col("frac_4N", end_frame, cc_names)
start_log_col <- resolve_frame_col("log_ratio_4N_2N", start_frame, cc_names)
end_log_col <- resolve_frame_col("log_ratio_4N_2N", end_frame, cc_names)
start_live_col <- resolve_frame_col("total_live", start_frame, cc_names)
end_live_col <- resolve_frame_col("total_live", end_frame, cc_names)
start_dead_col <- resolve_frame_col("n_dead", start_frame, cc_names)
end_dead_col <- resolve_frame_col("n_dead", end_frame, cc_names)
end_dead_frac_col <- resolve_frame_col("dead_fraction", end_frame, cc_names)

condition_change <- condition_change %>%
  mutate(
    delta_frac_4N = .data[[end_frac_col]] - .data[[start_frac_col]],
    delta_log_ratio = .data[[end_log_col]] - .data[[start_log_col]],
    live_fold_change = .data[[end_live_col]] / pmax(.data[[start_live_col]], 1e-8),
    dead_fold_change = .data[[end_dead_col]] / pmax(.data[[start_dead_col]], 1e-8)
  )

glucose_summary <- condition_change %>%
  group_by(glucose) %>%
  summarise(
    mean_delta_frac_4N = mean(delta_frac_4N, na.rm = TRUE),
    sd_delta_frac_4N = sd(delta_frac_4N, na.rm = TRUE),
    mean_delta_log_ratio = mean(delta_log_ratio, na.rm = TRUE),
    mean_live_fold_change = mean(live_fold_change, na.rm = TRUE),
    mean_dead_fraction_end = mean(.data[[end_dead_frac_col]], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(glucose)

trajectory_summary <- competition_df %>%
  group_by(glucose, frame) %>%
  summarise(
    mean_frac_4N = mean(frac_4N, na.rm = TRUE),
    mean_total_live = mean(total_live, na.rm = TRUE),
    mean_dead_fraction = mean(dead_fraction, na.rm = TRUE),
    .groups = "drop"
  )

peak_summary <- competition_df %>%
  group_by(rep, condition, glucose) %>%
  summarise(
    peak_total_live = max(total_live, na.rm = TRUE),
    peak_total_live_frame = frame[which.max(total_live)[1]],
    peak_frac_4N = max(frac_4N, na.rm = TRUE),
    peak_frac_4N_frame = frame[which.max(frac_4N)[1]],
    .groups = "drop"
  ) %>%
  arrange(glucose, condition, rep)

validated_states <- NULL
selection_strategy_note <- NULL
if (file.exists("data/stan_ready_data.Rds")) {
  stan_data <- readRDS("data/stan_ready_data.Rds")
  line_labels <- unique(as.character(stan_data$cellLine_per_well))
  validated_states <- line_labels[grepl("^SUM-159$", line_labels)]
  selection_strategy_note <- if (length(validated_states)) {
    "SUM-159 is present in stan_ready_data, so this dataset can be compared directly against fitted mixed-culture simulations for that line."
  } else {
    "SUM-159 was not detected in stan_ready_data, so this dataset currently supports empirical benchmarking and workflow design rather than direct fit-vs-data overlays."
  }
}
```

## Object Audit

```{r object-audit-table}
knitr::kable(object_audit, align = c("l", "l"))
```

The object is already in a useful long format. Each row is a count for one `frame x rep x condition x label` combination, where `label` distinguishes `2N`, `4N`, and `dead`.

For downstream use, the key derived variables are:

- `frac_4N = 4N / (2N + 4N)`
- `log_ratio_4N_2N = log(4N / 2N)`
- `dead_fraction = dead / (2N + 4N + dead)`

Those are the most natural targets for validating mixed-culture simulations.

## Initial Mixture Check

This table checks whether the experiment started close to a balanced co-culture in each condition.

```{r initial-mixture-table}
knitr::kable(initial_state, digits = 3)
```

The starting composition is close enough to 50:50 to make this a clean competition dataset rather than a strongly biased outgrowth assay.

## Composition Trajectories

```{r frac-trajectory-plot, fig.height=5}
ggplot(
  competition_df,
  aes(frame, frac_4N, group = interaction(rep, condition), color = factor(glucose))
) +
  geom_line(alpha = 0.35, linewidth = 0.7) +
  geom_point(alpha = 0.45, size = 1.2) +
  stat_summary(
    aes(group = glucose),
    fun = mean,
    geom = "line",
    linewidth = 1.2
  ) +
  scale_color_brewer(palette = "Dark2", name = "glucose") +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "4N Fraction Over Time",
    subtitle = "Thin lines show condition-replicate trajectories; thick lines show glucose-level means",
    x = "frame",
    y = "4N fraction among live cells"
  ) +
  theme_minimal(base_size = 11)
```

```{r log-ratio-trajectory-plot, fig.height=5}
ggplot(
  competition_df,
  aes(frame, log_ratio_4N_2N, group = interaction(rep, condition), color = factor(glucose))
) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.4) +
  geom_line(alpha = 0.35, linewidth = 0.7) +
  stat_summary(
    aes(group = glucose),
    fun = mean,
    geom = "line",
    linewidth = 1.2
  ) +
  scale_color_brewer(palette = "Dark2", name = "glucose") +
  labs(
    title = "Log Ratio of 4N to 2N Over Time",
    subtitle = "Positive values favor 4N; negative values favor 2N",
    x = "frame",
    y = "log(4N / 2N)"
  ) +
  theme_minimal(base_size = 11)
```

The log-ratio view is the most transferable summary if this dataset is later compared against the selection-strategy simulations, because that branch already reports final-state enrichment/depletion using a log ratio.

## Live and Dead Burden

```{r total-live-plot, fig.height=5}
ggplot(
  trajectory_summary,
  aes(frame, mean_total_live, color = factor(glucose))
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.5) +
  scale_color_brewer(palette = "Dark2", name = "glucose") +
  labs(
    title = "Mean Total Live Cells by Glucose",
    x = "frame",
    y = "mean total live cells"
  ) +
  theme_minimal(base_size = 11)
```

```{r dead-fraction-plot, fig.height=5}
ggplot(
  trajectory_summary,
  aes(frame, mean_dead_fraction, color = factor(glucose))
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.5) +
  scale_color_brewer(palette = "Dark2", name = "glucose") +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Mean Dead-Cell Fraction by Glucose",
    x = "frame",
    y = "dead fraction"
  ) +
  theme_minimal(base_size = 11)
```

These two plots matter because a mixed-culture model can be wrong in two qualitatively different ways:

- it can miss which ploidy state gains share,
- it can match composition while missing whether that outcome came from differential growth or differential death.

This dataset gives both signals.

## Start-to-End Summary

```{r glucose-summary-table}
glucose_summary_fmt <- glucose_summary %>%
  mutate(
    mean_delta_frac_4N = percent(mean_delta_frac_4N, accuracy = 0.1),
    sd_delta_frac_4N = percent(sd_delta_frac_4N, accuracy = 0.1),
    mean_dead_fraction_end = percent(mean_dead_fraction_end, accuracy = 0.1)
  )

knitr::kable(glucose_summary_fmt, digits = 3)
```

`mean_delta_frac_4N` and `mean_delta_log_ratio` are the simplest first-pass validation targets. A model that predicts the wrong sign at a given glucose is failing at the level of competitive direction, not just calibration.

## Suggested Use In The Project

This dataset should be used in the project in the following order:

1. As an empirical side branch that summarizes mixed-culture outcomes without changing the monoculture Stan pipeline.
2. As a validation target for the mixed-culture simulations in `R/selection_strategy_utils.R`.
3. As a model-ranking side metric: does a fitted model at least get the direction and rough magnitude of mixed-ploidy selection right for SUM-159?

The specific outputs worth standardizing from this notebook are:

- `frac_4N(frame)` by glucose
- `log_ratio_4N_2N(frame)` by glucose
- start-to-end `delta_log_ratio` by glucose
- total live-cell trajectory by glucose
- dead-cell fraction trajectory by glucose

```{r selection-branch-note, results='asis'}
if (!is.null(selection_strategy_note)) {
  cat(selection_strategy_note, "\n")
}
```

## Peak Summaries

These are useful if the project later wants a compact competition scorecard without carrying full trajectories.

```{r peak-summary-table}
knitr::kable(peak_summary, digits = 3)
```
