---
title: "Model-Free Ploidy Analysis"
author: "Codex"
date: "2026-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook reports a model-free ploidy analysis built directly from the curated `stan_ready_data` object rather than from fitted ODE parameters. The workflow has two stages:

1. extract per-condition features from observed count and glucose trajectories,
2. test whether a shared ploidy effect estimated from other lines improves held-out prediction of the missing ploidy state.

The notebook is a reporting layer. It expects the scripts below to have been run first:

```bash
Rscript scripts/extract_model_free_ploidy_features.R
Rscript scripts/evaluate_model_free_ploidy_transfer.R
```

## Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(scales)
library(patchwork)
library(forcats)

source("R/project_paths.R")
source("R/gpath_run_utils.R")
source("R/model_free_ploidy_utils.R")

analysis_root <- file.path("data", "model_free_ploidy")

paths <- list(
  stan_data = resolve_stan_data_path(),
  feature_panel = file.path(analysis_root, "feature_panel.Rds"),
  signature_panel = file.path(analysis_root, "signature_panel.Rds"),
  count_summary = file.path(analysis_root, "count_summary.Rds"),
  glucose_summary = file.path(analysis_root, "glucose_summary.Rds"),
  signature_effects = file.path(analysis_root, "signature_effects_long.Rds"),
  signature_pca_scores = file.path(analysis_root, "signature_pca_scores.Rds"),
  signature_pca_loadings = file.path(analysis_root, "signature_pca_loadings.Rds"),
  transfer_predictions = file.path(analysis_root, "transfer_predictions.Rds"),
  transfer_feature_summary = file.path(analysis_root, "transfer_feature_summary.Rds"),
  transfer_case_summary = file.path(analysis_root, "transfer_case_summary.Rds")
)

safe_read_rds <- function(path) {
  if (!file.exists(path)) {
    return(NULL)
  }
  readRDS(path)
}

feature_panel <- safe_read_rds(paths$feature_panel)
signature_panel <- safe_read_rds(paths$signature_panel)
count_summary <- safe_read_rds(paths$count_summary)
glucose_summary <- safe_read_rds(paths$glucose_summary)
signature_effects <- safe_read_rds(paths$signature_effects)
signature_pca_scores <- safe_read_rds(paths$signature_pca_scores)
signature_pca_loadings <- safe_read_rds(paths$signature_pca_loadings)
transfer_predictions <- safe_read_rds(paths$transfer_predictions)
transfer_feature_summary <- safe_read_rds(paths$transfer_feature_summary)
transfer_case_summary <- safe_read_rds(paths$transfer_case_summary)

cat("Working directory:", getwd(), "\n")
cat("stan_data path:", paths$stan_data, "\n")
cat("analysis root:", analysis_root, "\n")
```

## Output Check

```{r output-check}
output_status <- tibble(
  artifact = names(paths),
  path = unlist(paths),
  exists = file.exists(unlist(paths))
)

output_status
```

```{r maybe-stop, include=FALSE}
required_paths <- c(paths$feature_panel, paths$transfer_predictions, paths$transfer_feature_summary)
if (!all(file.exists(required_paths))) {
  stop("Required analysis outputs are missing. Run the extraction and transfer scripts before rendering this notebook.")
}
```

## Feature Inventory

```{r feature-inventory}
feature_panel %>%
  summarise(
    n_wells = n(),
    n_lines = n_distinct(cellLine),
    n_ploidy_states = n_distinct(paste(cellLine, ploidy_metric)),
    n_glucose_conditions = n_distinct(G0)
  )
```

```{r feature-preview}
feature_panel %>%
  select(cellLine, ploidy_metric, G0, live_peak, max_growth_rate, max_death_rate, glucose_drawdown, max_glucose_drawdown_rate) %>%
  arrange(cellLine, ploidy_metric, G0)
```

## Observed Trajectories

```{r trajectory-helper}
plot_one_trajectory <- function(cell_line, g0) {
  meta_sub <- feature_panel %>%
    filter(cellLine == cell_line, abs(G0 - g0) < 1e-12) %>%
    select(well_idx, ploidy_metric)

  cnt <- count_summary %>%
    semi_join(meta_sub, by = "well_idx") %>%
    select(well_idx, ploidy_metric, hours, live_cells, dead_cells) %>%
    pivot_longer(cols = c(live_cells, dead_cells), names_to = "state", values_to = "value")

  gluc <- glucose_summary %>%
    semi_join(meta_sub, by = "well_idx") %>%
    select(well_idx, ploidy_metric, hours, glucose_hat)

  p_count <- ggplot(cnt, aes(hours, value, color = factor(ploidy_metric))) +
    geom_line(linewidth = 0.9) +
    facet_wrap(~state, scales = "free_y") +
    scale_color_brewer(palette = "Set1", name = "ploidy") +
    labs(
      title = sprintf("%s | G0 = %s", cell_line, g0),
      x = "hours",
      y = "count"
    ) +
    theme_bw()

  p_gluc <- ggplot(gluc, aes(hours, glucose_hat, color = factor(ploidy_metric))) +
    geom_line(linewidth = 0.9) +
    scale_color_brewer(palette = "Set1", name = "ploidy") +
    labs(x = "hours", y = "estimated glucose") +
    theme_bw()

  p_count / p_gluc
}
```

```{r example-trajectory, fig.height=6, fig.width=8}
plot_one_trajectory("SNU668", 0.25)
```

## Glucose-Response Signatures

This notebook now distinguishes:

- per-condition trajectory features in `feature_panel`, and
- glucose-response signatures in `signature_panel`.

The signature layer is the transfer target. It uses bespoke summaries:

- growth: low-glucose median, high-glucose median, and glucose-response slope
- death: low-glucose median, high-glucose median, and glucose-response slope
- yield: regressions of alive AUC, dead AUC, total AUC, and peak total cells against `log1p(G0)`
- glucose handling: low-glucose exhaustion timing and high-glucose drawdown summaries

```{r top-effects}
signature_panel %>%
  select(cellLine, ploidy_metric, starts_with("growth_"), starts_with("death_"), starts_with("yield_alive_auc"), starts_with("yield_total_peak")) %>%
  arrange(cellLine, ploidy_metric)
```

## Direct Ploidy Effects On Signatures

The effect definition here is per-unit-ploidy change within each line:

`effect = (signature_high - signature_low) / (ploidy_high - ploidy_low)`

```{r top-effects-2}
effect_summary <- signature_effects %>%
  group_by(feature) %>%
  summarise(
    mean_effect = mean(effect_per_ploidy, na.rm = TRUE),
    sd_effect = sd(effect_per_ploidy, na.rm = TRUE),
    sign_consistency = mean(sign(effect_per_ploidy) == sign(mean(effect_per_ploidy, na.rm = TRUE)), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(sign_consistency), desc(abs(mean_effect)))

effect_summary %>% slice_head(n = 15)
```

```{r effect-heatmap, fig.height=8, fig.width=11}
plot_features <- transfer_feature_summary %>%
  arrange(desc(mean_abs_err_improvement)) %>%
  slice_head(n = 12) %>%
  pull(feature)

signature_effects %>%
  filter(feature %in% plot_features) %>%
  mutate(
    feature = forcats::fct_reorder(feature, effect_per_ploidy, .fun = median, na.rm = TRUE),
    line_label = cellLine
  ) %>%
  ggplot(aes(x = feature, y = line_label, fill = effect_per_ploidy)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b") +
  labs(
    title = "Empirical Ploidy Effects On Glucose-Response Signatures",
    x = NULL,
    y = NULL,
    fill = "effect\nper ploidy"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Shared Effect Structure

```{r pca-plot, fig.height=5, fig.width=8}
if (!is.null(signature_pca_scores) && nrow(signature_pca_scores)) {
  ggplot(signature_pca_scores, aes(PC1, PC2, color = cellLine)) +
    geom_hline(yintercept = 0, linewidth = 0.3, color = "grey70") +
    geom_vline(xintercept = 0, linewidth = 0.3, color = "grey70") +
    geom_point(alpha = 0.9, size = 3) +
    theme_bw() +
    labs(
      title = "PCA of Aggregated Ploidy-Effect Signatures",
      subtitle = "Each point is one cell line"
    )
}
```

```{r pca-loadings}
if (!is.null(signature_pca_loadings) && nrow(signature_pca_loadings)) {
  signature_pca_loadings %>%
    arrange(desc(abs(PC1))) %>%
    slice_head(n = 12)
}
```

## Transfer Evaluation

The transfer baseline is `null = observed ploidy state predicts missing signature with no change`. The transfer prediction adds the median cross-line ploidy effect for that signature.

```{r transfer-summary}
transfer_feature_summary %>%
  select(feature, n_cases, scaled_mae_null, scaled_mae_transfer, mean_scaled_err_improvement, transfer_win_rate, sign_accuracy) %>%
  arrange(desc(mean_scaled_err_improvement))
```

```{r transfer-bar, fig.height=7, fig.width=10}
transfer_feature_summary %>%
  slice_head(n = 15) %>%
  mutate(feature = forcats::fct_reorder(feature, mean_scaled_err_improvement)) %>%
  ggplot(aes(x = feature, y = mean_scaled_err_improvement, fill = transfer_win_rate)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Features Most Improved by Cross-Line Ploidy Transfer",
    x = NULL,
    y = "mean scaled error improvement",
    fill = "transfer\nwin rate"
  ) +
  theme_bw()
```

```{r case-summary}
transfer_case_summary %>%
  arrange(desc(mean_scaled_err_improvement))
```

```{r transfer-scatter, fig.height=6, fig.width=9}
top_transfer_features <- transfer_feature_summary %>%
  slice_head(n = 6) %>%
  pull(feature)

transfer_predictions %>%
  filter(feature %in% top_transfer_features) %>%
  mutate(feature = forcats::fct_reorder(feature, abs_err_improvement, .fun = median, na.rm = TRUE)) %>%
  ggplot(aes(x = true_value, y = pred_transfer - pred_null, color = direction)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_point(size = 2, alpha = 0.8) +
  facet_wrap(~feature, scales = "free") +
  theme_bw() +
  labs(
    title = "Transferred Adjustment Relative to Null",
    subtitle = "Positive values mean the shared ploidy effect pushes the prediction upward",
    x = "true held-out feature value",
    y = "transfer minus null prediction"
  )
```

## Notes

- Feature values are medians across replicate observations at each timepoint within a well-condition.
- Glucose is reconstructed from `lum_obs` using the fixed calibration parameters stored in `stan_ready_data`.
- The transfer layer now works on glucose-response signatures rather than raw `G0`-matched condition summaries.
- The cross-line effect is normalized per unit of `ploidy_metric`, which matters because `MDA-MB-231` has a smaller ploidy shift than the other paired lines.
