---
title: "Ecological Modelling"
author: "RJB"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook is now the reporting layer for the current `gpath` modelling workflow. It is not intended to recreate the full analysis from scratch. Instead it:

- documents the current modelling path,
- states the upstream scripts/commands that generate required artefacts where known,
- loads saved results from the current workflow,
- presents model selection, diagnostics, transfer results, and parameter-transfer summaries.

Where an upstream command is not known exactly from the repo history, that is stated explicitly.

## Setup

```{r libraries, include=FALSE}
library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(tibble)
library(posterior)
library(patchwork)

source("R/project_paths.R")
source(get_model_r_path("gpath", "v1"))
source("R/gpath_run_utils.R")
source("R/posterior_io_utils.R")
source("R/elpd_transfer_utils.R")
source("R/parameter_transfer_utils.R")
source("R/parameter_transfer_plot_utils.R")
source("R/transfer_cv_plot_utils.R")

focus_common_parameters <- c("ae[1]", "ah[1]", "Y_R[1]", "m")

normalize_focus_parameter <- function(parameter) {
  dplyr::case_when(
    parameter %in% c("ae", "ae[1]") ~ "ae[1]",
    parameter %in% c("ah", "ah[1]") ~ "ah[1]",
    parameter %in% c("Y_R", "Y_R[1]") ~ "Y_R[1]",
    parameter == "m" ~ "m",
    TRUE ~ parameter
  )
}

project_paths <- list(
  stan_data = resolve_stan_data_path(),
  assessment_run_ids = file.path("models", "gpath", "v1", "assessment_run_ids.txt"),
  optim_root = file.path("data", "runs", "gpath", "v1", "optim"),
  report_dataset_label = "gstarvation_v1",
  transfer_optim_root = "data/gpath_transfer_cv/gpath",
  transfer_nuts_root = "data/gpath_transfer_cv_nuts/gpath",
  transfer_nuts_report_root = file.path("data", "gpath_transfer_cv_nuts", "report"),
  report_export_root = file.path("data", "report_exports", "model_analysis"),
  agent_text_path = file.path("workflow", "model_analysis.txt")
)

cat("Current working directory:", getwd(), "\n")
cat("stan_data path:", project_paths$stan_data, "\n")
cat("assessment set:", project_paths$assessment_run_ids, "\n")
cat("report dataset:", project_paths$report_dataset_label, "\n")
```

```{r helpers, include=FALSE}
source("R/model_analysis_utils.R")
stan_data_ref <- readRDS(project_paths$stan_data)
line_label_map <- make_line_label_map(stan_data_ref$line_map)
```

Helper functions for this notebook now live in `R/model_analysis_utils.R`. That helper layer is currently too bloated and should be streamlined in a refactor ASAP.

```{r upstream-commands, include=FALSE}
artifact_commands <- tribble(
  ~artifact_group, ~path, ~command, ~status_note,
  "Stan data", project_paths$stan_data,
  "Rscript R/prepare_data.R",
  "Confirmed on 2026-03-20 by a byte-identical rerun of `stan_ready_data.Rds`.",
  "Assessment model set", project_paths$assessment_run_ids,
  "Tracked file read by scripts/make_gpath_optim_spec.R when generating rerun specs.",
  "Canonical model-id set for current `gpath v1` assessment.",
  "Versioned hierarchical optimization", "data/runs/gpath/v1/optim/<DATASET_LABEL>/<RUN_ID>/<RUN_LABEL>",
  "bash slurm/submit_gpath_optim_pipeline.sh <SPEC_FILE.tsv> ; specs can be generated with Rscript scripts/make_gpath_optim_spec.R <SPEC_FILE.tsv> <RUN_IDS>",
  "Trusted optimisation source. Exact historical submission command is not stored in the current run outputs; current run_id/run_label provenance is reported below.",
  "Versioned hierarchical NUTS", "not yet migrated",
  "No new versioned NUTS pipeline for legacy hierarchical fits yet.",
  "Transfer NUTS still seeds from the older legacy NUTS outputs for now.",
  "Directional transfer (optimization)", "data/gpath_transfer_cv/gpath/<RUN_ID>",
  "bash slurm/submit_gpath_transfer_pipeline.sh MODEL R P W C M all low_to_high,high_to_low null,transfer,oracle 1:K 500 1000 0.99 12 16 ecology/stan_ready_data.Rds data/gpath_transfer_cv 0",
  "Current optimization transfer path. Exact historical submission arguments are not stored in the saved summaries; current selected-model paths are reported below.",
  "Directional transfer (optimization summaries)", "data/gpath_transfer_cv/gpath/<RUN_ID>/*.Rds",
  "Rscript scripts/summarize_gpath_transfer_cv.R gpath <RUN_ID> data/gpath_transfer_cv",
  "Current summary path.",
  "Directional transfer (NUTS)", "data/gpath_transfer_cv_nuts/gpath/<RUN_ID>",
  "bash slurm/submit_gpath_transfer_nuts_pipeline.sh MODEL R P W C M all low_to_high,high_to_low null,transfer,oracle 1:4 500 1000 0.99 12 16 ecology/stan_ready_data.Rds data/gpath_transfer_cv_nuts",
  "Current NUTS transfer path. Exact historical submission arguments are not stored in the saved summaries; current selected-model paths are reported below.",
  "Directional transfer (NUTS summaries)", "data/gpath_transfer_cv_nuts/gpath/<RUN_ID>/*.Rds",
  "Rscript scripts/summarize_gpath_transfer_nuts.R gpath <RUN_ID> data/gpath_transfer_cv_nuts",
  "Current summary path."
)

print(as.data.frame(artifact_commands), row.names = FALSE)
```

## Available Outputs

```{r available-outputs, include=FALSE}
report_export_root <- ensure_report_export_root()
ranking_df <- build_model_ranking()
manifest_df <- summarize_manifest_status(ranking_df)
stability_df <- build_stability_summary(ranking_df)
selected_model_id_aic <- find_selected_model(ranking_df)
selected_model_id <- selected_model_id_aic
support_set_df <- build_support_set(ranking_df, stability_df)
shortlist_df <- tibble()
metric_glossary_df <- build_metric_glossary()
provenance_df <- build_current_provenance_table(
  ranking_df = ranking_df,
  selected_model_id = selected_model_id,
  project_paths = project_paths
)
model_screen_df <- tibble()
selected_model_summary_df <- tibble()
raw_theta_ploidy_global_df <- collect_best_raw_theta_ploidy_global(ranking_df)
raw_theta_ploidy_transfer_df <- collect_best_raw_theta_ploidy_transfer(ranking_df)
raw_theta_ploidy_df <- bind_rows(raw_theta_ploidy_global_df, raw_theta_ploidy_transfer_df)
common_raw_theta_parameters <- if (nrow(raw_theta_ploidy_global_df) && all(c("missing", "parameter", "model_id") %in% names(raw_theta_ploidy_global_df))) {
  raw_theta_ploidy_global_df %>%
    filter(!missing) %>%
    dplyr::count(parameter, name = "n_models_present") %>%
    filter(n_models_present == n_distinct(raw_theta_ploidy_global_df$model_id)) %>%
    pull(parameter)
} else {
  character(0)
}

audit_index_df <- bind_rows(
  write_audit_artifact(ranking_df, "ranking_df"),
  write_audit_artifact(manifest_df, "manifest_df"),
  write_audit_artifact(raw_theta_ploidy_global_df, "raw_theta_ploidy_global_df"),
  write_audit_artifact(raw_theta_ploidy_transfer_df, "raw_theta_ploidy_transfer_df"),
  write_audit_artifact(raw_theta_ploidy_df, "raw_theta_ploidy_df")
)
raw_theta_agreement_summary_df <- tibble()
raw_theta_agreement_model_df <- tibble()
raw_theta_agreement_direction_df <- tibble()
focused_hit_df <- tibble()

cat("Available trusted gpath model IDs:\n")
print(ranking_df$model_id)
cat("\nSelected model for downstream reporting:", selected_model_id, "\n")
cat("\nModels in the current Delta AIC <= 10 support set:\n")
print(support_set_df$model_id)
cat("\nAudit export root:", report_export_root, "\n")
```

```{r current-provenance, include=FALSE}
print(as.data.frame(provenance_df), row.names = FALSE)
```

```{r audit-artifact-index, include=FALSE}
print(as.data.frame(audit_index_df), row.names = FALSE)
```

```{r trusted-optimisation-runs, include=FALSE}
if (!nrow(manifest_df)) {
  announce_missing("No manifest-backed optimisation outputs found under %s", project_paths$optim_root)
} else {
  print(as.data.frame(manifest_df), row.names = FALSE)
}
```

## Model Selection

```{r model-selection-plot, fig.width=19, fig.height=6, out.width='100%'}
if (!nrow(ranking_df)) {
  announce_missing("Skipping model-selection plot because no ranking table was built.")
} else {
  plot_df <- ranking_df %>%
    distinct(model_id, .keep_all = TRUE) %>%
    left_join(
      stability_df %>%
        select(model_id, prop_within_1, n_within_1),
      by = "model_id"
    ) %>%
    mutate(
      prop_within_1 = if_else(is.na(prop_within_1), 0, prop_within_1),
      n_within_1 = if_else(is.na(n_within_1), 0L, as.integer(n_within_1))
    )

  alias_levels <- plot_df %>%
    arrange(delta_AIC) %>%
    pull(alias) %>%
    unique_preserve()

  alias_plot_map <- plot_df %>%
    distinct(alias, alias_plot) %>%
    tibble::deframe()

  plot_df <- plot_df %>%
    mutate(alias = factor(alias, levels = rev(alias_levels)))

  pareto_front <- plot_df %>%
    arrange(k, deviance) %>%
    filter(deviance == cummin(deviance)) %>%
    group_by(deviance) %>%
    slice_min(order_by = k, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(k)
  pareto_front_model_ids <- unique_preserve(as.character(pareto_front$model_id))
  pareto_front_alias_levels <- pareto_front %>%
    arrange(delta_AIC) %>%
    pull(alias_plot) %>%
    unique_preserve()

  p_aic <- ggplot(plot_df, aes(x = alias, y = delta_AIC)) +
    geom_col(fill = "grey35") +
    coord_flip() +
    scale_x_discrete(labels = function(x) unname(alias_plot_map[as.character(x)])) +
    theme_bw() +
    labs(x = "", y = expression(Delta ~ AIC))

  p_pareto <- ggplot(plot_df, aes(x = k, y = deviance)) +
    geom_point(color = "grey60", size = 2.5) +
    geom_line(data = pareto_front, color = "firebrick", linetype = "dashed") +
    geom_point(data = pareto_front, color = "firebrick", size = 3) +
    ggrepel::geom_text_repel(
      data = pareto_front,
      aes(label = alias_plot),
      color = "firebrick",
      direction = "both",
      box.padding = 0.35,
      point.padding = 0.2,
      min.segment.length = 0,
      max.overlaps = Inf,
      size = 3
    ) +
    theme_bw() +
    labs(x = "Effective parameters (k)", y = "Deviance")

  p_stability <- ggplot(plot_df, aes(x = alias, y = n_within_1)) +
    geom_col(fill = "steelblue4") +
    coord_flip() +
    scale_x_discrete(labels = function(x) unname(alias_plot_map[as.character(x)])) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    theme_bw() +
    labs(x = "", y = "Found optimum (/250)")

  p_aic + p_pareto + p_stability + plot_layout(ncol = 3)
}
```

```{r pareto-nuts-setup, include=FALSE}
pareto_nuts_focus_summary_df <- tibble()
pareto_nuts_focus_draws_df <- tibble()
pareto_nuts_focus_raw_theta_draws_df <- tibble()
pareto_nuts_growth_model_df <- tibble()
pareto_nuts_growth_delta_surface_df <- tibble()

if (exists("pareto_front_model_ids") && length(pareto_front_model_ids)) {
  source("R/nuts_consistency_utils.R")

  pareto_alias_df <- ranking_df %>%
    distinct(model_id, alias, alias_plot) %>%
    filter(model_id %in% pareto_front_model_ids)

  nuts_focus_draws_sidecar_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_focus_draws.Rds"))
  nuts_focus_raw_theta_draws_sidecar_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_focus_raw_theta_ploidy_draws.Rds"))
  nuts_focus_summary_sidecar_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_focus_summary.Rds"))
  nuts_growth_model_sidecar_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_growth_model_summary.Rds"))
  nuts_growth_delta_surface_sidecar_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_growth_delta_surface_summary.Rds"))

  if (is.null(nuts_focus_draws_sidecar_df)) nuts_focus_draws_sidecar_df <- tibble()
  if (is.null(nuts_focus_raw_theta_draws_sidecar_df)) nuts_focus_raw_theta_draws_sidecar_df <- tibble()
  if (is.null(nuts_focus_summary_sidecar_df)) nuts_focus_summary_sidecar_df <- tibble()
  if (is.null(nuts_growth_model_sidecar_df)) nuts_growth_model_sidecar_df <- tibble()
  if (is.null(nuts_growth_delta_surface_sidecar_df)) nuts_growth_delta_surface_sidecar_df <- tibble()

  pareto_nuts_focus_draws_df <- nuts_focus_draws_sidecar_df %>%
    filter(model_id %in% pareto_front_model_ids, parameter %in% focus_common_parameters)
  pareto_nuts_focus_raw_theta_draws_df <- nuts_focus_raw_theta_draws_sidecar_df %>%
    filter(model_id %in% pareto_front_model_ids, parameter %in% focus_common_parameters)

  pareto_nuts_focus_summary_df <- nuts_focus_summary_sidecar_df %>%
    filter(model_id %in% pareto_front_model_ids, parameter %in% focus_common_parameters)

  if (!nrow(pareto_nuts_focus_draws_df) || !all(focus_common_parameters %in% unique(pareto_nuts_focus_draws_df$parameter))) {
    focus_draw_df <- collect_oracle_focus_parameter_draws(
      model_ids = pareto_front_model_ids,
      stan_data = readRDS(project_paths$stan_data),
      focus_parameters = focus_common_parameters,
      transfer_nuts_root = project_paths$transfer_nuts_root,
      max_draws_per_model = 400L
    )
    pareto_nuts_focus_draws_df <- focus_draw_df
    pareto_nuts_focus_summary_df <- summarize_focus_parameter_posteriors(focus_draw_df)
  }

  pareto_nuts_focus_draws_df <- pareto_nuts_focus_draws_df %>%
    filter(model_id %in% pareto_front_model_ids, parameter %in% focus_common_parameters) %>%
    left_join(pareto_alias_df, by = "model_id")
  pareto_nuts_focus_raw_theta_draws_df <- pareto_nuts_focus_raw_theta_draws_df %>%
    filter(model_id %in% pareto_front_model_ids, parameter %in% focus_common_parameters) %>%
    left_join(pareto_alias_df, by = "model_id")

  pareto_nuts_focus_summary_df <- pareto_nuts_focus_summary_df %>%
    filter(model_id %in% pareto_front_model_ids, parameter %in% focus_common_parameters) %>%
    left_join(pareto_alias_df, by = "model_id")

  pareto_nuts_growth_model_df <- nuts_growth_model_sidecar_df %>%
    filter(model_id %in% pareto_front_model_ids) %>%
    left_join(pareto_alias_df, by = "model_id")

  pareto_nuts_growth_delta_surface_df <- nuts_growth_delta_surface_sidecar_df %>%
    filter(model_id %in% pareto_front_model_ids) %>%
    left_join(pareto_alias_df, by = "model_id")
}
```

## Pareto-Front NUTS Posterior Consistency

```{r pareto-nuts-focus-parameter-intervals, fig.height=8, fig.width=16, out.width='100%'}
if (!nrow(pareto_nuts_focus_summary_df)) {
  announce_missing("No Pareto-front oracle NUTS posterior summaries available for interval plots.")
} else {
  plot_df <- pareto_nuts_focus_summary_df %>%
    mutate(
      model_alias = factor(alias_plot, levels = rev(pareto_front_alias_levels)),
      parameter = factor(parameter, levels = focus_common_parameters),
      line_label = label_line_ids(line_id, line_label_map),
      ploidy_label = factor(ploidy_label, levels = c("low_ploidy", "high_ploidy"), labels = c("Low ploidy", "High ploidy"))
    )

  ggplot(plot_df, aes(y = model_alias, color = ploidy_label)) +
    geom_segment(aes(x = q025, xend = q975, yend = model_alias), linewidth = 0.7, position = position_dodge(width = 0.6)) +
    geom_segment(aes(x = q25, xend = q75, yend = model_alias), linewidth = 1.4, position = position_dodge(width = 0.6)) +
    geom_point(aes(x = median), size = 2.0, position = position_dodge(width = 0.6)) +
    facet_grid(line_label ~ parameter, scales = "free_x", drop = FALSE) +
    scale_color_manual(values = c("Low ploidy" = "#1b9e77", "High ploidy" = "#d95f02")) +
    scale_x_continuous(trans = scales::pseudo_log_trans(base = 10)) +
    theme_bw() +
    labs(x = "Posterior value", y = "Model", color = "")
}
```

```{r pareto-nuts-focus-parameter-ridgelines, fig.height=9, fig.width=16, out.width='100%'}
if (!nrow(pareto_nuts_focus_draws_df)) {
  announce_missing("No Pareto-front oracle NUTS focus-parameter draws available for ridgeline plots.")
} else {
  draw_plot_df <- pareto_nuts_focus_draws_df %>%
    mutate(
      model_alias = factor(alias_plot, levels = rev(pareto_front_alias_levels)),
      line_label = label_line_ids(line_id, line_label_map),
      parameter = factor(parameter, levels = focus_common_parameters),
      ploidy_label = factor(ploidy_label, levels = c("low_ploidy", "high_ploidy"), labels = c("Low ploidy", "High ploidy"))
    )

  ridge_df <- build_density_ridgeline_df(
    draw_df = draw_plot_df,
    y_var = "model_alias",
    y_levels = levels(draw_plot_df$model_alias),
    facet_vars = c("line_label", "parameter")
  )

  ggplot(ridge_df, aes(x = value, group = interaction(line_label, parameter, y_label, ploidy_label))) +
    geom_ribbon(aes(ymin = y_base, ymax = y_top, fill = ploidy_label), alpha = 0.28, color = NA) +
    geom_line(aes(y = y_top, color = ploidy_label), linewidth = 0.7) +
    facet_grid(line_label ~ parameter, scales = "free_x", drop = FALSE) +
    scale_y_continuous(
      breaks = seq_along(levels(draw_plot_df$model_alias)),
      labels = levels(draw_plot_df$model_alias),
      expand = expansion(mult = c(0.02, 0.05))
    ) +
    scale_fill_manual(values = c("Low ploidy" = "#1b9e77", "High ploidy" = "#d95f02")) +
    scale_color_manual(values = c("Low ploidy" = "#1b9e77", "High ploidy" = "#d95f02")) +
    scale_x_continuous(trans = scales::pseudo_log_trans(base = 10)) +
    theme_bw() +
    labs(x = "Posterior value", y = "Model", fill = "", color = "")
}
```

```{r pareto-nuts-focus-raw-theta-ridgelines, fig.height=6, fig.width=16, out.width='100%'}
if (!nrow(pareto_nuts_focus_raw_theta_draws_df)) {
  announce_missing("No Pareto-front oracle NUTS raw_theta_ploidy draws available for ridgeline plots.")
} else {
  draw_plot_df <- pareto_nuts_focus_raw_theta_draws_df %>%
    mutate(
      model_alias = factor(alias_plot, levels = rev(pareto_front_alias_levels)),
      parameter = factor(parameter, levels = focus_common_parameters),
      ploidy_label = factor("Ploidy effect", levels = "Ploidy effect")
    )

  ridge_df <- build_density_ridgeline_df(
    draw_df = draw_plot_df %>% mutate(panel = factor("raw_theta_ploidy")),
    y_var = "model_alias",
    y_levels = levels(draw_plot_df$model_alias),
    facet_vars = c("panel", "parameter")
  )

  ggplot(ridge_df, aes(x = value, group = interaction(panel, parameter, y_label, ploidy_label))) +
    geom_ribbon(aes(ymin = y_base, ymax = y_top, fill = ploidy_label), alpha = 0.32, color = NA) +
    geom_line(aes(y = y_top, color = ploidy_label), linewidth = 0.7) +
    facet_grid(. ~ parameter, scales = "free_x", drop = FALSE) +
    scale_y_continuous(
      breaks = seq_along(levels(draw_plot_df$model_alias)),
      labels = levels(draw_plot_df$model_alias),
      expand = expansion(mult = c(0.02, 0.05))
    ) +
    scale_fill_manual(values = c("Ploidy effect" = "#7570b3")) +
    scale_color_manual(values = c("Ploidy effect" = "#5b4d9a")) +
    theme_bw() +
    labs(x = "raw_theta_ploidy posterior value", y = "Model", fill = "", color = "")
}
```

```{r pareto-nuts-growth-curves-by-model, fig.height=10, fig.width=16, out.width='100%'}
if (!nrow(pareto_nuts_growth_model_df)) {
  announce_missing("No Pareto-front oracle NUTS growth-curve summaries available for model-level plotting.")
} else {
  plot_df <- pareto_nuts_growth_model_df %>%
    mutate(
      model_alias = factor(alias_plot, levels = pareto_front_alias_levels),
      line_label = label_line_ids(line_id, line_label_map),
      ploidy_label = factor(ploidy_label, levels = c("low_ploidy", "high_ploidy"), labels = c("Low ploidy", "High ploidy"))
    )

  ggplot(plot_df, aes(x = glucose, y = median_growth, color = ploidy_label, fill = ploidy_label)) +
    geom_ribbon(aes(ymin = q025_growth, ymax = q975_growth), alpha = 0.18, color = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.3) +
    facet_grid(line_label ~ model_alias, scales = "free_y") +
    scale_color_manual(values = c("Low ploidy" = "#1b9e77", "High ploidy" = "#d95f02")) +
    scale_fill_manual(values = c("Low ploidy" = "#1b9e77", "High ploidy" = "#d95f02")) +
    theme_bw() +
    scale_x_log10() +
    labs(
      x = "Initial glucose",
      y = "Instantaneous net growth",
      color = "",
      fill = ""
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank(),
      strip.background = element_rect(fill = "grey95")
    )
}
```

```{r pareto-nuts-growth-delta-heatmap-by-model, fig.height=10, fig.width=16, out.width='100%'}
if (!nrow(pareto_nuts_growth_delta_surface_df)) {
  announce_missing("No Pareto-front oracle NUTS delta-growth surface summaries available for heatmap plotting.")
} else {
  plot_df <- pareto_nuts_growth_delta_surface_df %>%
    mutate(
      model_alias = factor(alias_plot, levels = pareto_front_alias_levels),
      line_label = label_line_ids(line_id, line_label_map),
      resource2 = pmax(0, pmin(1, resource2))
    )

  fill_limit <- max(abs(plot_df$median_delta_growth), na.rm = TRUE)
  if (!is.finite(fill_limit) || fill_limit <= 0) {
    fill_limit <- 1e-6
  }

 ggplot(plot_df, aes(x = glucose, y = resource2, fill = median_delta_growth)) +
  geom_tile() +
  facet_grid(line_label ~ model_alias) +
  scale_x_log10() +
  scale_fill_gradientn(
    colours = c("#2166ac", "#67a9cf", "#f7f7f7", "#ef8a62", "#b2182b"),
    values = scales::rescale(c(-fill_limit, -0.15 * fill_limit, 0, 0.15 * fill_limit, fill_limit)),
    limits = c(-fill_limit, fill_limit),
    oob = scales::squish
  ) +
  theme_bw() +
  labs(x = "Initial glucose", y = "Resource 2", fill = "High - Low\nnet growth") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    strip.background = element_rect(fill = "grey95")
  )
}
```

```{r model-selection-table}
if (!nrow(ranking_df)) {
  announce_missing("No trusted optimisation outputs found under %s", project_paths$optim_root)
} else {
  print(as.data.frame(ranking_df %>% select(model_id, alias, best_lp, k, AIC, delta_AIC)), row.names = FALSE)
}
```

```{r model-support-set}
if (!nrow(support_set_df)) {
  announce_missing("No models fall within the current Delta AIC support threshold.")
} else {
  cat("Strict AIC support set (`delta_AIC <= 10`)\n\n")
  print(as.data.frame(support_set_df), row.names = FALSE)
}
```

```{r model-selection-takeaway, results='asis'}
if (nrow(ranking_df)) {
  best_row <- ranking_df %>% slice_head(n = 1)
  support_n <- nrow(support_set_df)
  knitr::asis_output(sprintf(
    "**Takeaway.** `%s` (%s) is the best pure-fit model by AIC, but downstream interpretation is no longer driven by that strict fit support set alone. The strict `delta_AIC <= 10` support set currently contains `%d` model(s), so the report now builds a broader aggregated shortlist using fit, optimisation stability, directional agreement, predictive transfer, and parameter transfer.\n",
    best_row$model_id[[1]],
    best_row$alias[[1]],
    support_n
  ))
}
```

## Best-Fit `raw_theta_ploidy` Coefficients

This section previews the `raw_theta_ploidy` vector from the best optimization start for each ranked `model_id`. The full audit table is exported.

```{r best-raw-theta-ploidy}
if (!nrow(raw_theta_ploidy_df)) {
  announce_missing("No best-fit raw_theta_ploidy coefficients could be extracted from ranked optimisation runs.")
} else {
  audit_index_df <- bind_rows(audit_index_df, write_audit_artifact(raw_theta_ploidy_df, "raw_theta_ploidy_df"))
  preview_df <- raw_theta_ploidy_df %>%
    filter(source == "global_optim", !missing) %>%
    select(model_id, parameter, raw_theta_ploidy, lp__) %>%
    arrange(model_id, parameter)
  print(preview_table(preview_df, n = 24), row.names = FALSE)
  emit_artifact_note("raw_theta_ploidy_df")
}
```

## Leave-One-Line-Out `raw_theta_ploidy` Fits

These rows are the best `transfer` fits for each held-out line and direction, with the corresponding `raw_theta_ploidy` values and fit scores attached.

```{r transfer-raw-theta-ploidy-table}
if (!nrow(raw_theta_ploidy_transfer_df)) {
  announce_missing("No leave-one-line-out transfer best-fit raw_theta_ploidy coefficients found under %s", project_paths$transfer_optim_root)
} else {
  cat("Full transfer-fit coefficient table is not printed here because it is too large to read in the report.\n\n")
  emit_artifact_note("raw_theta_ploidy_transfer_df")
}
```

```{r transfer-raw-theta-ploidy-plot, fig.height=12, fig.width=14}
if (!nrow(raw_theta_ploidy_transfer_df)) {
  announce_missing("No leave-one-line-out transfer best-fit raw_theta_ploidy coefficients available for plotting.")
} else {
  transfer_plot_df <- raw_theta_ploidy_transfer_df %>%
    filter(fit_type == "transfer", parameter %in% common_raw_theta_parameters, !missing)

  if (!nrow(transfer_plot_df)) {
    announce_missing("No transfer-fit raw_theta_ploidy coefficients available for common parameters.")
  } else {
  global_plot_df <- raw_theta_ploidy_global_df %>%
    filter(parameter %in% common_raw_theta_parameters, !missing) %>%
    transmute(
      model_id,
      alias,
      line_id = NA_integer_,
      parameter,
      raw_theta_ploidy
    )

  transfer_plot_df <- transfer_plot_df %>%
    transmute(
      model_id,
      alias,
      line_id,
      line_label = label_line_ids(line_id, line_label_map, as_factor = FALSE),
      direction,
      parameter,
      raw_theta_ploidy
    )

  line_levels <- label_line_ids(sort(unique(transfer_plot_df$line_id)), line_label_map, as_factor = FALSE)

  global_repeated_df <- tidyr::expand_grid(
    global_plot_df,
    line_label = line_levels
  )

  transfer_with_global_df <- transfer_plot_df %>%
    left_join(
      global_plot_df %>%
        select(model_id, parameter, global_raw_theta_ploidy = raw_theta_ploidy),
      by = c("model_id", "parameter")
    )

  line_df <- transfer_with_global_df %>%
    filter(!is.na(global_raw_theta_ploidy)) %>%
    transmute(
      parameter,
      model_id,
      line_label,
      direction,
      x = model_id,
      y = global_raw_theta_ploidy,
      xend = model_id,
      yend = raw_theta_ploidy
    )

  plot_df <- bind_rows(
    global_repeated_df %>% mutate(point_type = "global", direction = "global"),
    transfer_plot_df %>% mutate(point_type = "transfer")
  ) %>%
    mutate(
      direction = factor(direction, levels = c("global", "low_to_high", "high_to_low")),
      model_id = factor(model_id, levels = ranking_model_levels(ranking_df)),
      line_label = factor(
        line_label,
        levels = line_levels
      ),
      parameter = factor(parameter, levels = common_raw_theta_parameters)
    )

  ggplot(
    plot_df,
    aes(x = model_id, y = raw_theta_ploidy)
  ) +
    geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
    geom_segment(
      data = line_df,
      aes(x = x, y = y, xend = xend, yend = yend, color = direction),
      inherit.aes = FALSE,
      linewidth = 0.5,
      alpha = 0.45
    ) +
    geom_point(
      data = plot_df %>% filter(direction == "global"),
      shape = 21,
      size = 2.8,
      stroke = 0.7,
      fill = "white",
      color = "black"
    ) +
    geom_point(
      data = plot_df %>% filter(direction != "global"),
      aes(color = direction),
      position = position_jitter(width = 0.08, height = 0),
      alpha = 0.8,
      size = 2.2
    ) +
    facet_grid(parameter ~ line_label, scales = "free_y") +
    scale_color_manual(values = c(low_to_high = "#1b9e77", high_to_low = "#d95f02")) +
    theme_bw() +
    labs(
      x = "",
      y = "Best-fit raw_theta_ploidy",
      color = "Direction"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )
  }
}
```

```{r transfer-raw-theta-ploidy-summary}
if (!nrow(raw_theta_ploidy_transfer_df)) {
  announce_missing("No leave-one-line-out transfer best-fit raw_theta_ploidy coefficients available for directional-agreement summary.")
} else {
  global_sign_df <- raw_theta_ploidy_global_df %>%
    filter(parameter %in% focus_common_parameters, !missing) %>%
    transmute(
      model_id,
      parameter = normalize_focus_parameter(parameter),
      global_raw_theta_ploidy = raw_theta_ploidy,
      global_sign = sign(raw_theta_ploidy)
    )

  agreement_df <- raw_theta_ploidy_transfer_df %>%
    filter(fit_type == "transfer", !missing) %>%
    mutate(parameter = normalize_focus_parameter(parameter)) %>%
    filter(parameter %in% focus_common_parameters) %>%
    left_join(global_sign_df, by = c("model_id", "parameter")) %>%
    mutate(
      transfer_sign = sign(raw_theta_ploidy),
      directional_agreement = !is.na(global_sign) & !is.na(transfer_sign) & (global_sign == transfer_sign)
    )

  by_parameter_df <- agreement_df %>%
    group_by(parameter) %>%
    summarise(
      n_cases = dplyr::n(),
      prop_directional_agreement = mean(directional_agreement, na.rm = TRUE),
      mean_abs_shift_from_global = mean(abs(raw_theta_ploidy - global_raw_theta_ploidy), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(match(parameter, focus_common_parameters))
  raw_theta_agreement_summary_df <- by_parameter_df
  raw_theta_agreement_model_df <- agreement_df %>%
    group_by(model_id) %>%
    summarise(
      n_cases = dplyr::n(),
      prop_directional_agreement = mean(directional_agreement, na.rm = TRUE),
      mean_abs_shift_from_global = mean(abs(raw_theta_ploidy - global_raw_theta_ploidy), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(prop_directional_agreement), mean_abs_shift_from_global, model_id)
  raw_theta_agreement_direction_df <- agreement_df %>%
    group_by(direction, parameter) %>%
    summarise(
      n_cases = dplyr::n(),
      prop_directional_agreement = mean(directional_agreement, na.rm = TRUE),
      mean_abs_shift_from_global = mean(abs(raw_theta_ploidy - global_raw_theta_ploidy), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(direction, match(parameter, focus_common_parameters))

  cat("Directional agreement of transfer `raw_theta_ploidy` with the global all-data fit\n\n")
  print(as.data.frame(by_parameter_df), row.names = FALSE)

  cat("\nAgreement summary by model\n\n")
  print(as.data.frame(raw_theta_agreement_model_df), row.names = FALSE)

  cat("\nAgreement summary by transfer direction and parameter\n\n")
  print(as.data.frame(raw_theta_agreement_direction_df), row.names = FALSE)
}
```

```{r transfer-raw-theta-ploidy-takeaway, results='asis'}
if (nrow(raw_theta_agreement_summary_df)) {
  strongest <- raw_theta_agreement_summary_df %>%
    arrange(desc(prop_directional_agreement), desc(n_cases)) %>%
    head(3)
  txt <- paste(sprintf(
    "%s: %.2f agreement",
    strongest$parameter,
    strongest$prop_directional_agreement
  ), collapse = "; ")
  knitr::asis_output(sprintf(
    "**Takeaway.** Across leave-one-line-out transfer fits, directional agreement with the global all-data `raw_theta_ploidy` estimate is strongest for %s.\n",
    txt
  ))
}
```

## Optimisation Stability Across Trusted Models

```{r optimisation-stability-all-models}
if (!nrow(stability_df)) {
  announce_missing("No trusted optimisation outputs available for stability summary.")
} else {
  print(as.data.frame(stability_df), row.names = FALSE)
}
```

```{r metric-glossary, include=FALSE}
print(as.data.frame(metric_glossary_df), row.names = FALSE)
```

## Transfer Analysis: Predictive View

This section treats transfer as a predictive benchmark. It is useful, but the parameter-transfer section below is now closer to the biological question of interest.

```{r transfer-screening-setup}
screen_transfer_df <- collect_transfer_screening(ranking_model_levels(ranking_df))
screen_parameter_df <- collect_parameter_screening(ranking_model_levels(ranking_df))
focused_parameter_df <- collect_focused_parameter_transfer(ranking_model_levels(ranking_df))
family_parameter_df <- collect_parameter_family_transfer(ranking_model_levels(ranking_df))
family_parameter_summary_df <- summarize_parameter_family_transfer(family_parameter_df)
model_family_summary_df <- summarize_parameter_family_across_models(family_parameter_summary_df)
family_ranking_df <- rank_parameter_families(model_family_summary_df)
derived_transfer_df <- collect_derived_transfer_summaries(ranking_model_levels(ranking_df))
derived_transfer_cmp_df <- build_derived_transfer_comparison(derived_transfer_df)
derived_transfer_summary_df <- summarize_derived_transfer(derived_transfer_cmp_df)
derived_transfer_ranking_df <- rank_derived_transfer(derived_transfer_summary_df)
transfer_subset_diag_df <- build_transfer_subset_diagnostics(ranking_model_levels(ranking_df))
transfer_subset_summary_df <- summarize_transfer_subset_diagnostics(transfer_subset_diag_df)

nuts_qc_summary_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_qc_summary.Rds"))
nuts_focus_draws_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_focus_draws.Rds"))
nuts_focus_summary_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_focus_summary.Rds"))
nuts_growth_model_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_growth_model_summary.Rds"))
nuts_growth_delta_surface_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_growth_delta_surface_summary.Rds"))
nuts_growth_pooled_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_growth_pooled_summary.Rds"))
nuts_growth_consistency_df <- safe_read_rds(file.path(project_paths$transfer_nuts_report_root, "nuts_growth_consistency_summary.Rds"))

if (is.null(nuts_qc_summary_df)) nuts_qc_summary_df <- tibble()
if (is.null(nuts_focus_draws_df)) nuts_focus_draws_df <- tibble()
if (is.null(nuts_focus_summary_df)) nuts_focus_summary_df <- tibble()
if (is.null(nuts_growth_model_df)) nuts_growth_model_df <- tibble()
if (is.null(nuts_growth_delta_surface_df)) nuts_growth_delta_surface_df <- tibble()
if (is.null(nuts_growth_pooled_df)) nuts_growth_pooled_df <- tibble()
if (is.null(nuts_growth_consistency_df)) nuts_growth_consistency_df <- tibble()

if (!nrow(nuts_focus_draws_df) || !all(focus_common_parameters %in% unique(nuts_focus_draws_df$parameter))) {
  source("R/nuts_consistency_utils.R")
  nuts_focus_draws_df <- collect_oracle_focus_parameter_draws(
    model_ids = ranking_model_levels(ranking_df),
    stan_data = readRDS(project_paths$stan_data),
    focus_parameters = focus_common_parameters,
    transfer_nuts_root = project_paths$transfer_nuts_root,
    max_draws_per_model = 400L
  )
}

if (!nrow(nuts_focus_summary_df) || !all(focus_common_parameters %in% unique(nuts_focus_summary_df$parameter))) {
  source("R/nuts_consistency_utils.R")
  nuts_focus_summary_df <- summarize_focus_parameter_posteriors(nuts_focus_draws_df)
}

predictive_screen_summary <- summarize_predictive_screening(screen_transfer_df)
parameter_screen_summary <- summarize_parameter_screening(screen_parameter_df)
model_screen_df <- summarize_model_screen(
  ranking_df = ranking_df,
  stability_df = stability_df,
  directional_agreement_model_df = raw_theta_agreement_model_df,
  screen_transfer_df = screen_transfer_df,
  screen_parameter_df = screen_parameter_df
)
shortlist_df <- build_aggregated_shortlist(model_screen_df)
selection_pool_df <- if (nrow(shortlist_df)) shortlist_df else model_screen_df
selected_model_id <- selection_pool_df %>%
  arrange(mean_rank, delta_AIC, desc(prop_within_1), desc(mean_parameter_log_error_improvement)) %>%
  slice_head(n = 1) %>%
  pull(model_id)
selected_model_summary_df <- model_screen_df %>%
  filter(model_id == selected_model_id) %>%
  slice_head(n = 1)
provenance_df <- build_current_provenance_table(
  ranking_df = ranking_df,
  selected_model_id = selected_model_id,
  project_paths = project_paths
)
audit_index_df <- bind_rows(
  audit_index_df,
  write_audit_artifact(screen_transfer_df, "screen_transfer_df"),
  write_audit_artifact(screen_parameter_df, "screen_parameter_df"),
  write_audit_artifact(focused_parameter_df, "focused_parameter_df"),
  write_audit_artifact(family_parameter_df, "family_parameter_df"),
  write_audit_artifact(family_parameter_summary_df, "family_parameter_summary_df"),
  write_audit_artifact(model_family_summary_df, "model_family_summary_df"),
  write_audit_artifact(model_screen_df, "model_screen_df"),
  write_audit_artifact(derived_transfer_df, "derived_transfer_df"),
  write_audit_artifact(derived_transfer_cmp_df, "derived_transfer_cmp_df"),
  write_audit_artifact(derived_transfer_summary_df, "derived_transfer_summary_df"),
  write_audit_artifact(nuts_qc_summary_df, "nuts_qc_summary_df"),
  write_audit_artifact(nuts_focus_draws_df, "nuts_focus_draws_df"),
  write_audit_artifact(nuts_focus_summary_df, "nuts_focus_summary_df"),
  write_audit_artifact(nuts_growth_model_df, "nuts_growth_model_df"),
  write_audit_artifact(nuts_growth_delta_surface_df, "nuts_growth_delta_surface_df"),
  write_audit_artifact(nuts_growth_pooled_df, "nuts_growth_pooled_df"),
  write_audit_artifact(nuts_growth_consistency_df, "nuts_growth_consistency_df"),
  write_audit_artifact(transfer_subset_diag_df, "transfer_subset_diag_df"),
  write_audit_artifact(transfer_subset_summary_df, "transfer_subset_summary_df")
)
```

## Aggregated Shortlist

```{r aggregated-shortlist}
if (!nrow(shortlist_df)) {
  announce_missing("No aggregated shortlist could be built.")
} else {
  print(as.data.frame(shortlist_df), row.names = FALSE)
}
```

```{r aggregated-shortlist-takeaway, results='asis'}
if (nrow(shortlist_df)) {
  shortlist_txt <- paste(shortlist_df$model_id, collapse = ", ")
  knitr::asis_output(sprintf(
    "**Shortlist.** The report now treats `%s` as the primary shortlist for downstream interpretation. These are the top 3 models by mean rank after discarding models with very poor optimisation stability (`prop_within_1 < 0.05`).\n",
    shortlist_txt
  ))
}
```

## Representative Model Scorecard

```{r selected-model-scorecard}
if (!nrow(selected_model_summary_df)) {
  announce_missing("No selected-model scorecard could be built.")
} else {
  print(as.data.frame(selected_model_summary_df), row.names = FALSE)
  emit_artifact_note("model_screen_df", label = "Audit CSV (all-model scorecard)")
}
```

```{r model-rank-comparison-plot, fig.height=7, fig.width=12}
if (!nrow(model_screen_df)) {
  announce_missing("No all-model scorecard available for cross-metric rank comparison.")
} else {
  rank_plot_df <- model_screen_df %>%
    select(
      model_id,
      mean_rank,
      rank_delta_AIC,
      rank_stability,
      rank_directional_agreement,
      rank_predictive_transfer,
      rank_parameter_transfer
    ) %>%
    arrange(mean_rank, rank_delta_AIC) %>%
    mutate(model_id = factor(model_id, levels = unique_preserve(model_id))) %>%
    pivot_longer(
      cols = starts_with("rank_"),
      names_to = "metric",
      values_to = "rank"
    ) %>%
    mutate(
      metric = recode(
        metric,
        rank_delta_AIC = "Delta AIC",
        rank_stability = "Optimisation stability",
        rank_directional_agreement = "Directional agreement",
        rank_predictive_transfer = "Predictive transfer",
        rank_parameter_transfer = "Parameter transfer"
      )
    )

  ggplot(rank_plot_df, aes(x = model_id, y = rank, color = metric, group = metric)) +
    geom_line(linewidth = 0.8, alpha = 0.8) +
    geom_point(size = 2) +
    scale_y_reverse(breaks = scales::pretty_breaks()) +
    theme_bw() +
    labs(
      x = "",
      y = "Rank (1 = best)",
      color = "Metric"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )
}
```

```{r selected-model-scorecard-takeaway, results='asis'}
if (nrow(selected_model_summary_df)) {
  knitr::asis_output(sprintf("**Representative model.** %s This representative model is used for single-model summaries and plots, but the shortlist above is the primary object for interpretation.\n", build_selected_model_takeaway(selected_model_summary_df)))
}
```

## Transfer Screening Across Models

```{r screening-plots, fig.height=10, fig.width=12}
if (!nrow(screen_transfer_df) || !nrow(screen_parameter_df)) {
  announce_missing("Need both predictive and parameter-transfer screening outputs to draw compact model-screening plots.")
} else {
  screen_parameter_df <- screen_parameter_df %>%
    mutate(
      mean_log_error_improvement = mean_abs_log_null_holdout_ratio - mean_abs_log_holdout_ratio,
      direction = factor(direction, levels = c("low_to_high", "high_to_low"))
    )

  p_pred <- ggplot(
    screen_transfer_df,
    aes(x = reorder(model_id, transfer_gain, FUN = mean), y = transfer_gain, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Predictive transfer gain over null")

  p_param_log <- ggplot(
    screen_parameter_df,
    aes(x = reorder(model_id, mean_log_error_improvement, FUN = mean), y = mean_log_error_improvement, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Mean log-error improvement over null")

  p_param_win <- ggplot(
    screen_parameter_df,
    aes(x = reorder(model_id, prop_parameters_transfer_better_than_null, FUN = mean), y = prop_parameters_transfer_better_than_null, color = direction)
  ) +
    geom_hline(yintercept = 0.5, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    labs(x = "", y = "Proportion of parameters transfer beats null")

  (p_pred | p_param_log) / p_param_win
  }
```

## Transfer Fit Diagnostics On Common Well Subsets

These diagnostics evaluate each best `null`, `transfer`, and `oracle` fit on the same transfer-defined well subsets, rather than comparing each fit on its own training set.

Interpretation:

- `transfer - null > 0` means the transferred fit scores better than the null fit on the same subset.
- `transfer - oracle < 0` means the transferred fit is still worse than the oracle on the same subset, which is expected if oracle is the best attainable benchmark.
- holdout-well panels are the main predictive check; training-well panels mostly indicate whether transfer is overfitting or underfitting relative to the other fit types.

```{r transfer-subset-diagnostics, fig.height=10, fig.width=12}
if (!nrow(transfer_subset_summary_df)) {
  announce_missing("No common-subset transfer diagnostics could be built from saved best-start fits.")
} else {
  diag_df <- transfer_subset_summary_df %>%
    mutate(direction = factor(direction, levels = c("low_to_high", "high_to_low")))

  p1 <- ggplot(
    diag_df,
    aes(x = reorder(model_id, transfer_minus_oracle_train, FUN = mean), y = transfer_minus_oracle_train, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Transfer - Oracle on transfer-training wells")

  p2 <- ggplot(
    diag_df,
    aes(x = reorder(model_id, transfer_minus_null_train, FUN = mean), y = transfer_minus_null_train, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Transfer - Null on transfer-training wells")

  p3 <- ggplot(
    diag_df,
    aes(x = reorder(model_id, transfer_minus_oracle_holdout, FUN = mean), y = transfer_minus_oracle_holdout, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Transfer - Oracle on transfer-holdout wells")

  p4 <- ggplot(
    diag_df,
    aes(x = reorder(model_id, transfer_minus_null_holdout, FUN = mean), y = transfer_minus_null_holdout, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Transfer - Null on transfer-holdout wells")

  (p1 | p2) / (p3 | p4)
}
```

```{r transfer-direction-asymmetry, fig.height=6, fig.width=10}
if (!nrow(transfer_subset_summary_df)) {
  announce_missing("No common-subset transfer diagnostics available for direction-asymmetry plot.")
} else {
  asym_df <- transfer_subset_summary_df %>%
    select(model_id, line_id, direction, transfer_minus_oracle_train, transfer_minus_null_train) %>%
    tidyr::pivot_wider(
      names_from = direction,
      values_from = c(transfer_minus_oracle_train, transfer_minus_null_train)
    ) %>%
    mutate(
      asym_transfer_minus_oracle_train =
        transfer_minus_oracle_train_low_to_high - transfer_minus_oracle_train_high_to_low,
      asym_transfer_minus_null_train =
        transfer_minus_null_train_low_to_high - transfer_minus_null_train_high_to_low
    )

  p1 <- ggplot(
    asym_df,
    aes(x = reorder(model_id, asym_transfer_minus_oracle_train, FUN = mean), y = asym_transfer_minus_oracle_train)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2, color = "#1b9e77") +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Low->High minus High->Low\nfor Transfer - Oracle on training wells")

  p2 <- ggplot(
    asym_df,
    aes(x = reorder(model_id, asym_transfer_minus_null_train, FUN = mean), y = asym_transfer_minus_null_train)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2, color = "#d95f02") +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Low->High minus High->Low\nfor Transfer - Null on training wells")

  p1 | p2
}
```

```{r transfer-optim-summary}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for transfer summary.")
} else {
  optim_transfer <- load_transfer_outputs(selected_model_id, output_root = project_paths$transfer_optim_root)

  if (is.null(optim_transfer$transfer_summary)) {
    announce_missing("No optimisation transfer summary found for %s", selected_model_id)
  } else {
    optim_transfer$transfer_summary %>%
      select(line_id, direction, transfer_gain, transfer_regret, transfer, null, oracle) %>%
      arrange(direction, desc(transfer_gain)) %>%
      print()
  }
}
```

```{r transfer-nuts-summary}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for NUTS transfer summary.")
} else {
  nuts_summary_path <- file.path(project_paths$transfer_nuts_root, selected_model_id, "transfer_comparison_summary.Rds")
  nuts_transfer <- safe_read_rds(nuts_summary_path)

  if (is.null(nuts_transfer)) {
    announce_missing("No NUTS transfer summary found at %s", nuts_summary_path)
  } else {
    cat("NUTS transfer summary\n")
    print(nuts_transfer)
  }
}
```

## NUTS Posterior Consistency

This section reads precomputed summaries generated from the oracle NUTS fits under `data/gpath_transfer_cv_nuts/gpath/<MODEL_ID>/low_to_high/oracle/`. The expected inputs live under `data/gpath_transfer_cv_nuts/report/` and can be refreshed with `scripts/agentRrunner.sh scripts/summarize_gpath_nuts_consistency.R`. The focus remains the common mechanistic parameters `ae[1]`, `ah[1]`, `Y_R[1]`, and `m`, plus the implied instantaneous net growth curves across the observed glucose conditions when auxiliary resource 2 is fixed at 1 and waste is set to 0.

The model-level Pareto-front interval and growth-curve figures are shown near the top of the report immediately after model selection.

```{r nuts-qc-summary}
if (!nrow(nuts_qc_summary_df)) {
  announce_missing("No precomputed oracle NUTS QC summary found under %s", project_paths$transfer_nuts_report_root)
} else {
  cat("Oracle NUTS QC summary across screened models\n\n")
  print(
    as.data.frame(
      nuts_qc_summary_df %>%
        arrange(ranking_model_order(model_id, ranking_df)) %>%
        select(-oracle_dir)
    ),
    row.names = FALSE
  )
  emit_artifact_note("nuts_qc_summary_df")
}
```

```{r nuts-focus-parameter-summary,echo=FALSE,eval=FALSE}
if (!nrow(nuts_focus_summary_df)) {
  announce_missing("No precomputed oracle NUTS posterior summaries found for the common parameters.")
} else {
  cat("Reconstructed posterior summaries for common parameters across screened models, cell lines, and ploidy states\n\n")
  print(
    as.data.frame(
      nuts_focus_summary_df %>%
        mutate(
          parameter = factor(parameter, levels = focus_common_parameters),
          ploidy_label = factor(ploidy_label, levels = c("low_ploidy", "high_ploidy"))
        ) %>%
        arrange(line_id, ploidy_label, parameter, ranking_model_order(model_id, ranking_df))
    ),
    row.names = FALSE
  )
  emit_artifact_note("nuts_focus_summary_df")
}
```

```{r nuts-focus-parameter-intervals, fig.height=8, fig.width=14, eval=FALSE, include=FALSE}
if (!nrow(nuts_focus_summary_df)) {
  announce_missing("No precomputed oracle NUTS posterior summaries available for interval plots.")
} else {
  plot_df <- nuts_focus_summary_df %>%
    mutate(
      model_id = factor(model_id, levels = ranking_model_levels(ranking_df, reverse = TRUE)),
      parameter = factor(parameter, levels = focus_common_parameters),
      line_label = label_line_ids(line_id, line_label_map),
      ploidy_label = factor(ploidy_label, levels = c("low_ploidy", "high_ploidy"), labels = c("Low ploidy", "High ploidy"))
    )

  ggplot(plot_df, aes(y = model_id, color = ploidy_label)) +
    geom_segment(aes(x = q025, xend = q975, yend = model_id), linewidth = 0.7, position = position_dodge(width = 0.6)) +
    geom_segment(aes(x = q25, xend = q75, yend = model_id), linewidth = 1.4, position = position_dodge(width = 0.6)) +
    geom_point(aes(x = median), size = 2.0, position = position_dodge(width = 0.6)) +
    facet_grid(line_label ~ parameter, scales = "free_x", drop = FALSE) +
    scale_color_manual(values = c("Low ploidy" = "#1b9e77", "High ploidy" = "#d95f02")) +
    theme_bw() +
    scale_x_log10()+
    labs(x = "Posterior value", y = "Model ID", color = "")
}
```

```{r nuts-growth-consistency-table,echo=FALSE,eval=FALSE}
if (!nrow(nuts_growth_consistency_df)) {
  announce_missing("No precomputed oracle NUTS growth-curve summaries found.")
} else {
  growth_table_df <- nuts_growth_consistency_df %>%
    mutate(
      line_id = factor(line_id, levels = sort(unique(line_id))),
      ploidy_label = factor(ploidy_label, levels = c("low_ploidy", "high_ploidy"))
    ) %>%
    arrange(line_id, glucose, ploidy_label)

  cat("Cross-model consistency of oracle NUTS instantaneous net-growth curves\n\n")
  print(as.data.frame(growth_table_df), row.names = FALSE)
  emit_artifact_note("nuts_growth_consistency_df")
}
```

```{r nuts-growth-curves-by-model, fig.height=10, fig.width=16, eval=FALSE, include=FALSE}
if (!nrow(nuts_growth_model_df)) {
  announce_missing("No precomputed oracle NUTS growth-curve summaries available for model-level plotting.")
} else {
  plot_df <- nuts_growth_model_df %>%
    mutate(
      model_id = factor(model_id, levels = ranking_model_levels(ranking_df)),
      line_label = label_line_ids(line_id, line_label_map),
      ploidy_label = factor(ploidy_label, levels = c("low_ploidy", "high_ploidy"), labels = c("Low ploidy", "High ploidy"))
    )

  ggplot(plot_df, aes(x = glucose, y = median_growth, color = ploidy_label, fill = ploidy_label)) +
    geom_ribbon(aes(ymin = q025_growth, ymax = q975_growth), alpha = 0.18, color = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.3) +
    facet_grid(line_label ~ model_id, scales = "free_y") +
    scale_color_manual(values = c("Low ploidy" = "#1b9e77", "High ploidy" = "#d95f02")) +
    scale_fill_manual(values = c("Low ploidy" = "#1b9e77", "High ploidy" = "#d95f02")) +
    theme_bw() +
    scale_x_log10()+
    labs(
      x = "Initial glucose",
      y = "Instantaneous net growth",
      color = "",
      fill = ""
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank(),
      strip.background = element_rect(fill = "grey95")
    )
}
```

```{r nuts-growth-curves-pooled, fig.height=8, fig.width=12}
if (!nrow(nuts_growth_model_df) || !nrow(nuts_growth_pooled_df)) {
  announce_missing("No precomputed pooled oracle NUTS growth-curve summaries available for consistency plotting.")
} else {
  model_plot_df <- nuts_growth_model_df %>%
    mutate(
      line_label = label_line_ids(line_id, line_label_map),
      ploidy_label = factor(ploidy_label, levels = c("low_ploidy", "high_ploidy"), labels = c("Low ploidy", "High ploidy"))
    )
  pooled_plot_df <- nuts_growth_pooled_df %>%
    mutate(
      line_label = label_line_ids(line_id, line_label_map),
      ploidy_label = factor(ploidy_label, levels = c("low_ploidy", "high_ploidy"), labels = c("Low ploidy", "High ploidy"))
    )

  ggplot() +
    geom_line(
      data = model_plot_df,
      aes(x = glucose, y = median_growth, group = interaction(model_id, ploidy_label), color = ploidy_label),
      alpha = 0.22,
      linewidth = 0.6
    ) +
    geom_ribbon(
      data = pooled_plot_df,
      aes(x = glucose, ymin = pooled_q025_growth, ymax = pooled_q975_growth, fill = ploidy_label),
      alpha = 0.18,
      color = NA
    ) +
    geom_line(
      data = pooled_plot_df,
      aes(x = glucose, y = pooled_median_growth, color = ploidy_label),
      linewidth = 1.1
    ) +
    geom_point(
      data = pooled_plot_df,
      aes(x = glucose, y = pooled_median_growth, color = ploidy_label),
      size = 1.6
    ) +
    facet_wrap(~ line_label, scales = "free_y") +
    scale_color_manual(values = c("Low ploidy" = "#1b9e77", "High ploidy" = "#d95f02")) +
    scale_fill_manual(values = c("Low ploidy" = "#1b9e77", "High ploidy" = "#d95f02")) +
    theme_bw() +
    scale_x_log10()+
    labs(
      x = "Initial glucose",
      y = "Instantaneous net growth",
      color = "",
      fill = ""
    ) +
    theme(panel.grid.minor = element_blank())
}
```

```{r nuts-growth-takeaway, results='asis'}
if (nrow(nuts_growth_consistency_df)) {
  top_consistency <- nuts_growth_consistency_df %>%
    group_by(line_id, ploidy_label) %>%
    summarise(
      mean_between_model_sd = mean(between_model_sd, na.rm = TRUE),
      min_sign_agreement = min(ploidy_sign_agreement, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(mean_between_model_sd, desc(min_sign_agreement))

  txt <- paste(sprintf(
    "line %d %s: mean between-model SD %.3f, minimum ploidy sign agreement %.2f",
    top_consistency$line_id,
    gsub("_", " ", top_consistency$ploidy_label),
    top_consistency$mean_between_model_sd,
    top_consistency$min_sign_agreement
  ), collapse = "; ")

  knitr::asis_output(sprintf(
    "**NUTS consistency takeaway.** Across the screened models with oracle NUTS fits, the common-parameter posteriors and the implied instantaneous net-growth curves can now be compared directly. The growth-curve consistency summary is: %s.\n",
    txt
  ))
}
```

## Focused Common Parameters

These plots break out the four parameters that are present across the ranked model set: `ae[1]`, `ah[1]`, `Y_R[1]`, and `m`. The ratio view uses signed log-ratios to the oracle on the reconstructed held-out parameters, while the hit summaries below retain the same absolute-log-ratio win criteria used elsewhere in the notebook.


```{r focused-parameter-table}
if (!nrow(focused_parameter_df)) {
  announce_missing("No focused parameter-transfer comparison rows found for the common parameters.")
} else {
  parameter_summary_df <- focused_parameter_df %>%
    filter(!is.na(abs_log_null_ratio), !is.na(abs_log_transfer_ratio)) %>%
    group_by(parameter, direction) %>%
    summarise(
      n_cases = dplyr::n(),
      mean_ratio_error_improvement = mean(ratio_error_improvement, na.rm = TRUE),
      median_ratio_error_improvement = median(ratio_error_improvement, na.rm = TRUE),
      prop_transfer_better_than_null = mean(transfer_beats_null_ratio, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(parameter, direction)
  model_parameter_summary_df <- focused_parameter_df %>%
    filter(!is.na(abs_log_null_ratio), !is.na(abs_log_transfer_ratio)) %>%
    group_by(model_id, parameter) %>%
    summarise(
      mean_ratio_error_improvement = mean(ratio_error_improvement, na.rm = TRUE),
      prop_transfer_better_than_null = mean(transfer_beats_null_ratio, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(parameter, desc(mean_ratio_error_improvement), desc(prop_transfer_better_than_null), model_id)

  cat("Across-model summary by parameter and transfer direction\n\n")
  print(as.data.frame(parameter_summary_df), row.names = FALSE)
  cat("\nModel-level summary by parameter\n\n")
  print(as.data.frame(model_parameter_summary_df), row.names = FALSE)
  emit_artifact_note("focused_parameter_df")
}
```

```{r focused-parameter-ratio, fig.height=10, fig.width=14}
if (!nrow(focused_parameter_df %>% filter(!is.na(abs_log_null_ratio), !is.na(abs_log_transfer_ratio)))) {
  announce_missing("No positive-valued focused parameter rows available for ratio-based plotting.")
} else {
  plot_df <- focused_parameter_df %>%
    filter(!is.na(abs_log_null_ratio), !is.na(abs_log_transfer_ratio)) %>%
    mutate(
      parameter = factor(parameter, levels = focus_common_parameters),
      direction = factor(direction, levels = c("low_to_high", "high_to_low")),
      model_id = factor(model_id, levels = ranking_model_levels(ranking_df))
    )

  p1 <- ggplot(
    plot_df,
    aes(x = model_id)
  ) +
    geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
    geom_segment(
      aes(
        xend = model_id,
        y = log_null_ratio,
        yend = log_transfer_ratio,
        color = direction
      ),
      linewidth = 0.5,
      alpha = 0.45
    ) +
    geom_point(aes(y = log_null_ratio), shape = 21, size = 2.4, stroke = 0.7, fill = "white", color = "black") +
    geom_point(aes(y = log_transfer_ratio, color = direction), size = 2.2, alpha = 0.85) +
    facet_grid(parameter ~ line_direction, scales = "free_y") +
    scale_color_manual(values = c(low_to_high = "#1b9e77", high_to_low = "#d95f02")) +
    theme_bw() +
    labs(
      x = "",
      y = "log(fit / oracle) on reconstructed held-out parameter",
      color = "Direction"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )

  p2 <- ggplot(
    plot_df,
    aes(x = model_id, y = ratio_error_improvement, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.08, height = 0), size = 2.1, alpha = 0.85) +
    facet_grid(parameter ~ line_direction, scales = "free_y") +
    scale_color_manual(values = c(low_to_high = "#1b9e77", high_to_low = "#d95f02")) +
    theme_bw() +
    labs(
      x = "",
      y = "|log(null/oracle)| - |log(transfer/oracle)|",
      color = "Direction"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )

  p1 / p2
}
```

```{r focused-parameter-hit-summary, fig.height=8, fig.width=14}
if (!nrow(focused_parameter_df %>% filter(!is.na(abs_log_null_ratio), !is.na(abs_log_transfer_ratio)))) {
  announce_missing("No positive-valued focused parameter rows available for hit-summary plotting.")
} else {
  hit_df <- focused_parameter_df %>%
    filter(!is.na(abs_log_null_ratio), !is.na(abs_log_transfer_ratio)) %>%
    group_by(model_id, direction, parameter) %>%
    summarise(
      mean_log_error_improvement = mean(ratio_error_improvement, na.rm = TRUE),
      median_log_error_improvement = median(ratio_error_improvement, na.rm = TRUE),
      prop_transfer_better_than_null = mean(transfer_beats_null_ratio, na.rm = TRUE),
      n_cases = dplyr::n(),
      .groups = "drop"
    ) %>%
    mutate(
      parameter = factor(parameter, levels = focus_common_parameters),
      direction = factor(direction, levels = c("low_to_high", "high_to_low")),
      model_id = factor(model_id, levels = ranking_model_levels(ranking_df))
    )
  focused_hit_df <- hit_df
  audit_index_df <- bind_rows(audit_index_df, write_audit_artifact(hit_df, "focused_parameter_hit_df"))

  p1 <- ggplot(
    hit_df,
    aes(x = model_id, y = mean_log_error_improvement, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.08, height = 0), size = 2.2, alpha = 0.85) +
    facet_wrap(~ parameter, scales = "free_y", ncol = 2) +
    scale_color_manual(values = c(low_to_high = "#1b9e77", high_to_low = "#d95f02")) +
    theme_bw() +
    labs(
      x = "",
      y = "Mean log-error improvement over null",
      color = "Direction"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )

  p2 <- ggplot(
    hit_df,
    aes(x = model_id, y = prop_transfer_better_than_null, color = direction)
  ) +
    geom_hline(yintercept = 0.5, color = "grey75", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.08, height = 0), size = 2.2, alpha = 0.85) +
    facet_wrap(~ parameter, ncol = 2) +
    scale_color_manual(values = c(low_to_high = "#1b9e77", high_to_low = "#d95f02")) +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    labs(
      x = "",
      y = "Proportion transfer beats null",
      color = "Direction"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )

  p1 / p2
  emit_artifact_note("focused_parameter_hit_df")
}
```

```{r focused-parameter-takeaway, results='asis'}
if (nrow(focused_hit_df)) {
  top_hits <- focused_hit_df %>%
    group_by(parameter) %>%
    summarise(
      mean_log_error_improvement = mean(mean_log_error_improvement, na.rm = TRUE),
      mean_win_rate = mean(prop_transfer_better_than_null, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_log_error_improvement), desc(mean_win_rate))
  txt <- paste(sprintf(
    "%s (mean log-error improvement %.3f, win rate %.2f)",
    top_hits$parameter,
    top_hits$mean_log_error_improvement,
    top_hits$mean_win_rate
  ), collapse = "; ")
  knitr::asis_output(sprintf(
    "**Takeaway.** On reconstructed held-out parameter values, transfer remains much weaker than the raw-coefficient directional agreement: %s.\n",
    txt
  ))
}
```

```{r audit-artifact-index-final}
print(
  as.data.frame(audit_index_df %>% distinct(artifact, .keep_all = TRUE) %>% arrange(artifact)),
  row.names = FALSE
)
```

```{r agent-text-export, include=FALSE}
agent_text_path <- write_agent_text_report(
  ranking_df = ranking_df,
  support_set_df = if (exists("support_set_df", inherits = TRUE)) support_set_df else tibble(),
  shortlist_df = if (exists("shortlist_df", inherits = TRUE)) shortlist_df else tibble(),
  provenance_df = if (exists("provenance_df", inherits = TRUE)) provenance_df else tibble(),
  metric_glossary_df = if (exists("metric_glossary_df", inherits = TRUE)) metric_glossary_df else tibble(),
  selected_model_summary_df = if (exists("selected_model_summary_df", inherits = TRUE)) selected_model_summary_df else tibble(),
  raw_theta_agreement_df = raw_theta_agreement_summary_df,
  focused_hit_df = focused_hit_df,
  report_export_root = report_export_root
)
```

## Agent Text Artifact

This notebook also writes a lightweight text companion for local agents.

```{r agent-text-artifact, results='asis'}
knitr::asis_output(sprintf("Tracked text artifact: `%s`\n", normalizePath(project_paths$agent_text_path, winslash = "/", mustWork = FALSE)))
```

## Cross-Model Family Transfer

This section asks a narrower question than overall model transfer: whether particular parameter families repeatedly move closer to the oracle than the null baseline does across models, lines, and transfer directions.

```{r parameter-family-ranking, fig.height=8, fig.width=12}
if (!nrow(model_family_summary_df)) {
  announce_missing("No parameter-comparison outputs found for family-level transfer analysis.")
} else {
  plot_df <- family_ranking_df %>%
    mutate(
      direction = factor(direction, levels = c("low_to_high", "high_to_low")),
      family = reorder(family, mean_log_error_improvement)
    )

  p1 <- ggplot(plot_df, aes(x = family, y = mean_log_error_improvement, color = direction)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_point(size = 3) +
    coord_flip() +
    facet_wrap(~ direction, scales = "free_y") +
    theme_bw() +
    labs(x = "", y = "Cross-model mean log-error improvement")

  p2 <- ggplot(plot_df, aes(x = family, y = mean_win_rate, color = direction)) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey70") +
    geom_point(size = 3) +
    coord_flip() +
    facet_wrap(~ direction, scales = "free_y") +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    labs(x = "", y = "Cross-model mean family win rate")

  p1 | p2
}
```

```{r parameter-family-consistency-heatmaps, fig.height=10, fig.width=12}
if (!nrow(model_family_summary_df)) {
  announce_missing("No parameter-comparison outputs found for family-level heatmaps.")
} else {
  heat_df <- model_family_summary_df %>%
    mutate(direction = factor(direction, levels = c("low_to_high", "high_to_low")))

  p1 <- ggplot(heat_df, aes(x = family, y = model_id, fill = mean_family_log_error_improvement)) +
    geom_tile(color = "white") +
    facet_wrap(~ direction) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0) +
    theme_minimal() +
    labs(
      x = "Parameter family",
      y = "Model ID",
      fill = "Mean log-error\nimprovement"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p2 <- ggplot(heat_df, aes(x = family, y = model_id, fill = mean_family_win_rate)) +
    geom_tile(color = "white") +
    facet_wrap(~ direction) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0.5, limits = c(0, 1)) +
    theme_minimal() +
    labs(
      x = "Parameter family",
      y = "Model ID",
      fill = "Mean family\nwin rate"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p1 / p2
}
```

```{r parameter-family-line-heatmaps, fig.height=10, fig.width=12}
if (!nrow(family_parameter_summary_df)) {
  announce_missing("No parameter-comparison outputs found for line-level family heatmaps.")
} else {
  top_families <- family_ranking_df %>%
    group_by(family) %>%
    summarise(
      mean_log_error_improvement = mean(mean_log_error_improvement, na.rm = TRUE),
      mean_win_rate = mean(mean_win_rate, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_log_error_improvement), desc(mean_win_rate)) %>%
    slice_head(n = 4) %>%
    pull(family)

  line_heat_df <- family_parameter_summary_df %>%
    filter(family %in% top_families) %>%
    mutate(
      line_direction = sprintf("%s | %s", label_line_ids(line_id, line_label_map, as_factor = FALSE), direction),
      family = factor(family, levels = top_families)
    )

  p1 <- ggplot(line_heat_df, aes(x = family, y = line_direction, fill = mean_log_error_improvement)) +
    geom_tile(color = "white") +
    facet_wrap(~ model_id) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0) +
    theme_minimal() +
    labs(
      x = "Top parameter families",
      y = "",
      fill = "Mean log-error\nimprovement"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p2 <- ggplot(line_heat_df, aes(x = family, y = line_direction, fill = prop_transfer_better_than_null)) +
    geom_tile(color = "white") +
    facet_wrap(~ model_id) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0.5, limits = c(0, 1)) +
    theme_minimal() +
    labs(
      x = "Top parameter families",
      y = "",
      fill = "Family\nwin rate"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p1 / p2
}
```

## Cross-Model Derived Glucose Summaries

These plots summarize five glucose-focused derived quantities reconstructed from the parameter states:
`Umax_1 = ae1`, `Khalf_1 = ah1`, `Efflow_1 = ae1 / ah1`, `YieldCap_1 = Y_R1 * ae1`, and `NetGrow_1 = Y_R1 * ae1 - m`.

```{r derived-transfer-ranking, fig.height=8, fig.width=12}
if (!nrow(derived_transfer_summary_df)) {
  announce_missing("No derived transfer summaries could be built from parameter-transfer states.")
} else {
  plot_df <- derived_transfer_ranking_df %>%
    mutate(direction = factor(direction, levels = c("low_to_high", "high_to_low")))

  p1 <- ggplot(plot_df, aes(x = reorder(summary_name, mean_log_error_improvement), y = mean_log_error_improvement, color = direction)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_point(size = 3) +
    coord_flip() +
    facet_wrap(~ direction, scales = "free_y") +
    theme_bw() +
    labs(x = "", y = "Cross-model mean log-error improvement")

  p2 <- ggplot(plot_df, aes(x = reorder(summary_name, mean_win_rate), y = mean_win_rate, color = direction)) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey70") +
    geom_point(size = 3) +
    coord_flip() +
    facet_wrap(~ direction, scales = "free_y") +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    labs(x = "", y = "Cross-model mean win rate")

  p1 | p2
}
```

```{r derived-transfer-heatmaps, fig.height=10, fig.width=12}
if (!nrow(derived_transfer_summary_df)) {
  announce_missing("No derived transfer summaries available for cross-model heatmaps.")
} else {
  heat_df <- derived_transfer_summary_df %>%
    mutate(direction = factor(direction, levels = c("low_to_high", "high_to_low")))

  p1 <- ggplot(heat_df, aes(x = summary_name, y = model_id, fill = mean_log_error_improvement)) +
    geom_tile(color = "white") +
    facet_wrap(~ direction) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0) +
    theme_minimal() +
    labs(
      x = "Derived glucose summary",
      y = "Model ID",
      fill = "Mean log-error\nimprovement"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p2 <- ggplot(heat_df, aes(x = summary_name, y = model_id, fill = mean_win_rate)) +
    geom_tile(color = "white") +
    facet_wrap(~ direction) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0.5, limits = c(0, 1)) +
    theme_minimal() +
    labs(
      x = "Derived glucose summary",
      y = "Model ID",
      fill = "Mean\nwin rate"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p1 / p2
}
```

```{r derived-transfer-selected-model, fig.height=9, fig.width=10}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for derived transfer plots.")
} else {
  selected_df <- derived_transfer_cmp_df %>%
    filter(model_id == selected_model_id) %>%
    mutate(direction = factor(direction, levels = c("low_to_high", "high_to_low")))

  if (!nrow(selected_df)) {
    announce_missing("No derived transfer comparison rows found for %s", selected_model_id)
  } else {
    p1 <- ggplot(selected_df, aes(x = summary_name, y = log_error_improvement, color = direction)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
      geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
      stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
      facet_wrap(~ line_id) +
      theme_bw() +
      labs(x = "Derived glucose summary", y = "Line-level log-error improvement") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

    p2 <- ggplot(selected_df, aes(x = summary_name, y = as.numeric(transfer_better_than_null), color = direction)) +
      geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey70") +
      geom_point(position = position_jitter(width = 0.15, height = 0.03), alpha = 0.8, size = 2) +
      stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
      facet_wrap(~ line_id) +
      coord_cartesian(ylim = c(0, 1)) +
      theme_bw() +
      labs(x = "Derived glucose summary", y = "Transfer beats null") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

    p1 / p2
  }
}
```

## Transfer Analysis: Parameter View

This is currently the biologically primary transfer analysis. The null baseline is the observed-state parameterisation carried forward unchanged to the held-out ploidy state.

```{r parameter-transfer-summary}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer summary.")
} else {
  optim_transfer <- load_transfer_outputs(selected_model_id, output_root = project_paths$transfer_optim_root)

  if (is.null(optim_transfer$parameter_summary)) {
    announce_missing("No parameter transfer summary found for %s", selected_model_id)
  } else {
    optim_transfer$parameter_summary %>%
      mutate(
        mean_log_error_improvement = mean_abs_log_null_holdout_ratio - mean_abs_log_holdout_ratio
      ) %>%
      select(
        line_id,
        direction,
        mean_log_error_improvement,
        prop_parameters_transfer_better_than_null,
        mean_abs_log_holdout_ratio,
        mean_abs_log_null_holdout_ratio
      ) %>%
      arrange(direction, desc(mean_log_error_improvement)) %>%
      print()
  }
}
```

```{r parameter-transfer-plot-1, fig.height=6, fig.width=10}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer plots.")
} else {
  comparison_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_comparison.Rds")
  if (!file.exists(comparison_path)) {
    announce_missing("No parameter transfer comparison found at %s", comparison_path)
  } else {
    plot_parameter_transfer_improvement(
      model_id = selected_model_id,
      output_root = "data/gpath_transfer_cv",
      model_name = "gpath"
    )
  }
}
```

```{r parameter-transfer-plot-2, fig.height=8, fig.width=10}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer plots.")
} else {
    summary_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_summary.Rds")
    if (!file.exists(summary_path)) {
      announce_missing("No parameter transfer summary found at %s", summary_path)
    } else {
      plot_parameter_transfer_summary(
        model_id = selected_model_id,
        output_root = "data/gpath_transfer_cv",
        model_name = "gpath"
      )
    }
}
```

```{r parameter-transfer-plot-3, fig.height=9, fig.width=10}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer plots.")
} else {
    comparison_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_comparison.Rds")
    if (!file.exists(comparison_path)) {
      announce_missing("No parameter transfer comparison found at %s", comparison_path)
    } else {
      plot_parameter_transfer_family_summary(
        model_id = selected_model_id,
        output_root = "data/gpath_transfer_cv",
        model_name = "gpath"
      )
    }
}
```

## Transfer Fits For A Representative Line

This section overlays the best optimisation `null`, `transfer`, and `oracle` fits for one selected line. A line is chosen from the parameter-transfer summary by maximal mean log-error improvement over null; if that summary is absent, line 1 is used.

```{r transfer-overlay-plot, fig.height=6, fig.width=12}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for transfer overlay plot.")
} else {
  parameter_summary_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_summary.Rds")
  parameter_summary_df <- safe_read_rds(parameter_summary_path)

  target_line <- 1L
  if (!is.null(parameter_summary_df) && nrow(parameter_summary_df)) {
    score <- parameter_summary_df$mean_abs_log_null_holdout_ratio - parameter_summary_df$mean_abs_log_holdout_ratio
    target_line <- parameter_summary_df$line_id[which.max(score)]
  }

  best_start_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "transfer_best_start_summary.Rds")
  if (!file.exists(best_start_path)) {
    announce_missing("No transfer best-start summary found at %s", best_start_path)
  } else {
    overlay_data <- generate_transfer_overlay_data(
      model_id = selected_model_id,
      line_id = target_line,
      output_root = "data/gpath_transfer_cv",
      stan_data_path = project_paths$stan_data
    )

    plot_transfer_line_trajectories(
      transfer_data = overlay_data,
      line_id = target_line,
      model_id = selected_model_id
    )
  }
}
```

## Current Interpretation

Interpretation here is based on two distinct questions:

- predictive transfer: does the transferred ploidy effect improve held-out predictive fit relative to the null?
- parameter transfer: does the transferred ploidy effect move reconstructed held-out parameters closer to the oracle than the null does?

```{r current-interpretation-gloss}
metric_glossary_df %>%
  filter(metric %in% c("transfer_gain", "transfer_regret", "mean_log_error_improvement", "prop_parameters_transfer_better_than_null")) %>%
  print(row.names = FALSE)
```

```{r interpretation-table}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for interpretation summary.")
} else {
  optim_transfer <- load_transfer_outputs(selected_model_id, output_root = project_paths$transfer_optim_root)

  if (is.null(optim_transfer$parameter_summary) || is.null(optim_transfer$transfer_summary)) {
    announce_missing("Need both predictive and parameter-transfer summaries to build the interpretation table.")
  } else {
    interpretation_df <- optim_transfer$parameter_summary %>%
      mutate(
        mean_log_error_improvement = mean_abs_log_null_holdout_ratio - mean_abs_log_holdout_ratio
      ) %>%
      left_join(
        optim_transfer$transfer_summary %>% rename(model_direction = direction),
        by = c("line_id" = "line_id", "direction" = "model_direction")
      ) %>%
      select(
        line_id,
        direction,
        transfer,
        null,
        oracle,
        transfer_gain,
        transfer_regret,
        mean_log_error_improvement,
        prop_parameters_transfer_better_than_null,
        mean_abs_log_holdout_ratio,
        mean_abs_log_null_holdout_ratio
      )

    print(interpretation_df)
  }
}
```

```{r interpretation-takeaway, results='asis'}
if (nrow(selected_model_summary_df) || nrow(shortlist_df)) {
  shortlist_note <- if (nrow(shortlist_df)) {
    paste(shortlist_df$model_id, collapse = ", ")
  } else {
    "no aggregated shortlist"
  }
  shortlist_verdict <- if (nrow(shortlist_df)) {
    pos_pred <- shortlist_df %>% filter(!is.na(mean_transfer_gain), mean_transfer_gain > 0)
    pos_param <- shortlist_df %>% filter(!is.na(mean_parameter_log_error_improvement), mean_parameter_log_error_improvement > 0)
    if (!nrow(pos_pred) && !nrow(pos_param)) {
      "Across the shortlist, there is no model that is positive on average for either predictive transfer or parameter transfer."
    } else if (nrow(pos_pred) && !nrow(pos_param)) {
      "Across the shortlist, some models improve predictive transfer, but none show positive average parameter-transfer improvement."
    } else if (!nrow(pos_pred) && nrow(pos_param)) {
      "Across the shortlist, some models improve parameter transfer, but none show positive average predictive transfer gain."
    } else {
      "Across the shortlist, at least some models are positive on both predictive and parameter-transfer summaries."
    }
  } else {
    "No shortlist-level verdict could be formed."
  }
  knitr::asis_output(sprintf(
    paste(
      "**Interpretation.**",
      "Primary interpretation should now be based on the aggregated shortlist rather than a single winner.",
      sprintf("The current shortlist is: %s.", shortlist_note),
      shortlist_verdict,
      if (nrow(selected_model_summary_df)) sprintf("For single-model views, the representative model is `%s`.", selected_model_summary_df$model_id[[1]]) else "",
      "The predictive summaries remain a benchmark, but parameter transfer is still the more direct scientific readout."
    )
  ))
}
```

## Notes And Next Steps

At the time of writing, the modelling workflow supports four distinct questions:

1. Which `gpath` structure is best supported by the hierarchical optimisation fits?
2. Are selected models numerically stable enough to interpret?
3. Does a shared ploidy effect improve held-out prediction in directional transfer tests?
4. Does that same shared ploidy effect move held-out-line parameter values closer to the oracle fit than a null baseline does?

Recommended next steps:

1. Treat the selected `gpath` model as primary and stop comparing many model IDs throughout the downstream biological analysis.
2. Use the predictive transfer summaries as a benchmark, but make the parameter-transfer section the primary scientific result.
3. The current `oracle` transfer fits rerun the same all-data objective that the main global optimisation already solves when the model spec and ploidy mask are unchanged. That is redundant work and makes the workflow harder to follow; downstream comparisons should either reuse the global fit directly or state clearly why a separate all-data rerun is needed.
4. Add one compact comparison of optimisation-based versus NUTS-based transfer for the same selected model to quantify how much posterior averaging changes the conclusion.
5. If the parameter-transfer summaries are still hard to interpret, lean on the focused common-parameter plots and relative-error summaries before moving up to family-level averages.
6. Keep legacy or exploratory analysis blocks outside this notebook unless they are directly needed for the current `gpath` narrative.

## Appendix: Commands And Provenance

This notebook depends on artefacts created elsewhere. The table below documents the current known generation path.

```{r upstream-commands-appendix}
print(as.data.frame(artifact_commands), row.names = FALSE)
```

This section records the concrete current artifact paths used by this render, separate from the generic rerun commands above.

```{r current-provenance-appendix}
print(as.data.frame(provenance_df), row.names = FALSE)
```

## Appendix: Audit And Setup

Full audit tables are written during knitting under `data/report_exports/model_analysis/`. The report shows only compact previews.

```{r audit-artifact-index-appendix}
print(as.data.frame(audit_index_df), row.names = FALSE)
```

```{r trusted-optimisation-runs-appendix}
if (!nrow(manifest_df)) {
  announce_missing("No manifest-backed optimisation outputs found under %s", project_paths$optim_root)
} else {
  print(as.data.frame(manifest_df), row.names = FALSE)
}
```

```{r metric-glossary-appendix}
print(as.data.frame(metric_glossary_df), row.names = FALSE)
```
