---
title: "Ecological Modelling"
author: "RJB"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook is now the reporting layer for the current `gpath` modelling workflow. It is not intended to recreate the full analysis from scratch. Instead it:

- documents the current modelling path,
- states the upstream scripts/commands that generate required artefacts where known,
- loads saved results from the current workflow,
- presents model selection, diagnostics, transfer results, and parameter-transfer summaries.

Where an upstream command is not known exactly from the repo history, that is stated explicitly.

## Setup

```{r libraries}
library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(tibble)
library(posterior)
library(patchwork)

source("R/project_paths.R")
source(get_model_r_path("gpath", "v1"))
source("R/gpath_run_utils.R")
source("R/posterior_io_utils.R")
source("R/elpd_transfer_utils.R")
source("R/parameter_transfer_utils.R")
source("R/parameter_transfer_plot_utils.R")
source("R/transfer_cv_plot_utils.R")

project_paths <- list(
  stan_data = resolve_stan_data_path(),
  legacy_model_root = file.path("data", "runs", "gpath", "v1", "optim"),
  transfer_optim_root = "data/gpath_transfer_cv/gpath",
  transfer_nuts_root = "data/gpath_transfer_cv_nuts/gpath"
)

cat("Current working directory:", getwd(), "\n")
cat("stan_data path:", project_paths$stan_data, "\n")
```

```{r helpers}
safe_read_rds <- function(path) {
  if (!file.exists(path)) {
    return(NULL)
  }
  readRDS(path)
}

announce_missing <- function(...) {
  cat("SKIP:", sprintf(...), "\n")
}

discover_model_ids <- function(base_path) {
  if (!dir.exists(base_path)) {
    return(character(0))
  }
  run_dirs <- list.dirs(base_path, recursive = TRUE, full.names = TRUE)
  run_dirs[file.exists(file.path(run_dirs, "optim_lp_all.Rds"))]
}

build_model_ranking <- function(base_path = project_paths$legacy_model_root, stan_data_path = project_paths$stan_data) {
  run_dirs <- discover_model_ids(base_path)
  if (!length(run_dirs)) {
    return(tibble())
  }

  stan_data <- readRDS(stan_data_path)

  rows <- lapply(run_dirs, function(run_dir) {
    parts <- strsplit(normalizePath(run_dir, winslash = "/"), "/", fixed = TRUE)[[1]]
    n <- length(parts)
    run_label <- parts[n]
    run_id <- parts[n - 1]
    dataset_label <- parts[n - 2]

    lp <- readRDS(file.path(run_dir, "optim_lp_all.Rds"))
    lp <- lp[is.finite(lp)]
    if (!length(lp)) {
      return(NULL)
    }

    tibble(
      model_id = run_id,
      dataset_label = dataset_label,
      run_label = run_label,
      run_dir = run_dir,
      best_lp = max(lp),
      k = get_hierarchical_k(run_id, N_lines = stan_data$N_lines),
      alias = {
        run_parts <- parse_run_id(run_id)
        w_str <- if (run_parts$W > 0) sprintf(", %dW (%s)", run_parts$W, ifelse(run_parts$M == 1, "Add", "Mult")) else ""
        sprintf("%dR-%dP %s%s", run_parts$R, run_parts$P, ifelse(run_parts$C == 1, "Specialized", "Dense"), w_str)
      }
    )
  })

  bind_rows(rows) %>%
    group_by(model_id, alias, dataset_label, k) %>%
    slice_max(order_by = run_label, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      deviance = -2 * best_lp,
      AIC = 2 * k - 2 * best_lp,
      delta_AIC = AIC - min(AIC)
    ) %>%
    arrange(delta_AIC)
}

find_selected_model <- function(ranking_df) {
  if (!nrow(ranking_df)) {
    return(NA_character_)
  }

  ranking_df$model_id[which.min(ranking_df$delta_AIC)]
}

load_transfer_outputs <- function(model_id, output_root) {
  base_dir <- file.path(output_root, model_id)
  list(
    transfer_summary = safe_read_rds(file.path(base_dir, "transfer_comparison_summary.Rds")),
    parameter_summary = safe_read_rds(file.path(base_dir, "parameter_transfer_summary.Rds")),
    parameter_comparison = safe_read_rds(file.path(base_dir, "parameter_transfer_comparison.Rds")),
    best_start = safe_read_rds(file.path(base_dir, "transfer_best_start_summary.Rds"))
  )
}
```

## Upstream Commands

This notebook depends on artefacts created elsewhere. The table below documents the current known generation path.

```{r upstream-commands}
artifact_commands <- tribble(
  ~artifact_group, ~path, ~command, ~status_note,
  "Stan data", project_paths$stan_data,
  "Exact historical command not recorded in repo. Data are constructed by R/prepare_data.R.",
  "Known script, command not fixed in repo history.",
  "Versioned hierarchical optimization", "data/runs/gpath/v1/optim/<DATASET_LABEL>/<RUN_ID>/<RUN_LABEL>",
  "bash slurm/submit_gpath_optim_pipeline.sh <SPEC_FILE.tsv> ; specs can be generated with Rscript scripts/make_gpath_optim_spec.R <SPEC_FILE.tsv> <RUN_IDS>",
  "Current optimisation batch path with run manifest.",
  "Versioned hierarchical NUTS", "not yet migrated",
  "No new versioned NUTS pipeline for legacy hierarchical fits yet.",
  "Transfer NUTS still seeds from the older legacy NUTS outputs for now.",
  "Directional transfer (optimization)", "data/gpath_transfer_cv/gpath/<RUN_ID>",
  "bash slurm/submit_gpath_transfer_pipeline.sh MODEL R P W C M all low_to_high,high_to_low null,transfer,oracle 1:K 500 1000 0.99 12 16 ecology/stan_ready_data.Rds data/gpath_transfer_cv 0",
  "Current optimization transfer path.",
  "Directional transfer (optimization summaries)", "data/gpath_transfer_cv/gpath/<RUN_ID>/*.Rds",
  "Rscript scripts/summarize_gpath_transfer_cv.R gpath <RUN_ID> data/gpath_transfer_cv",
  "Current summary path.",
  "Directional transfer (NUTS)", "data/gpath_transfer_cv_nuts/gpath/<RUN_ID>",
  "bash slurm/submit_gpath_transfer_nuts_pipeline.sh MODEL R P W C M all low_to_high,high_to_low null,transfer,oracle 1:4 500 1000 0.99 12 16 ecology/stan_ready_data.Rds data/gpath_transfer_cv_nuts",
  "Current NUTS transfer path.",
  "Directional transfer (NUTS summaries)", "data/gpath_transfer_cv_nuts/gpath/<RUN_ID>/*.Rds",
  "Rscript scripts/summarize_gpath_transfer_nuts.R gpath <RUN_ID> data/gpath_transfer_cv_nuts",
  "Current summary path."
)

knitr::kable(artifact_commands)
```

## Available Outputs

```{r available-outputs}
ranking_df <- build_model_ranking()
selected_model_id <- find_selected_model(ranking_df)

cat("Available legacy gpath model IDs:\n")
print(ranking_df$model_id)
cat("\nSelected model for downstream reporting:", selected_model_id, "\n")
```

## Model Selection

```{r model-selection-table}
if (!nrow(ranking_df)) {
  announce_missing("No legacy optimisation outputs found under %s", project_paths$legacy_model_root)
} else {
  knitr::kable(ranking_df %>% select(model_id, alias, best_lp, k, AIC, delta_AIC))
}
```

```{r model-selection-plot}
if (!nrow(ranking_df)) {
  announce_missing("Skipping model-selection plot because no ranking table was built.")
} else {
  pareto_front <- ranking_df %>%
    arrange(k, deviance) %>%
    filter(deviance == cummin(deviance)) %>%
    group_by(deviance) %>%
    slice_min(order_by = k, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(k)

  p_aic <- ggplot(ranking_df, aes(x = reorder(alias, delta_AIC), y = delta_AIC)) +
    geom_col(fill = "grey35") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = expression(Delta ~ AIC))

  p_pareto <- ggplot(ranking_df, aes(x = k, y = deviance)) +
    geom_point(color = "grey60", size = 2.5) +
    geom_line(data = pareto_front, color = "firebrick", linetype = "dashed") +
    geom_point(data = pareto_front, color = "firebrick", size = 3) +
    theme_bw() +
    labs(x = "Effective parameters (k)", y = "Deviance")

  p_aic + p_pareto
}
```

## Optimisation Stability For The Selected Model

```{r selected-model-neighborhood}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available.")
} else {
  selected_run_dir <- ranking_df$run_dir[match(selected_model_id, ranking_df$model_id)][1]
  optim_obj <- list(
    lp = readRDS(file.path(selected_run_dir, "optim_lp_all.Rds")),
    draws = readRDS(file.path(selected_run_dir, "optim_draws_all.Rds"))
  )
  nbhd <- summarize_lp_neighborhood(optim_obj$lp, near_cut = 5)

  stability_df <- tibble(
    model_id = selected_model_id,
    n_total = length(optim_obj$lp),
    n_finite = sum(nbhd$finite_mask),
    best_lp = nbhd$best_lp,
    n_within_1 = sum(nbhd$finite_mask & nbhd$lp_diff < 1),
    n_within_5 = sum(nbhd$finite_mask & nbhd$lp_diff < 5),
    n_within_10 = sum(nbhd$finite_mask & nbhd$lp_diff < 10)
  )

  knitr::kable(stability_df)
}
```

## Transfer Analysis: Predictive View

This section treats transfer as a predictive benchmark. It is useful, but the parameter-transfer section below is now closer to the biological question of interest.

```{r transfer-optim-summary}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for transfer summary.")
} else {
  optim_transfer <- load_transfer_outputs(selected_model_id, output_root = project_paths$transfer_optim_root)

  if (is.null(optim_transfer$transfer_summary)) {
    announce_missing("No optimisation transfer summary found for %s", selected_model_id)
  } else {
    cat("Optimisation transfer summary\n")
    print(optim_transfer$transfer_summary)
  }
}
```

```{r transfer-nuts-summary}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for NUTS transfer summary.")
} else {
  nuts_summary_path <- file.path(project_paths$transfer_nuts_root, selected_model_id, "transfer_comparison_summary.Rds")
  nuts_transfer <- safe_read_rds(nuts_summary_path)

  if (is.null(nuts_transfer)) {
    announce_missing("No NUTS transfer summary found at %s", nuts_summary_path)
  } else {
    cat("NUTS transfer summary\n")
    print(nuts_transfer)
  }
}
```

## Transfer Analysis: Parameter View

This is currently the biologically primary transfer analysis. The null baseline is the observed-state parameterisation carried forward unchanged to the held-out ploidy state.

```{r parameter-transfer-summary}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer summary.")
} else {
  optim_transfer <- load_transfer_outputs(selected_model_id, output_root = project_paths$transfer_optim_root)

  if (is.null(optim_transfer$parameter_summary)) {
    announce_missing("No parameter transfer summary found for %s", selected_model_id)
  } else {
    print(optim_transfer$parameter_summary)
  }
}
```

```{r parameter-transfer-plot-1, fig.height=6, fig.width=10}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer plots.")
} else {
  comparison_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_comparison.Rds")
  if (!file.exists(comparison_path)) {
    announce_missing("No parameter transfer comparison found at %s", comparison_path)
  } else {
    plot_parameter_transfer_improvement(
      model_id = selected_model_id,
      output_root = "data/gpath_transfer_cv",
      model_name = "gpath"
    )
  }
}
```

```{r parameter-transfer-plot-2, fig.height=5, fig.width=9}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer plots.")
} else {
    summary_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_summary.Rds")
    if (!file.exists(summary_path)) {
      announce_missing("No parameter transfer summary found at %s", summary_path)
    } else {
      plot_parameter_transfer_summary(
        model_id = selected_model_id,
        output_root = "data/gpath_transfer_cv",
        model_name = "gpath"
      )
    }
}
```

## Transfer Fits For A Representative Line

This section overlays the best optimisation `null`, `transfer`, and `oracle` fits for one selected line. A line is chosen from the parameter-transfer summary by maximal mean improvement over null; if that summary is absent, line 1 is used.

```{r transfer-overlay-plot, fig.height=6, fig.width=12}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for transfer overlay plot.")
} else {
  parameter_summary_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_summary.Rds")
  parameter_summary_df <- safe_read_rds(parameter_summary_path)

  target_line <- 1L
  if (!is.null(parameter_summary_df) && nrow(parameter_summary_df)) {
    target_line <- parameter_summary_df$line_id[which.max(parameter_summary_df$mean_transfer_improvement_over_null)]
  }

  best_start_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "transfer_best_start_summary.Rds")
  if (!file.exists(best_start_path)) {
    announce_missing("No transfer best-start summary found at %s", best_start_path)
  } else {
    overlay_data <- generate_transfer_overlay_data(
      model_id = selected_model_id,
      line_id = target_line,
      output_root = "data/gpath_transfer_cv",
      stan_data_path = project_paths$stan_data
    )

    plot_transfer_line_trajectories(
      transfer_data = overlay_data,
      line_id = target_line,
      model_id = selected_model_id
    )
  }
}
```

## Current Interpretation

```{r interpretation-table}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for interpretation summary.")
} else {
  optim_transfer <- load_transfer_outputs(selected_model_id, output_root = project_paths$transfer_optim_root)

  if (is.null(optim_transfer$parameter_summary) || is.null(optim_transfer$transfer_summary)) {
    announce_missing("Need both predictive and parameter-transfer summaries to build the interpretation table.")
  } else {
    interpretation_df <- optim_transfer$parameter_summary %>%
      left_join(
        optim_transfer$transfer_summary %>% rename(model_direction = direction),
        by = c("line_id" = "line_id", "direction" = "model_direction")
      ) %>%
      select(
        line_id,
        direction,
        transfer,
        null,
        oracle,
        transfer_gain,
        transfer_regret,
        mean_transfer_improvement_over_null,
        prop_parameters_transfer_better_than_null
      )

    print(interpretation_df)
  }
}
```

## Notes And Next Steps

At the time of writing, the modelling workflow supports four distinct questions:

1. Which `gpath` structure is best supported by the hierarchical optimisation fits?
2. Are selected models numerically stable enough to interpret?
3. Does a shared ploidy effect improve held-out prediction in directional transfer tests?
4. Does that same shared ploidy effect move held-out-line parameter values closer to the oracle fit than a null baseline does?

Recommended next steps:

1. Treat the selected `gpath` model as primary and stop comparing many model IDs throughout the downstream biological analysis.
2. Use the predictive transfer summaries as a benchmark, but make the parameter-transfer section the primary scientific result.
3. Add one compact comparison of optimisation-based versus NUTS-based transfer for the same selected model to quantify how much posterior averaging changes the conclusion.
4. If the parameter-transfer summaries are still hard to interpret, add family-level summaries for `ae`, `ah`, `Y_R`, `A_mat`, and `m` rather than only per-parameter outputs.
5. Keep legacy or exploratory analysis blocks outside this notebook unless they are directly needed for the current `gpath` narrative.
