---
title: "Model-Free Ploidy Analysis"
author: "Codex"
date: "2026-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook reports a reduced model-free ploidy analysis built from `stan_ready_data` without fitting the ODE. The analysis now focuses on eight curated features only:

1. two growth features,
2. two death features,
3. two alive-yield features from `AUC_alive ~ log1p(G0)`,
4. two glucose-to-total-cell yield features from `peak_total_yield_per_glucose ~ log1p(G0)`.

The notebook expects these scripts to have been run first:

```bash
Rscript scripts/extract_model_free_ploidy_features.R
Rscript scripts/evaluate_model_free_ploidy_transfer.R
```

## Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(scales)
library(patchwork)
library(forcats)
library(purrr)

source("R/project_paths.R")
source("R/gpath_run_utils.R")
source("R/model_free_ploidy_utils.R")

analysis_root <- file.path("data", "model_free_ploidy")

paths <- list(
  stan_data = resolve_stan_data_path(),
  feature_panel = file.path(analysis_root, "feature_panel.Rds"),
  signature_panel = file.path(analysis_root, "signature_panel.Rds"),
  signature_effects = file.path(analysis_root, "signature_effects_long.Rds"),
  transfer_predictions = file.path(analysis_root, "transfer_predictions.Rds"),
  transfer_feature_summary = file.path(analysis_root, "transfer_feature_summary.Rds"),
  transfer_case_summary = file.path(analysis_root, "transfer_case_summary.Rds")
)

safe_read_rds <- function(path) {
  if (!file.exists(path)) {
    return(NULL)
  }
  readRDS(path)
}

feature_panel <- safe_read_rds(paths$feature_panel)
signature_panel <- safe_read_rds(paths$signature_panel)
signature_effects <- safe_read_rds(paths$signature_effects)
transfer_predictions <- safe_read_rds(paths$transfer_predictions)
transfer_feature_summary <- safe_read_rds(paths$transfer_feature_summary)
transfer_case_summary <- safe_read_rds(paths$transfer_case_summary)
feature_catalog <- get_model_free_feature_catalog()

required_paths <- unlist(paths[c("feature_panel", "signature_panel", "signature_effects", "transfer_predictions", "transfer_feature_summary")])
if (!all(file.exists(required_paths))) {
  stop("Required outputs are missing. Run the extraction and transfer scripts before rendering this notebook.")
}

cat("Working directory:", getwd(), "\n")
cat("stan_data path:", paths$stan_data, "\n")
cat("analysis root:", analysis_root, "\n")
```

## Output Check

```{r output-check}
tibble(
  artifact = names(paths),
  path = unlist(paths),
  exists = file.exists(unlist(paths))
)
```

## Curated Feature Set

```{r feature-catalog}
feature_catalog
```

## Core Per-Condition Quantities

Two yield quantities are central here:

- `live_auc`: trapezoidal area under the alive-cell curve
- `peak_total_yield_per_glucose`: `max(live + dead)` up to the last glucose measurement, divided by `initial_glucose - final_glucose`

```{r per-condition-preview}
feature_panel %>%
  select(
    cellLine, ploidy_metric, G0,
    max_growth_rate, max_death_rate,
    live_auc,
    total_peak_to_glucose_end,
    glucose_initial, glucose_final,
    peak_total_yield_per_glucose
  ) %>%
  arrange(cellLine, ploidy_metric, G0)
```

## Effect Overview

The old heatmap was not interpretable because feature scales differed too much. If we keep a heatmap at all, it should be feature-wise standardized.

```{r effect-heatmap, fig.height=5, fig.width=10}
signature_effects %>%
  left_join(feature_catalog, by = "feature") %>%
  group_by(feature) %>%
  mutate(effect_z = as.numeric(scale(effect_per_ploidy))) %>%
  ungroup() %>%
  ggplot(aes(x = short_label, y = cellLine, fill = effect_z)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b") +
  labs(
    title = "Standardized Ploidy Effects Across Curated Features",
    x = NULL,
    y = NULL,
    fill = "within-feature\nz-score"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Feature-by-Feature Analysis

```{r helpers}
plot_feature_scatter <- function(raw_col, title, ylab) {
  ggplot(feature_panel, aes(G0, .data[[raw_col]], color = factor(ploidy_metric))) +
    geom_point(size = 2) +
    geom_line(aes(group = interaction(cellLine, ploidy_metric)), alpha = 0.5) +
    facet_wrap(~cellLine, scales = "free_y") +
    scale_x_continuous(trans = pseudo_log_trans(base = 10)) +
    scale_color_brewer(palette = "Set1", name = "ploidy") +
    labs(title = title, x = "initial glucose (G0)", y = ylab) +
    theme_bw()
}

build_fit_df <- function(raw_col) {
  feature_panel %>%
    group_by(cellLine, ploidy_metric) %>%
    group_modify(~ {
      fit <- lm(reformulate("log1p(G0)", response = raw_col), data = .x)
      tibble(
        G0 = seq(0, 25, length.out = 100),
        fitted = predict(fit, newdata = data.frame(G0 = seq(0, 25, length.out = 100)))
      )
    }) %>%
    ungroup()
}

plot_feature_regression <- function(raw_col, title, ylab) {
  fit_df <- build_fit_df(raw_col)

  ggplot(feature_panel, aes(G0, .data[[raw_col]], color = factor(ploidy_metric))) +
    geom_point(size = 2) +
    geom_line(data = fit_df, aes(x = G0, y = fitted, color = factor(ploidy_metric)), linewidth = 0.8) +
    facet_wrap(~cellLine, scales = "free_y") +
    scale_x_continuous(trans = pseudo_log_trans(base = 10)) +
    scale_color_brewer(palette = "Set1", name = "ploidy") +
    labs(title = title, x = "initial glucose (G0)", y = ylab) +
    theme_bw()
}

plot_signature_values <- function(sig_col, title, ylab) {
  ggplot(signature_panel, aes(factor(ploidy_metric), .data[[sig_col]], fill = factor(ploidy_metric))) +
    geom_col() +
    facet_wrap(~cellLine, scales = "free_y") +
    scale_fill_brewer(palette = "Set1", name = "ploidy") +
    labs(title = title, x = "ploidy metric", y = ylab) +
    theme_bw()
}
```

### Growth Low G

```{r growth-lowg-text, results='asis'}
row <- feature_catalog %>% filter(feature == "growth_lowG_median")
cat("1. Rationale: ", row$rationale, "\n\n", sep = "")
cat("2. Computation: ", row$computation, "\n", sep = "")
```

```{r growth-lowg-plot, fig.height=6, fig.width=10}
plot_feature_scatter(
  raw_col = "max_growth_rate",
  title = "Per-Condition Max Growth Rate Across Initial Glucose",
  ylab = "max growth rate"
)
```

```{r growth-lowg-signature, fig.height=4, fig.width=10}
plot_signature_values(
  sig_col = "growth_lowG_median",
  title = "Aggregated Feature: Growth Low G Median",
  ylab = "median max growth rate at G0 <= 0.25"
)
```

### Growth High G

```{r growth-highg-text, results='asis'}
row <- feature_catalog %>% filter(feature == "growth_highG_median")
cat("1. Rationale: ", row$rationale, "\n\n", sep = "")
cat("2. Computation: ", row$computation, "\n", sep = "")
```

```{r growth-highg-plot, fig.height=6, fig.width=10}
plot_feature_scatter(
  raw_col = "max_growth_rate",
  title = "Per-Condition Max Growth Rate Across Initial Glucose",
  ylab = "max growth rate"
)
```

```{r growth-highg-signature, fig.height=4, fig.width=10}
plot_signature_values(
  sig_col = "growth_highG_median",
  title = "Aggregated Feature: Growth High G Median",
  ylab = "median max growth rate at G0 >= 1"
)
```

### Death Low G

```{r death-lowg-text, results='asis'}
row <- feature_catalog %>% filter(feature == "death_lowG_median")
cat("1. Rationale: ", row$rationale, "\n\n", sep = "")
cat("2. Computation: ", row$computation, "\n", sep = "")
```

```{r death-lowg-plot, fig.height=6, fig.width=10}
plot_feature_scatter(
  raw_col = "max_death_rate",
  title = "Per-Condition Max Death Rate Across Initial Glucose",
  ylab = "max death rate"
)
```

```{r death-lowg-signature, fig.height=4, fig.width=10}
plot_signature_values(
  sig_col = "death_lowG_median",
  title = "Aggregated Feature: Death Low G Median",
  ylab = "median max death rate at G0 <= 0.25"
)
```

### Death High G

```{r death-highg-text, results='asis'}
row <- feature_catalog %>% filter(feature == "death_highG_median")
cat("1. Rationale: ", row$rationale, "\n\n", sep = "")
cat("2. Computation: ", row$computation, "\n", sep = "")
```

```{r death-highg-plot, fig.height=6, fig.width=10}
plot_feature_scatter(
  raw_col = "max_death_rate",
  title = "Per-Condition Max Death Rate Across Initial Glucose",
  ylab = "max death rate"
)
```

```{r death-highg-signature, fig.height=4, fig.width=10}
plot_signature_values(
  sig_col = "death_highG_median",
  title = "Aggregated Feature: Death High G Median",
  ylab = "median max death rate at G0 >= 1"
)
```

### Alive AUC Baseline

```{r aliveauc-int-text, results='asis'}
row <- feature_catalog %>% filter(feature == "yield_alive_auc_intercept")
cat("1. Rationale: ", row$rationale, "\n\n", sep = "")
cat("2. Computation: ", row$computation, "\n", sep = "")
```

```{r aliveauc-reg-plot, fig.height=6, fig.width=10}
plot_feature_regression(
  raw_col = "live_auc",
  title = "Alive AUC Regressed Against Initial Glucose",
  ylab = "alive AUC"
)
```

```{r aliveauc-int-signature, fig.height=4, fig.width=10}
plot_signature_values(
  sig_col = "yield_alive_auc_intercept",
  title = "Aggregated Feature: Alive AUC Regression Intercept",
  ylab = "intercept of alive AUC ~ log1p(G0)"
)
```

### Alive AUC Glucose Response

```{r aliveauc-slope-text, results='asis'}
row <- feature_catalog %>% filter(feature == "yield_alive_auc_slope")
cat("1. Rationale: ", row$rationale, "\n\n", sep = "")
cat("2. Computation: ", row$computation, "\n", sep = "")
```

```{r aliveauc-slope-plot, fig.height=6, fig.width=10}
plot_feature_regression(
  raw_col = "live_auc",
  title = "Alive AUC Regressed Against Initial Glucose",
  ylab = "alive AUC"
)
```

```{r aliveauc-slope-signature, fig.height=4, fig.width=10}
plot_signature_values(
  sig_col = "yield_alive_auc_slope",
  title = "Aggregated Feature: Alive AUC Regression Slope",
  ylab = "slope of alive AUC ~ log1p(G0)"
)
```

### Peak Yield Baseline

```{r peakyield-int-text, results='asis'}
row <- feature_catalog %>% filter(feature == "peak_total_yield_intercept")
cat("1. Rationale: ", row$rationale, "\n\n", sep = "")
cat("2. Computation: ", row$computation, "\n", sep = "")
```

```{r peakyield-reg-plot, fig.height=6, fig.width=10}
plot_feature_regression(
  raw_col = "peak_total_yield_per_glucose",
  title = "Peak Total Yield Per Glucose Regressed Against Initial Glucose",
  ylab = "max(live + dead) / glucose drawdown"
)
```

```{r peakyield-int-signature, fig.height=4, fig.width=10}
plot_signature_values(
  sig_col = "peak_total_yield_intercept",
  title = "Aggregated Feature: Peak Yield Regression Intercept",
  ylab = "intercept of peak yield ~ log1p(G0)"
)
```

### Peak Yield Glucose Response

```{r peakyield-slope-text, results='asis'}
row <- feature_catalog %>% filter(feature == "peak_total_yield_slope")
cat("1. Rationale: ", row$rationale, "\n\n", sep = "")
cat("2. Computation: ", row$computation, "\n", sep = "")
```

```{r peakyield-slope-plot, fig.height=6, fig.width=10}
plot_feature_regression(
  raw_col = "peak_total_yield_per_glucose",
  title = "Peak Total Yield Per Glucose Regressed Against Initial Glucose",
  ylab = "max(live + dead) / glucose drawdown"
)
```

```{r peakyield-slope-signature, fig.height=4, fig.width=10}
plot_signature_values(
  sig_col = "peak_total_yield_slope",
  title = "Aggregated Feature: Peak Yield Regression Slope",
  ylab = "slope of peak yield ~ log1p(G0)"
)
```

## Transfer Summary

```{r transfer-summary}
transfer_feature_summary %>%
  left_join(feature_catalog, by = "feature") %>%
  select(short_label, n_cases, scaled_mae_null, scaled_mae_transfer, mean_scaled_err_improvement, transfer_win_rate, sign_accuracy) %>%
  arrange(desc(mean_scaled_err_improvement))
```

```{r transfer-bar, fig.height=5, fig.width=9}
transfer_feature_summary %>%
  left_join(feature_catalog, by = "feature") %>%
  mutate(short_label = forcats::fct_reorder(short_label, mean_scaled_err_improvement)) %>%
  ggplot(aes(x = short_label, y = mean_scaled_err_improvement, fill = transfer_win_rate)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Transfer Gain By Curated Feature",
    x = NULL,
    y = "mean scaled error improvement",
    fill = "transfer\nwin rate"
  ) +
  theme_bw()
```

```{r transfer-case-summary}
transfer_case_summary %>%
  arrange(desc(mean_scaled_err_improvement))
```

## Notes

- `peak_total_yield_per_glucose` uses `max(live + dead)` only up to the final glucose measurement time, as requested.
- Transfer is now evaluated only on the curated eight-feature panel.
- Cross-feature comparisons use scaled errors, so large-magnitude yield features no longer dominate the summary.
