---
title: "Ecological Modelling"
author: "RJB"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook is now the reporting layer for the current `gpath` modelling workflow. It is not intended to recreate the full analysis from scratch. Instead it:

- documents the current modelling path,
- states the upstream scripts/commands that generate required artefacts where known,
- loads saved results from the current workflow,
- presents model selection, diagnostics, transfer results, and parameter-transfer summaries.

Where an upstream command is not known exactly from the repo history, that is stated explicitly.

## Setup

```{r libraries}
library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(tibble)
library(posterior)
library(patchwork)

source("R/project_paths.R")
source(get_model_r_path("gpath", "v1"))
source("R/gpath_run_utils.R")
source("R/posterior_io_utils.R")
source("R/elpd_transfer_utils.R")
source("R/parameter_transfer_utils.R")
source("R/parameter_transfer_plot_utils.R")
source("R/transfer_cv_plot_utils.R")

focus_common_parameters <- c("ae[1]", "ah[1]", "Y_R[1]", "m")

project_paths <- list(
  stan_data = resolve_stan_data_path(),
  assessment_run_ids = file.path("models", "gpath", "v1", "assessment_run_ids.txt"),
  optim_root = file.path("data", "runs", "gpath", "v1", "optim"),
  transfer_optim_root = "data/gpath_transfer_cv/gpath",
  transfer_nuts_root = "data/gpath_transfer_cv_nuts/gpath"
)

cat("Current working directory:", getwd(), "\n")
cat("stan_data path:", project_paths$stan_data, "\n")
cat("assessment set:", project_paths$assessment_run_ids, "\n")
```

```{r helpers}
safe_read_rds <- function(path) {
  if (!file.exists(path)) {
    return(NULL)
  }
  readRDS(path)
}

announce_missing <- function(...) {
  cat("SKIP:", sprintf(...), "\n")
}

read_run_id_file <- function(path) {
  if (!file.exists(path)) {
    return(character(0))
  }

  vals <- readLines(path, warn = FALSE)
  vals <- trimws(vals)
  vals <- vals[nzchar(vals)]
  vals <- vals[!grepl("^#", vals)]
  vals
}

discover_model_runs <- function(base_path) {
  if (!dir.exists(base_path)) {
    return(character(0))
  }
  run_dirs <- list.dirs(base_path, recursive = TRUE, full.names = TRUE)
  run_dirs[
    file.exists(file.path(run_dirs, "optim_lp_all.Rds")) &
    file.exists(file.path(run_dirs, "optim_draws_all.Rds")) &
    file.exists(file.path(run_dirs, "run_manifest.json"))
  ]
}

build_model_ranking <- function(
  base_path = project_paths$optim_root,
  stan_data_path = project_paths$stan_data,
  assessment_ids = read_run_id_file(project_paths$assessment_run_ids)
) {
  run_dirs <- discover_model_runs(base_path)
  if (!length(run_dirs)) {
    return(tibble())
  }

  stan_data <- readRDS(stan_data_path)

  rows <- lapply(run_dirs, function(run_dir) {
    parts <- strsplit(normalizePath(run_dir, winslash = "/"), "/", fixed = TRUE)[[1]]
    n <- length(parts)
    run_label <- parts[n]
    run_id <- parts[n - 1]
    dataset_label <- parts[n - 2]

    lp <- readRDS(file.path(run_dir, "optim_lp_all.Rds"))
    lp <- lp[is.finite(lp)]
    if (!length(lp)) {
      return(NULL)
    }

    manifest_path <- file.path(run_dir, "run_manifest.json")
    manifest <- jsonlite::read_json(manifest_path, simplifyVector = TRUE)

    tibble(
      model_id = run_id,
      dataset_label = dataset_label,
      run_label = run_label,
      run_dir = run_dir,
      manifest_path = manifest_path,
      best_lp = max(lp),
      k = get_hierarchical_k(run_id, N_lines = stan_data$N_lines),
      manifest_stan_data = manifest$stan_data_path,
      manifest_stan_file = manifest$stan_file_path,
      alias = {
        run_parts <- parse_run_id(run_id)
        w_str <- if (run_parts$W > 0) sprintf(", %dW (%s)", run_parts$W, ifelse(run_parts$M == 1, "Add", "Mult")) else ""
        sprintf("%dR-%dP %s%s", run_parts$R, run_parts$P, ifelse(run_parts$C == 1, "Specialized", "Dense"), w_str)
      }
    )
  })

  bind_rows(rows) %>%
    filter(model_id %in% assessment_ids) %>%
    group_by(model_id, alias, dataset_label, k) %>%
    slice_max(order_by = run_label, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      deviance = -2 * best_lp,
      AIC = 2 * k - 2 * best_lp,
      delta_AIC = AIC - min(AIC)
    ) %>%
    arrange(delta_AIC)
}

summarize_manifest_status <- function(ranking_df) {
  if (!nrow(ranking_df)) {
    return(tibble())
  }

  ranking_df %>%
    transmute(
      model_id,
      run_label,
      run_dir,
      manifest_present = file.exists(manifest_path),
      optim_complete = file.exists(file.path(run_dir, "optim_lp_all.Rds")) &
        file.exists(file.path(run_dir, "optim_draws_all.Rds")),
      stan_data_matches_current = normalizePath(manifest_stan_data, winslash = "/", mustWork = FALSE) ==
        normalizePath(project_paths$stan_data, winslash = "/", mustWork = FALSE)
    ) %>%
    arrange(model_id)
}

find_selected_model <- function(ranking_df) {
  if (!nrow(ranking_df)) {
    return(NA_character_)
  }

  ranking_df$model_id[which.min(ranking_df$delta_AIC)]
}

extract_best_draw_vector <- function(run_dir) {
  lp <- readRDS(file.path(run_dir, "optim_lp_all.Rds"))
  draws <- readRDS(file.path(run_dir, "optim_draws_all.Rds"))
  rc_path <- file.path(run_dir, "optim_rc_all.Rds")
  rc <- if (file.exists(rc_path)) readRDS(rc_path) else rep(0L, length(lp))

  ok <- which(is.finite(lp) & rc == 0L)
  if (!length(ok)) {
    ok <- which(is.finite(lp))
  }
  if (!length(ok)) {
    return(NULL)
  }

  best_i <- ok[which.max(lp[ok])]
  best_draw <- draws[[best_i]]

  if (is.data.frame(best_draw)) {
    best_draw <- as.matrix(best_draw)
  }

  if (!is.matrix(best_draw) || !nrow(best_draw)) {
    stop(sprintf("Unsupported best-draw format in %s", run_dir))
  }

  out <- best_draw[1, , drop = TRUE]
  names(out) <- colnames(best_draw)
  out
}

collect_best_raw_theta_ploidy_global <- function(ranking_df) {
  if (!nrow(ranking_df)) {
    return(tibble())
  }

  rows <- lapply(seq_len(nrow(ranking_df)), function(i) {
    model_id <- ranking_df$model_id[[i]]
    run_dir <- ranking_df$run_dir[[i]]
    dims <- parse_run_id(model_id)
    param_labels <- get_param_names(
      R = dims$R,
      P = dims$P,
      W = dims$W,
      C = dims$C,
      M = dims$M,
      priors = base_priors
    )

    draw_vec <- tryCatch(extract_best_draw_vector(run_dir), error = function(e) NULL)
    if (is.null(draw_vec)) {
      return(NULL)
    }

    coef_names <- sprintf("raw_theta_ploidy[%d]", seq_along(param_labels))
    values <- unname(draw_vec[coef_names])

    tibble(
      source = "global_optim",
      model_id = model_id,
      alias = ranking_df$alias[[i]],
      run_label = ranking_df$run_label[[i]],
      line_id = NA_integer_,
      direction = NA_character_,
      fit_type = "all_data_best",
      run_tag = NA_character_,
      coefficient = coef_names,
      parameter = param_labels,
      raw_theta_ploidy = values,
      missing = is.na(values),
      return_code = NA_real_,
      lp__ = unname(draw_vec["lp__"]),
      elpd = NA_real_,
      ll_transfer_train = NA_real_,
      ll_transfer_holdout = NA_real_,
      holdout_wells = NA_integer_,
      holdout_obs = NA_integer_
    )
  })

  bind_rows(rows)
}

collect_best_raw_theta_ploidy_transfer <- function(
  ranking_df,
  output_root = project_paths$transfer_optim_root,
  stan_data_path = project_paths$stan_data
) {
  model_ids <- ranking_df$model_id
  if (!length(model_ids)) {
    return(tibble())
  }

  stan_data <- readRDS(resolve_stan_data_path(stan_data_path))
  all_wells <- seq_len(stan_data$N_wells)
  rows <- list()
  idx <- 1L

  for (model_id in model_ids) {
    base_dir <- file.path(output_root, model_id)
    best_path <- file.path(base_dir, "transfer_best_start_summary.Rds")
    if (!file.exists(best_path)) {
      next
    }

    best_df <- readRDS(best_path)
    if (is.null(best_df) || !nrow(best_df)) {
      next
    }

    dims <- parse_run_id(model_id)
    param_labels <- get_param_names(
      R = dims$R,
      P = dims$P,
      W = dims$W,
      C = dims$C,
      M = dims$M,
      priors = base_priors
    )
    coef_names <- sprintf("raw_theta_ploidy[%d]", seq_along(param_labels))

    for (j in seq_len(nrow(best_df))) {
      fit_row <- best_df[j, , drop = FALSE]
      fit_obj <- tryCatch(
        load_best_transfer_fit(
          model_id = model_id,
          line_id = fit_row$line_id[[1]],
          direction = fit_row$direction[[1]],
          fit_type = fit_row$fit_type[[1]],
          output_root = dirname(output_root),
          model_name = "gpath"
        ),
        error = function(e) NULL
      )

      if (is.null(fit_obj)) {
        next
      }

      draw_vec <- extract_draw_vector(fit_obj$draws)
      values <- unname(draw_vec[coef_names])
      holdout_wells <- fit_obj$split_meta$holdout_wells
      train_wells <- setdiff(all_wells, holdout_wells)

      rows[[idx]] <- tibble(
        source = "transfer_cv_best",
        model_id = model_id,
        alias = ranking_df$alias[match(model_id, ranking_df$model_id)],
        run_label = ranking_df$run_label[match(model_id, ranking_df$model_id)],
        line_id = as.integer(fit_row$line_id[[1]]),
        direction = as.character(fit_row$direction[[1]]),
        fit_type = as.character(fit_row$fit_type[[1]]),
        run_tag = as.character(fit_row$run_tag[[1]]),
        coefficient = coef_names,
        parameter = param_labels,
        raw_theta_ploidy = values,
        missing = is.na(values),
        return_code = as.numeric(fit_row$return_code[[1]]),
        lp__ = as.numeric(fit_row$lp__[[1]]),
        elpd = as.numeric(fit_row$elpd[[1]]),
        ll_transfer_train = compute_point_loglik_subset(fit_obj$draws, train_wells),
        ll_transfer_holdout = compute_point_loglik_subset(fit_obj$draws, holdout_wells),
        holdout_wells = as.integer(length(holdout_wells)),
        holdout_obs = as.integer(fit_row$holdout_obs[[1]])
      )
      idx <- idx + 1L
    }
  }

  bind_rows(rows)
}

load_transfer_outputs <- function(model_id, output_root) {
  base_dir <- file.path(output_root, model_id)
  list(
    transfer_summary = safe_read_rds(file.path(base_dir, "transfer_comparison_summary.Rds")),
    parameter_states = safe_read_rds(file.path(base_dir, "parameter_transfer_states.Rds")),
    parameter_summary = safe_read_rds(file.path(base_dir, "parameter_transfer_summary.Rds")),
    parameter_comparison = safe_read_rds(file.path(base_dir, "parameter_transfer_comparison.Rds")),
    best_start = safe_read_rds(file.path(base_dir, "transfer_best_start_summary.Rds"))
  )
}

collect_transfer_screening <- function(model_ids, output_root = project_paths$transfer_optim_root) {
  rows <- lapply(model_ids, function(model_id) {
    x <- load_transfer_outputs(model_id, output_root = output_root)
    ts <- x$transfer_summary

    if (is.null(ts) || !nrow(ts)) {
      return(NULL)
    }

    ts$model_id <- model_id
    ts
  })

  bind_rows(rows)
}

collect_parameter_screening <- function(model_ids, output_root = project_paths$transfer_optim_root) {
  rows <- lapply(model_ids, function(model_id) {
    x <- load_transfer_outputs(model_id, output_root = output_root)
    ps <- x$parameter_summary

    if (is.null(ps) || !nrow(ps)) {
      return(NULL)
    }

    ps$model_id <- model_id
    ps
  })

  bind_rows(rows)
}

collect_parameter_family_transfer <- function(model_ids, output_root = project_paths$transfer_optim_root) {
  rows <- lapply(model_ids, function(model_id) {
    x <- load_transfer_outputs(model_id, output_root = output_root)
    pc <- x$parameter_comparison

    if (is.null(pc) || !nrow(pc)) {
      return(NULL)
    }

    pc %>%
      mutate(
        model_id = model_id,
        family = vapply(parameter, get_parameter_family, character(1))
      ) %>%
      filter(!is.na(family)) %>%
      mutate(
        abs_null_err = pmax(abs(null_vs_oracle_diff), 1e-12),
        abs_transfer_err = pmax(abs(holdout_diff), 1e-12),
        log10_error_ratio = log10(abs_transfer_err / abs_null_err),
        log_error_improvement = log(abs_null_err) - log(abs_transfer_err),
        transfer_better_than_null = abs_transfer_err < abs_null_err
      )
  })

  bind_rows(rows)
}

summarize_predictive_screening <- function(screen_df) {
  if (!nrow(screen_df)) {
    return(tibble())
  }

  screen_df %>%
    group_by(model_id, direction) %>%
    summarise(
      mean_transfer_gain = mean(transfer_gain, na.rm = TRUE),
      median_transfer_gain = median(transfer_gain, na.rm = TRUE),
      mean_transfer_regret = mean(transfer_regret, na.rm = TRUE),
      prop_transfer_beats_null = mean(transfer > null, na.rm = TRUE),
      n_cases = dplyr::n(),
      .groups = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from = direction,
      values_from = c(
        mean_transfer_gain,
        median_transfer_gain,
        mean_transfer_regret,
        prop_transfer_beats_null,
        n_cases
      )
    ) %>%
    arrange(desc(mean_transfer_gain_low_to_high), desc(mean_transfer_gain_high_to_low))
}

summarize_parameter_screening <- function(screen_df) {
  if (!nrow(screen_df)) {
    return(tibble())
  }

  screen_df %>%
    mutate(
      mean_log_error_improvement = mean_abs_log_null_holdout_ratio - mean_abs_log_holdout_ratio
    ) %>%
    group_by(model_id, direction) %>%
    summarise(
      mean_log_error_improvement = mean(mean_log_error_improvement, na.rm = TRUE),
      median_log_error_improvement = median(mean_log_error_improvement, na.rm = TRUE),
      mean_prop_parameters_better = mean(prop_parameters_transfer_better_than_null, na.rm = TRUE),
      n_cases = dplyr::n(),
      .groups = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from = direction,
      values_from = c(
        mean_log_error_improvement,
        median_log_error_improvement,
        mean_prop_parameters_better,
        n_cases
      )
    ) %>%
    arrange(desc(mean_log_error_improvement_low_to_high), desc(mean_log_error_improvement_high_to_low))
}

summarize_parameter_family_transfer <- function(family_df) {
  if (!nrow(family_df)) {
    return(tibble())
  }

  family_df %>%
    group_by(model_id, line_id, direction, family) %>%
    summarise(
      median_log10_error_ratio = median(log10_error_ratio, na.rm = TRUE),
      mean_log_error_improvement = mean(log_error_improvement, na.rm = TRUE),
      prop_transfer_better_than_null = mean(transfer_better_than_null, na.rm = TRUE),
      n_parameters = dplyr::n(),
      .groups = "drop"
    )
}

collect_focused_parameter_transfer <- function(
  model_ids,
  focus_parameters = focus_common_parameters,
  output_root = project_paths$transfer_optim_root
) {
  rows <- lapply(model_ids, function(model_id) {
    x <- load_transfer_outputs(model_id, output_root = output_root)
    pc <- x$parameter_comparison

    if (is.null(pc) || !nrow(pc)) {
      return(NULL)
    }

    pc %>%
      filter(parameter %in% focus_parameters) %>%
      mutate(
        model_id = model_id,
        abs_null_err = abs(null_vs_oracle_diff),
        abs_transfer_err = abs(holdout_diff),
        abs_error_improvement = abs_null_err - abs_transfer_err,
        positive_triplet = null_holdout > 0 & transfer_holdout > 0 & oracle_holdout > 0,
        abs_log_null_ratio = ifelse(
          positive_triplet,
          abs(log(null_holdout / oracle_holdout)),
          NA_real_
        ),
        abs_log_transfer_ratio = ifelse(
          positive_triplet,
          abs(log(transfer_holdout / oracle_holdout)),
          NA_real_
        ),
        ratio_error_improvement = abs_log_null_ratio - abs_log_transfer_ratio,
        transfer_beats_null_abs = abs_transfer_err < abs_null_err,
        transfer_beats_null_ratio = abs_log_transfer_ratio < abs_log_null_ratio,
        line_direction = sprintf("line %d | %s", line_id, direction)
      )
  })

  bind_rows(rows)
}

summarize_parameter_family_across_models <- function(family_summary_df) {
  if (!nrow(family_summary_df)) {
    return(tibble())
  }

  family_summary_df %>%
    group_by(model_id, direction, family) %>%
    summarise(
      mean_family_log_error_improvement = mean(mean_log_error_improvement, na.rm = TRUE),
      median_family_log10_error_ratio = median(median_log10_error_ratio, na.rm = TRUE),
      mean_family_win_rate = mean(prop_transfer_better_than_null, na.rm = TRUE),
      n_line_cases = dplyr::n(),
      .groups = "drop"
    )
}

rank_parameter_families <- function(model_family_df) {
  if (!nrow(model_family_df)) {
    return(tibble())
  }

  model_family_df %>%
    group_by(direction, family) %>%
    summarise(
      mean_log_error_improvement = mean(mean_family_log_error_improvement, na.rm = TRUE),
      median_log10_error_ratio = median(median_family_log10_error_ratio, na.rm = TRUE),
      mean_win_rate = mean(mean_family_win_rate, na.rm = TRUE),
      prop_models_positive = mean(mean_family_log_error_improvement > 0, na.rm = TRUE),
      n_models = dplyr::n(),
      .groups = "drop"
    ) %>%
    arrange(direction, desc(mean_log_error_improvement), desc(mean_win_rate))
}

derive_glucose_transfer_summaries <- function(state_df) {
  if (is.null(state_df) || !nrow(state_df)) {
    return(tibble(
      line_id = integer(),
      direction = character(),
      fit_type = character(),
      state = character(),
      summary_name = character(),
      summary_value = numeric()
    ))
  }

  glucose_name_map <- c(
    "ae[1]" = "ae1",
    "ae" = "ae1",
    "ah[1]" = "ah1",
    "ah" = "ah1",
    "Y_R[1]" = "YR1",
    "Y_R" = "YR1",
    "m" = "m"
  )

  wide <- state_df %>%
    filter(parameter %in% names(glucose_name_map), fit_type %in% c("transfer", "oracle")) %>%
    mutate(parameter_key = unname(glucose_name_map[parameter])) %>%
    select(line_id, direction, fit_type, state, parameter_key, value) %>%
    tidyr::pivot_wider(names_from = parameter_key, values_from = value)

  needed <- c("ae1", "ah1", "YR1", "m")
  if (!all(needed %in% names(wide))) {
    return(tibble(
      line_id = integer(),
      direction = character(),
      fit_type = character(),
      state = character(),
      summary_name = character(),
      summary_value = numeric()
    ))
  }

  wide %>%
    transmute(
      line_id,
      direction,
      fit_type,
      state,
      Umax_1 = ae1,
      Khalf_1 = ah1,
      Efflow_1 = ae1 / ah1,
      YieldCap_1 = YR1 * ae1,
      NetGrow_1 = (YR1 * ae1) - m
    ) %>%
    tidyr::pivot_longer(
      cols = c(Umax_1, Khalf_1, Efflow_1, YieldCap_1, NetGrow_1),
      names_to = "summary_name",
      values_to = "summary_value"
    )
}

collect_derived_transfer_summaries <- function(model_ids, output_root = project_paths$transfer_optim_root) {
  rows <- lapply(model_ids, function(model_id) {
    x <- load_transfer_outputs(model_id, output_root = output_root)
    ds <- derive_glucose_transfer_summaries(x$parameter_states)

    if (is.null(ds) || !nrow(ds)) {
      return(NULL)
    }

    ds$model_id <- model_id
    ds
  })

  out <- bind_rows(rows)
  if (!nrow(out)) {
    return(tibble(
      line_id = integer(),
      direction = character(),
      fit_type = character(),
      state = character(),
      summary_name = character(),
      summary_value = numeric(),
      model_id = character()
    ))
  }
  out
}

build_derived_transfer_comparison <- function(derived_df) {
  if (!nrow(derived_df)) {
    return(tibble(
      model_id = character(),
      line_id = integer(),
      direction = character(),
      summary_name = character(),
      observed_transfer = numeric(),
      transfer_holdout = numeric(),
      oracle_holdout = numeric(),
      observed_oracle = numeric(),
      null_holdout = numeric(),
      null_vs_oracle_diff = numeric(),
      holdout_diff = numeric(),
      abs_null_err = numeric(),
      abs_transfer_err = numeric(),
      log10_error_ratio = numeric(),
      log_error_improvement = numeric(),
      transfer_better_than_null = logical(),
      oracle_shift = numeric(),
      transfer_shift = numeric(),
      shift_diff = numeric()
    ))
  }

  transfer_obs <- derived_df %>%
    filter(fit_type == "transfer", state == "observed") %>%
    select(model_id, line_id, direction, summary_name, observed_transfer = summary_value)

  transfer_holdout <- derived_df %>%
    filter(fit_type == "transfer", state == "holdout") %>%
    select(model_id, line_id, direction, summary_name, transfer_holdout = summary_value)

  oracle_holdout <- derived_df %>%
    filter(fit_type == "oracle", state == "holdout") %>%
    select(model_id, line_id, direction, summary_name, oracle_holdout = summary_value)

  oracle_obs <- derived_df %>%
    filter(fit_type == "oracle", state == "observed") %>%
    select(model_id, line_id, direction, summary_name, observed_oracle = summary_value)

  transfer_obs %>%
    left_join(transfer_holdout, by = c("model_id", "line_id", "direction", "summary_name")) %>%
    left_join(oracle_holdout, by = c("model_id", "line_id", "direction", "summary_name")) %>%
    left_join(oracle_obs, by = c("model_id", "line_id", "direction", "summary_name")) %>%
    mutate(
      null_holdout = observed_transfer,
      null_vs_oracle_diff = null_holdout - oracle_holdout,
      holdout_diff = transfer_holdout - oracle_holdout,
      abs_null_err = pmax(abs(null_vs_oracle_diff), 1e-12),
      abs_transfer_err = pmax(abs(holdout_diff), 1e-12),
      log10_error_ratio = log10(abs_transfer_err / abs_null_err),
      log_error_improvement = log(abs_null_err) - log(abs_transfer_err),
      transfer_better_than_null = abs_transfer_err < abs_null_err,
      oracle_shift = oracle_holdout - observed_oracle,
      transfer_shift = transfer_holdout - observed_transfer,
      shift_diff = transfer_shift - oracle_shift
    )
}

summarize_derived_transfer <- function(derived_cmp_df) {
  if (!nrow(derived_cmp_df)) {
    return(tibble(
      model_id = character(),
      direction = character(),
      summary_name = character(),
      mean_log_error_improvement = numeric(),
      median_log10_error_ratio = numeric(),
      mean_win_rate = numeric(),
      prop_lines_positive = numeric(),
      mean_abs_shift_diff = numeric(),
      n_line_cases = integer()
    ))
  }

  derived_cmp_df %>%
    group_by(model_id, direction, summary_name) %>%
    summarise(
      mean_log_error_improvement = mean(log_error_improvement, na.rm = TRUE),
      median_log10_error_ratio = median(log10_error_ratio, na.rm = TRUE),
      mean_win_rate = mean(transfer_better_than_null, na.rm = TRUE),
      prop_lines_positive = mean(log_error_improvement > 0, na.rm = TRUE),
      mean_abs_shift_diff = mean(abs(shift_diff), na.rm = TRUE),
      n_line_cases = dplyr::n(),
      .groups = "drop"
    )
}

rank_derived_transfer <- function(derived_summary_df) {
  if (!nrow(derived_summary_df)) {
    return(tibble(
      direction = character(),
      summary_name = character(),
      mean_log_error_improvement = numeric(),
      mean_win_rate = numeric(),
      prop_models_positive = numeric(),
      mean_abs_shift_diff = numeric(),
      n_models = integer()
    ))
  }

  derived_summary_df %>%
    group_by(direction, summary_name) %>%
    summarise(
      mean_log_error_improvement = mean(mean_log_error_improvement, na.rm = TRUE),
      mean_win_rate = mean(mean_win_rate, na.rm = TRUE),
      prop_models_positive = mean(mean_log_error_improvement > 0, na.rm = TRUE),
      mean_abs_shift_diff = mean(mean_abs_shift_diff, na.rm = TRUE),
      n_models = dplyr::n(),
      .groups = "drop"
    ) %>%
    arrange(direction, desc(mean_log_error_improvement), desc(mean_win_rate))
}

compute_point_loglik_subset <- function(draws, well_idx) {
  ll <- get_well_loglik_draws(draws, well_idx)
  if (!length(ll)) {
    return(NA_real_)
  }
  as.numeric(ll[1])
}

build_transfer_subset_diagnostics <- function(
  model_ids,
  output_root = project_paths$transfer_optim_root,
  stan_data_path = project_paths$stan_data
) {
  if (!length(model_ids)) {
    return(tibble(
      model_id = character(),
      line_id = integer(),
      direction = character(),
      fit_type = character(),
      return_code = numeric(),
      lp__ = numeric(),
      ll_transfer_train = numeric(),
      ll_transfer_holdout = numeric(),
      n_transfer_train_wells = integer(),
      n_transfer_holdout_wells = integer()
    ))
  }

  stan_data <- readRDS(resolve_stan_data_path(stan_data_path))
  all_wells <- seq_len(stan_data$N_wells)
  rows <- list()
  idx <- 1L

  for (model_id in model_ids) {
    best_path <- file.path(output_root, model_id, "transfer_best_start_summary.Rds")
    if (!file.exists(best_path)) {
      next
    }

    best_df <- readRDS(best_path)
    keys <- unique(best_df[, c("line_id", "direction")])

    for (k in seq_len(nrow(keys))) {
      line_id <- keys$line_id[k]
      direction <- keys$direction[k]
      split_meta <- get_directional_transfer_split(stan_data, line_id, direction)
      transfer_train_wells <- setdiff(all_wells, split_meta$holdout_wells)
      transfer_holdout_wells <- split_meta$holdout_wells

      for (fit_type in c("null", "transfer", "oracle")) {
        fit_obj <- tryCatch(
          load_transfer_best_fit(
            model_id = model_id,
            line_id = line_id,
            direction = direction,
            fit_type = fit_type,
            output_root = dirname(output_root),
            model_name = "gpath"
          ),
          error = function(e) NULL
        )

        if (is.null(fit_obj)) {
          next
        }

        rows[[idx]] <- data.frame(
          model_id = model_id,
          line_id = as.integer(line_id),
          direction = direction,
          fit_type = fit_type,
          return_code = as.numeric(fit_obj$summary$return_code[[1]]),
          lp__ = as.numeric(fit_obj$summary$lp__[[1]]),
          ll_transfer_train = compute_point_loglik_subset(fit_obj$draws, transfer_train_wells),
          ll_transfer_holdout = compute_point_loglik_subset(fit_obj$draws, transfer_holdout_wells),
          n_transfer_train_wells = length(transfer_train_wells),
          n_transfer_holdout_wells = length(transfer_holdout_wells),
          stringsAsFactors = FALSE
        )
        idx <- idx + 1L
      }
    }
  }

  if (!length(rows)) {
    return(tibble(
      model_id = character(),
      line_id = integer(),
      direction = character(),
      fit_type = character(),
      return_code = numeric(),
      lp__ = numeric(),
      ll_transfer_train = numeric(),
      ll_transfer_holdout = numeric(),
      n_transfer_train_wells = integer(),
      n_transfer_holdout_wells = integer()
    ))
  }

  bind_rows(rows)
}

summarize_transfer_subset_diagnostics <- function(diag_df) {
  if (!nrow(diag_df)) {
    return(tibble(
      model_id = character(),
      line_id = integer(),
      direction = character(),
      transfer_minus_oracle_train = numeric(),
      transfer_minus_null_train = numeric(),
      transfer_minus_oracle_holdout = numeric(),
      transfer_minus_null_holdout = numeric()
    ))
  }

  wide <- diag_df %>%
    select(model_id, line_id, direction, fit_type, ll_transfer_train, ll_transfer_holdout) %>%
    tidyr::pivot_wider(
      names_from = fit_type,
      values_from = c(ll_transfer_train, ll_transfer_holdout)
    )

  wide %>%
    transmute(
      model_id,
      line_id,
      direction,
      transfer_minus_oracle_train = ll_transfer_train_transfer - ll_transfer_train_oracle,
      transfer_minus_null_train = ll_transfer_train_transfer - ll_transfer_train_null,
      transfer_minus_oracle_holdout = ll_transfer_holdout_transfer - ll_transfer_holdout_oracle,
      transfer_minus_null_holdout = ll_transfer_holdout_transfer - ll_transfer_holdout_null
    )
}
```

## Upstream Commands

This notebook depends on artefacts created elsewhere. The table below documents the current known generation path.

```{r upstream-commands}
artifact_commands <- tribble(
  ~artifact_group, ~path, ~command, ~status_note,
  "Stan data", project_paths$stan_data,
  "Exact historical command not recorded in repo. Data are constructed by R/prepare_data.R.",
  "Known script, command not fixed in repo history.",
  "Assessment model set", project_paths$assessment_run_ids,
  "Tracked file read by scripts/make_gpath_optim_spec.R when generating rerun specs.",
  "Canonical model-id set for current `gpath v1` assessment.",
  "Versioned hierarchical optimization", "data/runs/gpath/v1/optim/<DATASET_LABEL>/<RUN_ID>/<RUN_LABEL>",
  "bash slurm/submit_gpath_optim_pipeline.sh <SPEC_FILE.tsv> ; specs can be generated with Rscript scripts/make_gpath_optim_spec.R <SPEC_FILE.tsv> <RUN_IDS>",
  "Trusted optimisation source. This notebook now ignores older non-manifest results.",
  "Versioned hierarchical NUTS", "not yet migrated",
  "No new versioned NUTS pipeline for legacy hierarchical fits yet.",
  "Transfer NUTS still seeds from the older legacy NUTS outputs for now.",
  "Directional transfer (optimization)", "data/gpath_transfer_cv/gpath/<RUN_ID>",
  "bash slurm/submit_gpath_transfer_pipeline.sh MODEL R P W C M all low_to_high,high_to_low null,transfer,oracle 1:K 500 1000 0.99 12 16 ecology/stan_ready_data.Rds data/gpath_transfer_cv 0",
  "Current optimization transfer path.",
  "Directional transfer (optimization summaries)", "data/gpath_transfer_cv/gpath/<RUN_ID>/*.Rds",
  "Rscript scripts/summarize_gpath_transfer_cv.R gpath <RUN_ID> data/gpath_transfer_cv",
  "Current summary path.",
  "Directional transfer (NUTS)", "data/gpath_transfer_cv_nuts/gpath/<RUN_ID>",
  "bash slurm/submit_gpath_transfer_nuts_pipeline.sh MODEL R P W C M all low_to_high,high_to_low null,transfer,oracle 1:4 500 1000 0.99 12 16 ecology/stan_ready_data.Rds data/gpath_transfer_cv_nuts",
  "Current NUTS transfer path.",
  "Directional transfer (NUTS summaries)", "data/gpath_transfer_cv_nuts/gpath/<RUN_ID>/*.Rds",
  "Rscript scripts/summarize_gpath_transfer_nuts.R gpath <RUN_ID> data/gpath_transfer_cv_nuts",
  "Current summary path."
)

knitr::kable(artifact_commands)
```

## Available Outputs

```{r available-outputs}
ranking_df <- build_model_ranking()
manifest_df <- summarize_manifest_status(ranking_df)
selected_model_id <- find_selected_model(ranking_df)
raw_theta_ploidy_global_df <- collect_best_raw_theta_ploidy_global(ranking_df)
raw_theta_ploidy_transfer_df <- collect_best_raw_theta_ploidy_transfer(ranking_df)
raw_theta_ploidy_df <- bind_rows(raw_theta_ploidy_global_df, raw_theta_ploidy_transfer_df)
common_raw_theta_parameters <- raw_theta_ploidy_global_df %>%
  filter(!missing) %>%
  dplyr::count(parameter, name = "n_models_present") %>%
  filter(n_models_present == n_distinct(raw_theta_ploidy_global_df$model_id)) %>%
  pull(parameter)

cat("Available trusted gpath model IDs:\n")
print(ranking_df$model_id)
cat("\nSelected model for downstream reporting:", selected_model_id, "\n")
```

## Trusted Optimisation Runs

```{r trusted-optimisation-runs}
if (!nrow(manifest_df)) {
  announce_missing("No manifest-backed optimisation outputs found under %s", project_paths$optim_root)
} else {
  knitr::kable(manifest_df)
}
```

## Best-Fit `raw_theta_ploidy` Coefficients

This table loads the `raw_theta_ploidy` vector from the best optimization start for each ranked `model_id` and relabels each coefficient with the corresponding model parameter.

```{r best-raw-theta-ploidy}
if (!nrow(raw_theta_ploidy_df)) {
  announce_missing("No best-fit raw_theta_ploidy coefficients could be extracted from ranked optimisation runs.")
} else {
  raw_theta_ploidy_df %>%
    select(source, model_id, alias, line_id, direction, fit_type, parameter, raw_theta_ploidy, lp__, elpd, ll_transfer_holdout, missing, run_label, run_tag) %>%
    arrange(source, model_id, line_id, direction, fit_type, parameter) %>%
    print(n = Inf)
}
```

## Leave-One-Line-Out `raw_theta_ploidy` Fits

These rows are the best `transfer` fits for each held-out line and direction, with the corresponding `raw_theta_ploidy` values and fit scores attached.

```{r transfer-raw-theta-ploidy-table}
if (!nrow(raw_theta_ploidy_transfer_df)) {
  announce_missing("No leave-one-line-out transfer best-fit raw_theta_ploidy coefficients found under %s", project_paths$transfer_optim_root)
} else {
  raw_theta_ploidy_transfer_df %>%
    filter(fit_type == "transfer") %>%
    select(model_id, alias, line_id, direction, fit_type, parameter, raw_theta_ploidy, lp__, elpd, ll_transfer_train, ll_transfer_holdout, holdout_obs, run_tag) %>%
    arrange(model_id, line_id, direction, fit_type, parameter) %>%
    print(n = Inf)
}
```

```{r transfer-raw-theta-ploidy-plot, fig.height=12, fig.width=14}
if (!nrow(raw_theta_ploidy_transfer_df)) {
  announce_missing("No leave-one-line-out transfer best-fit raw_theta_ploidy coefficients available for plotting.")
} else {
  transfer_plot_df <- raw_theta_ploidy_transfer_df %>%
    filter(fit_type == "transfer", parameter %in% common_raw_theta_parameters, !missing)

  if (!nrow(transfer_plot_df)) {
    announce_missing("No transfer-fit raw_theta_ploidy coefficients available for common parameters.")
  } else {
  global_plot_df <- raw_theta_ploidy_global_df %>%
    filter(parameter %in% common_raw_theta_parameters, !missing) %>%
    transmute(
      model_id,
      alias,
      line_id = NA_integer_,
      parameter,
      raw_theta_ploidy
    )

  transfer_plot_df <- transfer_plot_df %>%
    transmute(
      model_id,
      alias,
      line_id,
      line_label = sprintf("line %d", line_id),
      direction,
      parameter,
      raw_theta_ploidy
    )

  line_levels <- sprintf("line %d", sort(unique(transfer_plot_df$line_id)))

  global_repeated_df <- tidyr::expand_grid(
    global_plot_df,
    line_label = line_levels
  )

  transfer_with_global_df <- transfer_plot_df %>%
    left_join(
      global_plot_df %>%
        select(model_id, parameter, global_raw_theta_ploidy = raw_theta_ploidy),
      by = c("model_id", "parameter")
    )

  line_df <- transfer_with_global_df %>%
    filter(!is.na(global_raw_theta_ploidy)) %>%
    transmute(
      parameter,
      model_id,
      line_label,
      direction,
      x = model_id,
      y = global_raw_theta_ploidy,
      xend = model_id,
      yend = raw_theta_ploidy
    )

  plot_df <- bind_rows(
    global_repeated_df %>% mutate(point_type = "global", direction = "global"),
    transfer_plot_df %>% mutate(point_type = "transfer")
  ) %>%
    mutate(
      direction = factor(direction, levels = c("global", "low_to_high", "high_to_low")),
      model_id = factor(model_id, levels = ranking_df$model_id),
      line_label = factor(
        line_label,
        levels = line_levels
      ),
      parameter = factor(parameter, levels = common_raw_theta_parameters)
    )

  ggplot(
    plot_df,
    aes(x = model_id, y = raw_theta_ploidy)
  ) +
    geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
    geom_segment(
      data = line_df,
      aes(x = x, y = y, xend = xend, yend = yend, color = direction),
      inherit.aes = FALSE,
      linewidth = 0.5,
      alpha = 0.45
    ) +
    geom_point(
      data = plot_df %>% filter(direction == "global"),
      shape = 21,
      size = 2.8,
      stroke = 0.7,
      fill = "white",
      color = "black"
    ) +
    geom_point(
      data = plot_df %>% filter(direction != "global"),
      aes(color = direction),
      position = position_jitter(width = 0.08, height = 0),
      alpha = 0.8,
      size = 2.2
    ) +
    facet_grid(parameter ~ line_label, scales = "free_y") +
    scale_color_manual(values = c(low_to_high = "#1b9e77", high_to_low = "#d95f02")) +
    theme_bw() +
    labs(
      x = "",
      y = "Best-fit raw_theta_ploidy",
      color = "Direction"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )
  }
}
```

## Model Selection

```{r model-selection-table}
if (!nrow(ranking_df)) {
  announce_missing("No trusted optimisation outputs found under %s", project_paths$optim_root)
} else {
  knitr::kable(ranking_df %>% select(model_id, alias, best_lp, k, AIC, delta_AIC))
}
```

```{r model-selection-plot}
if (!nrow(ranking_df)) {
  announce_missing("Skipping model-selection plot because no ranking table was built.")
} else {
  pareto_front <- ranking_df %>%
    arrange(k, deviance) %>%
    filter(deviance == cummin(deviance)) %>%
    group_by(deviance) %>%
    slice_min(order_by = k, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(k)

  p_aic <- ggplot(ranking_df, aes(x = reorder(alias, delta_AIC), y = delta_AIC)) +
    geom_col(fill = "grey35") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = expression(Delta ~ AIC))

  p_pareto <- ggplot(ranking_df, aes(x = k, y = deviance)) +
    geom_point(color = "grey60", size = 2.5) +
    geom_line(data = pareto_front, color = "firebrick", linetype = "dashed") +
    geom_point(data = pareto_front, color = "firebrick", size = 3) +
    theme_bw() +
    labs(x = "Effective parameters (k)", y = "Deviance")

  p_aic + p_pareto
}
```

## Optimisation Stability Across Trusted Models

```{r optimisation-stability-all-models}
if (!nrow(ranking_df)) {
  announce_missing("No trusted optimisation outputs available for stability summary.")
} else {
  stability_df <- purrr::map_dfr(seq_len(nrow(ranking_df)), function(i) {
    run_dir <- ranking_df$run_dir[i]
    lp <- readRDS(file.path(run_dir, "optim_lp_all.Rds"))
    nbhd <- summarize_lp_neighborhood(lp, near_cut = 5)

    tibble(
      model_id = ranking_df$model_id[i],
      alias = ranking_df$alias[i],
      run_label = ranking_df$run_label[i],
      n_total = length(lp),
      n_finite = sum(nbhd$finite_mask),
      best_lp = nbhd$best_lp,
      n_within_1 = sum(nbhd$finite_mask & nbhd$lp_diff < 1),
      n_within_5 = sum(nbhd$finite_mask & nbhd$lp_diff < 5),
      n_within_10 = sum(nbhd$finite_mask & nbhd$lp_diff < 10),
      prop_within_1 = mean(nbhd$finite_mask & nbhd$lp_diff < 1),
      prop_within_5 = mean(nbhd$finite_mask & nbhd$lp_diff < 5),
      prop_within_10 = mean(nbhd$finite_mask & nbhd$lp_diff < 10),
      delta_AIC = ranking_df$delta_AIC[i]
    )
  }) %>%
    arrange(desc(prop_within_5), desc(prop_within_1), delta_AIC)

  knitr::kable(stability_df)
}
```

## Transfer Analysis: Predictive View

This section treats transfer as a predictive benchmark. It is useful, but the parameter-transfer section below is now closer to the biological question of interest.

```{r transfer-screening-setup}
screen_transfer_df <- collect_transfer_screening(ranking_df$model_id)
screen_parameter_df <- collect_parameter_screening(ranking_df$model_id)
focused_parameter_df <- collect_focused_parameter_transfer(ranking_df$model_id)
family_parameter_df <- collect_parameter_family_transfer(ranking_df$model_id)
family_parameter_summary_df <- summarize_parameter_family_transfer(family_parameter_df)
model_family_summary_df <- summarize_parameter_family_across_models(family_parameter_summary_df)
family_ranking_df <- rank_parameter_families(model_family_summary_df)
derived_transfer_df <- collect_derived_transfer_summaries(ranking_df$model_id)
derived_transfer_cmp_df <- build_derived_transfer_comparison(derived_transfer_df)
derived_transfer_summary_df <- summarize_derived_transfer(derived_transfer_cmp_df)
derived_transfer_ranking_df <- rank_derived_transfer(derived_transfer_summary_df)
transfer_subset_diag_df <- build_transfer_subset_diagnostics(ranking_df$model_id)
transfer_subset_summary_df <- summarize_transfer_subset_diagnostics(transfer_subset_diag_df)

predictive_screen_summary <- summarize_predictive_screening(screen_transfer_df)
parameter_screen_summary <- summarize_parameter_screening(screen_parameter_df)
```

## Transfer Screening Across Models

```{r screening-plots, fig.height=10, fig.width=12}
if (!nrow(screen_transfer_df) || !nrow(screen_parameter_df)) {
  announce_missing("Need both predictive and parameter-transfer screening outputs to draw compact model-screening plots.")
} else {
  screen_parameter_df <- screen_parameter_df %>%
    mutate(
      mean_log_error_improvement = mean_abs_log_null_holdout_ratio - mean_abs_log_holdout_ratio,
      direction = factor(direction, levels = c("low_to_high", "high_to_low"))
    )

  p_pred <- ggplot(
    screen_transfer_df,
    aes(x = reorder(model_id, transfer_gain, FUN = mean), y = transfer_gain, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Predictive transfer gain over null")

  p_param_log <- ggplot(
    screen_parameter_df,
    aes(x = reorder(model_id, mean_log_error_improvement, FUN = mean), y = mean_log_error_improvement, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Mean log-error improvement over null")

  p_param_win <- ggplot(
    screen_parameter_df,
    aes(x = reorder(model_id, prop_parameters_transfer_better_than_null, FUN = mean), y = prop_parameters_transfer_better_than_null, color = direction)
  ) +
    geom_hline(yintercept = 0.5, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    labs(x = "", y = "Proportion of parameters transfer beats null")

  (p_pred | p_param_log) / p_param_win
  }
```

## Transfer Fit Diagnostics On Common Well Subsets

These diagnostics evaluate each best `null`, `transfer`, and `oracle` fit on the same transfer-defined well subsets, rather than comparing each fit on its own training set.

```{r transfer-subset-diagnostics, fig.height=10, fig.width=12}
if (!nrow(transfer_subset_summary_df)) {
  announce_missing("No common-subset transfer diagnostics could be built from saved best-start fits.")
} else {
  diag_df <- transfer_subset_summary_df %>%
    mutate(direction = factor(direction, levels = c("low_to_high", "high_to_low")))

  p1 <- ggplot(
    diag_df,
    aes(x = reorder(model_id, transfer_minus_oracle_train, FUN = mean), y = transfer_minus_oracle_train, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Transfer - Oracle on transfer-training wells")

  p2 <- ggplot(
    diag_df,
    aes(x = reorder(model_id, transfer_minus_null_train, FUN = mean), y = transfer_minus_null_train, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Transfer - Null on transfer-training wells")

  p3 <- ggplot(
    diag_df,
    aes(x = reorder(model_id, transfer_minus_oracle_holdout, FUN = mean), y = transfer_minus_oracle_holdout, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Transfer - Oracle on transfer-holdout wells")

  p4 <- ggplot(
    diag_df,
    aes(x = reorder(model_id, transfer_minus_null_holdout, FUN = mean), y = transfer_minus_null_holdout, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Transfer - Null on transfer-holdout wells")

  (p1 | p2) / (p3 | p4)
}
```

```{r transfer-direction-asymmetry, fig.height=6, fig.width=10}
if (!nrow(transfer_subset_summary_df)) {
  announce_missing("No common-subset transfer diagnostics available for direction-asymmetry plot.")
} else {
  asym_df <- transfer_subset_summary_df %>%
    select(model_id, line_id, direction, transfer_minus_oracle_train, transfer_minus_null_train) %>%
    tidyr::pivot_wider(
      names_from = direction,
      values_from = c(transfer_minus_oracle_train, transfer_minus_null_train)
    ) %>%
    mutate(
      asym_transfer_minus_oracle_train =
        transfer_minus_oracle_train_low_to_high - transfer_minus_oracle_train_high_to_low,
      asym_transfer_minus_null_train =
        transfer_minus_null_train_low_to_high - transfer_minus_null_train_high_to_low
    )

  p1 <- ggplot(
    asym_df,
    aes(x = reorder(model_id, asym_transfer_minus_oracle_train, FUN = mean), y = asym_transfer_minus_oracle_train)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2, color = "#1b9e77") +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Low->High minus High->Low\nfor Transfer - Oracle on training wells")

  p2 <- ggplot(
    asym_df,
    aes(x = reorder(model_id, asym_transfer_minus_null_train, FUN = mean), y = asym_transfer_minus_null_train)
  ) +
    geom_hline(yintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2, color = "#d95f02") +
    stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
    coord_flip() +
    theme_bw() +
    labs(x = "", y = "Low->High minus High->Low\nfor Transfer - Null on training wells")

  p1 | p2
}
```

```{r transfer-optim-summary}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for transfer summary.")
} else {
  optim_transfer <- load_transfer_outputs(selected_model_id, output_root = project_paths$transfer_optim_root)

  if (is.null(optim_transfer$transfer_summary)) {
    announce_missing("No optimisation transfer summary found for %s", selected_model_id)
  } else {
    optim_transfer$transfer_summary %>%
      select(line_id, direction, transfer_gain, transfer_regret, transfer, null, oracle) %>%
      arrange(direction, desc(transfer_gain)) %>%
      print()
  }
}
```

```{r transfer-nuts-summary}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for NUTS transfer summary.")
} else {
  nuts_summary_path <- file.path(project_paths$transfer_nuts_root, selected_model_id, "transfer_comparison_summary.Rds")
  nuts_transfer <- safe_read_rds(nuts_summary_path)

  if (is.null(nuts_transfer)) {
    announce_missing("No NUTS transfer summary found at %s", nuts_summary_path)
  } else {
    cat("NUTS transfer summary\n")
    print(nuts_transfer)
  }
}
```

## Focused Common Parameters

These plots break out the four parameters that are present across the ranked model set: `ae[1]`, `ah[1]`, `Y_R[1]`, and `m`. The first view uses the current absolute-difference criterion against the oracle. The second uses multiplicative discrepancy, via absolute log-ratios to the oracle, on the reconstructed held-out parameters.

```{r focused-parameter-table}
if (!nrow(focused_parameter_df)) {
  announce_missing("No focused parameter-transfer comparison rows found for the common parameters.")
} else {
  focused_parameter_df %>%
    select(
      model_id,
      parameter,
      line_id,
      direction,
      null_holdout,
      transfer_holdout,
      oracle_holdout,
      abs_null_err,
      abs_transfer_err,
      abs_error_improvement,
      abs_log_null_ratio,
      abs_log_transfer_ratio,
      ratio_error_improvement
    ) %>%
    arrange(parameter, model_id, line_id, direction) %>%
    print(n = Inf)
}
```

```{r focused-parameter-absolute, fig.height=10, fig.width=14}
if (!nrow(focused_parameter_df)) {
  announce_missing("No focused parameter-transfer rows available for absolute-error plotting.")
} else {
  plot_df <- focused_parameter_df %>%
    mutate(
      parameter = factor(parameter, levels = focus_common_parameters),
      direction = factor(direction, levels = c("low_to_high", "high_to_low")),
      model_id = factor(model_id, levels = ranking_df$model_id),
      fit_target = factor("oracle", levels = "oracle")
    )

  p1 <- ggplot(
    plot_df,
    aes(x = model_id)
  ) +
    geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
    geom_segment(
      aes(
        xend = model_id,
        y = abs_null_err,
        yend = abs_transfer_err,
        color = direction
      ),
      linewidth = 0.5,
      alpha = 0.45
    ) +
    geom_point(aes(y = abs_null_err), shape = 21, size = 2.4, stroke = 0.7, fill = "white", color = "black") +
    geom_point(aes(y = abs_transfer_err, color = direction), size = 2.2, alpha = 0.85) +
    facet_grid(parameter ~ line_direction, scales = "free_y") +
    scale_color_manual(values = c(low_to_high = "#1b9e77", high_to_low = "#d95f02")) +
    theme_bw() +
    labs(
      x = "",
      y = "|fit - oracle| on reconstructed held-out parameter",
      color = "Direction"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )

  p2 <- ggplot(
    plot_df,
    aes(x = model_id, y = abs_error_improvement, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.08, height = 0), size = 2.1, alpha = 0.85) +
    facet_grid(parameter ~ line_direction, scales = "free_y") +
    scale_color_manual(values = c(low_to_high = "#1b9e77", high_to_low = "#d95f02")) +
    theme_bw() +
    labs(
      x = "",
      y = "|null-oracle| - |transfer-oracle|",
      color = "Direction"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )

  p1 / p2
}
```

```{r focused-parameter-ratio, fig.height=10, fig.width=14}
if (!nrow(focused_parameter_df %>% filter(!is.na(abs_log_null_ratio), !is.na(abs_log_transfer_ratio)))) {
  announce_missing("No positive-valued focused parameter rows available for ratio-based plotting.")
} else {
  plot_df <- focused_parameter_df %>%
    filter(!is.na(abs_log_null_ratio), !is.na(abs_log_transfer_ratio)) %>%
    mutate(
      parameter = factor(parameter, levels = focus_common_parameters),
      direction = factor(direction, levels = c("low_to_high", "high_to_low")),
      model_id = factor(model_id, levels = ranking_df$model_id)
    )

  p1 <- ggplot(
    plot_df,
    aes(x = model_id)
  ) +
    geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
    geom_segment(
      aes(
        xend = model_id,
        y = abs_log_null_ratio,
        yend = abs_log_transfer_ratio,
        color = direction
      ),
      linewidth = 0.5,
      alpha = 0.45
    ) +
    geom_point(aes(y = abs_log_null_ratio), shape = 21, size = 2.4, stroke = 0.7, fill = "white", color = "black") +
    geom_point(aes(y = abs_log_transfer_ratio, color = direction), size = 2.2, alpha = 0.85) +
    facet_grid(parameter ~ line_direction, scales = "free_y") +
    scale_color_manual(values = c(low_to_high = "#1b9e77", high_to_low = "#d95f02")) +
    theme_bw() +
    labs(
      x = "",
      y = "|log(fit / oracle)| on reconstructed held-out parameter",
      color = "Direction"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )

  p2 <- ggplot(
    plot_df,
    aes(x = model_id, y = ratio_error_improvement, color = direction)
  ) +
    geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
    geom_point(position = position_jitter(width = 0.08, height = 0), size = 2.1, alpha = 0.85) +
    facet_grid(parameter ~ line_direction, scales = "free_y") +
    scale_color_manual(values = c(low_to_high = "#1b9e77", high_to_low = "#d95f02")) +
    theme_bw() +
    labs(
      x = "",
      y = "|log(null/oracle)| - |log(transfer/oracle)|",
      color = "Direction"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )

  p1 / p2
}
```

## Cross-Model Family Transfer

This section asks a narrower question than overall model transfer: whether particular parameter families repeatedly move closer to the oracle than the null baseline does across models, lines, and transfer directions.

```{r parameter-family-ranking, fig.height=8, fig.width=12}
if (!nrow(model_family_summary_df)) {
  announce_missing("No parameter-comparison outputs found for family-level transfer analysis.")
} else {
  plot_df <- family_ranking_df %>%
    mutate(
      direction = factor(direction, levels = c("low_to_high", "high_to_low")),
      family = reorder(family, mean_log_error_improvement)
    )

  p1 <- ggplot(plot_df, aes(x = family, y = mean_log_error_improvement, color = direction)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_point(size = 3) +
    coord_flip() +
    facet_wrap(~ direction, scales = "free_y") +
    theme_bw() +
    labs(x = "", y = "Cross-model mean log-error improvement")

  p2 <- ggplot(plot_df, aes(x = family, y = mean_win_rate, color = direction)) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey70") +
    geom_point(size = 3) +
    coord_flip() +
    facet_wrap(~ direction, scales = "free_y") +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    labs(x = "", y = "Cross-model mean family win rate")

  p1 | p2
}
```

```{r parameter-family-consistency-heatmaps, fig.height=10, fig.width=12}
if (!nrow(model_family_summary_df)) {
  announce_missing("No parameter-comparison outputs found for family-level heatmaps.")
} else {
  heat_df <- model_family_summary_df %>%
    mutate(direction = factor(direction, levels = c("low_to_high", "high_to_low")))

  p1 <- ggplot(heat_df, aes(x = family, y = model_id, fill = mean_family_log_error_improvement)) +
    geom_tile(color = "white") +
    facet_wrap(~ direction) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0) +
    theme_minimal() +
    labs(
      x = "Parameter family",
      y = "Model ID",
      fill = "Mean log-error\nimprovement"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p2 <- ggplot(heat_df, aes(x = family, y = model_id, fill = mean_family_win_rate)) +
    geom_tile(color = "white") +
    facet_wrap(~ direction) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0.5, limits = c(0, 1)) +
    theme_minimal() +
    labs(
      x = "Parameter family",
      y = "Model ID",
      fill = "Mean family\nwin rate"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p1 / p2
}
```

```{r parameter-family-line-heatmaps, fig.height=10, fig.width=12}
if (!nrow(family_parameter_summary_df)) {
  announce_missing("No parameter-comparison outputs found for line-level family heatmaps.")
} else {
  top_families <- family_ranking_df %>%
    group_by(family) %>%
    summarise(
      mean_log_error_improvement = mean(mean_log_error_improvement, na.rm = TRUE),
      mean_win_rate = mean(mean_win_rate, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_log_error_improvement), desc(mean_win_rate)) %>%
    slice_head(n = 4) %>%
    pull(family)

  line_heat_df <- family_parameter_summary_df %>%
    filter(family %in% top_families) %>%
    mutate(
      line_direction = sprintf("line %d | %s", line_id, direction),
      family = factor(family, levels = top_families)
    )

  p1 <- ggplot(line_heat_df, aes(x = family, y = line_direction, fill = mean_log_error_improvement)) +
    geom_tile(color = "white") +
    facet_wrap(~ model_id) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0) +
    theme_minimal() +
    labs(
      x = "Top parameter families",
      y = "",
      fill = "Mean log-error\nimprovement"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p2 <- ggplot(line_heat_df, aes(x = family, y = line_direction, fill = prop_transfer_better_than_null)) +
    geom_tile(color = "white") +
    facet_wrap(~ model_id) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0.5, limits = c(0, 1)) +
    theme_minimal() +
    labs(
      x = "Top parameter families",
      y = "",
      fill = "Family\nwin rate"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p1 / p2
}
```

## Cross-Model Derived Glucose Summaries

These plots summarize five glucose-focused derived quantities reconstructed from the parameter states:
`Umax_1 = ae1`, `Khalf_1 = ah1`, `Efflow_1 = ae1 / ah1`, `YieldCap_1 = Y_R1 * ae1`, and `NetGrow_1 = Y_R1 * ae1 - m`.

```{r derived-transfer-ranking, fig.height=8, fig.width=12}
if (!nrow(derived_transfer_summary_df)) {
  announce_missing("No derived transfer summaries could be built from parameter-transfer states.")
} else {
  plot_df <- derived_transfer_ranking_df %>%
    mutate(direction = factor(direction, levels = c("low_to_high", "high_to_low")))

  p1 <- ggplot(plot_df, aes(x = reorder(summary_name, mean_log_error_improvement), y = mean_log_error_improvement, color = direction)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_point(size = 3) +
    coord_flip() +
    facet_wrap(~ direction, scales = "free_y") +
    theme_bw() +
    labs(x = "", y = "Cross-model mean log-error improvement")

  p2 <- ggplot(plot_df, aes(x = reorder(summary_name, mean_win_rate), y = mean_win_rate, color = direction)) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey70") +
    geom_point(size = 3) +
    coord_flip() +
    facet_wrap(~ direction, scales = "free_y") +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    labs(x = "", y = "Cross-model mean win rate")

  p1 | p2
}
```

```{r derived-transfer-heatmaps, fig.height=10, fig.width=12}
if (!nrow(derived_transfer_summary_df)) {
  announce_missing("No derived transfer summaries available for cross-model heatmaps.")
} else {
  heat_df <- derived_transfer_summary_df %>%
    mutate(direction = factor(direction, levels = c("low_to_high", "high_to_low")))

  p1 <- ggplot(heat_df, aes(x = summary_name, y = model_id, fill = mean_log_error_improvement)) +
    geom_tile(color = "white") +
    facet_wrap(~ direction) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0) +
    theme_minimal() +
    labs(
      x = "Derived glucose summary",
      y = "Model ID",
      fill = "Mean log-error\nimprovement"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p2 <- ggplot(heat_df, aes(x = summary_name, y = model_id, fill = mean_win_rate)) +
    geom_tile(color = "white") +
    facet_wrap(~ direction) +
    scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#1b9e77", midpoint = 0.5, limits = c(0, 1)) +
    theme_minimal() +
    labs(
      x = "Derived glucose summary",
      y = "Model ID",
      fill = "Mean\nwin rate"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(color = "grey85", fill = NA)
    )

  p1 / p2
}
```

```{r derived-transfer-selected-model, fig.height=9, fig.width=10}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for derived transfer plots.")
} else {
  selected_df <- derived_transfer_cmp_df %>%
    filter(model_id == selected_model_id) %>%
    mutate(direction = factor(direction, levels = c("low_to_high", "high_to_low")))

  if (!nrow(selected_df)) {
    announce_missing("No derived transfer comparison rows found for %s", selected_model_id)
  } else {
    p1 <- ggplot(selected_df, aes(x = summary_name, y = log_error_improvement, color = direction)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
      geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.8, size = 2) +
      stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
      facet_wrap(~ line_id) +
      theme_bw() +
      labs(x = "Derived glucose summary", y = "Line-level log-error improvement") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

    p2 <- ggplot(selected_df, aes(x = summary_name, y = as.numeric(transfer_better_than_null), color = direction)) +
      geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey70") +
      geom_point(position = position_jitter(width = 0.15, height = 0.03), alpha = 0.8, size = 2) +
      stat_summary(fun = mean, geom = "point", shape = 95, size = 7, color = "black") +
      facet_wrap(~ line_id) +
      coord_cartesian(ylim = c(0, 1)) +
      theme_bw() +
      labs(x = "Derived glucose summary", y = "Transfer beats null") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

    p1 / p2
  }
}
```

## Transfer Analysis: Parameter View

This is currently the biologically primary transfer analysis. The null baseline is the observed-state parameterisation carried forward unchanged to the held-out ploidy state.

```{r parameter-transfer-summary}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer summary.")
} else {
  optim_transfer <- load_transfer_outputs(selected_model_id, output_root = project_paths$transfer_optim_root)

  if (is.null(optim_transfer$parameter_summary)) {
    announce_missing("No parameter transfer summary found for %s", selected_model_id)
  } else {
    optim_transfer$parameter_summary %>%
      mutate(
        mean_log_error_improvement = mean_abs_log_null_holdout_ratio - mean_abs_log_holdout_ratio
      ) %>%
      select(
        line_id,
        direction,
        mean_log_error_improvement,
        prop_parameters_transfer_better_than_null,
        mean_abs_log_holdout_ratio,
        mean_abs_log_null_holdout_ratio
      ) %>%
      arrange(direction, desc(mean_log_error_improvement)) %>%
      print()
  }
}
```

```{r parameter-transfer-plot-1, fig.height=6, fig.width=10}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer plots.")
} else {
  comparison_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_comparison.Rds")
  if (!file.exists(comparison_path)) {
    announce_missing("No parameter transfer comparison found at %s", comparison_path)
  } else {
    plot_parameter_transfer_improvement(
      model_id = selected_model_id,
      output_root = "data/gpath_transfer_cv",
      model_name = "gpath"
    )
  }
}
```

```{r parameter-transfer-plot-2, fig.height=8, fig.width=10}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer plots.")
} else {
    summary_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_summary.Rds")
    if (!file.exists(summary_path)) {
      announce_missing("No parameter transfer summary found at %s", summary_path)
    } else {
      plot_parameter_transfer_summary(
        model_id = selected_model_id,
        output_root = "data/gpath_transfer_cv",
        model_name = "gpath"
      )
    }
}
```

```{r parameter-transfer-plot-3, fig.height=9, fig.width=10}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for parameter transfer plots.")
} else {
    comparison_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_comparison.Rds")
    if (!file.exists(comparison_path)) {
      announce_missing("No parameter transfer comparison found at %s", comparison_path)
    } else {
      plot_parameter_transfer_family_summary(
        model_id = selected_model_id,
        output_root = "data/gpath_transfer_cv",
        model_name = "gpath"
      )
    }
}
```

## Transfer Fits For A Representative Line

This section overlays the best optimisation `null`, `transfer`, and `oracle` fits for one selected line. A line is chosen from the parameter-transfer summary by maximal mean log-error improvement over null; if that summary is absent, line 1 is used.

```{r transfer-overlay-plot, fig.height=6, fig.width=12}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for transfer overlay plot.")
} else {
  parameter_summary_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "parameter_transfer_summary.Rds")
  parameter_summary_df <- safe_read_rds(parameter_summary_path)

  target_line <- 1L
  if (!is.null(parameter_summary_df) && nrow(parameter_summary_df)) {
    score <- parameter_summary_df$mean_abs_log_null_holdout_ratio - parameter_summary_df$mean_abs_log_holdout_ratio
    target_line <- parameter_summary_df$line_id[which.max(score)]
  }

  best_start_path <- file.path(project_paths$transfer_optim_root, selected_model_id, "transfer_best_start_summary.Rds")
  if (!file.exists(best_start_path)) {
    announce_missing("No transfer best-start summary found at %s", best_start_path)
  } else {
    overlay_data <- generate_transfer_overlay_data(
      model_id = selected_model_id,
      line_id = target_line,
      output_root = "data/gpath_transfer_cv",
      stan_data_path = project_paths$stan_data
    )

    plot_transfer_line_trajectories(
      transfer_data = overlay_data,
      line_id = target_line,
      model_id = selected_model_id
    )
  }
}
```

## Current Interpretation

```{r interpretation-table}
if (is.na(selected_model_id)) {
  announce_missing("No selected model available for interpretation summary.")
} else {
  optim_transfer <- load_transfer_outputs(selected_model_id, output_root = project_paths$transfer_optim_root)

  if (is.null(optim_transfer$parameter_summary) || is.null(optim_transfer$transfer_summary)) {
    announce_missing("Need both predictive and parameter-transfer summaries to build the interpretation table.")
  } else {
    interpretation_df <- optim_transfer$parameter_summary %>%
      mutate(
        mean_log_error_improvement = mean_abs_log_null_holdout_ratio - mean_abs_log_holdout_ratio
      ) %>%
      left_join(
        optim_transfer$transfer_summary %>% rename(model_direction = direction),
        by = c("line_id" = "line_id", "direction" = "model_direction")
      ) %>%
      select(
        line_id,
        direction,
        transfer,
        null,
        oracle,
        transfer_gain,
        transfer_regret,
        mean_log_error_improvement,
        prop_parameters_transfer_better_than_null,
        mean_abs_log_holdout_ratio,
        mean_abs_log_null_holdout_ratio
      )

    print(interpretation_df)
  }
}
```

## Notes And Next Steps

At the time of writing, the modelling workflow supports four distinct questions:

1. Which `gpath` structure is best supported by the hierarchical optimisation fits?
2. Are selected models numerically stable enough to interpret?
3. Does a shared ploidy effect improve held-out prediction in directional transfer tests?
4. Does that same shared ploidy effect move held-out-line parameter values closer to the oracle fit than a null baseline does?

Recommended next steps:

1. Treat the selected `gpath` model as primary and stop comparing many model IDs throughout the downstream biological analysis.
2. Use the predictive transfer summaries as a benchmark, but make the parameter-transfer section the primary scientific result.
3. The current `oracle` transfer fits rerun the same all-data objective that the main global optimisation already solves when the model spec and ploidy mask are unchanged. That is redundant work and makes the workflow harder to follow; downstream comparisons should either reuse the global fit directly or state clearly why a separate all-data rerun is needed.
4. Add one compact comparison of optimisation-based versus NUTS-based transfer for the same selected model to quantify how much posterior averaging changes the conclusion.
5. If the parameter-transfer summaries are still hard to interpret, lean on the focused common-parameter plots and relative-error summaries before moving up to family-level averages.
6. Keep legacy or exploratory analysis blocks outside this notebook unless they are directly needed for the current `gpath` narrative.
