---
title: "Ecological Modelling"
author: "RJB"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(tibble)
  library(patchwork)
  library(tidyr)
  library(posterior)
})


plot_text_size <- 14
model_analysis_fig_dir <- file.path("dev-figs", "model_analysis")
dir.create(model_analysis_fig_dir, recursive = TRUE, showWarnings = FALSE)

theme_model_analysis <- function(...) {
  theme_bw(base_size = plot_text_size) +
    theme(
      panel.grid.minor = element_blank(),
      strip.background = element_rect(fill = "grey95"),
      ...
    )
}

focussed_parameter_alias_map <- c(
  "ae[1]" = "v1",
  "ah[1]" = "K1",
  "Y_R[1]" = "y1",
  "m" = "m"
)

apply_focussed_parameter_alias <- function(parameter, alias_map = focussed_parameter_alias_map) {
  out <- unname(alias_map[parameter])
  out[is.na(out) | !nzchar(out)] <- parameter[is.na(out) | !nzchar(out)]
  out
}

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = paste0(model_analysis_fig_dir, "/")
)
knitr::opts_knit$set(root.dir = normalizePath("../"))


```

```{r}
source("R/model_analysis_utils.R")
```


## Purpose

This notebook is now the reporting layer for the current `gpath` modelling workflow. It is not intended to recreate the full analysis from scratch. Instead it:

- documents the current modelling path,
- states the upstream scripts/commands that generate required artefacts where known,
- loads saved results from the current workflow,
- presents model selection, diagnostics, transfer results, and parameter-transfer summaries.

Where an upstream command is not known exactly from the repo history, that is stated explicitly.


Core datasets:


```{r}
source("R/validation_utils.R")
core_datasets <- c("reproducibility/gpath_v1_optim_batch1/")
hashes_check <- sapply(core_datasets,check_hashes)
spec_paths <- sapply(core_datasets,get_spec_path)

```
```{r summarize-optim-objects}
source("R/optim_utils.R")
optim_objects_path <- file.path("data", "report_exports", "model_analysis", "optim_objects_from_spec.Rds")
dir.create(dirname(optim_objects_path), recursive = TRUE, showWarnings = FALSE)

if (file.exists(optim_objects_path)) {
  x <- readRDS(optim_objects_path)
} else {
  x <- build_optim_objects_from_spec(spec_path = spec_paths[1])
  saveRDS(x, optim_objects_path)
}
```

Oracle NUTS note: this chunk now treats the full-data oracle job definition in [`config/jobs/gpath_full_oracle_assessment.json`](/share/lab_crd/lab_crd/HighPloidy_CostBenefits/data/GlucoseStarvation/config/jobs/gpath_full_oracle_assessment.json) as the source of truth for which model IDs should be checked and where their standard per-model NUTS output directories should live. The deployment path for that job is: materialize per-task configs with [`scripts/prepare_jobs.R`](/share/lab_crd/lab_crd/HighPloidy_CostBenefits/data/GlucoseStarvation/scripts/prepare_jobs.R), submit the resulting manifest with [`slurm/submit_gpath_full_oracle_job.sh`](/share/lab_crd/lab_crd/HighPloidy_CostBenefits/data/GlucoseStarvation/slurm/submit_gpath_full_oracle_job.sh), dispatch array tasks through [`slurm/jobs/gpath_full_oracle_array.sh`](/share/lab_crd/lab_crd/HighPloidy_CostBenefits/data/GlucoseStarvation/slurm/jobs/gpath_full_oracle_array.sh), and execute each task via the R entrypoint [`scripts/run_job.R`](/share/lab_crd/lab_crd/HighPloidy_CostBenefits/data/GlucoseStarvation/scripts/run_job.R), which sources [`R/job_config_utils.R`](/share/lab_crd/lab_crd/HighPloidy_CostBenefits/data/GlucoseStarvation/R/job_config_utils.R) and then calls the generic sampler in [`R/nuts_utils.R`](/share/lab_crd/lab_crd/HighPloidy_CostBenefits/data/GlucoseStarvation/R/nuts_utils.R).

```{r nuts-output-binding-diagnostic}
source("R/job_config_utils.R")
source("R/nuts_utils.R")

oracle_job_config_path <- "config/jobs/gpath_full_oracle_assessment.json"
oracle_job_info <- read_job_info(oracle_job_config_path)
diagnostic_roots <- oracle_job_info$output_root
oracle_subdir <- ""
rhat_threshold <- 1.01
ess_threshold <- 400

model_ids <- oracle_job_info$run_ids
available_model_ids <- model_ids[vapply(
  model_ids,
  function(model_id) dir.exists(get_job_output_dir(oracle_job_info$job_cfg, model_id)),
  logical(1)
)]

batch_summary_df <- dplyr::bind_rows(lapply(available_model_ids, function(model_id) {
  dir_path <- get_job_output_dir(oracle_job_info$job_cfg, model_id)
  available_run_tags <- nuts_list_run_tags(dir_path)
  bound_path <- file.path(dir_path, "bound_nuts_draws.Rds")
  diag_path <- file.path(dir_path, "nuts_chain_diagnostics.Rds")

  if (file.exists(bound_path) && file.exists(diag_path)) {
    bound_obj <- readRDS(bound_path)
    chain_diag_df <- readRDS(diag_path)
  } else {
    bound_obj <- nuts_bind_run(dir_path, run_tag = available_run_tags[[1]])
    chain_diag_df <- nuts_summarize_draw_diagnostics(bound_obj$draws)
    saveRDS(bound_obj, bound_path)
    saveRDS(chain_diag_df, diag_path)
  }
  draws_matrix <- posterior::as_draws_matrix(bound_obj$draws)

  tibble::tibble(
    model_id = model_id,
    selected_run_tag = if (!is.null(bound_obj$run_tag)) bound_obj$run_tag else available_run_tags[[1]],
    n_available_oracle_run_tags = length(available_run_tags),
    n_draws = nrow(draws_matrix),
    n_parameters = nrow(chain_diag_df),
    n_chains = length(bound_obj$chain_ids),
    n_bad_rhat = sum(chain_diag_df$rhat > rhat_threshold, na.rm = TRUE),
    n_bad_ess_bulk = sum(chain_diag_df$ess_bulk < ess_threshold, na.rm = TRUE),
    n_bad_ess_tail = sum(chain_diag_df$ess_tail < ess_threshold, na.rm = TRUE),
    max_rhat = max(chain_diag_df$rhat, na.rm = TRUE),
    min_ess_bulk = min(chain_diag_df$ess_bulk, na.rm = TRUE),
    min_ess_tail = min(chain_diag_df$ess_tail, na.rm = TRUE)
  )
}))

cat(sprintf(
  "Processed %d/%d model ids from %s\nOutput root: %s\nThresholds: rhat > %.2f, ess < %d\n",
  length(available_model_ids),
  length(model_ids),
  oracle_job_config_path,
  oracle_job_info$output_root,
  rhat_threshold,
  ess_threshold
))
print(as.data.frame(batch_summary_df), row.names = FALSE)
```

```{r nuts-instantaneous-growth-export}
source("R/project_paths.R")
source(get_model_r_path("gpath", "v1"))
source("R/gpath_run_utils.R")
source("R/nuts_consistency_utils.R")
regenerate <- TRUE
stan_data <- readRDS(resolve_stan_data_path())
line_ids <- sort(unique(as.integer(stan_data$line_id)))
glucose_grid <- exp(seq(log(0.0001), log(25), length.out = 25L))
resource2_grid <- exp(seq(log(0.0001), log(1), length.out = 25L))
state_grid <- expand.grid(
  glucose = glucose_grid,
  resource2 = resource2_grid,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

growth_export_summary_df <- dplyr::bind_rows(lapply(model_ids, function(model_id) {
  dir_path <- file.path(diagnostic_roots, model_id, oracle_subdir)
  bound_obj <- readRDS(file.path(dir_path, "bound_nuts_draws.Rds"))
  growth_draws_path <- file.path(dir_path, "nuts_instantaneous_growth_draws.Rds")
  growth_summary_path <- file.path(dir_path, "nuts_instantaneous_growth_summary.Rds")

  if (file.exists(growth_draws_path) && file.exists(growth_summary_path) &!regenerate) {
    instantaneous_growth_df <- readRDS(growth_draws_path)
    growth_summary_df <- readRDS(growth_summary_path)
  } else {
    draws_matrix <- posterior::as_draws_matrix(bound_obj$draws)
    draw_idx <- sort(sample.int(nrow(draws_matrix), size = min(20L, nrow(draws_matrix))))

    instantaneous_growth_df <- dplyr::bind_rows(lapply(line_ids, function(line_id) {
      states <- get_line_ploidy_states(stan_data, line_id = line_id)

      dplyr::bind_rows(lapply(draw_idx, function(i_draw) {
        draw_vec <- draws_matrix[i_draw, , drop = TRUE]
        names(draw_vec) <- colnames(draws_matrix)

        parms_low <- reconstruct_draw_parameters(
          draw_vec = draw_vec,
          model_id = model_id,
          line_id = line_id,
          ploidy_metric = states$low_value
        )
        parms_high <- reconstruct_draw_parameters(
          draw_vec = draw_vec,
          model_id = model_id,
          line_id = line_id,
          ploidy_metric = states$high_value
        )

        low_growth <- vapply(
          seq_len(nrow(state_grid)),
          function(i) compute_instantaneous_net_growth_with_resource2(
            parms_low,
            glucose = state_grid$glucose[[i]],
            resource2_value = state_grid$resource2[[i]],
            extra_resource_value = 1.0
          ),
          numeric(1)
        )
        high_growth <- vapply(
          seq_len(nrow(state_grid)),
          function(i) compute_instantaneous_net_growth_with_resource2(
            parms_high,
            glucose = state_grid$glucose[[i]],
            resource2_value = state_grid$resource2[[i]],
            extra_resource_value = 1.0
          ),
          numeric(1)
        )

        tibble::tibble(
          model_id = model_id,
          line_id = line_id,
          .draw = i_draw,
          glucose = state_grid$glucose,
          resource2 = state_grid$resource2,
          low_ploidy_growth = low_growth,
          high_ploidy_growth = high_growth,
          delta_growth = high_growth - low_growth
        )
      }))
    }))

    growth_summary_df <- instantaneous_growth_df %>%
      dplyr::group_by(.data$model_id, .data$line_id, .data$glucose, .data$resource2) %>%
      dplyr::summarise(
        n_draws = dplyr::n(),
        median_delta_growth = stats::median(.data$delta_growth, na.rm = TRUE),
        q025_delta_growth = stats::quantile(.data$delta_growth, probs = 0.025, na.rm = TRUE, names = FALSE),
        q975_delta_growth = stats::quantile(.data$delta_growth, probs = 0.975, na.rm = TRUE, names = FALSE),
        .groups = "drop"
      )

    saveRDS(instantaneous_growth_df, growth_draws_path)
    saveRDS(growth_summary_df, growth_summary_path)
  }

  tibble::tibble(
    model_id = model_id,
    n_lines = dplyr::n_distinct(instantaneous_growth_df$line_id),
    n_draws_sampled = dplyr::n_distinct(instantaneous_growth_df$.draw),
    n_surface_rows = nrow(growth_summary_df)
  )
}))

print(as.data.frame(growth_export_summary_df), row.names = FALSE)
```



## Model Selection

```{r spec-model-selection-inputs, include=FALSE}
library(dplyr)
library(ggplot2)
library(tibble)
library(patchwork)

source("R/project_paths.R")
source(get_model_r_path("gpath", "v1"))
source("R/gpath_run_utils.R")
source("R/model_analysis_utils.R")

project_paths <- list(
  stan_data = resolve_stan_data_path(),
  assessment_run_ids = file.path("models", "gpath", "v1", "assessment_run_ids.txt"),
  optim_root = file.path("data", "runs", "gpath", "v1", "optim"),
  report_dataset_label = "gstarvation_v1",
  transfer_optim_root = "data/gpath_transfer_cv/gpath",
  transfer_nuts_root = "data/gpath_transfer_cv_nuts/gpath",
  transfer_nuts_report_root = file.path("data", "gpath_transfer_cv_nuts", "report"),
  report_export_root = file.path("data", "report_exports", "model_analysis"),
  agent_text_path = file.path("workflow", "model_analysis.txt")
)

stan_data_ref <- readRDS(project_paths$stan_data)
line_label_map <- make_line_label_map(stan_data_ref$line_map)

summarize_stan_data_condition <- function(stan_data_path, tol = 1e-12) {
  stan_data <- readRDS(stan_data_path)

  if (!all(c("ploidy_metric", "line_id") %in% names(stan_data))) {
    stop("stan_data must contain `ploidy_metric` and `line_id`.")
  }

  ploidy_metric <- as.numeric(stan_data$ploidy_metric)
  line_id <- as.integer(stan_data$line_id)
  is_train <- if ("is_train" %in% names(stan_data)) as.integer(stan_data$is_train) else rep(1L, length(line_id))
  split_meta <- attr(stan_data, "transfer_split")

  finite_ploidy <- ploidy_metric[is.finite(ploidy_metric)]
  is_null <- length(finite_ploidy) > 0L && all(abs(finite_ploidy) < tol)

  holdout_idx <- which(is_train == 0L)
  holdout_lines <- if (!is.null(split_meta$line_id)) {
    as.integer(split_meta$line_id)
  } else {
    sort(unique(line_id[holdout_idx]))
  }
  holdout_ploidies <- if (!is.null(split_meta$holdout_value)) {
    as.numeric(split_meta$holdout_value)
  } else {
    sort(unique(ploidy_metric[holdout_idx]))
  }

  if (!length(holdout_lines) || !length(holdout_ploidies)) {
    meta_path <- file.path(dirname(stan_data_path), "dataset_meta.Rds")
    if (file.exists(meta_path)) {
      dataset_meta <- readRDS(meta_path)
      if (!length(holdout_lines) && !is.null(dataset_meta$line_id)) {
        holdout_lines <- as.integer(dataset_meta$line_id)
      }
      if (!length(holdout_ploidies) && !is.null(dataset_meta$holdout_value)) {
        holdout_ploidies <- as.numeric(dataset_meta$holdout_value)
      }
    }
  }

  holdout_lines <- holdout_lines[is.finite(holdout_lines)]
  holdout_ploidies <- holdout_ploidies[is.finite(holdout_ploidies)]

  if (length(holdout_lines) > 1L) {
    warning(sprintf(
      "Multiple holdout lines in %s: %s",
      stan_data_path,
      paste(holdout_lines, collapse = ", ")
    ))
  }

  if (length(holdout_ploidies) > 1L) {
    warning(sprintf(
      "Multiple holdout ploidies in %s: %s",
      stan_data_path,
      paste(format(holdout_ploidies, trim = TRUE), collapse = ", ")
    ))
  }

  data.frame(
    stan_data_path = stan_data_path,
    is_null = is_null,
    holdout_line = if (length(holdout_lines)) holdout_lines[[1]] else NA_integer_,
    holdout_ploidy = if (length(holdout_ploidies)) holdout_ploidies[[1]] else NA_real_,
    stringsAsFactors = FALSE
  )
}

stan_data_summary_df <- bind_rows(lapply(unique(x$summary_df$stan_data_path), summarize_stan_data_condition))

spec_summary_df <- x$summary_df %>%
  left_join(stan_data_summary_df, by = "stan_data_path") %>%
  mutate(
    model_id = run_id,
    fit_type = case_when(
      !is.na(is_null) & is_null & !is.na(holdout_line) ~ "null",
      !is.na(is_null) & !is_null & !is.na(holdout_line) ~ "transfer",
      !is.na(is_null) & !is_null & is.na(holdout_line) & is.na(holdout_ploidy) ~ "oracle",
      TRUE ~ "other"
    )
  )

transfer_predictive_df <- spec_summary_df %>%
  filter(fit_type == "transfer") %>%
  transmute(
    model_id,
    holdout_line = as.integer(holdout_line),
    holdout_ploidy = as.numeric(holdout_ploidy),
    spec_row_id_transfer = spec_row_id,
    log_lik_train_transfer = log_lik_train,
    log_lik_holdout_transfer = log_lik_holdout
  )

null_predictive_df <- spec_summary_df %>%
  filter(fit_type == "null") %>%
  transmute(
    model_id,
    holdout_line = as.integer(holdout_line),
    holdout_ploidy = as.numeric(holdout_ploidy),
    spec_row_id_null = spec_row_id,
    log_lik_train_null = log_lik_train,
    log_lik_holdout_null = log_lik_holdout
  )

predictive_transfer_df <- transfer_predictive_df %>%
  full_join(
    null_predictive_df,
    by = c("model_id", "holdout_line", "holdout_ploidy")
  ) %>%
  mutate(
    transfer_gain = log_lik_holdout_transfer - log_lik_holdout_null
  ) %>%
  arrange(model_id, holdout_line, holdout_ploidy)

model_selection_df <- spec_summary_df %>%
  filter(is.na(holdout_ploidy)) %>%
  mutate(
    run_dir = optim_output_dir,
    manifest_path = file.path(optim_output_dir, "run_manifest.json"),
    manifest_stan_data = stan_data_path,
    manifest_stan_file = NA_character_,
    alias = vapply(run_id, build_model_alias, character(1), format = "text"),
    alias_plot = vapply(run_id, build_model_alias, character(1), format = "plotmath"),
    n_total = n_starts_total,
    n_finite = n_starts_valid,
    prop_within_1 = if_else(n_total > 0, n_within_1 / n_total, NA_real_),
    prop_within_5 = if_else(n_total > 0, n_within_5 / n_total, NA_real_),
    deviance = -2 * best_lp,
    AIC = 2 * k - 2 * best_lp
  ) %>%
  group_by(model_id, alias, alias_plot, dataset_label, k) %>%
  slice_max(order_by = run_label, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(delta_AIC = AIC - min(AIC, na.rm = TRUE)) %>%
  arrange(delta_AIC)

selected_model_id_aic <- if (nrow(model_selection_df)) model_selection_df$model_id[[1]] else NA_character_
selected_model_id <- selected_model_id_aic
selected_model_row <- if (nrow(model_selection_df)) {
  model_selection_df %>% slice_head(n = 1)
} else {
  tibble()
}
selected_model_best_draw_vec <- NULL
if (nrow(selected_model_row)) {
  selected_draw_idx <- match(selected_model_row$spec_row_id[[1]], x$summary_df$spec_row_id)
  if (is.finite(selected_draw_idx) && !is.na(selected_draw_idx)) {
    selected_model_best_draw_vec <- x$best_draws[[selected_draw_idx]]
  }
}
```

```{r model-selection-plot, fig.width=19, fig.height=6, out.width='100%'}
if (!nrow(model_selection_df)) {
  announce_missing("Skipping model-selection plot because no spec-backed all-data ranking table was built.")
} else {
  plot_df <- model_selection_df %>%
    distinct(model_id, .keep_all = TRUE) %>%
    mutate(
      n_within_1 = if_else(is.na(n_within_1), 0L, as.integer(n_within_1)),
      prop_within_1 = if_else(is.na(prop_within_1), 0, prop_within_1)
    )

  alias_levels <- plot_df %>%
    arrange(delta_AIC) %>%
    pull(alias) %>%
    unique_preserve()

  alias_plot_map <- plot_df %>%
    distinct(alias, alias_plot) %>%
    tibble::deframe()

  plot_df <- plot_df %>%
    mutate(alias = factor(alias, levels = rev(alias_levels)))

  pareto_front <- plot_df %>%
    arrange(k, deviance) %>%
    filter(deviance == cummin(deviance)) %>%
    group_by(deviance) %>%
    slice_min(order_by = k, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(k)

  plot_df <- plot_df %>%
    mutate(
      is_pareto_front = model_id %in% pareto_front$model_id,
      pareto_fill = if_else(is_pareto_front, "firebrick", "grey35"),
      stability_fill = if_else(is_pareto_front, "firebrick", "steelblue4")
    )

  p_aic <- ggplot(plot_df, aes(x = alias, y = delta_AIC)) +
    geom_col(aes(fill = pareto_fill)) +
    coord_flip() +
    scale_fill_identity() +
    scale_x_discrete(labels = function(x) unname(alias_plot_map[as.character(x)])) +
    theme_model_analysis() +
    labs(x = "", y = expression(Delta ~ AIC))

  p_pareto <- ggplot(plot_df, aes(x = k, y = deviance)) +
    geom_point(color = "grey60", size = 2.5) +
    geom_line(data = pareto_front, color = "firebrick", linetype = "dashed") +
    geom_point(data = pareto_front, color = "firebrick", size = 3) +
    ggrepel::geom_label_repel(
      data = pareto_front,
      aes(label = alias_plot),
      color = "firebrick",
      fill = "white",
      label.size = 0.15,
      direction = "both",
      box.padding = 0.35,
      point.padding = 0.2,
      min.segment.length = 0,
      max.overlaps = Inf,
      size = plot_text_size / ggplot2::.pt
    ) +
    theme_model_analysis() +
    labs(x = "Effective parameters (k)", y = "Deviance")

  p_stability <- ggplot(plot_df, aes(x = alias, y = n_within_1)) +
    geom_col(aes(fill = stability_fill)) +
    coord_flip() +
    scale_fill_identity() +
    scale_x_discrete(labels = function(x) unname(alias_plot_map[as.character(x)])) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    theme_model_analysis() +
    labs(x = "", y = "Found optimum (/500)")

  p_aic + p_pareto + p_stability + plot_layout(ncol = 3)
}
```

```{r selected-global-fit-overlay, fig.width=12, fig.height=6, out.width='100%'}
if (!exists("pareto_front") || !nrow(pareto_front)) {
  announce_missing("Skipping selected global-fit overlays because no Pareto-front models are available.")
} else {
  source("R/plot_utils.R")
  source("R/sim_utils.R")

  regenerate_selected_global_fit_overlays <- FALSE
  pareto_front_model_ids <- unique_preserve(as.character(pareto_front$model_id))

  fit_rows <- model_selection_df %>%
    distinct(model_id, .keep_all = TRUE) %>%
    filter(model_id %in% pareto_front_model_ids) %>%
    arrange(match(model_id, pareto_front_model_ids))

  draw_idx <- match(fit_rows$spec_row_id, x$summary_df$spec_row_id)
  fit_rows$best_draw_vec <- x$best_draws[draw_idx]
  keep_fit <- !is.na(draw_idx) & vapply(fit_rows$best_draw_vec, Negate(is.null), logical(1))
  fit_rows <- fit_rows[keep_fit, , drop = FALSE]

  if (!nrow(fit_rows)) {
    announce_missing("Could not resolve any Pareto-front optimisation draws for the current ranking table.")
  } else {
    stan_data_overlay <- add_group_structure(readRDS(project_paths$stan_data))
    sim_times <- seq(0, max(stan_data_overlay$t_grid), by = 0.5)
    output_dir <- file.path("dev-figs", "model_analysis", "pareto_global_fit_overlays")
    dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

    alias_levels <- fit_rows$alias
    palette_vals <- setNames(
      grDevices::hcl.colors(length(alias_levels), palette = "Dark 3"),
      alias_levels
    )

    all_line_ids <- sort(unique(as.integer(stan_data_overlay$line_id)))

    overlay_manifest_df <- bind_rows(lapply(all_line_ids, function(target_line_id) {
      line_name_hit <- names(stan_data_overlay$line_map)[as.integer(unname(stan_data_overlay$line_map)) == target_line_id]
      target_line_name <- if (length(line_name_hit)) line_name_hit[[1]] else sprintf("line %d", target_line_id)
      out_path <- file.path(output_dir, sprintf("line_%02d_pareto_overlay.png", target_line_id))
      existed_before <- file.exists(out_path)

      if (isTRUE(regenerate_selected_global_fit_overlays) || !existed_before) {
        sim_df <- bind_rows(lapply(seq_len(nrow(fit_rows)), function(i) {
          simulate_line_from_draw(
            model_id = fit_rows$model_id[[i]],
            model_r_path = get_model_r_path("gpath", "v1"),
            draw_vec = fit_rows$best_draw_vec[[i]],
            stan_data = stan_data_overlay,
            line_id = target_line_id,
            times = sim_times,
            line_name = target_line_name,
            extra_cols = list(
              group_1 = fit_rows$alias[[i]],
              run_label = fit_rows$run_label[[i]]
            )
          )
        }))

        obs_df <- build_obs_df_from_stan_data(
          stan_data = stan_data_overlay,
          line_id = target_line_id,
          line_name = target_line_name
        )

        plt <- plot_fit_overlays(
          sim_df = sim_df,
          obs_df = obs_df,
          color_by = "group_1",
          color_values = palette_vals,
          color_label = "Model",
          title = target_line_name,
          print_plot = FALSE
        )

        ggsave(
          filename = out_path,
          plot = plt,
          width = 12,
          height = 6,
          units = "in",
          dpi = 200
        )

        status <- if (existed_before) "regenerated" else "generated"
      } else {
        status <- "cached"
      }

      tibble(
        line_id = target_line_id,
        line_name = target_line_name,
        output_path = out_path,
        status = status
      )
    }))

    print(as.data.frame(overlay_manifest_df), row.names = FALSE)
  }
}
```

```{r nuts-focused-raw-theta-ploidy-posteriors, fig.height=8, fig.width=16, out.width='100%'}
source("R/project_paths.R")
source(get_model_r_path("gpath", "v1"))

extract_focus_raw_theta_from_bound <- function(bound_obj, model_id, focus_parameters = c("ae[1]", "ah[1]", "Y_R[1]", "m")) {
  dims <- parse_run_id(model_id)
  param_labels <- get_param_names(dims$R, dims$P, dims$W, dims$C, dims$M)
  keep_idx <- which(param_labels %in% focus_parameters)
  if (!length(keep_idx)) return(tibble::tibble())

  draws_matrix <- posterior::as_draws_matrix(bound_obj$draws)
  coef_names <- sprintf("raw_theta_ploidy[%d]", keep_idx)
  coef_df <- tibble::as_tibble(draws_matrix[, coef_names, drop = FALSE])
  names(coef_df) <- param_labels[keep_idx]

  dplyr::bind_cols(
    tibble::tibble(
      model_id = model_id,
      .draw = seq_len(nrow(draws_matrix))
    ),
    coef_df
  ) %>%
    tidyr::pivot_longer(
      cols = -c(.data$model_id, .data$.draw),
      names_to = "parameter",
      values_to = "value"
    )
}

focus_parameters <- c("ae[1]", "ah[1]", "Y_R[1]", "m")
raw_theta_focus_df <- dplyr::bind_rows(lapply(model_ids, function(model_id) {
  bound_obj <- readRDS(file.path(diagnostic_roots, model_id, oracle_subdir, "bound_nuts_draws.Rds"))
  extract_focus_raw_theta_from_bound(bound_obj, model_id, focus_parameters = focus_parameters)
}))

plot_df <- raw_theta_focus_df %>%
  dplyr::mutate(
    alias = vapply(.data$model_id, build_model_alias, character(1), format = "text"),
    alias = factor(alias, levels = rev(unique_preserve(vapply(model_ids, build_model_alias, character(1), format = "text")))),
    parameter = factor(
      apply_focussed_parameter_alias(.data$parameter),
      levels = apply_focussed_parameter_alias(focus_parameters)
    )
  )

ggplot(plot_df, aes(x = value, y = alias)) +
  geom_vline(color="red",xintercept = 0)+
  geom_violin(fill = "#9ecae1", color = "#4a6d8c", scale = "width", linewidth = 0.4) +
  facet_grid(. ~ parameter, scales = "free_x", drop = FALSE) +
  theme_model_analysis() +
  labs(x = "raw_theta_ploidy posterior value", y = "Model")
```

VERY IMPORTANT: Sandy said we should think harder about interpreting results in terms of differences between cell lines. The `nuts-line-parameter-posteriors-demo` block below is important for that, because it exposes the line-specific posterior parameter structure directly rather than only the shared ploidy-effect coefficients. A plausible next step is to operationalize this in holdout tests by asking whether ploidy-effect transfer works better between cell lines with more similar line-specific parameter sets.

```{r nuts-line-parameter-posteriors-demo, fig.height=24, fig.width=20, out.width='100%'}
source("R/parameter_transfer_utils.R")

order_gpath_parameter_names <- function(parameter_names) {
  family_rank <- dplyr::case_when(
    grepl("^ae\\[", parameter_names) ~ 1L,
    grepl("^ah\\[", parameter_names) ~ 2L,
    grepl("^Y_R\\[", parameter_names) ~ 3L,
    grepl("^A_mat\\[", parameter_names) ~ 4L,
    grepl("^K_mat\\[", parameter_names) ~ 5L,
    parameter_names == "m" ~ 6L,
    grepl("^waste_mech\\[", parameter_names) ~ 7L,
    parameter_names %in% c("R", "P", "W") ~ 8L,
    TRUE ~ 99L
  )

  order_df <- tibble::tibble(
    parameter = parameter_names,
    family_rank = family_rank
  ) %>%
    dplyr::distinct() %>%
    dplyr::arrange(.data$family_rank, .data$parameter)

  order_df$parameter
}

if (!exists("pareto_front") || !nrow(pareto_front)) {
  announce_missing("Skipping line-specific reconstructed parameter posteriors because no Pareto-front models are available.")
} else {
  line_ids <- sort(unique(as.integer(stan_data$line_id)))
  reconstructed_rows <- list()
  parameter_levels <- character(0)
  model_ids_demo <- head(pareto_front$model_id, 5L)

  for (demo_model_id in model_ids_demo) {
    bound_obj <- readRDS(file.path(diagnostic_roots, demo_model_id, oracle_subdir, "bound_nuts_draws.Rds"))
    draws_matrix <- posterior::as_draws_matrix(bound_obj$draws)
    draws_matrix <- thin_draws_matrix(draws_matrix, max_draws = 300L)

    for (line_id in line_ids) {
      states <- tryCatch(
        get_line_ploidy_states(stan_data, line_id = line_id),
        error = function(e) NULL
      )
      if (is.null(states)) {
        next
      }

      state_tbl <- tibble::tribble(
        ~ploidy_label, ~ploidy_metric,
        "Low ploidy", states$low_value,
        "High ploidy", states$high_value
      )

      for (draw_idx in seq_len(nrow(draws_matrix))) {
        draw_vec <- draws_matrix[draw_idx, , drop = TRUE]
        names(draw_vec) <- colnames(draws_matrix)

        for (state_i in seq_len(nrow(state_tbl))) {
          parms <- reconstruct_draw_parameters(
            draw_vec = draw_vec,
            model_id = demo_model_id,
            line_id = line_id,
            ploidy_metric = state_tbl$ploidy_metric[[state_i]]
          )

          flat_vals <- flatten_gpath_reconstructed_parms(parms)
          keep <- !(names(flat_vals) %in% c("R", "P", "W")) & !grepl("^waste_mech", names(flat_vals))
          flat_vals <- flat_vals[keep]
          parameter_levels <- unique(c(parameter_levels, names(flat_vals)))

          reconstructed_rows[[length(reconstructed_rows) + 1L]] <- tibble::tibble(
            model_id = demo_model_id,
            line_id = line_id,
            ploidy_label = state_tbl$ploidy_label[[state_i]],
            .draw = draw_idx,
            parameter = names(flat_vals),
            value = as.numeric(flat_vals)
          )
        }
      }
    }
  }

  posterior_df <- dplyr::bind_rows(reconstructed_rows) %>%
    dplyr::mutate(
      alias = vapply(.data$model_id, build_model_alias, character(1), format = "text"),
      alias = factor(
        alias,
        levels = unique_preserve(vapply(model_ids_demo, build_model_alias, character(1), format = "text"))
      ),
      parameter = factor(.data$parameter, levels = order_gpath_parameter_names(parameter_levels)),
      line_label = factor(
        label_line_ids(.data$line_id, line_label_map),
        levels = rev(unique_preserve(label_line_ids(sort(unique(.data$line_id)), line_label_map, as_factor = FALSE)))
      ),
      ploidy_label = factor(.data$ploidy_label, levels = c("Low ploidy", "High ploidy"))
    )

  reconstructed_parameter_name_df <- posterior_df %>%
    dplyr::distinct(.data$model_id, .data$alias, .data$parameter) %>%
    dplyr::arrange(.data$alias, .data$parameter)

  ppost <- ggplot(posterior_df, aes(x = value, y = line_label, fill = ploidy_label)) +
    geom_violin(position = "identity", alpha = 0.6, color = NA, scale = "width", trim = TRUE) +
    facet_grid(alias ~ parameter, scales = "free_x") +
    scale_fill_manual(values = c("Low ploidy" = "#9ecae1", "High ploidy" = "#fdae6b")) +
    theme_model_analysis(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank()
    ) +
    scale_x_log10()+
    labs(
      title = "Reconstructed oracle-NUTS parameter posteriors across Pareto-front models",
      x = "Reconstructed parameter value",
      y = "Cell line",
      fill = "Ploidy state"
    )
}
```

```{r nuts-line-parameter-posterior-names}
if (!exists("reconstructed_parameter_name_df") || !nrow(reconstructed_parameter_name_df)) {
  announce_missing("No reconstructed parameter-name diagnostics available for Pareto-front oracle NUTS models.")
} else {
  parameter_name_list <- split(
    as.character(reconstructed_parameter_name_df$parameter),
    reconstructed_parameter_name_df$alias
  )

  for (alias_name in names(parameter_name_list)) {
    cat(alias_name, "\n", sep = "")
    cat(paste(parameter_name_list[[alias_name]], collapse = ", "), "\n\n", sep = "")
  }
}
```

```{r nuts-reconstruct-parms-debug}
source("R/parameter_transfer_utils.R")

debug_model_id <- "2R_2P_1W_C0_M0"

if (!dir.exists(file.path(diagnostic_roots, debug_model_id, oracle_subdir))) {
  announce_missing("Skipping reconstruct_parms debug chunk because %s oracle NUTS draws were not found.", debug_model_id)
} else {
  debug_bound_obj <- readRDS(file.path(diagnostic_roots, debug_model_id, oracle_subdir, "bound_nuts_draws.Rds"))
  debug_draws_matrix <- posterior::as_draws_matrix(debug_bound_obj$draws)
  debug_draw_vec <- debug_draws_matrix[1, , drop = TRUE]
  names(debug_draw_vec) <- colnames(debug_draws_matrix)

  debug_line_id <- sort(unique(as.integer(stan_data$line_id)))[[1]]
  debug_states <- get_line_ploidy_states(stan_data, line_id = debug_line_id)
  debug_dims <- parse_run_id(debug_model_id)
  debug_L <- 3L * debug_dims$R + (debug_dims$P - 1L) * debug_dims$R + debug_dims$W * debug_dims$R + 1L

  debug_parms <- reconstruct_parms(
    R = debug_dims$R,
    P = debug_dims$P,
    W = debug_dims$W,
    strict_spec = (debug_dims$C == 1L),
    M = debug_dims$M,
    base_priors = base_priors,
    raw_theta_line = debug_draw_vec[sprintf("raw_theta_line[%d,%d]", seq_len(debug_L), debug_line_id)],
    raw_theta_ploidy = debug_draw_vec[sprintf("raw_theta_ploidy[%d]", seq_len(debug_L))],
    ploidy_metric = debug_states$low_value
  )

  debug_flat <- flatten_gpath_reconstructed_parms(debug_parms)

  cat(sprintf(
    "Model: %s | alias: %s | draw_row: 1 | line_id: %d | ploidy_metric: %s\n\n",
    debug_model_id,
    build_model_alias(debug_model_id, format = "text"),
    debug_line_id,
    format(debug_states$low_value, digits = 6)
  ))

  cat("Output of reconstruct_parms():\n")
  print(debug_parms)
  cat("\nOutput of flatten_gpath_reconstructed_parms():\n")
  print(debug_flat)
}
```

```{r nuts-instantaneous-growth-heatmap, fig.height=10, fig.width=16, out.width='100%'}
growth_surface_df <- dplyr::bind_rows(lapply(model_ids, function(model_id) {
  readRDS(file.path(diagnostic_roots, model_id, oracle_subdir, "nuts_instantaneous_growth_summary.Rds"))
}))

growth_draw_surface_df <- dplyr::bind_rows(lapply(model_ids, function(model_id) {
  readRDS(file.path(diagnostic_roots, model_id, oracle_subdir, "nuts_instantaneous_growth_draws.Rds")) %>%
    dplyr::group_by(.data$model_id, .data$line_id, .data$glucose, .data$resource2) %>%
    dplyr::summarise(
      median_low_ploidy_growth = stats::median(.data$low_ploidy_growth, na.rm = TRUE),
      median_high_ploidy_growth = stats::median(.data$high_ploidy_growth, na.rm = TRUE),
      .groups = "drop"
    )
}))

if (!exists("pareto_front") || !nrow(pareto_front)) {
  announce_missing("Skipping NUTS instantaneous-growth heatmap because no Pareto-front models are available.")
} else {
  plot_df <- growth_surface_df %>%
    dplyr::filter(.data$model_id %in% pareto_front$model_id) %>%
    dplyr::mutate(
      alias = vapply(.data$model_id, build_model_alias, character(1), format = "text"),
      alias = factor(
        alias,
        levels = unique_preserve(vapply(pareto_front$model_id, build_model_alias, character(1), format = "text"))
      ),
      line_label = label_line_ids(.data$line_id, line_label_map)
    )

  delta_fill_limit <- max(abs(plot_df$median_delta_growth), na.rm = TRUE)
  if (!is.finite(delta_fill_limit) || delta_fill_limit <= 0) delta_fill_limit <- 1e-6

  ggplot(plot_df, aes(x = glucose, y = resource2, fill = median_delta_growth)) +
    geom_tile() +
    facet_grid(line_label ~ alias) +
    scale_x_log10() +
    scale_y_log10() +
    scale_fill_gradientn(
      colours = c("#2166ac", "#67a9cf", "#f7f7f7", "#ef8a62", "#b2182b"),
      values = scales::rescale(c(-delta_fill_limit, -0.15 * delta_fill_limit, 0, 0.15 * delta_fill_limit, delta_fill_limit)),
      limits = c(-delta_fill_limit, delta_fill_limit),
      oob = scales::squish
    ) +
    theme_model_analysis(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank()
    ) +
    labs(x = "Glucose", y = "Resource 2", fill = "High - Low\nnet growth")
}
```

```{r nuts-instantaneous-growth-heatmap-low, fig.height=10, fig.width=16, out.width='100%'}
if (!exists("pareto_front") || !nrow(pareto_front)) {
  announce_missing("Skipping low-ploidy NUTS instantaneous-growth heatmap because no Pareto-front models are available.")
} else {
  absolute_growth_df <- growth_draw_surface_df %>%
    dplyr::filter(.data$model_id %in% pareto_front$model_id) %>%
    dplyr::mutate(
      alias = vapply(.data$model_id, build_model_alias, character(1), format = "text"),
      alias = factor(
        alias,
        levels = unique_preserve(vapply(pareto_front$model_id, build_model_alias, character(1), format = "text"))
      ),
      line_label = label_line_ids(.data$line_id, line_label_map)
    )

  absolute_fill_limits <- range(
    c(absolute_growth_df$median_low_ploidy_growth, absolute_growth_df$median_high_ploidy_growth),
    na.rm = TRUE
  )
  if (!all(is.finite(absolute_fill_limits)) || diff(absolute_fill_limits) <= 0) {
    absolute_fill_limits <- c(-1e-6, 1e-6)
  }

  plot_df <- absolute_growth_df %>%
    dplyr::transmute(
      alias = .data$alias,
      line_label = .data$line_label,
      glucose = .data$glucose,
      resource2 = .data$resource2,
      median_growth = .data$median_low_ploidy_growth
    )

  p_lo_gro <- ggplot(plot_df, aes(x = glucose, y = resource2, fill = median_growth)) +
    geom_tile() +
    facet_grid(line_label ~ alias) +
    scale_x_log10() +
    scale_fill_viridis_c(option = "C", limits = absolute_fill_limits, oob = scales::squish) +
    theme_model_analysis(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank()
    ) +
    labs(x = "Glucose", y = "Resource 2", fill = "Low-ploidy\nmedian net growth")
}
p_lo_gro
```

```{r nuts-instantaneous-growth-heatmap-high, fig.height=10, fig.width=16, out.width='100%'}
if (!exists("pareto_front") || !nrow(pareto_front)) {
  announce_missing("Skipping high-ploidy NUTS instantaneous-growth heatmap because no Pareto-front models are available.")
} else {
  absolute_growth_df <- growth_draw_surface_df %>%
    dplyr::filter(.data$model_id %in% pareto_front$model_id) %>%
    dplyr::mutate(
      alias = vapply(.data$model_id, build_model_alias, character(1), format = "text"),
      alias = factor(
        alias,
        levels = unique_preserve(vapply(pareto_front$model_id, build_model_alias, character(1), format = "text"))
      ),
      line_label = label_line_ids(.data$line_id, line_label_map)
    )

  absolute_fill_limits <- range(
    c(absolute_growth_df$median_low_ploidy_growth, absolute_growth_df$median_high_ploidy_growth),
    na.rm = TRUE
  )
  if (!all(is.finite(absolute_fill_limits)) || diff(absolute_fill_limits) <= 0) {
    absolute_fill_limits <- c(-1e-6, 1e-6)
  }

  plot_df <- absolute_growth_df %>%
    dplyr::transmute(
      alias = .data$alias,
      line_label = .data$line_label,
      glucose = .data$glucose,
      resource2 = .data$resource2,
      median_growth = .data$median_high_ploidy_growth
    )

  p_hi_gro <- ggplot(plot_df, aes(x = glucose, y = resource2, fill = median_growth)) +
    geom_tile() +
    facet_grid(line_label ~ alias) +
    scale_x_log10() +
    scale_y_log10() +
    scale_fill_viridis_c(option = "C", limits = absolute_fill_limits, oob = scales::squish) +
    theme_model_analysis(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank()
    ) +
    labs(x = "Glucose", y = "Resource 2", fill = "High-ploidy\nmedian net growth")
}

p_hi_gro
```

```{r predictive-transfer-heatmap, fig.width=12, fig.height=7, out.width='100%'}
if (!nrow(predictive_transfer_df)) {
  announce_missing("Skipping predictive-transfer heatmap because no transfer/null pairings were built from x.")
} else {
  positive_gains <- predictive_transfer_df$transfer_gain[predictive_transfer_df$transfer_gain > 0]
  if (length(positive_gains)) {
    fill_cap <- 1.5 * max(positive_gains, na.rm = TRUE)
  } else {
    fill_cap <- max(abs(predictive_transfer_df$transfer_gain), na.rm = TRUE)
  }

  plot_df <- predictive_transfer_df %>%
    mutate(
      alias = vapply(model_id, build_model_alias, character(1), format = "text"),
      alias = factor(
        alias,
        levels = rev(unique_preserve(vapply(model_selection_df$model_id, build_model_alias, character(1), format = "text")))
      ),
      holdout_line_label = label_line_ids(holdout_line, line_label_map),
      holdout_panel = case_when(
        is.na(holdout_ploidy) ~ NA_character_,
        abs(holdout_ploidy) < 1e-12 ~ "Held out low",
        TRUE ~ "Held out high"
      ),
      holdout_panel = factor(holdout_panel, levels = c("Held out low", "Held out high"))
    )

  ggplot(plot_df, aes(x = holdout_line_label, y = alias, fill = transfer_gain)) +
    geom_tile(color = "white", linewidth = 0.6) +
    facet_grid(. ~ holdout_panel, drop = FALSE) +
    scale_fill_gradient2(
      low = "#3b4cc0",
      mid = "white",
      high = "#b40426",
      midpoint = 0,
      limits = c(-fill_cap, fill_cap),
      oob = scales::squish,
      na.value = "grey90"
    ) +
    theme_model_analysis(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    labs(
      x = "Holdout line",
      y = "Model",
      fill = sprintf("Transfer gain\n(saturated beyond +/- %.0f)", fill_cap)
    )
}
```

```{r example-holdout-transfer-plot, fig.width=12, fig.height=6, out.width='100%'}
source("R/transfer_cv_plot_utils.R")

example_model_id <- if (!is.na(selected_model_id)) {
  selected_model_id
} else if (nrow(model_selection_df)) {
  model_selection_df$model_id[[1]]
} else {
  NA_character_
}

if (is.na(example_model_id)) {
  announce_missing("Skipping example holdout transfer plot because no selected model is available.")
} else {
  example_case_df <- predictive_transfer_df %>%
    filter(
      model_id == example_model_id,
      !is.na(holdout_line),
      !is.na(holdout_ploidy),
      is.finite(transfer_gain)
    ) %>%
    arrange(desc(transfer_gain), holdout_line, holdout_ploidy)

  if (!nrow(example_case_df)) {
    announce_missing("Skipping example holdout transfer plot because no finite holdout transfer cases were found for %s.", example_model_id)
  } else {
    example_case <- example_case_df[1, , drop = FALSE]
    example_direction <- if (abs(example_case$holdout_ploidy[[1]]) < 1e-12) {
      "high_to_low"
    } else {
      "low_to_high"
    }

    example_plot_data <- build_transfer_overlay_plot_data(
      model_id = example_model_id,
      line_id = example_case$holdout_line[[1]],
      directions = example_direction,
      fit_types = c("null", "transfer", "oracle"),
      output_root = "data/gpath_transfer_cv",
      stan_data_path = project_paths$stan_data,
      include_score_text = FALSE
    )

    holdout_label <- if (abs(example_case$holdout_ploidy[[1]]) < 1e-12) {
      "held out low-ploidy wells"
    } else {
      "held out high-ploidy wells"
    }

    metric_line <- sprintf(
      "%s | transfer gain = %.2f | transfer log-lik = %.2f | null log-lik = %.2f",
      example_direction,
      example_case$transfer_gain[[1]],
      example_case$log_lik_holdout_transfer[[1]],
      example_case$log_lik_holdout_null[[1]]
    )

    plot_fit_overlays(
      sim_df = example_plot_data$sim_df,
      obs_df = example_plot_data$obs_df,
      color_by = "group_1",
      color_values = c(null = "#6f6f6f", transfer = "#1b9e77", oracle = "#d95f02"),
      color_label = "Fit",
      title = sprintf(
        "Example holdout transfer fit | %s | %s",
        label_line_ids(example_case$holdout_line[[1]], line_label_map, as_factor = FALSE),
        holdout_label
      ),
      subtitle = metric_line
    )
  }
}
```

```{r parameter-transfer-from-x, include=FALSE}
source("R/parameter_transfer_utils.R")

extract_raw_theta_ploidy_df <- function(draw_vec, model_id) {
  dims <- parse_run_id(model_id)
  parameter_labels <- get_param_names(
    R = dims$R,
    P = dims$P,
    W = dims$W,
    C = dims$C,
    M = dims$M,
    priors = base_priors
  )
  coefficient_names <- sprintf("raw_theta_ploidy[%d]", seq_along(parameter_labels))
  values <- unname(draw_vec[coefficient_names])

  tibble(
    parameter = parameter_labels,
    coefficient = coefficient_names,
    value = values,
    missing = is.na(values)
  )
}

if (!exists("line_label_map")) {
  stan_data_for_labels <- readRDS(project_paths$stan_data)
  line_label_map <- make_line_label_map(stan_data_for_labels$line_map)
}

parameter_transfer_base_df <- spec_summary_df %>%
  filter(fit_type %in% c("null", "transfer", "oracle")) %>%
  transmute(
    model_id,
    fit_type,
    holdout_line = as.integer(holdout_line),
    holdout_ploidy = as.numeric(holdout_ploidy),
    spec_row_id,
    run_label,
    draw_idx = match(spec_row_id, x$summary_df$spec_row_id),
    draw_vec = x$best_draws[draw_idx]
  ) %>%
  filter(
    (fit_type %in% c("null", "transfer") & !is.na(holdout_line) & !is.na(holdout_ploidy)) |
      (fit_type == "oracle")
  ) %>%
  group_by(model_id, fit_type, holdout_line, holdout_ploidy) %>%
  slice_max(order_by = run_label, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-run_label)

oracle_raw_theta_df <- parameter_transfer_base_df %>%
  filter(fit_type == "oracle") %>%
  rowwise() %>%
  mutate(compare_parameter_df = list(
    if (is.null(draw_vec)) {
      NULL
    } else {
      extract_raw_theta_ploidy_df(draw_vec = draw_vec, model_id = model_id)
    }
  )) %>%
  ungroup() %>%
  select(model_id, compare_parameter_df) %>%
  tidyr::unnest(compare_parameter_df)

common_raw_theta_parameters_from_x <- oracle_raw_theta_df %>%
  filter(!missing) %>%
  group_by(parameter) %>%
  summarise(n_models_present = n_distinct(model_id), .groups = "drop") %>%
  filter(n_models_present == n_distinct(oracle_raw_theta_df$model_id)) %>%
  pull(parameter)

oracle_parameter_df <- parameter_transfer_base_df %>%
  filter(fit_type == "oracle") %>%
  select(model_id, spec_row_id_oracle = spec_row_id, draw_vec_oracle = draw_vec)

parameter_transfer_long_df <- parameter_transfer_base_df %>%
  filter(fit_type %in% c("null", "transfer")) %>%
  left_join(oracle_parameter_df, by = "model_id") %>%
  rowwise() %>%
  mutate(
    compare_parameter_df = list(
      if (is.null(draw_vec) || is.null(draw_vec_oracle) || is.na(holdout_line) || is.na(holdout_ploidy)) {
        NULL
      } else {
        fit_df <- extract_raw_theta_ploidy_df(
          draw_vec = draw_vec,
          model_id = model_id
        ) %>%
          rename(fit_value = value)

        oracle_df <- extract_raw_theta_ploidy_df(
          draw_vec = draw_vec_oracle,
          model_id = model_id
        ) %>%
          rename(oracle_value = value)

        left_join(
          fit_df %>% select(parameter, fit_value),
          oracle_df %>% select(parameter, oracle_value),
          by = "parameter"
        )
      }
    )
  ) %>%
  ungroup()

parameter_transfer_summary_df <- parameter_transfer_long_df %>%
  select(
    model_id,
    fit_type,
    holdout_line,
    holdout_ploidy,
    spec_row_id,
    spec_row_id_oracle,
    compare_parameter_df
  ) %>%
  tidyr::unnest(compare_parameter_df) %>%
  tidyr::pivot_wider(
    id_cols = c(model_id, holdout_line, holdout_ploidy, parameter, spec_row_id_oracle, oracle_value),
    names_from = fit_type,
    values_from = c(spec_row_id, fit_value),
    names_glue = "{.value}_{fit_type}"
  ) %>%
  transmute(
    model_id,
    holdout_line,
    holdout_ploidy,
    parameter,
    spec_row_id_null,
    spec_row_id_transfer,
    spec_row_id_oracle,
    null_value = fit_value_null,
    transfer_value = fit_value_transfer,
    oracle_value = oracle_value,
    null_vs_oracle_diff = null_value - oracle_value,
    transfer_vs_oracle_diff = transfer_value - oracle_value,
    abs_null_vs_oracle_diff = abs(null_value - oracle_value),
    abs_transfer_vs_oracle_diff = abs(transfer_value - oracle_value),
    transfer_improvement_over_null = abs(null_value - oracle_value) - abs(transfer_value - oracle_value)
  ) %>%
  arrange(model_id, holdout_line, holdout_ploidy, match(parameter, common_raw_theta_parameters_from_x))
```

```{r parameter-transfer-from-x-plot, fig.height=10, fig.width=14}
parameter_plot_model_levels <- if (exists("model_selection_df") && nrow(model_selection_df)) {
  unique_preserve(model_selection_df$model_id)
} else {
  unique_preserve(parameter_transfer_summary_df$model_id)
}

plot_parameter_transfer_df <- parameter_transfer_summary_df %>%
  filter(
    parameter %in% common_raw_theta_parameters_from_x,
    !is.na(null_value),
    !is.na(transfer_value),
    !is.na(oracle_value)
  ) %>%
  mutate(
    holdout_panel = case_when(
      is.na(holdout_ploidy) ~ NA_character_,
      abs(holdout_ploidy) < 1e-12 ~ "Held out low",
      TRUE ~ "Held out high"
    ),
    holdout_line_label = label_line_ids(holdout_line, line_label_map, as_factor = FALSE),
    parameter_label = apply_focussed_parameter_alias(parameter),
    parameter_label = factor(
      parameter_label,
      levels = apply_focussed_parameter_alias(common_raw_theta_parameters_from_x)
    ),
    holdout_panel = factor(holdout_panel, levels = c("Held out low", "Held out high")),
    holdout_line_label = factor(
      holdout_line_label,
      levels = unique_preserve(label_line_ids(sort(unique(holdout_line)), line_label_map, as_factor = FALSE))
    ),
    alias = vapply(model_id, build_model_alias, character(1), format = "text"),
    alias = factor(
      alias,
      levels = unique_preserve(vapply(parameter_plot_model_levels, build_model_alias, character(1), format = "text"))
    )
  )

if (!nrow(plot_parameter_transfer_df)) {
  announce_missing("No raw_theta_ploidy comparison rows available for plotting from x.")
} else {
  p1 <- ggplot(
    plot_parameter_transfer_df,
    aes(x = alias)
  ) +
    geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
    geom_segment(
      aes(
        xend = alias,
        y = null_value,
        yend = transfer_value
      ),
      color = "grey55",
      linewidth = 0.5,
      alpha = 0.45
    ) +
    geom_point(aes(y = oracle_value), shape = 21, size = 2.8, stroke = 0.7, fill = "white", color = "black") +
    #geom_point(aes(y = null_value), shape = 4, size = 2.4, stroke = 0.8, color = "black") +
    geom_jitter(aes(y = transfer_value, color = holdout_line_label), size = 2.2, alpha = 0.85, height = 0, width = 0.2) +
    facet_grid(parameter_label ~ holdout_panel, scales = "free_y", drop = FALSE) +
    theme_model_analysis(
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    labs(
      x = "",
      y = "Best-fit raw ploidy-effect coefficient\n(open circle = oracle, colored point = transfer)",
      color = "Held-out cell line"
    )

  p1
}
```

```{r parameter-transfer-from-x-heatmaps, fig.height=10, fig.width=14}
if (!exists("plot_parameter_transfer_df") || !nrow(plot_parameter_transfer_df)) {
  announce_missing("No raw_theta_ploidy comparison rows available for heatmap summaries from x.")
} else {
  heatmap_df <- plot_parameter_transfer_df %>%
    group_by(model_id, alias, parameter_label, holdout_panel) %>%
    summarise(
      mean_improvement = mean(transfer_improvement_over_null, na.rm = TRUE),
      frac_lines_improved = mean(transfer_improvement_over_null > 0, na.rm = TRUE),
      n_lines = dplyr::n_distinct(holdout_line_label),
      .groups = "drop"
    )

  mean_cap <- max(abs(heatmap_df$mean_improvement), na.rm = TRUE)
  if (!is.finite(mean_cap) || mean_cap <= 0) {
    mean_cap <- 1
  }

  p_mean <- ggplot(
    heatmap_df,
    aes(x = parameter_label, y = alias, fill = mean_improvement)
  ) +
    geom_tile(color = "white", linewidth = 0.6) +
    facet_grid(. ~ holdout_panel, drop = FALSE) +
    scale_fill_gradient2(
      low = "#3b4cc0",
      mid = "white",
      high = "#b40426",
      midpoint = 0,
      limits = c(-mean_cap, mean_cap),
      oob = scales::squish,
      na.value = "grey90"
    ) +
    theme_model_analysis(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank()
    ) +
    labs(
      x = "",
      y = "Model",
      fill = "Mean improvement\n|null-oracle| - |transfer-oracle|"
    )

  p_frac <- ggplot(
    heatmap_df,
    aes(x = parameter_label, y = alias, fill = frac_lines_improved)
  ) +
    geom_tile(color = "white", linewidth = 0.6) +
    facet_grid(. ~ holdout_panel, drop = FALSE) +
    scale_fill_gradient(
      low = "white",
      high = "#1b9e77",
      limits = c(0, 1),
      na.value = "grey90"
    ) +
    theme_model_analysis(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank()
    ) +
    labs(
      x = "",
      y = "Model",
      fill = "Fraction of held-out\ncell lines improved"
    )

  p_mean / p_frac
}
```
