---
title: "Selection Strategy Analysis"
author: "Codex"
date: "2026-03-21"
output: html_document
params:
  config_path: "scripts/selection_strategy_config.R"
  output_root: !r file.path("data", "selection_strategy", "gpath", "v1", "gstarvation_v1")
  include_detailed: false
  competition_path: "data/counts/competition/SUM-159-fuse.Rds"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook reports the main selection-strategy result from the mixed-ploidy simulations: which 6-day media schedules most strongly select for or against high ploidy by day 6.

Each strategy is a three-digit code:

- digit 1: initial glucose condition at day 0,
- digit 2: action at day 2 (`0` = no refresh, `1:7` = refresh to one of the tested glucose conditions),
- digit 3: action at day 4 (`0` = no refresh, `1:7` = refresh to one of the tested glucose conditions).

The primary outcome is the day-6 log ratio:

`log_ratio_final = log(N_high / N_low)`

More negative values mean stronger selection against high ploidy. More positive values mean stronger selection for high ploidy.

## Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(patchwork)

source("R/selection_strategy_utils.R")

cfg <- load_selection_config(params$config_path)
cfg$output_root <- dirname(params$output_root)
cfg$dataset_label <- basename(params$output_root)

stan_data <- resolve_selection_stan(cfg)
line_map <- get_line_label_map(stan_data)
glucose_values <- resolve_glucose_values(stan_data, cfg)
strategy_grid <- build_strategy_grid(glucose_values)

outputs <- collect_selection_outputs(cfg)
final_summary <- outputs$final_summary
detail_summary <- outputs$detail

if (!nrow(final_summary)) {
  stop("No selection-strategy outputs were found. Run scripts/run_gpath_selection_strategy_batch.R first.")
}

final_summary <- final_summary %>%
  left_join(line_map, by = "line_id") %>%
  left_join(
    strategy_grid %>%
      select(strategy_code, g0_day0, day2_action, g0_day2, day4_action, g0_day4),
    by = "strategy_code"
  ) %>%
  mutate(
    cell_line = if_else(is.na(cell_line) | !nzchar(cell_line), sprintf("line_%02d", as.integer(line_id)), cell_line),
    context_label = if_else(fit_context == "global", "global", paste(direction, fit_type, sep = " | "))
  )

agent_text_path <- file.path("workflow", "selection_strategy_analysis.txt")
competition_line_id <- 5L
competition_frame_hours <- 8

competition_available <- file.exists(params$competition_path)

context_key <- function(df) {
  ifelse(
    df$fit_context == "global",
    paste(df$model_id, df$fit_context, sep = "||"),
    paste(df$model_id, df$fit_context, df$direction, df$fit_type, sep = "||")
  )
}

load_competition_fit_context <- function(row, config) {
  if (identical(row$fit_context[[1]], "global")) {
    load_best_global_fit(config = config, model_id = row$model_id[[1]])
  } else {
    load_best_transfer_fit_context(
      config = config,
      model_id = row$model_id[[1]],
      line_id = row$line_id[[1]],
      direction = row$direction[[1]],
      fit_type = row$fit_type[[1]]
    )
  }
}

simulate_competition_validation_context <- function(task_row, config, stan_data, init_df, total_hours, time_step_hours = 1) {
  task_row <- as.data.frame(task_row, stringsAsFactors = FALSE)
  fit_obj <- load_competition_fit_context(task_row, config)
  line_pair <- get_line_ploidy_pair(stan_data, line_id = task_row$line_id[[1]])

  comp <- reconstruct_mixture_components(
    draw_vec = fit_obj$draw_vec,
    model_id = task_row$model_id[[1]],
    line_id = task_row$line_id[[1]],
    low_ploidy = line_pair$low,
    high_ploidy = line_pair$high,
    ploidy_effect_mask = fit_obj$ploidy_effect_mask
  )

  pieces <- vector("list", nrow(init_df))
  for (i in seq_len(nrow(init_df))) {
    init_row <- init_df[i, , drop = FALSE]
    state0 <- make_initial_mix_state(
      parms_mix = comp$parms_mix,
      g0 = init_row$glucose[[1]],
      seed_total_n = init_row$total_live_start[[1]],
      initial_high_fraction = init_row$high_fraction_start[[1]]
    )

    sim <- simulate_mixture_segment(
      parms_mix = comp$parms_mix,
      state0 = state0,
      hours = total_hours,
      time_step_hours = time_step_hours,
      detailed = TRUE
    )

    sim$glucose <- init_row$glucose[[1]]
    sim$time_hours <- sim$time
    sim$model_id <- task_row$model_id[[1]]
    sim$line_id <- task_row$line_id[[1]]
    sim$fit_context <- task_row$fit_context[[1]]
    sim$direction <- task_row$direction[[1]]
    sim$fit_type <- task_row$fit_type[[1]]
    pieces[[i]] <- sim
  }

  bind_rows(pieces)
}

competition_raw <- NULL
competition_df <- data.frame()
competition_mean_df <- data.frame()
competition_init_df <- data.frame()
competition_validation_summary <- data.frame()
competition_overlay_df <- data.frame()
competition_best_contexts <- data.frame()

if (competition_available) {
  competition_raw <- readRDS(params$competition_path) %>%
    mutate(
      frame = as.integer(frame),
      time_hours = as.integer(frame) * competition_frame_hours,
      label = as.character(label),
      rep = as.character(rep),
      condition = as.character(condition),
      glucose = as.numeric(glucose),
      ncells = as.numeric(ncells)
    )

  competition_live <- competition_raw %>%
    filter(label %in% c("2N", "4N")) %>%
    select(time_hours, frame, rep, condition, glucose, label, ncells) %>%
    pivot_wider(names_from = label, values_from = ncells, values_fill = 0) %>%
    rename(N_low = `2N`, N_high = `4N`) %>%
    mutate(
      total_live = N_low + N_high,
      high_fraction = if_else(total_live > 0, N_high / total_live, NA_real_),
      log_ratio = log((N_high + 1e-8) / (N_low + 1e-8))
    )

  competition_dead <- competition_raw %>%
    filter(label == "dead") %>%
    transmute(time_hours, frame, rep, condition, glucose, n_dead = ncells)

  competition_df <- competition_live %>%
    left_join(competition_dead, by = c("time_hours", "frame", "rep", "condition", "glucose")) %>%
    mutate(
      n_dead = coalesce(n_dead, 0),
      total_cells = total_live + n_dead,
      dead_fraction = if_else(total_cells > 0, n_dead / total_cells, NA_real_)
    )

  competition_mean_df <- competition_df %>%
    group_by(glucose, time_hours) %>%
    summarise(
      N_low = mean(N_low, na.rm = TRUE),
      N_high = mean(N_high, na.rm = TRUE),
      total_live = mean(total_live, na.rm = TRUE),
      high_fraction = mean(high_fraction, na.rm = TRUE),
      log_ratio = mean(log_ratio, na.rm = TRUE),
      dead_fraction = mean(dead_fraction, na.rm = TRUE),
      .groups = "drop"
    )

  competition_init_df <- competition_df %>%
    filter(time_hours == min(time_hours, na.rm = TRUE)) %>%
    group_by(glucose) %>%
    summarise(
      high_fraction_start = mean(high_fraction, na.rm = TRUE),
      total_live_start = mean(total_live, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(glucose)

  competition_total_hours <- max(competition_df$time_hours, na.rm = TRUE)
  competition_times <- sort(unique(competition_df$time_hours))

  validation_tasks <- final_summary %>%
    filter(
      line_id == competition_line_id,
      day2_action == "carry",
      day4_action == "carry",
      g0_day0 %in% competition_init_df$glucose
    ) %>%
    distinct(model_id, line_id, fit_context, direction, fit_type, context_label)

  validation_task_rows <- split(validation_tasks, seq_len(nrow(validation_tasks)))
  validation_sims <- lapply(validation_task_rows, function(task_df) {
    simulate_competition_validation_context(
      task_row = task_df,
      config = cfg,
      stan_data = stan_data,
      init_df = competition_init_df,
      total_hours = competition_total_hours,
      time_step_hours = 1
    )
  })

  validation_sim_df <- bind_rows(validation_sims) %>%
    mutate(
      context_label = if_else(fit_context == "global", "global", paste(direction, fit_type, sep = " | ")),
      high_fraction = if_else(total_live > 0, N_high / total_live, NA_real_),
      log_ratio = log((N_high + 1e-8) / (N_low + 1e-8))
    ) %>%
    filter(time_hours %in% competition_times)

  validation_compare_df <- validation_sim_df %>%
    left_join(
      competition_mean_df %>%
        rename(
          obs_total_live = total_live,
          obs_high_fraction = high_fraction,
          obs_log_ratio = log_ratio,
          obs_dead_fraction = dead_fraction
        ),
      by = c("glucose", "time_hours")
    ) %>%
    mutate(
      abs_err_log_ratio = abs(log_ratio - obs_log_ratio),
      abs_err_high_fraction = abs(high_fraction - obs_high_fraction),
      abs_err_log_total_live = abs(log1p(total_live) - log1p(obs_total_live))
    )

  validation_endpoints <- validation_compare_df %>%
    group_by(model_id, context_label, fit_context, direction, fit_type, glucose) %>%
    summarise(
      sim_start_log_ratio = log_ratio[which.min(time_hours)],
      sim_end_log_ratio = log_ratio[which.max(time_hours)],
      obs_start_log_ratio = obs_log_ratio[which.min(time_hours)],
      obs_end_log_ratio = obs_log_ratio[which.max(time_hours)],
      sim_end_high_fraction = high_fraction[which.max(time_hours)],
      obs_end_high_fraction = obs_high_fraction[which.max(time_hours)],
      .groups = "drop"
    ) %>%
    mutate(
      sim_delta_log_ratio = sim_end_log_ratio - sim_start_log_ratio,
      obs_delta_log_ratio = obs_end_log_ratio - obs_start_log_ratio,
      end_sign_match = sign(sim_delta_log_ratio) == sign(obs_delta_log_ratio)
    )

  competition_validation_summary <- validation_compare_df %>%
    group_by(model_id, context_label, fit_context, direction, fit_type) %>%
    summarise(
      mae_log_ratio = mean(abs_err_log_ratio, na.rm = TRUE),
      mae_high_fraction = mean(abs_err_high_fraction, na.rm = TRUE),
      mae_log_total_live = mean(abs_err_log_total_live, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(
      validation_endpoints %>%
        group_by(model_id, context_label, fit_context, direction, fit_type) %>%
        summarise(
          glucose_sign_match = mean(end_sign_match, na.rm = TRUE),
          mean_abs_end_log_ratio_error = mean(abs(sim_end_log_ratio - obs_end_log_ratio), na.rm = TRUE),
          mean_abs_end_high_fraction_error = mean(abs(sim_end_high_fraction - obs_end_high_fraction), na.rm = TRUE),
          .groups = "drop"
        ),
      by = c("model_id", "context_label", "fit_context", "direction", "fit_type")
    ) %>%
    mutate(
      validation_score = mae_log_ratio + mae_high_fraction + 0.5 * mae_log_total_live,
      context_family = factor(
        context_label,
        levels = c("global", "high_to_low | transfer", "low_to_high | transfer")
      )
    ) %>%
    arrange(desc(glucose_sign_match), validation_score, mae_log_ratio)

  competition_best_contexts <- competition_validation_summary %>%
    group_by(context_label) %>%
    slice_min(order_by = validation_score, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(validation_label = paste(model_id, context_label, sep = " | "))

  competition_overlay_df <- validation_sim_df %>%
    semi_join(
      competition_best_contexts %>% select(model_id, context_label, fit_context, direction, fit_type),
      by = c("model_id", "context_label", "fit_context", "direction", "fit_type")
    ) %>%
    mutate(
      validation_label = paste(model_id, context_label, sep = " | ")
    )
}
```

## Objects

The relevant artefacts are:

- `final_summary`: one row per `model x fit-context x line x strategy`, with the day-6 outcome in `log_ratio_final`.
- `detail.Rds`: optional time-resolved trajectories for rerun strategies, including `N_low`, `N_high`, glucose, and any extra resource / waste states present in that model.

On disk these live under:

- `data/selection_strategy/gpath/v1/<dataset>/<model>/<context>/line_##/final_summary.Rds`
- `data/selection_strategy/gpath/v1/<dataset>/<model>/<context>/line_##/detail.Rds`

## Global Strategy Ordering

This view ignores model identity and asks a simpler question: across all simulated contexts, how consistently does each strategy push selection in one direction?

For each strategy:

- `frac_for` = fraction of contexts with `log_ratio_final > 0`
- `frac_against` = fraction of contexts with `log_ratio_final < 0`
- `signed_consistency = frac_for - frac_against`

So:

- values near `+1` mean the strategy almost always selects for high ploidy
- values near `-1` mean the strategy almost always selects against high ploidy
- values near `0` mean the strategy has no clear directional tendency

```{r strategy-ordering-prep}
strategy_sign_summary <- final_summary %>%
  group_by(strategy_code, g0_day0, day2_action, g0_day2, day4_action, g0_day4) %>%
  summarise(
    n_contexts = n(),
    frac_for = mean(log_ratio_final > 0, na.rm = TRUE),
    frac_against = mean(log_ratio_final < 0, na.rm = TRUE),
    frac_neutral = mean(log_ratio_final == 0, na.rm = TRUE),
    signed_consistency = frac_for - frac_against,
    abs_consistency = pmax(frac_for, frac_against),
    mean_log_ratio = mean(log_ratio_final, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(signed_consistency, desc(abs_consistency), mean_log_ratio)

strategy_levels <- strategy_sign_summary$strategy_code

strategy_score_plot_df <- strategy_sign_summary %>%
  mutate(strategy_code = factor(strategy_code, levels = strategy_levels))

strategy_extreme_codes <- c(
  head(strategy_sign_summary$strategy_code, 25),
  tail(strategy_sign_summary$strategy_code, 25)
)
strategy_extreme_codes <- unique(strategy_extreme_codes)

strategy_heatmap_summary <- strategy_sign_summary %>%
  filter(strategy_code %in% strategy_extreme_codes) %>%
  mutate(strategy_code = factor(strategy_code, levels = strategy_extreme_codes))

strategy_effect_heatmap <- final_summary %>%
  filter(strategy_code %in% strategy_extreme_codes) %>%
  group_by(strategy_code) %>%
  summarise(
    q10 = quantile(log_ratio_final, probs = 0.10, na.rm = TRUE, names = FALSE),
    median = median(log_ratio_final, na.rm = TRUE),
    q90 = quantile(log_ratio_final, probs = 0.90, na.rm = TRUE, names = FALSE),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(cols = c(q10, median, q90), names_to = "summary_stat", values_to = "value") %>%
  mutate(
    strategy_code = factor(strategy_code, levels = strategy_extreme_codes),
    summary_stat = factor(summary_stat, levels = c("q10", "median", "q90"))
  )

effect_limit <- max(abs(strategy_effect_heatmap$value), na.rm = TRUE)
if (!is.finite(effect_limit) || effect_limit <= 0) {
  effect_limit <- 1
}

strategy_design_long <- bind_rows(
  strategy_sign_summary %>%
    transmute(strategy_code, stage = "day0", value = g0_day0, action_label = "seed"),
  strategy_sign_summary %>%
    transmute(strategy_code, stage = "day2", value = g0_day2, action_label = if_else(day2_action == "carry", "carry", "refresh")),
  strategy_sign_summary %>%
    transmute(strategy_code, stage = "day4", value = g0_day4, action_label = if_else(day4_action == "carry", "carry", "refresh"))
) %>%
  mutate(
    glucose_label = if_else(action_label == "carry", "carry", as.character(value)),
    stage = factor(stage, levels = c("day0", "day2", "day4")),
    glucose_label = factor(glucose_label, levels = c("carry", as.character(sort(glucose_values))))
  )

strategy_design_heatmap <- strategy_design_long %>%
  filter(strategy_code %in% strategy_extreme_codes) %>%
  mutate(strategy_code = factor(strategy_code, levels = strategy_extreme_codes))
```

```{r strategy-consistency-line, fig.height=5}
ggplot(strategy_sign_summary, aes(seq_along(signed_consistency), signed_consistency)) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.4) +
  geom_line(linewidth = 0.8, color = "black") +
  labs(
    title = "Per-Strategy Directional Consistency",
    subtitle = "Ordered from consistently against high ploidy to consistently for high ploidy",
    x = "strategies ordered by signed consistency",
    y = "signed consistency = frac_for - frac_against"
  ) +
  theme_minimal(base_size = 11)
```

```{r strategy-ordering-heatmap, fig.height=16}
p_score_heat <- ggplot(strategy_heatmap_summary, aes(x = 1, y = strategy_code, fill = signed_consistency)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-1, 1),
    name = "signed\nconsistency"
  ) +
  labs(
    title = "Most Consistent Strategies",
    subtitle = "Bottom 25 and top 25 strategies by signed consistency",
    x = NULL,
    y = "strategy code"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

p_effect <- ggplot(strategy_effect_heatmap, aes(summary_stat, strategy_code, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-effect_limit, effect_limit),
    name = "log ratio"
  ) +
  labs(
    title = "Outcome Summary",
    subtitle = "q10, median, q90 across all contexts",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11)

p_design <- ggplot(strategy_design_heatmap, aes(stage, strategy_code, fill = glucose_label)) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c(
      "carry" = "#bdbdbd",
      "0" = "#f7fbff",
      "0.1" = "#deebf7",
      "0.25" = "#c6dbef",
      "0.5" = "#9ecae1",
      "1" = "#6baed6",
      "5" = "#3182bd",
      "25" = "#08519c"
    ),
    drop = FALSE,
    name = "glucose"
  ) +
  labs(
    title = "Strategy Design",
    subtitle = "Day-0 seed, then day-2/day-4 action",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11)

p_score_heat + p_effect + p_design + patchwork::plot_layout(widths = c(0.45, 0.8, 1))
```

## Main Result: Robust Strategies That Select Against High Ploidy

Within each `model x fit-context x line` context, rank strategies by `log_ratio_final` from most negative to most positive. Then identify the strategies with the best mean within-context rank across the entire result set.

```{r against-prep}
against_ranked <- final_summary %>%
  group_by(model_id, line_id, cell_line, context_label) %>%
  mutate(against_rank = min_rank(log_ratio_final)) %>%
  ungroup()

against_strategy_summary <- against_ranked %>%
  group_by(strategy_code) %>%
  summarise(
    mean_rank = mean(against_rank, na.rm = TRUE),
    top5_fraction = mean(against_rank <= 5, na.rm = TRUE),
    favorable_fraction = mean(log_ratio_final < 0, na.rm = TRUE),
    mean_log_ratio = mean(log_ratio_final, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(mean_rank, desc(top5_fraction), mean_log_ratio)

against_top_codes <- against_strategy_summary %>%
  slice_head(n = 8) %>%
  pull(strategy_code)

against_plot_df <- against_ranked %>%
  filter(strategy_code %in% against_top_codes) %>%
  mutate(strategy_code = factor(strategy_code, levels = rev(against_top_codes)))

against_limit <- max(abs(against_plot_df$log_ratio_final), na.rm = TRUE)
if (!is.finite(against_limit) || against_limit <= 0) {
  against_limit <- 1
}
```

```{r against-summary}
print(against_strategy_summary %>% slice_head(n = 8), n = 8)
```

```{r against-heatmap, fig.height=14}
ggplot(against_plot_df, aes(model_id, strategy_code, fill = log_ratio_final)) +
  geom_tile(color = "white") +
  facet_grid(cell_line ~ context_label) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-against_limit, against_limit),
    name = "Day-6 log(N_high / N_low)"
  ) +
  labs(
    title = "Robust Strategies For Selecting Against High Ploidy",
    subtitle = "Same strategy set shown across all models, cell lines, and fit contexts",
    x = "model",
    y = "strategy code"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Main Result: Robust Strategies That Select For High Ploidy

Repeat the same ranking logic, but now rank strategies from most positive to most negative `log_ratio_final`.

```{r for-prep}
for_ranked <- final_summary %>%
  group_by(model_id, line_id, cell_line, context_label) %>%
  mutate(for_rank = min_rank(desc(log_ratio_final))) %>%
  ungroup()

for_strategy_summary <- for_ranked %>%
  group_by(strategy_code) %>%
  summarise(
    mean_rank = mean(for_rank, na.rm = TRUE),
    top5_fraction = mean(for_rank <= 5, na.rm = TRUE),
    favorable_fraction = mean(log_ratio_final > 0, na.rm = TRUE),
    mean_log_ratio = mean(log_ratio_final, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(mean_rank, desc(top5_fraction), desc(mean_log_ratio))

for_top_codes <- for_strategy_summary %>%
  slice_head(n = 8) %>%
  pull(strategy_code)

for_plot_df <- for_ranked %>%
  filter(strategy_code %in% for_top_codes) %>%
  mutate(strategy_code = factor(strategy_code, levels = rev(for_top_codes)))

for_limit <- max(abs(for_plot_df$log_ratio_final), na.rm = TRUE)
if (!is.finite(for_limit) || for_limit <= 0) {
  for_limit <- 1
}
```

```{r for-summary}
print(for_strategy_summary %>% slice_head(n = 8), n = 8)
```

```{r for-heatmap, fig.height=14}
ggplot(for_plot_df, aes(model_id, strategy_code, fill = log_ratio_final)) +
  geom_tile(color = "white") +
  facet_grid(cell_line ~ context_label) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-for_limit, for_limit),
    name = "Day-6 log(N_high / N_low)"
  ) +
  labs(
    title = "Robust Strategies For Selecting For High Ploidy",
    subtitle = "Same strategy set shown across all models, cell lines, and fit contexts",
    x = "model",
    y = "strategy code"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Empirical Competition Check: SUM-159

This section compares the mixed-culture simulation branch against the empirical SUM-159 co-culture experiment in `data/counts/competition/SUM-159-fuse.Rds`.

The alignment used here is:

- SUM-159 corresponds to `line_id = 5`
- `2N` is mapped to `N_low`
- `4N` is mapped to `N_high`
- each frame is treated as an `8 h` interval
- the experiment is treated as a no-refresh mixed-culture trajectory over `112 h`
- simulations are initialized from the empirical frame-0 live total and high-ploidy fraction at each glucose, then evolved using the existing fitted contexts

So this is not a new fitting procedure. It is a validation rerun that uses the already-fitted mixed-culture dynamics to generate detailed trajectories on the empirical time grid.

```{r competition-availability, results='asis'}
if (!competition_available) {
  cat("Competition dataset not found at `", params$competition_path, "`. Skipping empirical validation section.\n", sep = "")
}
```

```{r competition-initial-state, eval=competition_available}
competition_init_fmt <- competition_init_df %>%
  mutate(
    high_fraction_start = percent(high_fraction_start, accuracy = 0.1),
    total_live_start = round(total_live_start, 1)
  )

knitr::kable(competition_init_fmt)
```

```{r competition-validation-summary, eval=competition_available}
competition_validation_fmt <- competition_validation_summary %>%
  select(
    model_id, context_label, glucose_sign_match, mae_log_ratio,
    mae_high_fraction, mae_log_total_live, validation_score
  ) %>%
  mutate(
    glucose_sign_match = percent(glucose_sign_match, accuracy = 1),
    mae_high_fraction = percent(mae_high_fraction, accuracy = 0.1)
  )

knitr::kable(competition_validation_fmt, digits = 3)
```

`glucose_sign_match` is the fraction of empirical glucose levels for which the simulation gets the sign of the start-to-end log-ratio shift correct. That is the most basic directional check. The error columns then ask how close the simulated trajectories stay to the empirical mean trajectories over time.

```{r competition-validation-heatmap, fig.height=7, eval=competition_available}
heat_df <- competition_validation_summary %>%
  mutate(
    context_label = factor(
      context_label,
      levels = c("global", "high_to_low | transfer", "low_to_high | transfer")
    ),
    model_id = factor(model_id, levels = unique(model_id))
  )

ggplot(heat_df, aes(model_id, context_label, fill = glucose_sign_match)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", validation_score)), size = 3) +
  scale_fill_gradient(
    low = "#f7fbff",
    high = "#08519c",
    limits = c(0, 1),
    labels = percent_format(accuracy = 1),
    name = "glucose\nsign match"
  ) +
  labs(
    title = "SUM-159 Competition Validation Across Existing Fitted Contexts",
    subtitle = "Tile color shows directional agreement; overlaid text shows composite validation score (lower is better)",
    x = "model",
    y = "fit context"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r competition-log-ratio-overlay, fig.height=12, eval=competition_available}
overlay_empirical <- competition_df %>%
  mutate(validation_label = "empirical")

overlay_plot_df <- bind_rows(
  competition_overlay_df %>%
    transmute(
      glucose,
      time_hours,
      rep = NA_character_,
      validation_label,
      source = "simulation",
      high_fraction,
      log_ratio,
      total_live
    ),
  competition_df %>%
    transmute(
      glucose,
      time_hours,
      rep,
      validation_label = "empirical",
      source = "empirical_replicate",
      high_fraction,
      log_ratio,
      total_live
    ),
  competition_mean_df %>%
    transmute(
      glucose,
      time_hours,
      rep = NA_character_,
      validation_label = "empirical",
      source = "empirical_mean",
      high_fraction,
      log_ratio,
      total_live
    )
)

ggplot() +
  geom_line(
    data = overlay_plot_df %>% filter(source == "empirical_replicate"),
    aes(time_hours / 24, log_ratio, group = interaction(glucose, rep)),
    color = "grey75",
    linewidth = 0.5,
    alpha = 0.6
  ) +
  geom_line(
    data = overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, log_ratio),
    color = "black",
    linewidth = 1
  ) +
  geom_point(
    data = overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, log_ratio),
    color = "black",
    size = 1.4
  ) +
  geom_line(
    data = overlay_plot_df %>% filter(source == "simulation"),
    aes(time_hours / 24, log_ratio, color = validation_label),
    linewidth = 1
  ) +
  geom_hline(yintercept = 0, color = "grey55", linewidth = 0.4) +
  facet_grid(factor(glucose) ~ validation_label, scales = "free_y") +
  labs(
    title = "Observed vs Simulated Competition Log-Ratio Trajectories",
    subtitle = "Black = empirical mean, grey = empirical replicates, color = best fitted context within each fit family",
    x = "days",
    y = "log(N_high / N_low)",
    color = "best context"
  ) +
  theme_minimal(base_size = 11)
```

```{r competition-high-fraction-overlay, fig.height=12, eval=competition_available}
ggplot() +
  geom_line(
    data = overlay_plot_df %>% filter(source == "empirical_replicate"),
    aes(time_hours / 24, high_fraction, group = interaction(glucose, rep)),
    color = "grey75",
    linewidth = 0.5,
    alpha = 0.6
  ) +
  geom_line(
    data = overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, high_fraction),
    color = "black",
    linewidth = 1
  ) +
  geom_point(
    data = overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, high_fraction),
    color = "black",
    size = 1.4
  ) +
  geom_line(
    data = overlay_plot_df %>% filter(source == "simulation"),
    aes(time_hours / 24, high_fraction, color = validation_label),
    linewidth = 1
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  facet_grid(factor(glucose) ~ validation_label) +
  labs(
    title = "Observed vs Simulated High-Ploidy Fraction",
    subtitle = "Same comparison as above, on the directly interpretable composition scale",
    x = "days",
    y = "high-ploidy fraction among live cells",
    color = "best context"
  ) +
  theme_minimal(base_size = 11)
```

```{r competition-total-live-overlay, fig.height=12, eval=competition_available}
ggplot() +
  geom_line(
    data = overlay_plot_df %>% filter(source == "empirical_replicate"),
    aes(time_hours / 24, total_live, group = interaction(glucose, rep)),
    color = "grey75",
    linewidth = 0.5,
    alpha = 0.6
  ) +
  geom_line(
    data = overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, total_live),
    color = "black",
    linewidth = 1
  ) +
  geom_point(
    data = overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, total_live),
    color = "black",
    size = 1.4
  ) +
  geom_line(
    data = overlay_plot_df %>% filter(source == "simulation"),
    aes(time_hours / 24, total_live, color = validation_label),
    linewidth = 1
  ) +
  facet_grid(factor(glucose) ~ validation_label, scales = "free_y") +
  labs(
    title = "Observed vs Simulated Total Live Burden",
    subtitle = "Simulations are initialized at the empirical frame-0 live total for each glucose",
    x = "days",
    y = "total live cells",
    color = "best context"
  ) +
  theme_minimal(base_size = 11)
```

This validation should be read as a credibility check on the mixed-culture branch, not as a replacement objective for the strategy ranking. If a context gets the sign of competitive shift wrong across glucose, the corresponding strategy recommendations should be treated cautiously for SUM-159 even if the simulated ranking itself looks internally stable.

## Optional Detailed Follow-Up

This section reruns a very small number of representative top strategies in detailed mode. It is secondary to the heatmaps above.

```{r representative-selection, eval=params$include_detailed}
rep_model_ids <- cfg$representative_model_ids
if (is.null(rep_model_ids) || !length(rep_model_ids)) {
  rep_model_ids <- unique(head(as.character(against_plot_df$model_id), 2L))
}

rep_rows <- bind_rows(
  against_plot_df %>%
    filter(model_id %in% rep_model_ids, strategy_code == against_top_codes[[1]]) %>%
    group_by(model_id, cell_line, context_label) %>%
    slice_head(n = 1L) %>%
    ungroup() %>%
    mutate(selection_objective = "against_high"),
  for_plot_df %>%
    filter(model_id %in% rep_model_ids, strategy_code == for_top_codes[[1]]) %>%
    group_by(model_id, cell_line, context_label) %>%
    slice_head(n = 1L) %>%
    ungroup() %>%
    mutate(selection_objective = "for_high")
)

detail_df <- rerun_selected_strategies_detailed(
  config = cfg,
  selected_rows = rep_rows,
  time_step_hours = cfg$detailed_time_step_hours
)

if (nrow(detail_df)) {
  detail_df <- detail_df %>%
    left_join(line_map, by = "line_id") %>%
    mutate(
      cell_line = if_else(is.na(cell_line) | !nzchar(cell_line), sprintf("line_%02d", as.integer(line_id)), cell_line),
      context_label = if_else(fit_context == "global", "global", paste(direction, fit_type, sep = " | "))
    )
}
```

```{r detailed-ratio, fig.height=10, eval=params$include_detailed}
if (nrow(detail_df)) {
  detail_ratio <- detail_df %>%
    mutate(log_ratio = log((N_high + 1e-8) / (N_low + 1e-8)))

  ggplot(detail_ratio, aes(time_global / 24, log_ratio, color = strategy_code, group = interaction(model_id, cell_line, context_label, strategy_code, selection_objective))) +
    geom_line(linewidth = 0.9) +
    facet_grid(cell_line ~ model_id + context_label, scales = "free_y") +
    labs(
      title = "Representative Detailed Log-Ratio Trajectories",
      x = "days",
      y = "log(N_high / N_low)",
      color = "strategy"
    ) +
    theme_minimal(base_size = 11)
}
```

## Agent Summary

```{r agent-summary, include=FALSE}
against_text <- against_strategy_summary %>%
  slice_head(n = 5) %>%
  transmute(
    line = sprintf(
      "%s | mean_rank=%.2f | top5=%.2f | frac_against=%.2f | mean_log_ratio=%.3f",
      strategy_code, mean_rank, top5_fraction, favorable_fraction, mean_log_ratio
    )
  ) %>%
  pull(line)

for_text <- for_strategy_summary %>%
  slice_head(n = 5) %>%
  transmute(
    line = sprintf(
      "%s | mean_rank=%.2f | top5=%.2f | frac_for=%.2f | mean_log_ratio=%.3f",
      strategy_code, mean_rank, top5_fraction, favorable_fraction, mean_log_ratio
    )
  ) %>%
  pull(line)

consistency_extremes <- strategy_sign_summary %>%
  slice(unique(c(seq_len(min(5, nrow(strategy_sign_summary))), seq.int(max(1, nrow(strategy_sign_summary) - 4), nrow(strategy_sign_summary))))) %>%
  transmute(
    line = sprintf(
      "%s | signed_consistency=%.3f | frac_for=%.2f | frac_against=%.2f | mean_log_ratio=%.3f",
      strategy_code, signed_consistency, frac_for, frac_against, mean_log_ratio
    )
  ) %>%
  pull(line)

against_best <- against_strategy_summary %>% slice_head(n = 1)
for_best <- for_strategy_summary %>% slice_head(n = 1)
competition_text <- character(0)

if (competition_available && nrow(competition_validation_summary)) {
  competition_text <- competition_validation_summary %>%
    slice_head(n = 5) %>%
    transmute(
      line = sprintf(
        "%s | %s | sign_match=%.2f | mae_log_ratio=%.3f | mae_high_fraction=%.3f | validation_score=%.3f",
        model_id, context_label, glucose_sign_match, mae_log_ratio, mae_high_fraction, validation_score
      )
    ) %>%
    pull(line)
}

verdict_lines <- c(
  sprintf(
    "best_against_high\t%s | mean_rank=%.2f | frac_against=%.2f | mean_log_ratio=%.3f",
    against_best$strategy_code[[1]],
    against_best$mean_rank[[1]],
    against_best$favorable_fraction[[1]],
    against_best$mean_log_ratio[[1]]
  ),
  sprintf(
    "best_for_high\t%s | mean_rank=%.2f | frac_for=%.2f | mean_log_ratio=%.3f",
    for_best$strategy_code[[1]],
    for_best$mean_rank[[1]],
    for_best$favorable_fraction[[1]],
    for_best$mean_log_ratio[[1]]
  ),
  sprintf(
    "verdict\tRobust selection FOR high ploidy is substantially stronger than robust selection AGAINST high ploidy in this strategy space. Best-for strategies reach frac_for around %.2f-%.2f, whereas best-against strategies only reach frac_against around %.2f-%.2f.",
    min(head(for_strategy_summary$favorable_fraction, 5)),
    max(head(for_strategy_summary$favorable_fraction, 5)),
    min(head(against_strategy_summary$favorable_fraction, 5)),
    max(head(against_strategy_summary$favorable_fraction, 5))
  )
)

summary_lines <- c(
  "SELECTION_STRATEGY_ANALYSIS",
  sprintf("generated\t%s", as.character(Sys.time())),
  sprintf("config\t%s", params$config_path),
  sprintf("output_root\t%s", params$output_root),
  sprintf("n_models\t%d", dplyr::n_distinct(final_summary$model_id)),
  sprintf("n_contexts\t%d", dplyr::n_distinct(paste(final_summary$model_id, final_summary$line_id, final_summary$context_label, sep = '||'))),
  sprintf("n_strategies\t%d", dplyr::n_distinct(final_summary$strategy_code)),
  "",
  "VERDICT",
  verdict_lines,
  "",
  "TOP_AGAINST_HIGH",
  against_text,
  "",
  "TOP_FOR_HIGH",
  for_text,
  "",
  "CONSISTENCY_EXTREMES",
  consistency_extremes,
  "",
  "COMPETITION_VALIDATION",
  competition_text
)

writeLines(summary_lines, agent_text_path)
```
