---
title: "Selection Strategy Analysis"
author: "Codex"
date: "2026-03-21"
output: html_document
params:
  config_path: "scripts/selection_strategy_config.R"
  output_root: !r file.path("data", "selection_strategy", "gpath", "v1", "gstarvation_v1")
  include_detailed: false
  competition_path: "data/counts/competition/SUM-159-fuse.Rds"
  competition_fit_dataset_label: "gstarvation_v1_single_line"
  competition_fit_stan_data_path: "data/inputs/stan/gstarvation_v1_single_line/line_05/stan_ready_data.Rds"
  competition_fit_run_label: "line05_multistart"
  competition_export_root: !r file.path("data", "selection_strategy", "gpath", "v1", "gstarvation_v1_single_line_competition_validation")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook reports the main selection-strategy result from the mixed-ploidy simulations: which 6-day media schedules most strongly select for or against high ploidy by day 6.

Each strategy is a three-digit code:

- digit 1: initial glucose condition at day 0,
- digit 2: action at day 2 (`0` = no refresh, `1:7` = refresh to one of the tested glucose conditions),
- digit 3: action at day 4 (`0` = no refresh, `1:7` = refresh to one of the tested glucose conditions).

The primary outcome is the day-6 log ratio:

`log_ratio_final = log(N_high / N_low)`

More negative values mean stronger selection against high ploidy. More positive values mean stronger selection for high ploidy.

## Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(patchwork)

source("R/selection_strategy_utils.R")

cfg <- load_selection_config(params$config_path)
cfg$output_root <- dirname(params$output_root)
cfg$dataset_label <- basename(params$output_root)

stan_data <- resolve_selection_stan(cfg)
line_map <- get_line_label_map(stan_data)
glucose_values <- resolve_glucose_values(stan_data, cfg)
strategy_grid <- build_strategy_grid(glucose_values)

outputs <- collect_selection_outputs(cfg)
final_summary <- outputs$final_summary
detail_summary <- outputs$detail

if (!nrow(final_summary)) {
  stop("No selection-strategy outputs were found. Run scripts/run_gpath_selection_strategy_batch.R first.")
}

final_summary <- final_summary %>%
  left_join(line_map, by = "line_id") %>%
  left_join(
    strategy_grid %>%
      select(strategy_code, g0_day0, day2_action, g0_day2, day4_action, g0_day4),
    by = "strategy_code"
  ) %>%
  mutate(
    cell_line = if_else(is.na(cell_line) | !nzchar(cell_line), sprintf("line_%02d", as.integer(line_id)), cell_line),
    context_label = if_else(fit_context == "global", "global", paste(direction, fit_type, sep = " | "))
  )

agent_text_path <- file.path("workflow", "selection_strategy_analysis.txt")
competition_line_id <- 5L
competition_frame_hours <- 8

competition_eval <- evaluate_single_line_competition_validation(
  config = cfg,
  competition_path = params$competition_path,
  fit_dataset_label = params$competition_fit_dataset_label,
  fit_stan_data_path = params$competition_fit_stan_data_path,
  run_label = params$competition_fit_run_label,
  competition_frame_hours = competition_frame_hours,
  time_step_hours = 1,
  export_root = params$competition_export_root
)

competition_available <- isTRUE(competition_eval$competition_available)
competition_df <- competition_eval$competition_df
competition_mean_df <- competition_eval$competition_mean_df
competition_init_df <- competition_eval$competition_init_df

load_competition_fit_context <- function(row, config) {
  if (identical(row$fit_context[[1]], "global")) {
    load_best_global_fit(config = config, model_id = row$model_id[[1]])
  } else {
    load_best_transfer_fit_context(
      config = config,
      model_id = row$model_id[[1]],
      line_id = row$line_id[[1]],
      direction = row$direction[[1]],
      fit_type = row$fit_type[[1]]
    )
  }
}

simulate_competition_validation_context <- function(task_row, config, stan_data, init_df, total_hours, time_step_hours = 1) {
  task_row <- as.data.frame(task_row, stringsAsFactors = FALSE)
  fit_obj <- load_competition_fit_context(task_row, config)
  line_pair <- get_line_ploidy_pair(stan_data, line_id = task_row$line_id[[1]])

  comp <- reconstruct_mixture_components(
    draw_vec = fit_obj$draw_vec,
    model_id = task_row$model_id[[1]],
    line_id = task_row$line_id[[1]],
    low_ploidy = line_pair$low,
    high_ploidy = line_pair$high,
    ploidy_effect_mask = fit_obj$ploidy_effect_mask
  )

  pieces <- vector("list", nrow(init_df))
  for (i in seq_len(nrow(init_df))) {
    init_row <- init_df[i, , drop = FALSE]
    state0 <- make_initial_mix_state(
      parms_mix = comp$parms_mix,
      g0 = init_row$glucose[[1]],
      seed_total_n = init_row$total_live_start[[1]],
      initial_high_fraction = init_row$high_fraction_start[[1]]
    )

    sim <- simulate_mixture_segment(
      parms_mix = comp$parms_mix,
      state0 = state0,
      hours = total_hours,
      time_step_hours = time_step_hours,
      detailed = TRUE
    )

    sim$glucose <- init_row$glucose[[1]]
    sim$time_hours <- sim$time
    sim$model_id <- task_row$model_id[[1]]
    sim$line_id <- task_row$line_id[[1]]
    sim$fit_context <- task_row$fit_context[[1]]
    sim$direction <- task_row$direction[[1]]
    sim$fit_type <- task_row$fit_type[[1]]
    pieces[[i]] <- sim
  }

  bind_rows(pieces)
}

competition_validation_summary <- data.frame()
competition_overlay_df <- data.frame()
competition_best_contexts <- data.frame()
competition_global_summary <- data.frame()
competition_global_overlay_df <- data.frame()
competition_transfer_summary <- data.frame()
competition_transfer_overlay_df <- data.frame()

if (competition_available) {
  competition_total_hours <- competition_eval$competition_total_hours
  competition_times <- competition_eval$competition_times

  validation_tasks <- final_summary %>%
    filter(
      line_id == competition_line_id,
      day2_action == "carry",
      day4_action == "carry",
      g0_day0 %in% competition_init_df$glucose
    ) %>%
    distinct(model_id, line_id, fit_context, direction, fit_type, context_label)

  validation_task_rows <- split(validation_tasks, seq_len(nrow(validation_tasks)))
  validation_sims <- lapply(validation_task_rows, function(task_df) {
    simulate_competition_validation_context(
      task_row = task_df,
      config = cfg,
      stan_data = stan_data,
      init_df = competition_init_df,
      total_hours = competition_total_hours,
      time_step_hours = 1
    )
  })

  validation_sim_df <- bind_rows(validation_sims) %>%
    mutate(
      context_label = if_else(fit_context == "global", "global", paste(direction, fit_type, sep = " | ")),
      high_fraction = if_else(total_live > 0, N_high / total_live, NA_real_),
      log_ratio = log((N_high + 1e-8) / (N_low + 1e-8))
    ) %>%
    filter(time_hours %in% competition_times)

  validation_compare_df <- validation_sim_df %>%
    left_join(
      competition_mean_df %>%
        rename(
          obs_total_live = total_live,
          obs_high_fraction = high_fraction,
          obs_log_ratio = log_ratio,
          obs_dead_fraction = dead_fraction
        ),
      by = c("glucose", "time_hours")
    ) %>%
    mutate(
      abs_err_log_ratio = abs(log_ratio - obs_log_ratio),
      abs_err_high_fraction = abs(high_fraction - obs_high_fraction),
      abs_err_log_total_live = abs(log1p(total_live) - log1p(obs_total_live))
    )

  validation_endpoints <- validation_compare_df %>%
    group_by(model_id, context_label, fit_context, direction, fit_type, glucose) %>%
    summarise(
      sim_start_log_ratio = log_ratio[which.min(time_hours)],
      sim_end_log_ratio = log_ratio[which.max(time_hours)],
      obs_start_log_ratio = obs_log_ratio[which.min(time_hours)],
      obs_end_log_ratio = obs_log_ratio[which.max(time_hours)],
      sim_end_high_fraction = high_fraction[which.max(time_hours)],
      obs_end_high_fraction = obs_high_fraction[which.max(time_hours)],
      .groups = "drop"
    ) %>%
    mutate(
      sim_delta_log_ratio = sim_end_log_ratio - sim_start_log_ratio,
      obs_delta_log_ratio = obs_end_log_ratio - obs_start_log_ratio,
      end_sign_match = sign(sim_delta_log_ratio) == sign(obs_delta_log_ratio)
    )

  competition_validation_summary <- validation_compare_df %>%
    group_by(model_id, context_label, fit_context, direction, fit_type) %>%
    summarise(
      mae_log_ratio = mean(abs_err_log_ratio, na.rm = TRUE),
      mae_high_fraction = mean(abs_err_high_fraction, na.rm = TRUE),
      mae_log_total_live = mean(abs_err_log_total_live, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(
      validation_endpoints %>%
        group_by(model_id, context_label, fit_context, direction, fit_type) %>%
        summarise(
          glucose_sign_match = mean(end_sign_match, na.rm = TRUE),
          mean_abs_end_log_ratio_error = mean(abs(sim_end_log_ratio - obs_end_log_ratio), na.rm = TRUE),
          mean_abs_end_high_fraction_error = mean(abs(sim_end_high_fraction - obs_end_high_fraction), na.rm = TRUE),
          .groups = "drop"
        ),
      by = c("model_id", "context_label", "fit_context", "direction", "fit_type")
    ) %>%
    mutate(
      validation_score = mae_log_ratio + mae_high_fraction + 0.5 * mae_log_total_live,
      context_family = factor(
        context_label,
        levels = c("global", "high_to_low | transfer", "low_to_high | transfer")
      )
    ) %>%
    arrange(desc(glucose_sign_match), validation_score, mae_log_ratio)

  competition_best_contexts <- competition_validation_summary %>%
    group_by(context_label) %>%
    slice_min(order_by = validation_score, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(validation_label = paste(model_id, context_label, sep = " | "))

  competition_overlay_df <- validation_sim_df %>%
    semi_join(
      competition_best_contexts %>% select(model_id, context_label, fit_context, direction, fit_type),
      by = c("model_id", "context_label", "fit_context", "direction", "fit_type")
    ) %>%
    mutate(
      validation_label = paste(model_id, context_label, sep = " | ")
    )

  competition_global_summary <- competition_validation_summary %>%
    filter(fit_context == "global") %>%
    arrange(desc(glucose_sign_match), validation_score, mae_log_ratio)

  competition_global_overlay_df <- validation_sim_df %>%
    semi_join(
      competition_global_summary %>%
        select(model_id, context_label, fit_context, direction, fit_type),
      by = c("model_id", "context_label", "fit_context", "direction", "fit_type")
    ) %>%
    mutate(validation_label = model_id)

  competition_transfer_summary <- competition_validation_summary %>%
    filter(fit_context == "transfer") %>%
    arrange(direction, desc(glucose_sign_match), validation_score, mae_log_ratio)

  competition_transfer_overlay_df <- validation_sim_df %>%
    semi_join(
      competition_transfer_summary %>%
        select(model_id, context_label, fit_context, direction, fit_type),
      by = c("model_id", "context_label", "fit_context", "direction", "fit_type")
    ) %>%
    mutate(validation_label = paste(model_id, direction, sep = " | "))
}

singleline_competition_validation_summary <- competition_eval$summary_df
singleline_competition_overlay_df <- competition_eval$overlay_df
singleline_competition_overlay_models <- competition_eval$overlay_models
```

## Objects

The relevant artefacts are:

- `final_summary`: one row per `model x fit-context x line x strategy`, with the day-6 outcome in `log_ratio_final`.
- `detail.Rds`: optional time-resolved trajectories for rerun strategies, including `N_low`, `N_high`, glucose, and any extra resource / waste states present in that model.

On disk these live under:

- `data/selection_strategy/gpath/v1/<dataset>/<model>/<context>/line_##/final_summary.Rds`
- `data/selection_strategy/gpath/v1/<dataset>/<model>/<context>/line_##/detail.Rds`

## Strategy Questions

The strategy-ranking summaries below are split by question rather than pooled across all fit contexts. `global` and `transfer` answer different questions and are now reported separately.

```{r strategy-question-prep}
build_strategy_question <- function(df, glucose_values, top_n = 8L, extreme_n = 25L) {
  sign_summary <- df %>%
    group_by(strategy_code, g0_day0, day2_action, g0_day2, day4_action, g0_day4) %>%
    summarise(
      n_contexts = n(),
      frac_for = mean(log_ratio_final > 0, na.rm = TRUE),
      frac_against = mean(log_ratio_final < 0, na.rm = TRUE),
      signed_consistency = frac_for - frac_against,
      abs_consistency = pmax(frac_for, frac_against),
      mean_log_ratio = mean(log_ratio_final, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(signed_consistency, desc(abs_consistency), mean_log_ratio)

  extreme_codes <- unique(c(head(sign_summary$strategy_code, extreme_n), tail(sign_summary$strategy_code, extreme_n)))

  effect_heatmap <- df %>%
    filter(strategy_code %in% extreme_codes) %>%
    group_by(strategy_code) %>%
    summarise(
      q10 = quantile(log_ratio_final, probs = 0.10, na.rm = TRUE, names = FALSE),
      median = median(log_ratio_final, na.rm = TRUE),
      q90 = quantile(log_ratio_final, probs = 0.90, na.rm = TRUE, names = FALSE),
      .groups = "drop"
    ) %>%
    tidyr::pivot_longer(cols = c(q10, median, q90), names_to = "summary_stat", values_to = "value") %>%
    mutate(
      strategy_code = factor(strategy_code, levels = extreme_codes),
      summary_stat = factor(summary_stat, levels = c("q10", "median", "q90"))
    )

  design_heatmap <- bind_rows(
    sign_summary %>% transmute(strategy_code, stage = "day0", value = g0_day0, action_label = "seed"),
    sign_summary %>% transmute(strategy_code, stage = "day2", value = g0_day2, action_label = if_else(day2_action == "carry", "carry", "refresh")),
    sign_summary %>% transmute(strategy_code, stage = "day4", value = g0_day4, action_label = if_else(day4_action == "carry", "carry", "refresh"))
  ) %>%
    mutate(
      glucose_label = if_else(action_label == "carry", "carry", as.character(value)),
      stage = factor(stage, levels = c("day0", "day2", "day4")),
      glucose_label = factor(glucose_label, levels = c("carry", as.character(sort(glucose_values)))),
      strategy_code = factor(strategy_code, levels = extreme_codes)
    ) %>%
    filter(as.character(strategy_code) %in% extreme_codes)

  summarize_rank <- function(descending = FALSE, favorable_sign = c("negative", "positive")) {
    favorable_sign <- match.arg(favorable_sign)
    ranked <- df %>%
      group_by(model_id, line_id, cell_line, context_label) %>%
      mutate(rank_value = if (descending) min_rank(desc(log_ratio_final)) else min_rank(log_ratio_final)) %>%
      ungroup()

    summary_df <- ranked %>%
      group_by(strategy_code) %>%
      summarise(
        mean_rank = mean(rank_value, na.rm = TRUE),
        top5_fraction = mean(rank_value <= 5, na.rm = TRUE),
        favorable_fraction = if (favorable_sign == "negative") {
          mean(log_ratio_final < 0, na.rm = TRUE)
        } else {
          mean(log_ratio_final > 0, na.rm = TRUE)
        },
        mean_log_ratio = mean(log_ratio_final, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      arrange(mean_rank, desc(top5_fraction), if (descending) desc(mean_log_ratio) else mean_log_ratio)

    top_codes <- summary_df %>% slice_head(n = top_n) %>% pull(strategy_code)
    plot_df <- ranked %>%
      filter(strategy_code %in% top_codes) %>%
      mutate(strategy_code = factor(strategy_code, levels = rev(top_codes)))

    list(
      summary = summary_df,
      plot_df = plot_df,
      limit = max(1, max(abs(plot_df$log_ratio_final), na.rm = TRUE))
    )
  }

  list(
    sign_summary = sign_summary,
    effect_heatmap = effect_heatmap,
    design_heatmap = design_heatmap,
    effect_limit = max(1, max(abs(effect_heatmap$value), na.rm = TRUE)),
    against = summarize_rank(descending = FALSE, favorable_sign = "negative"),
    for_high = summarize_rank(descending = TRUE, favorable_sign = "positive")
  )
}

global_strategy <- build_strategy_question(final_summary %>% filter(fit_context == "global"), glucose_values)
transfer_strategy <- build_strategy_question(final_summary %>% filter(fit_context == "transfer"), glucose_values)

strategy_metric_glossary <- data.frame(
  column = c("mean_rank", "top5_fraction", "favorable_fraction", "mean_log_ratio"),
  definition = c(
    "Within the question-specific subset, rank all strategies separately inside each model x line x context block; then average that rank for the strategy across those blocks.",
    "Within the same question-specific subset, fraction of model x line x context blocks where the strategy ranked in the top 5 by the relevant ordering.",
    "Within the same question-specific subset, fraction of rows where the strategy had the desired sign: log_ratio_final < 0 for against-high, log_ratio_final > 0 for for-high.",
    "Arithmetic mean of log_ratio_final across all rows in the question-specific subset for that strategy."
  ),
  stringsAsFactors = FALSE
)
```

## Question 1: Which strategies look most robust within the global fits?

This section uses `fit_context == "global"` only.

```{r strategy-global-consistency-line, fig.height=5}
ggplot(global_strategy$sign_summary, aes(seq_along(signed_consistency), signed_consistency)) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.4) +
  geom_line(linewidth = 0.8, color = "black") +
  labs(title = "Global Directional Consistency", x = "strategies ordered by signed consistency", y = "frac_for - frac_against") +
  theme_minimal(base_size = 11)
```

```{r strategy-global-ordering-heatmap, fig.height=16}
global_sign_plot <- global_strategy$sign_summary %>%
  filter(strategy_code %in% levels(global_strategy$effect_heatmap$strategy_code)) %>%
  mutate(strategy_code = factor(strategy_code, levels = levels(global_strategy$effect_heatmap$strategy_code)))

global_score_heat <- ggplot(global_sign_plot, aes(x = 1, y = strategy_code, fill = signed_consistency)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, limits = c(-1, 1), name = "signed consistency") +
  labs(title = "Global: Consistency", x = NULL, y = "strategy code") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "bottom")

global_effect <- ggplot(global_strategy$effect_heatmap, aes(summary_stat, strategy_code, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, limits = c(-global_strategy$effect_limit, global_strategy$effect_limit), name = "log ratio") +
  labs(title = "Global: Outcome Summary", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

global_design <- ggplot(global_strategy$design_heatmap, aes(stage, strategy_code, fill = glucose_label)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("carry" = "#bdbdbd", "0" = "#f7fbff", "0.1" = "#deebf7", "0.25" = "#c6dbef", "0.5" = "#9ecae1", "1" = "#6baed6", "5" = "#3182bd", "25" = "#08519c"), drop = FALSE, name = "strategy legend") +
  labs(title = "Global: Design", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

global_score_heat + global_effect + global_design + patchwork::plot_layout(widths = c(0.45, 0.8, 1), guides = "collect") & theme(legend.position = "bottom")
```

Within each `model x fit-context x line` context, rank strategies by `log_ratio_final` from most negative to most positive. Then identify the strategies with the best mean within-context rank across the global subset.

```{r strategy-metric-glossary}
knitr::kable(strategy_metric_glossary)
```

```{r against-global-summary}
print(global_strategy$against$summary %>% slice_head(n = 8), n = 8)
```

```{r against-global-heatmap, fig.height=12}
ggplot(global_strategy$against$plot_df, aes(model_id, strategy_code, fill = log_ratio_final)) +
  geom_tile(color = "white") +
  facet_grid(cell_line ~ context_label) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, limits = c(-global_strategy$against$limit, global_strategy$against$limit), name = "day-6 log ratio") +
  labs(title = "Global: Against High Ploidy", x = "model", y = "strategy code") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```

```{r for-global-summary}
print(global_strategy$for_high$summary %>% slice_head(n = 8), n = 8)
```

```{r for-global-heatmap, fig.height=12}
ggplot(global_strategy$for_high$plot_df, aes(model_id, strategy_code, fill = log_ratio_final)) +
  geom_tile(color = "white") +
  facet_grid(cell_line ~ context_label) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, limits = c(-global_strategy$for_high$limit, global_strategy$for_high$limit), name = "day-6 log ratio") +
  labs(title = "Global: For High Ploidy", x = "model", y = "strategy code") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```

## Question 2: Which strategies look most robust within the transfer fits?

This section uses `fit_context == "transfer"` only.

```{r strategy-transfer-consistency-line, fig.height=5}
ggplot(transfer_strategy$sign_summary, aes(seq_along(signed_consistency), signed_consistency)) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.4) +
  geom_line(linewidth = 0.8, color = "black") +
  labs(title = "Transfer Directional Consistency", x = "strategies ordered by signed consistency", y = "frac_for - frac_against") +
  theme_minimal(base_size = 11)
```

```{r strategy-transfer-ordering-heatmap, fig.height=16}
transfer_sign_plot <- transfer_strategy$sign_summary %>%
  filter(strategy_code %in% levels(transfer_strategy$effect_heatmap$strategy_code)) %>%
  mutate(strategy_code = factor(strategy_code, levels = levels(transfer_strategy$effect_heatmap$strategy_code)))

transfer_score_heat <- ggplot(transfer_sign_plot, aes(x = 1, y = strategy_code, fill = signed_consistency)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, limits = c(-1, 1), name = "signed consistency") +
  labs(title = "Transfer: Consistency", x = NULL, y = "strategy code") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "bottom")

transfer_effect <- ggplot(transfer_strategy$effect_heatmap, aes(summary_stat, strategy_code, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, limits = c(-transfer_strategy$effect_limit, transfer_strategy$effect_limit), name = "log ratio") +
  labs(title = "Transfer: Outcome Summary", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

transfer_design <- ggplot(transfer_strategy$design_heatmap, aes(stage, strategy_code, fill = glucose_label)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("carry" = "#bdbdbd", "0" = "#f7fbff", "0.1" = "#deebf7", "0.25" = "#c6dbef", "0.5" = "#9ecae1", "1" = "#6baed6", "5" = "#3182bd", "25" = "#08519c"), drop = FALSE, name = "strategy legend") +
  labs(title = "Transfer: Design", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

transfer_score_heat + transfer_effect + transfer_design + patchwork::plot_layout(widths = c(0.45, 0.8, 1), guides = "collect") & theme(legend.position = "bottom")
```

Within each `model x fit-context x line` context, rank strategies by `log_ratio_final` from most negative to most positive. Then identify the strategies with the best mean within-context rank across the transfer subset.

```{r against-transfer-summary}
print(transfer_strategy$against$summary %>% slice_head(n = 8), n = 8)
```

```{r against-transfer-heatmap, fig.height=12}
ggplot(transfer_strategy$against$plot_df, aes(model_id, strategy_code, fill = log_ratio_final)) +
  geom_tile(color = "white") +
  facet_grid(cell_line ~ context_label) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, limits = c(-transfer_strategy$against$limit, transfer_strategy$against$limit), name = "day-6 log ratio") +
  labs(title = "Transfer: Against High Ploidy", x = "model", y = "strategy code") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```

```{r for-transfer-summary}
print(transfer_strategy$for_high$summary %>% slice_head(n = 8), n = 8)
```

```{r for-transfer-heatmap, fig.height=12}
ggplot(transfer_strategy$for_high$plot_df, aes(model_id, strategy_code, fill = log_ratio_final)) +
  geom_tile(color = "white") +
  facet_grid(cell_line ~ context_label) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, limits = c(-transfer_strategy$for_high$limit, transfer_strategy$for_high$limit), name = "day-6 log ratio") +
  labs(title = "Transfer: For High Ploidy", x = "model", y = "strategy code") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```

## Empirical Competition Check: Existing Fitted Contexts

This section compares the mixed-culture simulation branch against the empirical SUM-159 co-culture experiment in `data/counts/competition/SUM-159-fuse.Rds`, but now splits the interpretation into two separate questions.

The shared alignment used throughout is:

- SUM-159 corresponds to `line_id = 5`
- `2N` is mapped to `N_low`
- `4N` is mapped to `N_high`
- each frame is treated as an `8 h` interval
- the experiment is treated as a no-refresh mixed-culture trajectory over `112 h`
- simulations are initialized from the empirical frame-0 live total and high-ploidy fraction at each glucose

So this is not a new fitting procedure. It is a validation rerun that uses already-fitted mixed-culture dynamics to generate detailed trajectories on the empirical time grid.

```{r competition-availability, results='asis'}
if (!competition_available) {
  cat("Competition dataset not found at `", params$competition_path, "`. Skipping empirical validation section.\n", sep = "")
}
```

```{r competition-initial-state, eval=competition_available}
competition_init_fmt <- competition_init_df %>%
  mutate(
    high_fraction_start = percent(high_fraction_start, accuracy = 0.1),
    total_live_start = round(total_live_start, 1)
  )

knitr::kable(competition_init_fmt)
```

### Question 1: If line 5 was included in the hierarchical fit, how well does the model explain the competition assay?

This is the `global` question. It is an in-panel explanatory check: line 5 contributed to the hierarchical fit, so success here means the fitted model family can reconcile the monoculture data and the independent competition data for SUM-159.

```{r competition-global-summary, eval=competition_available}
competition_global_fmt <- competition_global_summary %>%
  select(
    model_id, context_label, glucose_sign_match, mae_log_ratio,
    mae_high_fraction, mae_log_total_live, validation_score
  ) %>%
  mutate(
    glucose_sign_match = percent(glucose_sign_match, accuracy = 1),
    mae_high_fraction = percent(mae_high_fraction, accuracy = 0.1)
  )

knitr::kable(competition_global_fmt, digits = 3)
```

`glucose_sign_match` is the fraction of empirical glucose levels for which the simulation gets the sign of the start-to-end log-ratio shift correct. That is the most basic directional check. The error columns then ask how close the simulated trajectories stay to the empirical mean trajectories over time.

```{r competition-global-heatmap, fig.height=7, eval=competition_available}
global_heat_df <- competition_global_summary %>%
  mutate(model_id = factor(model_id, levels = rev(model_id)))

ggplot(global_heat_df, aes(x = 1, y = model_id, fill = glucose_sign_match)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", validation_score)), size = 3) +
  scale_fill_gradient(
    low = "#f7fbff",
    high = "#08519c",
    limits = c(0, 1),
    labels = percent_format(accuracy = 1),
    name = "glucose\nsign match"
  ) +
  labs(
    title = "SUM-159 Competition Validation: Global Contexts",
    subtitle = "Line 5 was included in the hierarchical fit; text shows composite validation score (lower is better)",
    x = NULL,
    y = "model"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

```{r competition-global-log-ratio-overlay, fig.width=16, fig.height=10, eval=competition_available}
global_empirical_reps <- tidyr::crossing(
  validation_label = sort(unique(competition_global_overlay_df$validation_label)),
  competition_df %>%
    transmute(glucose, time_hours, rep, log_ratio, total_live)
)

global_empirical_mean <- tidyr::crossing(
  validation_label = sort(unique(competition_global_overlay_df$validation_label)),
  competition_mean_df %>%
    transmute(glucose, time_hours, rep = NA_character_, log_ratio, total_live)
)

global_overlay_plot_df <- bind_rows(
  competition_global_overlay_df %>%
    transmute(
      glucose,
      time_hours,
      rep = NA_character_,
      validation_label,
      source = "simulation",
      log_ratio,
      total_live
    ),
  global_empirical_reps %>%
    transmute(
      glucose,
      time_hours,
      rep,
      validation_label,
      source = "empirical_replicate",
      log_ratio,
      total_live
    ),
  global_empirical_mean %>%
    transmute(
      glucose,
      time_hours,
      rep = NA_character_,
      validation_label,
      source = "empirical_mean",
      log_ratio,
      total_live
    )
)

ggplot() +
  geom_point(
    data = global_overlay_plot_df %>% filter(source == "empirical_replicate"),
    aes(time_hours / 24, log_ratio),
    color = "grey75",
    size = 0.9,
    alpha = 0.6
  ) +
  geom_point(
    data = global_overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, log_ratio),
    color = "black",
    size = 1.4
  ) +
  geom_line(
    data = global_overlay_plot_df %>% filter(source == "simulation"),
    aes(time_hours / 24, log_ratio, group = validation_label),
    color = "#2166ac",
    linewidth = 0.9
  ) +
  geom_hline(yintercept = 0, color = "grey55", linewidth = 0.4) +
  facet_grid(factor(glucose) ~ validation_label, scales = "free_y") +
  labs(
    title = "Global Fits: Competition Log-Ratio",
    x = "days",
    y = "log(N_high / N_low)"
  ) +
  theme_minimal(base_size = 11)
```

```{r competition-global-total-live-overlay, fig.width=16, fig.height=10, eval=competition_available}
ggplot() +
  geom_point(
    data = global_overlay_plot_df %>% filter(source == "empirical_replicate"),
    aes(time_hours / 24, total_live),
    color = "grey75",
    size = 0.9,
    alpha = 0.6
  ) +
  geom_point(
    data = global_overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, total_live),
    color = "black",
    size = 1.4
  ) +
  geom_line(
    data = global_overlay_plot_df %>% filter(source == "simulation"),
    aes(time_hours / 24, total_live, group = validation_label),
    color = "#2166ac",
    linewidth = 0.9
  ) +
  facet_grid(factor(glucose) ~ validation_label, scales = "free_y") +
  labs(
    title = "Global Fits: Competition Total Live Burden",
    x = "days",
    y = "total live cells"
  ) +
  theme_minimal(base_size = 11)
```

### Question 2: If the relevant ploidy state for line 5 was held out, how much competition-prediction ability remains?

This is the transfer question. Here the model is still hierarchical and still trained on the wider panel, but the relevant line-5 ploidy state is withheld via the directional transfer setup. So this is closer to a held-out-state generalization check than to the in-panel global check above.

```{r competition-transfer-summary, eval=competition_available}
competition_transfer_fmt <- competition_transfer_summary %>%
  select(
    model_id, direction, context_label, glucose_sign_match, mae_log_ratio,
    mae_high_fraction, mae_log_total_live, validation_score
  ) %>%
  mutate(
    glucose_sign_match = percent(glucose_sign_match, accuracy = 1),
    mae_high_fraction = percent(mae_high_fraction, accuracy = 0.1)
  )

knitr::kable(competition_transfer_fmt, digits = 3)
```

```{r competition-transfer-heatmap, fig.height=8, eval=competition_available}
transfer_heat_df <- competition_transfer_summary %>%
  mutate(
    direction = factor(direction, levels = c("high_to_low", "low_to_high")),
    model_id = factor(model_id, levels = unique(model_id))
  )

ggplot(transfer_heat_df, aes(model_id, direction, fill = glucose_sign_match)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", validation_score)), size = 3) +
  scale_fill_gradient(
    low = "#f7fbff",
    high = "#08519c",
    limits = c(0, 1),
    labels = percent_format(accuracy = 1),
    name = "glucose\nsign match"
  ) +
  labs(
    title = "SUM-159 Competition Validation: Transfer Contexts",
    subtitle = "Directional holdout check; text shows composite validation score (lower is better)",
    x = "model",
    y = "transfer direction"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r competition-transfer-log-ratio-overlay, fig.width=16, fig.height=12, eval=competition_available}
transfer_empirical_reps <- tidyr::crossing(
  validation_label = sort(unique(competition_transfer_overlay_df$validation_label)),
  competition_df %>%
    transmute(glucose, time_hours, rep, log_ratio, total_live)
)

transfer_empirical_mean <- tidyr::crossing(
  validation_label = sort(unique(competition_transfer_overlay_df$validation_label)),
  competition_mean_df %>%
    transmute(glucose, time_hours, rep = NA_character_, log_ratio, total_live)
)

transfer_overlay_plot_df <- bind_rows(
  competition_transfer_overlay_df %>%
    transmute(
      model_id,
      direction,
      glucose,
      time_hours,
      rep = NA_character_,
      validation_label,
      source = "simulation",
      log_ratio,
      total_live
    ),
  transfer_empirical_reps %>%
    transmute(
      model_id = sub(" \\| .*", "", validation_label),
      direction = NA_character_,
      glucose,
      time_hours,
      rep,
      validation_label,
      source = "empirical_replicate",
      log_ratio,
      total_live
    ),
  transfer_empirical_mean %>%
    transmute(
      model_id = sub(" \\| .*", "", validation_label),
      direction = NA_character_,
      glucose,
      time_hours,
      rep = NA_character_,
      validation_label,
      source = "empirical_mean",
      log_ratio,
      total_live
    )
)

ggplot() +
  geom_point(
    data = transfer_overlay_plot_df %>% filter(source == "empirical_replicate"),
    aes(time_hours / 24, log_ratio),
    color = "grey75",
    size = 0.9,
    alpha = 0.6
  ) +
  geom_point(
    data = transfer_overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, log_ratio),
    color = "black",
    size = 1.4
  ) +
  geom_line(
    data = transfer_overlay_plot_df %>% filter(source == "simulation"),
    aes(time_hours / 24, log_ratio, color = direction, group = interaction(model_id, direction)),
    linewidth = 0.9
  ) +
  geom_hline(yintercept = 0, color = "grey55", linewidth = 0.4) +
  facet_grid(factor(glucose) ~ model_id, scales = "free_y") +
  scale_color_manual(values = c("high_to_low" = "#2166ac", "low_to_high" = "#b2182b"), name = "transfer direction") +
  labs(
    title = "Transfer Fits: Competition Log-Ratio",
    x = "days",
    y = "log(N_high / N_low)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r competition-transfer-total-live-overlay, fig.width=16, fig.height=12, eval=competition_available}
ggplot() +
  geom_point(
    data = transfer_overlay_plot_df %>% filter(source == "empirical_replicate"),
    aes(time_hours / 24, total_live),
    color = "grey75",
    size = 0.9,
    alpha = 0.6
  ) +
  geom_point(
    data = transfer_overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, total_live),
    color = "black",
    size = 1.4
  ) +
  geom_line(
    data = transfer_overlay_plot_df %>% filter(source == "simulation"),
    aes(time_hours / 24, total_live, color = direction, group = interaction(model_id, direction)),
    linewidth = 0.9
  ) +
  facet_grid(factor(glucose) ~ model_id, scales = "free_y") +
  scale_color_manual(values = c("high_to_low" = "#2166ac", "low_to_high" = "#b2182b"), name = "transfer direction") +
  labs(
    title = "Transfer Fits: Competition Total Live Burden",
    x = "days",
    y = "total live cells"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Taken together, these two subsections separate in-panel explanatory adequacy (`global`) from held-out-state transfer adequacy (`transfer`). Both are useful, but they answer different questions and should not be read as interchangeable evidence.

## Supplementary: Single-Line Fit Competition Check

This supplementary section asks a narrower question: among the models fit only to the `line_id = 5` monoculture wells, which ones best describe the independent SUM-159 competition dataset?

The alignment to the empirical assay is the same as above. The only change is the model source: each model is loaded from the best completed single-line optimization run under `data/runs/gpath/v1/optim/gstarvation_v1_single_line/...`, using the line-5 subset input.

```{r competition-singleline-summary, eval=competition_available}
singleline_competition_validation_fmt <- singleline_competition_validation_summary %>%
  select(
    model_id, glucose_sign_match, mae_log_ratio,
    mae_high_fraction, mae_log_total_live, mae_dead_fraction, validation_score
  ) %>%
  mutate(
    glucose_sign_match = percent(glucose_sign_match, accuracy = 1),
    mae_high_fraction = percent(mae_high_fraction, accuracy = 0.1),
    mae_dead_fraction = percent(mae_dead_fraction, accuracy = 0.1)
  )

knitr::kable(singleline_competition_validation_fmt, digits = 3)
```

```{r competition-singleline-heatmap, fig.height=7, eval=competition_available}
singleline_heat_df <- singleline_competition_validation_summary %>%
  mutate(model_id = factor(model_id, levels = rev(model_id)))

ggplot(singleline_heat_df, aes(x = 1, y = model_id, fill = glucose_sign_match)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", validation_score)), size = 3) +
  scale_fill_gradient(
    low = "#f7fbff",
    high = "#08519c",
    limits = c(0, 1),
    labels = percent_format(accuracy = 1),
    name = "glucose\nsign match"
  ) +
  labs(
    title = "SUM-159 Competition Validation Across Single-Line Fits",
    subtitle = "Tile color shows directional agreement; overlaid text shows composite validation score (lower is better)",
    x = NULL,
    y = "model"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

```{r competition-singleline-log-ratio-overlay, fig.width=16, fig.height=12, eval=competition_available}
singleline_empirical_reps <- tidyr::crossing(
  validation_label = sort(unique(singleline_competition_overlay_df$validation_label)),
  competition_df %>%
    transmute(glucose, time_hours, rep, log_ratio, total_live)
)

singleline_empirical_mean <- tidyr::crossing(
  validation_label = sort(unique(singleline_competition_overlay_df$validation_label)),
  competition_mean_df %>%
    transmute(glucose, time_hours, rep = NA_character_, log_ratio, total_live)
)

singleline_overlay_plot_df <- bind_rows(
  singleline_competition_overlay_df %>%
    transmute(
      glucose,
      time_hours,
      rep = NA_character_,
      validation_label,
      source = "simulation",
      log_ratio,
      total_live
    ),
  singleline_empirical_reps %>%
    transmute(
      glucose,
      time_hours,
      rep,
      validation_label,
      source = "empirical_replicate",
      log_ratio,
      total_live
    ),
  singleline_empirical_mean %>%
    transmute(
      glucose,
      time_hours,
      rep = NA_character_,
      validation_label,
      source = "empirical_mean",
      log_ratio,
      total_live
    )
)

ggplot() +
  geom_point(
    data = singleline_overlay_plot_df %>% filter(source == "empirical_replicate"),
    aes(time_hours / 24, log_ratio),
    color = "grey75",
    size = 0.9,
    alpha = 0.6
  ) +
  geom_point(
    data = singleline_overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, log_ratio),
    color = "black",
    size = 1.4
  ) +
  geom_line(
    data = singleline_overlay_plot_df %>% filter(source == "simulation"),
    aes(time_hours / 24, log_ratio, group = validation_label),
    color = "#2166ac",
    linewidth = 0.9
  ) +
  geom_hline(yintercept = 0, color = "grey55", linewidth = 0.4) +
  facet_grid(factor(glucose) ~ validation_label, scales = "free_y") +
  labs(
    title = "Single-Line Fits: Competition Log-Ratio",
    x = "days",
    y = "log(N_high / N_low)"
  ) +
  theme_minimal(base_size = 11)
```

```{r competition-singleline-total-live-overlay, fig.width=16, fig.height=12, eval=competition_available}
ggplot() +
  geom_point(
    data = singleline_overlay_plot_df %>% filter(source == "empirical_replicate"),
    aes(time_hours / 24, total_live),
    color = "grey75",
    size = 0.9,
    alpha = 0.6
  ) +
  geom_point(
    data = singleline_overlay_plot_df %>% filter(source == "empirical_mean"),
    aes(time_hours / 24, total_live),
    color = "black",
    size = 1.4
  ) +
  geom_line(
    data = singleline_overlay_plot_df %>% filter(source == "simulation"),
    aes(time_hours / 24, total_live, group = validation_label),
    color = "#2166ac",
    linewidth = 0.9
  ) +
  facet_grid(factor(glucose) ~ validation_label, scales = "free_y") +
  labs(
    title = "Single-Line Fits: Competition Total Live Burden",
    x = "days",
    y = "total live cells"
  ) +
  theme_minimal(base_size = 11)
```

This supplementary validation is intended to answer a narrower model-selection question: whether any model, when fit only to the line-5 monoculture wells, can also explain the independent SUM-159 competition assay.

## Optional Detailed Follow-Up

This section reruns a very small number of representative top strategies in detailed mode. It is secondary to the heatmaps above.

```{r representative-selection, eval=params$include_detailed}
rep_model_ids <- cfg$representative_model_ids
if (is.null(rep_model_ids) || !length(rep_model_ids)) {
  rep_model_ids <- unique(head(as.character(global_strategy$against$plot_df$model_id), 2L))
}

rep_rows <- bind_rows(
  global_strategy$against$plot_df %>%
    filter(model_id %in% rep_model_ids, strategy_code == levels(global_strategy$against$plot_df$strategy_code)[[length(levels(global_strategy$against$plot_df$strategy_code))]]) %>%
    group_by(model_id, cell_line, context_label) %>%
    slice_head(n = 1L) %>%
    ungroup() %>%
    mutate(selection_objective = "against_high"),
  global_strategy$for_high$plot_df %>%
    filter(model_id %in% rep_model_ids, strategy_code == levels(global_strategy$for_high$plot_df$strategy_code)[[length(levels(global_strategy$for_high$plot_df$strategy_code))]]) %>%
    group_by(model_id, cell_line, context_label) %>%
    slice_head(n = 1L) %>%
    ungroup() %>%
    mutate(selection_objective = "for_high")
)

detail_df <- rerun_selected_strategies_detailed(
  config = cfg,
  selected_rows = rep_rows,
  time_step_hours = cfg$detailed_time_step_hours
)

if (nrow(detail_df)) {
  detail_df <- detail_df %>%
    left_join(line_map, by = "line_id") %>%
    mutate(
      cell_line = if_else(is.na(cell_line) | !nzchar(cell_line), sprintf("line_%02d", as.integer(line_id)), cell_line),
      context_label = if_else(fit_context == "global", "global", paste(direction, fit_type, sep = " | "))
    )
}
```

```{r detailed-ratio, fig.height=10, eval=params$include_detailed}
if (nrow(detail_df)) {
  detail_ratio <- detail_df %>%
    mutate(log_ratio = log((N_high + 1e-8) / (N_low + 1e-8)))

  ggplot(detail_ratio, aes(time_global / 24, log_ratio, color = strategy_code, group = interaction(model_id, cell_line, context_label, strategy_code, selection_objective))) +
    geom_line(linewidth = 0.9) +
    facet_grid(cell_line ~ model_id + context_label, scales = "free_y") +
    labs(
      title = "Representative Detailed Log-Ratio Trajectories",
      x = "days",
      y = "log(N_high / N_low)",
      color = "strategy"
    ) +
    theme_minimal(base_size = 11)
}
```

## Agent Summary

```{r agent-summary, include=FALSE}
against_text <- global_strategy$against$summary %>%
  slice_head(n = 5) %>%
  transmute(
    line = sprintf(
      "%s | mean_rank=%.2f | top5=%.2f | frac_against=%.2f | mean_log_ratio=%.3f",
      strategy_code, mean_rank, top5_fraction, favorable_fraction, mean_log_ratio
    )
  ) %>%
  pull(line)

for_text <- global_strategy$for_high$summary %>%
  slice_head(n = 5) %>%
  transmute(
    line = sprintf(
      "%s | mean_rank=%.2f | top5=%.2f | frac_for=%.2f | mean_log_ratio=%.3f",
      strategy_code, mean_rank, top5_fraction, favorable_fraction, mean_log_ratio
    )
  ) %>%
  pull(line)

consistency_extremes <- global_strategy$sign_summary %>%
  slice(unique(c(seq_len(min(5, nrow(global_strategy$sign_summary))), seq.int(max(1, nrow(global_strategy$sign_summary) - 4), nrow(global_strategy$sign_summary))))) %>%
  transmute(
    line = sprintf(
      "%s | signed_consistency=%.3f | frac_for=%.2f | frac_against=%.2f | mean_log_ratio=%.3f",
      strategy_code, signed_consistency, frac_for, frac_against, mean_log_ratio
    )
  ) %>%
  pull(line)

against_best <- global_strategy$against$summary %>% slice_head(n = 1)
for_best <- global_strategy$for_high$summary %>% slice_head(n = 1)
transfer_against_text <- transfer_strategy$against$summary %>%
  slice_head(n = 5) %>%
  transmute(
    line = sprintf(
      "%s | mean_rank=%.2f | top5=%.2f | frac_against=%.2f | mean_log_ratio=%.3f",
      strategy_code, mean_rank, top5_fraction, favorable_fraction, mean_log_ratio
    )
  ) %>%
  pull(line)
transfer_for_text <- transfer_strategy$for_high$summary %>%
  slice_head(n = 5) %>%
  transmute(
    line = sprintf(
      "%s | mean_rank=%.2f | top5=%.2f | frac_for=%.2f | mean_log_ratio=%.3f",
      strategy_code, mean_rank, top5_fraction, favorable_fraction, mean_log_ratio
    )
  ) %>%
  pull(line)
competition_text <- character(0)
competition_singleline_text <- character(0)

if (competition_available && nrow(competition_validation_summary)) {
  competition_text <- competition_validation_summary %>%
    slice_head(n = 5) %>%
    transmute(
      line = sprintf(
        "%s | %s | sign_match=%.2f | mae_log_ratio=%.3f | mae_high_fraction=%.3f | validation_score=%.3f",
        model_id, context_label, glucose_sign_match, mae_log_ratio, mae_high_fraction, validation_score
      )
    ) %>%
    pull(line)
}

if (competition_available && nrow(singleline_competition_validation_summary)) {
  competition_singleline_text <- singleline_competition_validation_summary %>%
    slice_head(n = 5) %>%
    transmute(
      line = sprintf(
        "%s | sign_match=%.2f | mae_log_ratio=%.3f | mae_high_fraction=%.3f | mae_dead_fraction=%.3f | validation_score=%.3f",
        model_id, glucose_sign_match, mae_log_ratio, mae_high_fraction, mae_dead_fraction, validation_score
      )
    ) %>%
    pull(line)
}

verdict_lines <- c(
  sprintf(
    "best_against_high\t%s | mean_rank=%.2f | frac_against=%.2f | mean_log_ratio=%.3f",
    against_best$strategy_code[[1]],
    against_best$mean_rank[[1]],
    against_best$favorable_fraction[[1]],
    against_best$mean_log_ratio[[1]]
  ),
  sprintf(
    "best_for_high\t%s | mean_rank=%.2f | frac_for=%.2f | mean_log_ratio=%.3f",
    for_best$strategy_code[[1]],
    for_best$mean_rank[[1]],
    for_best$favorable_fraction[[1]],
    for_best$mean_log_ratio[[1]]
  ),
  sprintf(
    "verdict\tRobust selection FOR high ploidy is substantially stronger than robust selection AGAINST high ploidy in this strategy space. Best-for strategies reach frac_for around %.2f-%.2f, whereas best-against strategies only reach frac_against around %.2f-%.2f.",
    min(head(global_strategy$for_high$summary$favorable_fraction, 5)),
    max(head(global_strategy$for_high$summary$favorable_fraction, 5)),
    min(head(global_strategy$against$summary$favorable_fraction, 5)),
    max(head(global_strategy$against$summary$favorable_fraction, 5))
  )
)

summary_lines <- c(
  "SELECTION_STRATEGY_ANALYSIS",
  sprintf("generated\t%s", as.character(Sys.time())),
  sprintf("config\t%s", params$config_path),
  sprintf("output_root\t%s", params$output_root),
  sprintf("n_models\t%d", dplyr::n_distinct(final_summary$model_id)),
  sprintf("n_contexts\t%d", dplyr::n_distinct(paste(final_summary$model_id, final_summary$line_id, final_summary$context_label, sep = '||'))),
  sprintf("n_strategies\t%d", dplyr::n_distinct(final_summary$strategy_code)),
  "",
  "VERDICT",
  verdict_lines,
  "",
  "TOP_AGAINST_HIGH",
  against_text,
  "",
  "TOP_FOR_HIGH",
  for_text,
  "",
  "TOP_AGAINST_HIGH_TRANSFER",
  transfer_against_text,
  "",
  "TOP_FOR_HIGH_TRANSFER",
  transfer_for_text,
  "",
  "CONSISTENCY_EXTREMES",
  consistency_extremes,
  "",
  "COMPETITION_VALIDATION",
  competition_text,
  "",
  "COMPETITION_VALIDATION_SINGLE_LINE",
  competition_singleline_text
)

writeLines(summary_lines, agent_text_path)
```
