---
title: "Selection Strategy Analysis"
author: "Codex"
date: "2026-03-21"
output: html_document
params:
  config_path: "scripts/selection_strategy_config.R"
  output_root: !r file.path("data", "selection_strategy", "gpath", "v1", "gstarvation_v1")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook reports the main selection-strategy result from the mixed-ploidy simulations: which 6-day media schedules most strongly select for or against high ploidy by day 6.

Each strategy is a three-digit code:

- digit 1: initial glucose condition at day 0,
- digit 2: action at day 2 (`0` = no refresh, `1:7` = refresh to one of the tested glucose conditions),
- digit 3: action at day 4 (`0` = no refresh, `1:7` = refresh to one of the tested glucose conditions).

The primary outcome is the day-6 log ratio:

`log_ratio_final = log(N_high / N_low)`

More negative values mean stronger selection against high ploidy. More positive values mean stronger selection for high ploidy.

## Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

source("R/selection_strategy_utils.R")

cfg <- load_selection_config(params$config_path)
cfg$output_root <- dirname(params$output_root)
cfg$dataset_label <- basename(params$output_root)

stan_data <- resolve_selection_stan(cfg)
line_map <- get_line_label_map(stan_data)
glucose_values <- resolve_glucose_values(stan_data, cfg)
strategy_grid <- build_strategy_grid(glucose_values)

outputs <- collect_selection_outputs(cfg)
final_summary <- outputs$final_summary
interval_summary <- outputs$interval_summary
task_meta <- outputs$task_meta

if (!nrow(final_summary)) {
  stop("No selection-strategy outputs were found. Run scripts/run_gpath_selection_strategy_batch.R first.")
}

final_summary <- final_summary %>%
  left_join(line_map, by = "line_id") %>%
  left_join(
    strategy_grid %>%
      select(strategy_code, g0_day0, day2_action, g0_day2, day4_action, g0_day4),
    by = "strategy_code"
  ) %>%
  mutate(
    context_label = if_else(fit_context == "global", "global", paste(direction, fit_type, sep = " | ")),
    select_against_high_score = -log_ratio_final,
    select_for_high_score = log_ratio_final
  )

task_meta <- task_meta %>%
  left_join(line_map, by = "line_id")
```

## Objects

The main objects used below are:

- `final_summary`: one row per `model x fit_context x line x strategy`, containing the day-6 outcome and ranking metrics.
- `interval_summary`: one row per interval end (`48h`, `96h`, `144h`) for the same simulations.
- `task_meta`: one row per simulated fit context, including the fitted run path and the inferred mixed-culture starting density.

The on-disk artefacts are:

- `data/selection_strategy/gpath/v1/<dataset>/<model>/<context>/line_##/final_summary.Rds`
- `data/selection_strategy/gpath/v1/<dataset>/<model>/<context>/line_##/interval_summary.Rds`
- `data/selection_strategy/gpath/v1/<dataset>/<model>/<context>/line_##/detail.Rds` when detailed mode was requested
- `data/selection_strategy/gpath/v1/<dataset>/<model>/<context>/line_##/task_meta.Rds`

```{r task-summary}
print(
  task_meta %>%
    select(model_id, cell_line, fit_context, direction, fit_type, run_label, seed_total_n) %>%
    arrange(model_id, cell_line, fit_context, direction, fit_type),
  n = 20
)
```

## Main Result: Select Against High Ploidy

For each `model x fit-context x cell-line` combination, take the single best strategy for selecting against high ploidy, using the day-6 log ratio directly.

```{r against-heatmap-prep}
against_top <- final_summary %>%
  group_by(model_id, cell_line, context_label) %>%
  slice_min(order_by = log_ratio_final, n = 1, with_ties = FALSE) %>%
  ungroup()
```

```{r against-heatmap, fig.height=12}
ggplot(against_top, aes(model_id, cell_line, fill = log_ratio_final)) +
  geom_tile(color = "white") +
  geom_text(aes(label = strategy_code), size = 3) +
  facet_wrap(~ context_label, ncol = 1) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    name = "Day-6 log(N_high / N_low)"
  ) +
  labs(
    title = "Best Day-6 Strategies For Selecting Against High Ploidy",
    subtitle = "Tile color is the achieved day-6 log ratio; tile label is the winning strategy code",
    x = "model",
    y = "cell line"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Main Result: Select For High Ploidy

Now do the same, but take the best strategy for selecting for high ploidy.

```{r for-heatmap-prep}
for_top <- final_summary %>%
  group_by(model_id, cell_line, context_label) %>%
  slice_max(order_by = log_ratio_final, n = 1, with_ties = FALSE) %>%
  ungroup()
```

```{r for-heatmap, fig.height=12}
ggplot(for_top, aes(model_id, cell_line, fill = log_ratio_final)) +
  geom_tile(color = "white") +
  geom_text(aes(label = strategy_code), size = 3) +
  facet_wrap(~ context_label, ncol = 1) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    name = "Day-6 log(N_high / N_low)"
  ) +
  labs(
    title = "Best Day-6 Strategies For Selecting For High Ploidy",
    subtitle = "Tile color is the achieved day-6 log ratio; tile label is the winning strategy code",
    x = "model",
    y = "cell line"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Consistency Of Winning Strategy Codes

This view ignores effect magnitude and just asks whether the same strategy codes recur across models and fit contexts.

```{r strategy-frequency, fig.height=8}
strategy_freq <- bind_rows(
  against_top %>% mutate(objective = "against_high"),
  for_top %>% mutate(objective = "for_high")
) %>%
  dplyr::count(objective, strategy_code, sort = TRUE)

ggplot(strategy_freq, aes(reorder(strategy_code, n), n, fill = objective)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "How Often Winning Strategy Codes Recur",
    x = "strategy code",
    y = "winning contexts"
  ) +
  theme_minimal(base_size = 11)
```

## Optional Detailed Follow-Up

This section reruns a very small number of representative top strategies in detailed mode. It is secondary to the two heatmaps above.

```{r representative-selection}
rep_model_ids <- cfg$representative_model_ids
if (is.null(rep_model_ids) || !length(rep_model_ids)) {
  rep_model_ids <- unique(head(against_top$model_id, 2L))
}

rep_rows <- bind_rows(
  against_top %>%
    filter(model_id %in% rep_model_ids) %>%
    group_by(model_id, context_label) %>%
    slice_head(n = 1L) %>%
    ungroup() %>%
    mutate(selection_objective = "against_high"),
  for_top %>%
    filter(model_id %in% rep_model_ids) %>%
    group_by(model_id, context_label) %>%
    slice_head(n = 1L) %>%
    ungroup() %>%
    mutate(selection_objective = "for_high")
)

detail_df <- rerun_selected_strategies_detailed(
  config = cfg,
  selected_rows = rep_rows,
  time_step_hours = cfg$detailed_time_step_hours
)

detail_df <- detail_df %>%
  left_join(line_map, by = "line_id")
```

```{r detailed-ratio, fig.height=10}
if (nrow(detail_df)) {
  detail_ratio <- detail_df %>%
    mutate(log_ratio = log((N_high + 1e-8) / (N_low + 1e-8)))

  ggplot(detail_ratio, aes(time_global / 24, log_ratio, color = selection_objective, group = interaction(model_id, cell_line, strategy_code, selection_objective))) +
    geom_line(linewidth = 0.9) +
    facet_grid(cell_line ~ model_id, scales = "free_y") +
    labs(
      title = "Representative Detailed Log-Ratio Trajectories",
      x = "days",
      y = "log(N_high / N_low)",
      color = "objective"
    ) +
    theme_minimal(base_size = 11)
}
```

```{r detailed-glucose, fig.height=10}
if (nrow(detail_df) && "G1" %in% names(detail_df)) {
  ggplot(detail_df, aes(time_global / 24, G1, color = selection_objective, group = interaction(model_id, cell_line, strategy_code, selection_objective))) +
    geom_line(linewidth = 0.9) +
    facet_grid(cell_line ~ model_id, scales = "free_y") +
    labs(
      title = "Representative Detailed Glucose Trajectories",
      x = "days",
      y = "glucose",
      color = "objective"
    ) +
    theme_minimal(base_size = 11)
}
```
