---
title: "Selection Strategy Analysis"
author: "Codex"
date: "2026-03-21"
output: html_document
params:
  config_path: "scripts/selection_strategy_config.R"
  output_root: !r file.path("data", "selection_strategy", "gpath", "v1", "gstarvation_v1")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook summarizes the mixed-ploidy selection-strategy simulations. Each strategy is a three-digit code:

- digit 1: initial glucose condition at day 0,
- digit 2: action at day 2 (`0` = no refresh, `1:7` = refresh to one of the tested glucose conditions),
- digit 3: action at day 4 (`0` = no refresh, `1:7` = refresh to one of the tested glucose conditions).

The simulation resets media variables at refresh and rescales the mixed culture back to the fitted average starting density for the low- and high-ploidy states of that line, while preserving the current high/low composition.

## Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(purrr)
library(scales)

source("R/selection_strategy_utils.R")

cfg <- load_selection_config(params$config_path)
cfg$output_root <- dirname(params$output_root)
cfg$dataset_label <- basename(params$output_root)

stan_data <- resolve_selection_stan(cfg)
line_map <- get_line_label_map(stan_data)
glucose_values <- resolve_glucose_values(stan_data, cfg)
strategy_grid <- build_strategy_grid(glucose_values)

outputs <- collect_selection_outputs(cfg)
final_summary <- outputs$final_summary
interval_summary <- outputs$interval_summary
task_meta <- outputs$task_meta

if (!nrow(final_summary)) {
  stop("No selection-strategy outputs were found. Run scripts/run_gpath_selection_strategy_batch.R first.")
}

final_summary <- final_summary %>%
  left_join(line_map, by = "line_id")

interval_summary <- interval_summary %>%
  left_join(line_map, by = "line_id")

top_by_group <- final_summary %>%
  group_by(model_id, line_id, cell_line, fit_context, direction, fit_type) %>%
  group_modify(~ select_top_strategies(.x, n_top = 3L)) %>%
  ungroup() %>%
  left_join(
    strategy_grid %>%
      select(strategy_code, g0_day0, day2_action, g0_day2, day4_action, g0_day4),
    by = "strategy_code"
  )
```

## Coverage

```{r coverage}
coverage_tbl <- final_summary %>%
  distinct(model_id, line_id, cell_line, fit_context, direction, fit_type) %>%
  dplyr::count(model_id, fit_context, direction, fit_type, name = "n_line_contexts") %>%
  arrange(model_id, fit_context, direction, fit_type)

print(coverage_tbl, n = 20)
```

## Basic Criteria

The ranking logic is intentionally simple:

- `select_against_high_score`: maximize depletion of high ploidy relative to low ploidy at day 6.
- `against_high_viable`: same target, but only keep strategies where low ploidy at least maintained its own starting abundance.
- `select_for_high_score` and `for_high_viable`: the analogous enrichment criteria for high ploidy.

```{r criteria-preview}
criteria_tbl <- top_by_group %>%
  select(
    model_id, cell_line, fit_context, direction, fit_type,
    selection_objective, rank, strategy_code, g0_day0, day2_action, g0_day2, day4_action, g0_day4,
    select_against_high_score, against_high_viable, select_for_high_score, for_high_viable,
    total_live, N_low, N_high, high_fraction_final
  ) %>%
  arrange(model_id, cell_line, fit_context, direction, fit_type, selection_objective, rank)

print(head(criteria_tbl, 24), n = 24)
```

## Top Strategies

```{r top-heatmap, fig.height=10}
against_top <- top_by_group %>%
  filter(selection_objective == "select_against_high_viable", rank == 1) %>%
  mutate(context_label = if_else(fit_context == "global", "global", paste(direction, fit_type, sep = " | ")))

ggplot(against_top, aes(model_id, cell_line, fill = select_against_high_score)) +
  geom_tile(color = "white") +
  geom_text(aes(label = strategy_code), size = 3) +
  facet_wrap(~ context_label, ncol = 1) +
  scale_fill_viridis_c(option = "C", name = "Against-high score") +
  labs(
    title = "Top Viable Strategies That Select Against High Ploidy",
    x = "model",
    y = "cell line"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r top-table}
print(
  against_top %>%
    select(model_id, cell_line, fit_context, direction, fit_type, strategy_code, g0_day0, day2_action, g0_day2, day4_action, g0_day4, select_against_high_score, total_live, high_fraction_final) %>%
    arrange(model_id, cell_line, fit_context, direction, fit_type),
  n = 20
)
```

## Stability Against Global Ranking

```{r stability-summary}
global_top <- top_by_group %>%
  filter(selection_objective == "select_against_high_viable", rank == 1, fit_context == "global") %>%
  select(model_id, line_id, global_strategy_code = strategy_code)

transfer_rank_tbl <- final_summary %>%
  filter(fit_context == "transfer") %>%
  group_by(model_id, line_id, direction, fit_type) %>%
  group_modify(~ {
    ranked <- rank_strategy_table(.x, score_col = "against_high_viable", decreasing = TRUE)
    ranked %>% select(strategy_code, rank)
  }) %>%
  ungroup() %>%
  left_join(global_top, by = c("model_id", "line_id")) %>%
  filter(strategy_code == global_strategy_code)

stability_tbl <- transfer_rank_tbl %>%
  left_join(line_map, by = "line_id") %>%
  group_by(model_id, cell_line, fit_type) %>%
  summarise(
    mean_rank_of_global_top = mean(rank, na.rm = TRUE),
    max_rank_of_global_top = max(rank, na.rm = TRUE),
    n_transfer_contexts = n(),
    .groups = "drop"
  ) %>%
  arrange(mean_rank_of_global_top, max_rank_of_global_top)

print(stability_tbl, n = 20)
```

## Representative Detailed Trajectories

```{r representative-selection}
rep_model_ids <- cfg$representative_model_ids
if (is.null(rep_model_ids) || !length(rep_model_ids)) {
  rep_model_ids <- against_top %>%
    distinct(model_id) %>%
    head(2L) %>%
    pull(model_id)
}

rep_rows <- top_by_group %>%
  filter(
    selection_objective == "select_against_high_viable",
    rank == 1,
    fit_context %in% cfg$representative_fit_contexts,
    model_id %in% rep_model_ids
  ) %>%
  group_by(model_id, fit_context, direction, fit_type) %>%
  slice_head(n = 3L) %>%
  ungroup()

detail_df <- rerun_selected_strategies_detailed(
  config = cfg,
  selected_rows = rep_rows,
  time_step_hours = cfg$detailed_time_step_hours
)
detail_df <- detail_df %>%
  left_join(line_map, by = "line_id")
```

```{r detailed-live, fig.height=10}
if (nrow(detail_df)) {
  ggplot(detail_df, aes(time_global / 24, N_low, color = "low ploidy")) +
    geom_line(linewidth = 0.8) +
    geom_line(aes(y = N_high, color = "high ploidy"), linewidth = 0.8) +
    facet_grid(cell_line ~ model_id, scales = "free_y") +
    scale_color_manual(values = c("low ploidy" = "#1b9e77", "high ploidy" = "#d95f02"), name = NULL) +
    labs(
      title = "Detailed Live-Cell Trajectories For Top Against-High Strategies",
      subtitle = "One top viable strategy per representative model and selected line contexts",
      x = "days",
      y = "live cells"
    ) +
    theme_minimal(base_size = 11)
}
```

```{r detailed-fraction, fig.height=10}
if (nrow(detail_df)) {
  frac_df <- detail_df %>%
    mutate(high_fraction = N_high / pmax(N_low + N_high, 1e-8))

  ggplot(frac_df, aes(time_global / 24, high_fraction, color = strategy_code, group = interaction(strategy_code, cell_line, model_id))) +
    geom_line(linewidth = 0.8) +
    facet_grid(cell_line ~ model_id) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = "High-Ploidy Fraction Under Selected Strategies",
      x = "days",
      y = "high-ploidy fraction",
      color = "strategy"
    ) +
    theme_minimal(base_size = 11)
}
```

```{r detailed-resource, fig.height=10}
if (nrow(detail_df) && "G1" %in% names(detail_df)) {
  ggplot(detail_df, aes(time_global / 24, G1, color = strategy_code, group = interaction(strategy_code, cell_line, model_id))) +
    geom_line(linewidth = 0.8) +
    facet_grid(cell_line ~ model_id, scales = "free_y") +
    labs(
      title = "Glucose Trajectories For Selected Strategies",
      x = "days",
      y = "glucose",
      color = "strategy"
    ) +
    theme_minimal(base_size = 11)
}
```

## Key Tables

```{r interval-end-table}
interval_end_tbl <- interval_summary %>%
  filter(segment_end_hours %in% c(48, 96, 144)) %>%
  select(model_id, cell_line, fit_context, direction, fit_type, strategy_code, segment_end_hours, N_low, N_high, total_live, high_fraction) %>%
  arrange(model_id, cell_line, fit_context, direction, fit_type, strategy_code, segment_end_hours)

print(head(interval_end_tbl, 24), n = 24)
```
