---
title: "Selection Strategy Analysis"
author: "Codex"
date: "2026-03-21"
output: html_document
params:
  config_path: "scripts/selection_strategy_config.R"
  output_root: !r file.path("data", "selection_strategy", "gpath", "v1", "gstarvation_v1")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook reports the main selection-strategy result from the mixed-ploidy simulations: which 6-day media schedules most strongly select for or against high ploidy by day 6.

Each strategy is a three-digit code:

- digit 1: initial glucose condition at day 0,
- digit 2: action at day 2 (`0` = no refresh, `1:7` = refresh to one of the tested glucose conditions),
- digit 3: action at day 4 (`0` = no refresh, `1:7` = refresh to one of the tested glucose conditions).

The primary outcome is the day-6 log ratio:

`log_ratio_final = log(N_high / N_low)`

More negative values mean stronger selection against high ploidy. More positive values mean stronger selection for high ploidy.

## Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(patchwork)

source("R/selection_strategy_utils.R")

cfg <- load_selection_config(params$config_path)
cfg$output_root <- dirname(params$output_root)
cfg$dataset_label <- basename(params$output_root)

stan_data <- resolve_selection_stan(cfg)
line_map <- get_line_label_map(stan_data)
glucose_values <- resolve_glucose_values(stan_data, cfg)
strategy_grid <- build_strategy_grid(glucose_values)

outputs <- collect_selection_outputs(cfg)
final_summary <- outputs$final_summary
detail_summary <- outputs$detail

if (!nrow(final_summary)) {
  stop("No selection-strategy outputs were found. Run scripts/run_gpath_selection_strategy_batch.R first.")
}

final_summary <- final_summary %>%
  left_join(line_map, by = "line_id") %>%
  left_join(
    strategy_grid %>%
      select(strategy_code, g0_day0, day2_action, g0_day2, day4_action, g0_day4),
    by = "strategy_code"
  ) %>%
  mutate(
    cell_line = if_else(is.na(cell_line) | !nzchar(cell_line), sprintf("line_%02d", as.integer(line_id)), cell_line),
    context_label = if_else(fit_context == "global", "global", paste(direction, fit_type, sep = " | "))
  )
```

## Objects

The relevant artefacts are:

- `final_summary`: one row per `model x fit-context x line x strategy`, with the day-6 outcome in `log_ratio_final`.
- `detail.Rds`: optional time-resolved trajectories for rerun strategies, including `N_low`, `N_high`, glucose, and any extra resource / waste states present in that model.

On disk these live under:

- `data/selection_strategy/gpath/v1/<dataset>/<model>/<context>/line_##/final_summary.Rds`
- `data/selection_strategy/gpath/v1/<dataset>/<model>/<context>/line_##/detail.Rds`

## Global Strategy Ordering

This view ignores model identity and asks a simpler question: across all simulated contexts, how consistently does each strategy push selection in one direction?

For each strategy:

- `frac_for` = fraction of contexts with `log_ratio_final > 0`
- `frac_against` = fraction of contexts with `log_ratio_final < 0`
- `signed_consistency = frac_for - frac_against`

So:

- values near `+1` mean the strategy almost always selects for high ploidy
- values near `-1` mean the strategy almost always selects against high ploidy
- values near `0` mean the strategy has no clear directional tendency

```{r strategy-ordering-prep}
strategy_sign_summary <- final_summary %>%
  group_by(strategy_code, g0_day0, day2_action, g0_day2, day4_action, g0_day4) %>%
  summarise(
    n_contexts = n(),
    frac_for = mean(log_ratio_final > 0, na.rm = TRUE),
    frac_against = mean(log_ratio_final < 0, na.rm = TRUE),
    frac_neutral = mean(log_ratio_final == 0, na.rm = TRUE),
    signed_consistency = frac_for - frac_against,
    abs_consistency = pmax(frac_for, frac_against),
    mean_log_ratio = mean(log_ratio_final, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(signed_consistency, desc(abs_consistency), mean_log_ratio)

strategy_levels <- strategy_sign_summary$strategy_code

strategy_score_plot_df <- strategy_sign_summary %>%
  mutate(strategy_code = factor(strategy_code, levels = strategy_levels))

strategy_design_long <- bind_rows(
  strategy_sign_summary %>%
    transmute(strategy_code, stage = "day0", value = g0_day0, action_label = "seed"),
  strategy_sign_summary %>%
    transmute(strategy_code, stage = "day2", value = g0_day2, action_label = if_else(day2_action == "carry", "carry", "refresh")),
  strategy_sign_summary %>%
    transmute(strategy_code, stage = "day4", value = g0_day4, action_label = if_else(day4_action == "carry", "carry", "refresh"))
) %>%
  mutate(
    strategy_code = factor(strategy_code, levels = strategy_levels),
    stage = factor(stage, levels = c("day0", "day2", "day4")),
    glucose_plot = if_else(is.na(value), -1, value)
  )
```

```{r strategy-ordering, fig.height=18}
p_score <- ggplot(strategy_score_plot_df, aes(x = signed_consistency, y = strategy_code, fill = mean_log_ratio)) +
  geom_col(width = 0.8) +
  geom_vline(xintercept = 0, color = "grey40", linewidth = 0.4) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    name = "Mean log ratio"
  ) +
  labs(
    title = "Per-Strategy Directional Consistency",
    subtitle = "Ordered from consistently against high ploidy to consistently for high ploidy",
    x = "signed consistency = frac_for - frac_against",
    y = "strategy code"
  ) +
  theme_minimal(base_size = 11)

p_design <- ggplot(strategy_design_long, aes(stage, strategy_code, fill = glucose_plot)) +
  geom_tile(color = "white") +
  geom_text(aes(label = if_else(action_label == "carry", "carry", scales::number(value, accuracy = 0.1))), size = 2.6) +
  scale_fill_gradientn(
    colours = c("#bdbdbd", "#f7fbff", "#6baed6", "#08519c"),
    values = scales::rescale(c(-1, min(glucose_values), median(glucose_values), max(glucose_values))),
    name = "glucose / carry",
    breaks = c(-1, glucose_values),
    labels = c("carry", as.character(glucose_values))
  ) +
  labs(
    title = "Strategy Design",
    subtitle = "Seed glucose at day 0, then day-2 and day-4 actions",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11)

p_score + p_design + patchwork::plot_layout(widths = c(1.6, 1))
```

## Main Result: Robust Strategies That Select Against High Ploidy

Within each `model x fit-context x line` context, rank strategies by `log_ratio_final` from most negative to most positive. Then identify the strategies with the best mean within-context rank across the entire result set.

```{r against-prep}
against_ranked <- final_summary %>%
  group_by(model_id, line_id, cell_line, context_label) %>%
  mutate(against_rank = min_rank(log_ratio_final)) %>%
  ungroup()

against_strategy_summary <- against_ranked %>%
  group_by(strategy_code) %>%
  summarise(
    mean_rank = mean(against_rank, na.rm = TRUE),
    top5_fraction = mean(against_rank <= 5, na.rm = TRUE),
    favorable_fraction = mean(log_ratio_final < 0, na.rm = TRUE),
    mean_log_ratio = mean(log_ratio_final, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(mean_rank, desc(top5_fraction), mean_log_ratio)

against_top_codes <- against_strategy_summary %>%
  slice_head(n = 8) %>%
  pull(strategy_code)

against_plot_df <- against_ranked %>%
  filter(strategy_code %in% against_top_codes) %>%
  mutate(strategy_code = factor(strategy_code, levels = rev(against_top_codes)))

against_limit <- max(abs(against_plot_df$log_ratio_final), na.rm = TRUE)
if (!is.finite(against_limit) || against_limit <= 0) {
  against_limit <- 1
}
```

```{r against-summary}
print(against_strategy_summary %>% slice_head(n = 8), n = 8)
```

```{r against-heatmap, fig.height=14}
ggplot(against_plot_df, aes(model_id, strategy_code, fill = log_ratio_final)) +
  geom_tile(color = "white") +
  facet_grid(cell_line ~ context_label) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-against_limit, against_limit),
    name = "Day-6 log(N_high / N_low)"
  ) +
  labs(
    title = "Robust Strategies For Selecting Against High Ploidy",
    subtitle = "Same strategy set shown across all models, cell lines, and fit contexts",
    x = "model",
    y = "strategy code"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Main Result: Robust Strategies That Select For High Ploidy

Repeat the same ranking logic, but now rank strategies from most positive to most negative `log_ratio_final`.

```{r for-prep}
for_ranked <- final_summary %>%
  group_by(model_id, line_id, cell_line, context_label) %>%
  mutate(for_rank = min_rank(desc(log_ratio_final))) %>%
  ungroup()

for_strategy_summary <- for_ranked %>%
  group_by(strategy_code) %>%
  summarise(
    mean_rank = mean(for_rank, na.rm = TRUE),
    top5_fraction = mean(for_rank <= 5, na.rm = TRUE),
    favorable_fraction = mean(log_ratio_final > 0, na.rm = TRUE),
    mean_log_ratio = mean(log_ratio_final, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(mean_rank, desc(top5_fraction), desc(mean_log_ratio))

for_top_codes <- for_strategy_summary %>%
  slice_head(n = 8) %>%
  pull(strategy_code)

for_plot_df <- for_ranked %>%
  filter(strategy_code %in% for_top_codes) %>%
  mutate(strategy_code = factor(strategy_code, levels = rev(for_top_codes)))

for_limit <- max(abs(for_plot_df$log_ratio_final), na.rm = TRUE)
if (!is.finite(for_limit) || for_limit <= 0) {
  for_limit <- 1
}
```

```{r for-summary}
print(for_strategy_summary %>% slice_head(n = 8), n = 8)
```

```{r for-heatmap, fig.height=14}
ggplot(for_plot_df, aes(model_id, strategy_code, fill = log_ratio_final)) +
  geom_tile(color = "white") +
  facet_grid(cell_line ~ context_label) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0,
    limits = c(-for_limit, for_limit),
    name = "Day-6 log(N_high / N_low)"
  ) +
  labs(
    title = "Robust Strategies For Selecting For High Ploidy",
    subtitle = "Same strategy set shown across all models, cell lines, and fit contexts",
    x = "model",
    y = "strategy code"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Optional Detailed Follow-Up

This section reruns a very small number of representative top strategies in detailed mode. It is secondary to the heatmaps above.

```{r representative-selection}
rep_model_ids <- cfg$representative_model_ids
if (is.null(rep_model_ids) || !length(rep_model_ids)) {
  rep_model_ids <- unique(head(as.character(against_plot_df$model_id), 2L))
}

rep_rows <- bind_rows(
  against_plot_df %>%
    filter(model_id %in% rep_model_ids, strategy_code == against_top_codes[[1]]) %>%
    group_by(model_id, cell_line, context_label) %>%
    slice_head(n = 1L) %>%
    ungroup() %>%
    mutate(selection_objective = "against_high"),
  for_plot_df %>%
    filter(model_id %in% rep_model_ids, strategy_code == for_top_codes[[1]]) %>%
    group_by(model_id, cell_line, context_label) %>%
    slice_head(n = 1L) %>%
    ungroup() %>%
    mutate(selection_objective = "for_high")
)

detail_df <- rerun_selected_strategies_detailed(
  config = cfg,
  selected_rows = rep_rows,
  time_step_hours = cfg$detailed_time_step_hours
)

if (nrow(detail_df)) {
  detail_df <- detail_df %>%
    left_join(line_map, by = "line_id") %>%
    mutate(
      cell_line = if_else(is.na(cell_line) | !nzchar(cell_line), sprintf("line_%02d", as.integer(line_id)), cell_line),
      context_label = if_else(fit_context == "global", "global", paste(direction, fit_type, sep = " | "))
    )
}
```

```{r detailed-ratio, fig.height=10}
if (nrow(detail_df)) {
  detail_ratio <- detail_df %>%
    mutate(log_ratio = log((N_high + 1e-8) / (N_low + 1e-8)))

  ggplot(detail_ratio, aes(time_global / 24, log_ratio, color = strategy_code, group = interaction(model_id, cell_line, context_label, strategy_code, selection_objective))) +
    geom_line(linewidth = 0.9) +
    facet_grid(cell_line ~ model_id + context_label, scales = "free_y") +
    labs(
      title = "Representative Detailed Log-Ratio Trajectories",
      x = "days",
      y = "log(N_high / N_low)",
      color = "strategy"
    ) +
    theme_minimal(base_size = 11)
}
```
