---
title: "Transfer Trajectory Spline First Attempt"
author: "Codex"
date: "2026-03-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Purpose

This notebook is a first pass at fitting smooth empirical trajectories for each cell line in a layout similar to the transfer CV overlays in `R/transfer_cv_plot_utils.R::plot_transfer_line_trajectories`.

The goals for this first attempt are:

1. keep the fitted curves fairly smooth rather than highly adaptive,
2. preserve positivity by fitting on a `log1p` scale and back-transforming,
3. show where the approach is likely useful and where it may need stronger shape constraints.

## Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(scales)
library(patchwork)
library(mgcv)

source("R/project_paths.R")
source("R/gpath_run_utils.R")
source("R/model_free_ploidy_utils.R")
```

```{r data}
tables <- build_model_free_tables()

line_levels <- tables$well_meta %>%
  distinct(cellLine) %>%
  arrange(cellLine) %>%
  pull(cellLine)

excluded_censored_wells <- tables$glucose_obs %>%
  group_by(well_idx) %>%
  summarise(any_censored = any(is_censored == 1L), .groups = "drop") %>%
  filter(any_censored) %>%
  pull(well_idx)

included_wells <- tables$well_meta %>%
  filter(!well_idx %in% excluded_censored_wells) %>%
  pull(well_idx)

count_df <- tables$count_obs %>%
  filter(well_idx %in% included_wells) %>%
  left_join(
    tables$well_meta %>%
      select(well_idx, cellLine, line_id, ploidy_abs, ploidy_metric, G0),
    by = "well_idx"
  ) %>%
  transmute(
    cellLine,
    line_id,
    ploidy_abs,
    ploidy_metric,
    G0,
    well_idx,
    hours,
    variable = "N",
    value = total_cells
  )

glucose_df <- tables$glucose_summary %>%
  filter(well_idx %in% included_wells, any_glucose_censored == 0L) %>%
  transmute(
    cellLine,
    line_id,
    ploidy_abs,
    ploidy_metric,
    G0,
    well_idx,
    hours,
    variable = "R1",
    value = glucose_hat
  )

obs_df <- bind_rows(count_df, glucose_df) %>%
  mutate(
    value = pmax(value, 0),
    cellLine = factor(cellLine, levels = line_levels),
    variable = factor(variable, levels = c("N", "R1"))
  )

line_counts <- obs_df %>%
  distinct(cellLine, ploidy_abs, G0, variable, well_idx) %>%
  count(cellLine, variable, name = "n_wells")

count_point_counts <- count_df %>%
  count(cellLine, ploidy_abs, G0, well_idx, hours, name = "n_points_at_time")

line_counts
```

```{r censor-diagnostics}
tibble(
  n_total_wells = nrow(tables$well_meta),
  n_excluded_censored_wells = length(excluded_censored_wells),
  n_included_wells = length(included_wells)
)
```

## Spline Strategy

The first pass pooled all wells inside a panel, which makes the cell-count fits too blunt when replicate wells diverge. This revision instead fits each well separately, which is closer to the original transfer trajectory overlays.

The plotting now shows all raw `N_obs` count observations as individual points. The spline fitting still uses the same well-level table being plotted, so `N` fits now see the raw count replicates rather than the earlier per-time medians.

The fitting rules are:

1. fit on the `log1p(value)` scale and back-transform with `expm1` to preserve positivity,
2. use a lighter smoothing level for `N` than for `R1`,
3. keep `cv = FALSE` so the fits remain fairly regular rather than chasing every point.

```{r spline-helpers}
default_spar_by_variable <- c(
  N = 0.45,
  R1 = 0.70
)

fit_positive_spline <- function(df, spar = 0.72, min_unique_x = 4L, n_eval = 200L) {
  df <- df %>%
    filter(is.finite(hours), is.finite(value)) %>%
    arrange(hours)

  unique_x <- sort(unique(df$hours))
  if (nrow(df) < min_unique_x || length(unique_x) < min_unique_x) {
    return(NULL)
  }

  fit <- try(
    stats::smooth.spline(
      x = df$hours,
      y = log1p(pmax(df$value, 0)),
      spar = spar,
      cv = FALSE
    ),
    silent = TRUE
  )

  if (inherits(fit, "try-error")) {
    return(NULL)
  }

  eval_hours <- seq(min(df$hours), max(df$hours), length.out = n_eval)
  pred <- predict(fit, x = eval_hours)
  deriv <- predict(fit, x = eval_hours, deriv = 1)

  fit_value <- pmax(expm1(pred$y), 0)
  fit_deriv <- pmax(exp(pred$y), 1) * deriv$y

  tibble(
    hours = eval_hours,
    fit_value = fit_value,
    fit_deriv = fit_deriv,
    spar = spar,
    n_obs = nrow(df),
    n_unique_hours = length(unique_x)
  )
}

fit_panel_splines <- function(df, spar_map = default_spar_by_variable) {
  nested <- df %>%
    group_by(cellLine, line_id, ploidy_abs, ploidy_metric, G0, variable, well_idx) %>%
    nest()

  fits <- nested %>%
    mutate(
      fit_spar = unname(spar_map[as.character(variable)]),
      fit_spar = ifelse(is.na(fit_spar), 0.65, fit_spar),
      fit = map2(data, fit_spar, fit_positive_spline),
      fit_ok = map_lgl(fit, ~ !is.null(.x))
    )

  fit_df <- fits %>%
    filter(fit_ok) %>%
    select(-data, -fit_ok) %>%
    unnest(fit)

  fit_qc <- fits %>%
    transmute(
      cellLine,
      ploidy_abs,
      G0,
      variable,
      well_idx,
      fit_spar,
      n_obs = map_int(data, nrow),
      n_unique_hours = map_int(data, ~ length(unique(.x$hours))),
      fit_ok
    ) %>%
    arrange(cellLine, variable, ploidy_abs, G0)

  list(
    fit_df = fit_df,
    fit_qc = fit_qc
  )
}

spline_results <- fit_panel_splines(obs_df)
fit_df <- spline_results$fit_df
fit_qc <- spline_results$fit_qc

fit_qc %>%
  count(variable, fit_ok)
```

```{r rate-helpers}
build_rate_relationships <- function(fit_df, min_n = 1) {
  n_fit <- fit_df %>%
    filter(variable == "N") %>%
    transmute(
      cellLine,
      line_id,
      ploidy_abs,
      ploidy_metric,
      G0,
      well_idx,
      hours,
      N_fit = fit_value,
      dN_dt = fit_deriv
    )

  r1_fit <- fit_df %>%
    filter(variable == "R1") %>%
    transmute(
      cellLine,
      line_id,
      ploidy_abs,
      ploidy_metric,
      G0,
      well_idx,
      hours,
      R1_fit = fit_value,
      dR1_dt = fit_deriv
    )

  inner_join(
    n_fit,
    r1_fit,
    by = c("cellLine", "line_id", "ploidy_abs", "ploidy_metric", "G0", "well_idx", "hours")
  ) %>%
    filter(is.finite(N_fit), is.finite(R1_fit), N_fit >= min_n) %>%
    mutate(
      uptake_per_cell = dR1_dt / pmax(N_fit, min_n),
      growth_per_cell = dN_dt / pmax(N_fit, min_n),
      ploidy_label = factor(as.character(ploidy_abs), levels = sort(unique(as.character(ploidy_abs))))
    )
}

rate_df <- build_rate_relationships(fit_df, min_n = 1)

cumtrapz_vec <- function(x, y) {
  n <- length(x)
  out <- rep(0, n)
  if (n < 2L) {
    return(out)
  }
  for (i in 2:n) {
    dx <- x[i] - x[i - 1L]
    out[i] <- out[i - 1L] + dx * (y[i] + y[i - 1L]) / 2
  }
  out
}

rate_df <- rate_df %>%
  mutate(
    log_R1 = log10(pmax(R1_fit, 1e-6)),
    log_N = log10(pmax(N_fit, 1)),
    log_G0 = log10(pmax(G0, 1e-4)),
    gluc_drawdown = pmax(G0 - R1_fit, 0),
    frac_gluc_remaining = if_else(G0 > 0, pmin(pmax(R1_fit / G0, 0), 5), NA_real_),
    well_factor = factor(well_idx)
  ) %>%
  group_by(cellLine, ploidy_abs, G0, well_idx) %>%
  arrange(hours, .by_group = TRUE) %>%
  mutate(
    lag_log_R1 = lag(log_R1),
    lag_log_N = lag(log_N),
    lag_uptake_per_cell = lag(uptake_per_cell),
    lag_growth_per_cell = lag(growth_per_cell),
    cum_R1_auc = cumtrapz_vec(hours, R1_fit),
    cum_drawdown_auc = cumtrapz_vec(hours, gluc_drawdown),
    cum_N_auc = cumtrapz_vec(hours, N_fit)
  ) %>%
  ungroup()

```

```{r fit-qc}
fit_qc %>%
  filter(!fit_ok) %>%
  arrange(cellLine, variable, ploidy_abs, G0)
```

```{r residual-model-screen}
`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0L) y else x
}

lag1_resid_cor <- function(df, resid_col = "resid") {
  vals <- df[[resid_col]]
  if (length(vals) < 3L || sd(vals, na.rm = TRUE) == 0 || any(!is.finite(vals))) {
    return(NA_real_)
  }
  stats::cor(vals[-length(vals)], vals[-1L])
}

rate_model_formulas <- function() {
  list(
    baseline = y ~ s(log_R1, k = 6),
    plus_logN = y ~ s(log_R1, k = 6) + s(log_N, k = 6),
    plus_time = y ~ s(log_R1, k = 6) + s(hours, k = 6),
    plus_logN_time = y ~ s(log_R1, k = 6) + s(log_N, k = 6) + s(hours, k = 6),
    plus_logN_time_well = y ~ s(log_R1, k = 6) + s(log_N, k = 6) + s(hours, k = 6) + s(well_factor, bs = "re"),
    plus_logG0 = y ~ s(log_R1, k = 6) + s(log_G0, k = 6),
    plus_drawdown = y ~ s(log_R1, k = 6) + s(gluc_drawdown, k = 6),
    plus_cum_drawdown_auc = y ~ s(log_R1, k = 6) + s(cum_drawdown_auc, k = 6),
    plus_cum_N_auc = y ~ s(log_R1, k = 6) + s(cum_N_auc, k = 6),
    plus_cum_history_both = y ~ s(log_R1, k = 6) + s(cum_drawdown_auc, k = 6) + s(cum_N_auc, k = 6),
    plus_lag_state = y ~ s(log_R1, k = 6) + s(log_N, k = 6) + s(lag_log_R1, k = 6) + s(lag_log_N, k = 6),
    plus_integrated_state = y ~ s(log_R1, k = 6) + s(log_N, k = 6) + s(cum_drawdown_auc, k = 6) + s(cum_N_auc, k = 6),
    plus_mech_history = y ~ s(log_R1, k = 6) + s(log_N, k = 6) + s(lag_log_R1, k = 6) + s(cum_drawdown_auc, k = 6)
  )
}

fit_rate_model <- function(df, response, model_name) {
  dat <- df %>%
    ungroup() %>%
    transmute(
      y = .data[[response]],
      G0,
      log_R1,
      log_N,
      hours,
      log_G0,
      gluc_drawdown,
      lag_log_R1,
      lag_log_N,
      cum_drawdown_auc,
      cum_N_auc,
      well_factor,
      well_idx,
      cellLine,
      ploidy_abs
    ) %>%
    filter(
      is.finite(y),
      is.finite(G0),
      is.finite(log_R1),
      is.finite(log_N),
      is.finite(hours),
      is.finite(log_G0),
      is.finite(gluc_drawdown),
      is.finite(lag_log_R1),
      is.finite(lag_log_N),
      is.finite(cum_drawdown_auc),
      is.finite(cum_N_auc)
    )

  fit <- mgcv::gam(rate_model_formulas()[[model_name]], data = dat, method = "REML")
  resid <- residuals(fit, type = "response")
  pred <- fitted(fit)

  acf_by_well <- dat %>%
    mutate(resid = resid) %>%
    group_by(well_idx) %>%
    summarise(acf1 = lag1_resid_cor(pick(everything()), "resid"), .groups = "drop")

  list(
    fit = fit,
    augmented = dat %>%
      mutate(
        pred = pred,
        resid = resid,
        response = response,
        model_name = model_name
      ),
    metrics = tibble(
      model_name = model_name,
      rmse = sqrt(mean((dat$y - pred)^2)),
      mae = mean(abs(dat$y - pred)),
      dev_expl = summary(fit)$dev.expl %||% NA_real_,
      r_sq = summary(fit)$r.sq %||% NA_real_,
      aic = AIC(fit),
      median_abs_acf1 = median(abs(acf_by_well$acf1), na.rm = TRUE),
      mean_abs_acf1 = mean(abs(acf_by_well$acf1), na.rm = TRUE),
      n = nrow(dat),
      n_wells = n_distinct(dat$well_idx)
    )
  )
}

candidate_models <- names(rate_model_formulas())
group_keys <- rate_df %>%
  distinct(cellLine, ploidy_abs) %>%
  arrange(cellLine, ploidy_abs)

model_fit_tbl <- bind_rows(lapply(seq_len(nrow(group_keys)), function(i) {
  cell_line <- group_keys$cellLine[[i]]
  ploidy_value <- group_keys$ploidy_abs[[i]]
  df_group <- rate_df %>%
    filter(cellLine == cell_line, ploidy_abs == ploidy_value)

  bind_rows(lapply(c("uptake_per_cell", "growth_per_cell"), function(response_name) {
    bind_rows(lapply(candidate_models, function(model_name) {
      fit_obj <- fit_rate_model(df_group, response_name, model_name)
      tibble(
        cellLine = cell_line,
        ploidy_abs = ploidy_value,
        response = response_name,
        fit_obj = list(fit_obj$fit),
        augmented = list(fit_obj$augmented)
      ) %>%
        bind_cols(fit_obj$metrics)
    }))
  }))
}))

model_screen_summary <- model_fit_tbl %>%
  group_by(response, model_name) %>%
  summarise(
    n_groups = n(),
    median_rmse = median(rmse, na.rm = TRUE),
    median_mae = median(mae, na.rm = TRUE),
    median_dev_expl = median(dev_expl, na.rm = TRUE),
    median_r_sq = median(r_sq, na.rm = TRUE),
    median_abs_acf1 = median(median_abs_acf1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(response, median_rmse)

baseline_vs_candidates <- model_fit_tbl %>%
  select(cellLine, ploidy_abs, response, model_name, rmse, dev_expl, median_abs_acf1) %>%
  pivot_wider(
    names_from = model_name,
    values_from = c(rmse, dev_expl, median_abs_acf1),
    values_fn = dplyr::first
  )

model_improvement_summary <- baseline_vs_candidates %>%
  transmute(
    cellLine,
    ploidy_abs,
    response,
    rmse_reduction_logN_time = 1 - rmse_plus_logN_time / rmse_baseline,
    rmse_reduction_logN_time_well = 1 - rmse_plus_logN_time_well / rmse_baseline,
    dev_gain_logN_time = dev_expl_plus_logN_time - dev_expl_baseline,
    dev_gain_logN_time_well = dev_expl_plus_logN_time_well - dev_expl_baseline,
    acf_reduction_logN_time = median_abs_acf1_baseline - median_abs_acf1_plus_logN_time,
    acf_reduction_logN_time_well = median_abs_acf1_baseline - median_abs_acf1_plus_logN_time_well
  )

model_best_per_group <- model_fit_tbl %>%
  group_by(cellLine, ploidy_abs, response) %>%
  arrange(rmse, desc(dev_expl), median_abs_acf1) %>%
  slice(1L) %>%
  ungroup()

model_screen_summary
```

## Plotting Helpers

```{r plotting-helpers}
format_g0_label <- function(x) {
  factor(
    paste0("G[0]=", x),
    levels = paste0("G[0]=", sort(unique(x)))
  )
}

plot_one_line_variable <- function(line_name, variable_name) {
  sub_obs <- obs_df %>%
    filter(cellLine == line_name, variable == variable_name) %>%
    mutate(G0_label = format_g0_label(G0))

  sub_fit <- fit_df %>%
    filter(cellLine == line_name, variable == variable_name) %>%
    mutate(G0_label = format_g0_label(G0))

  title_str <- if (identical(as.character(variable_name), "N")) {
    "N (total cells)"
  } else {
    "R1 (estimated glucose)"
  }

  ggplot() +
    geom_point(
      data = sub_obs,
      aes(x = hours, y = value, group = well_idx),
      color = "black",
      alpha = 0.45,
      size = 1.1
    ) +
    geom_line(
      data = sub_fit,
      aes(x = hours, y = fit_value, group = well_idx),
      color = "#1b9e77",
      linewidth = 1.0
    ) +
    facet_grid(rows = vars(G0_label), cols = vars(ploidy_abs), scales = "free_y", labeller = label_parsed) +
    scale_y_continuous(labels = label_number(big.mark = ",")) +
    labs(
      title = paste(line_name, "|", title_str),
      x = "Time (hours)",
      y = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      strip.background = element_rect(fill = "grey90", color = NA),
      strip.text = element_text(face = "bold"),
      panel.border = element_rect(color = "grey80", fill = NA),
      plot.title = element_text(face = "bold")
    )
}

plot_one_line_ploidy_overlay <- function(line_name, variable_name) {
  sub_fit <- fit_df %>%
    filter(cellLine == line_name, variable == variable_name) %>%
    mutate(
      G0_label = format_g0_label(G0),
      ploidy_label = factor(as.character(ploidy_abs), levels = sort(unique(as.character(ploidy_abs))))
    )

  title_str <- if (identical(as.character(variable_name), "N")) {
    "N (total cells)"
  } else {
    "R1 (estimated glucose)"
  }

  ggplot(
    sub_fit,
    aes(x = hours, y = fit_value, color = ploidy_label, group = interaction(ploidy_label, well_idx))
  ) +
    geom_line(linewidth = 0.95, alpha = 0.9) +
    facet_wrap(vars(G0_label), scales = "free_y", labeller = label_parsed) +
    scale_color_manual(
      values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a"),
      name = "Ploidy"
    ) +
    scale_y_continuous(labels = label_number(big.mark = ",")) +
    labs(
      title = paste(line_name, "|", title_str, "| spline-only ploidy overlay"),
      x = "Time (hours)",
      y = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      strip.background = element_rect(fill = "grey90", color = NA),
      strip.text = element_text(face = "bold"),
      panel.border = element_rect(color = "grey80", fill = NA),
      plot.title = element_text(face = "bold")
    )
}

```

## Example Cell Lines

Below I show all cell lines detected in `stan_ready_data`. The black points are observed summaries per well-time; the green curves are well-level positive splines grouped by `well_idx`.

```{r line-plots, fig.width=14, fig.height=8}
line_plots <- map(
  line_levels,
  function(line_name) {
    left <- plot_one_line_variable(line_name, "N")
    right <- plot_one_line_variable(line_name, "R1")
    left + right
  }
)

walk(line_plots, print)
```

## Spline-Only Ploidy Overlays

These plots remove the observed data and instead overlay the fitted low- and high-ploidy well trajectories within each cell line. Duplicate wells remain separate trajectories because the line grouping uses `well_idx`.

```{r ploidy-overlay-plots, fig.width=14, fig.height=8}
overlay_plots <- map(
  line_levels,
  function(line_name) {
    left <- plot_one_line_ploidy_overlay(line_name, "N")
    right <- plot_one_line_ploidy_overlay(line_name, "R1")
    left + right
  }
)

walk(overlay_plots, print)
```

## Rate Relationships From The Fitted Splines

These plots use the existing well-level spline fits to evaluate derivatives on a shared time grid for `N` and `R1`, remove any well that has a censored glucose observation anywhere, then scatter all resulting pointwise pairs on a log glucose axis.

1. `1/N * dR1/dt` as a function of `R1`,
2. `1/N * dN/dt` as a function of `R1`.

```{r rate-relationship-plots, fig.width=14, fig.height=8}
plot_rate_relationship <- function(line_name, response = c("uptake_per_cell", "growth_per_cell")) {
  response <- match.arg(response)

  sub_df <- rate_df %>%
    filter(cellLine == line_name) %>%
    mutate(
      ploidy_panel = factor(as.character(ploidy_abs), levels = sort(unique(as.character(ploidy_abs)))),
      y_value = .data[[response]]
    )

  title_str <- if (identical(response, "uptake_per_cell")) {
    "1/N * dR1/dt vs R1"
  } else {
    "1/N * dN/dt vs R1"
  }

  y_lab <- if (identical(response, "uptake_per_cell")) {
    "1/N * dR1/dt"
  } else {
    "1/N * dN/dt"
  }

  ggplot(
    sub_df,
    aes(
      x = R1_fit,
      y = y_value,
      color = ploidy_label
    )
  ) +
    geom_point(size = 1.7) +
    facet_wrap(vars(ploidy_panel), scales = "fixed") +
    scale_color_manual(
      values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a"),
      name = "Ploidy"
    ) +
    scale_x_log10() +
    labs(
      title = paste(line_name, "|", title_str),
      subtitle = "Points: all uncensored spline-evaluation pairs.",
      x = "R1 (estimated glucose)",
      y = y_lab
    ) +
    theme_minimal(base_size = 11) +
    theme(
      strip.background = element_rect(fill = "grey90", color = NA),
      strip.text = element_text(face = "bold"),
      panel.border = element_rect(color = "grey80", fill = NA),
      plot.title = element_text(face = "bold")
    )
}

rate_plots <- map(line_levels, function(line_name) {
  plot_rate_relationship(line_name, "uptake_per_cell") +
    plot_rate_relationship(line_name, "growth_per_cell")
})

walk(rate_plots, print)
```

## Residual-Structure Discovery

To test whether the `R1`-only relationships are missing additional structure, I screened a small family of GAMs on the uncensored spline-evaluation points within each `cellLine x ploidy` group.

This screen should not be read as a direct criticism of the main `gpath` workflow. The active `gpath` models already contain latent resource / waste style state variables, and those are exactly the sort of slow hidden states that could plausibly account for the residual time/history structure detected here. The purpose of this section is narrower: to ask whether a naive reduced view, `rate ~ R1`, is adequate on its own, and if not, what kinds of additional state summaries seem to matter empirically.

The candidate models were:

1. `baseline`: smooth of `log10(R1)` only,
2. `plus_logN`: add `log10(N)` as a density / confluence proxy,
3. `plus_time`: add time,
4. `plus_logN_time`: add both density and time,
5. `plus_logN_time_well`: add density, time, and a well random effect,
6. `plus_logG0`: add initial glucose,
7. `plus_drawdown`: add glucose drawdown `G0 - R1`,
8. `plus_cum_drawdown_auc`: add cumulative glucose drawdown only,
9. `plus_cum_N_auc`: add cumulative `N` only,
10. `plus_cum_history_both`: add cumulative glucose drawdown and cumulative `N` together,
11. `plus_lag_state`: add lagged `R1` and lagged `N`,
12. `plus_integrated_state`: combine current `R1`, current `N`, and both cumulative-history terms,
13. `plus_mech_history`: combine current `R1`, current `N`, lagged `R1`, and cumulative drawdown.

The working interpretation is:

1. `logN` tests density-dependent or confluence-like latent structure,
2. `time` tests adaptation / history / slowly evolving state,
3. lagged / integrated terms test whether the time signal can be replaced by mechanistic state history,
4. `well` tests persistent well-specific intrinsic or extrinsic states.

In that sense, the useful comparison is:

1. if `R1`-only fails badly, that supports the need for hidden state in some form,
2. if lagged or integrated summaries replace time, that points toward a concrete mechanistic surrogate for the hidden state,
3. if time still wins, the hidden state is likely more complex than the simple proxies screened here,
4. and if well effects are still large, there is persistent heterogeneity that a pooled low-dimensional rate curve is not capturing.

```{r model-screen-summary}
model_screen_summary
```

```{r model-screen-improvements}
model_improvement_summary %>%
  group_by(response) %>%
  summarise(
    median_rmse_reduction_logN_time = median(rmse_reduction_logN_time, na.rm = TRUE),
    median_rmse_reduction_logN_time_well = median(rmse_reduction_logN_time_well, na.rm = TRUE),
    median_dev_gain_logN_time = median(dev_gain_logN_time, na.rm = TRUE),
    median_dev_gain_logN_time_well = median(dev_gain_logN_time_well, na.rm = TRUE),
    median_acf_reduction_logN_time = median(acf_reduction_logN_time, na.rm = TRUE),
    median_acf_reduction_logN_time_well = median(acf_reduction_logN_time_well, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r time-vs-history-summary}
model_screen_summary %>%
  filter(model_name %in% c(
    "baseline",
    "plus_time",
    "plus_cum_drawdown_auc",
    "plus_cum_N_auc",
    "plus_cum_history_both",
    "plus_logN_time",
    "plus_lag_state",
    "plus_integrated_state",
    "plus_mech_history",
    "plus_logN_time_well"
  )) %>%
  arrange(response, median_rmse)
```

```{r model-screen-plot, fig.width=12, fig.height=6}
plot_metric_by_model <- function(metric_col, y_lab, response_name) {
  model_fit_tbl %>%
    filter(response == response_name) %>%
    ggplot(aes(x = reorder(model_name, .data[[metric_col]], FUN = median), y = .data[[metric_col]])) +
    geom_boxplot(outlier.alpha = 0.25, fill = "#b8d8c8", color = "#2f4f4f") +
    coord_flip() +
    theme_minimal(base_size = 11) +
    labs(
      title = paste(response_name, "|", y_lab),
      x = NULL,
      y = y_lab
    )
}

(
  plot_metric_by_model("rmse", "RMSE", "uptake_per_cell") +
    plot_metric_by_model("rmse", "RMSE", "growth_per_cell")
) /
(
  plot_metric_by_model("median_abs_acf1", "Median |lag-1 residual cor|", "uptake_per_cell") +
    plot_metric_by_model("median_abs_acf1", "Median |lag-1 residual cor|", "growth_per_cell")
)
```

```{r best-model-counts}
model_best_per_group %>%
  count(response, model_name, sort = TRUE)
```

```{r rmse-raster-by-condition, fig.width=14, fig.height=10}
rmse_raster_df <- bind_rows(lapply(seq_len(nrow(group_keys)), function(i) {
  cell_line <- group_keys$cellLine[[i]]
  ploidy_value <- group_keys$ploidy_abs[[i]]
  df_group <- rate_df %>%
    filter(cellLine == cell_line, ploidy_abs == ploidy_value)

  bind_rows(lapply(c("uptake_per_cell", "growth_per_cell"), function(response_name) {
    bind_rows(lapply(c("baseline", "plus_integrated_state"), function(model_name) {
      dat <- df_group %>%
        ungroup() %>%
        transmute(
          y = .data[[response_name]],
          G0,
          log_R1,
          log_N,
          hours,
          log_G0,
          gluc_drawdown,
          lag_log_R1,
          lag_log_N,
          cum_drawdown_auc,
          cum_N_auc
        ) %>%
        filter(
          is.finite(y),
          is.finite(G0),
          is.finite(log_R1),
          is.finite(log_N),
          is.finite(hours),
          is.finite(log_G0),
          is.finite(gluc_drawdown),
          is.finite(lag_log_R1),
          is.finite(lag_log_N),
          is.finite(cum_drawdown_auc),
          is.finite(cum_N_auc)
        )

      fit <- mgcv::gam(rate_model_formulas()[[model_name]], data = dat, method = "REML")
      pred <- fitted(fit)

      tibble(G0 = dat$G0, resid = dat$y - pred) %>%
        group_by(G0) %>%
        summarise(mean_rmse = sqrt(mean(resid^2)), .groups = "drop") %>%
        mutate(
          cellLine = cell_line,
          ploidy_abs = ploidy_value,
          response = response_name,
          model_name = model_name,
          .before = 1
        )
    }))
  }))
}))

y_levels <- rmse_raster_df %>%
  distinct(cellLine, ploidy_abs) %>%
  arrange(cellLine, ploidy_abs) %>%
  transmute(line_ploidy = paste(cellLine, "|", ploidy_abs)) %>%
  pull(line_ploidy)

plot_rmse_raster_by_response <- function(response_key, title_stub) {
  rmse_raster_df %>%
    filter(response == response_key) %>%
    mutate(
      line_ploidy = factor(paste(cellLine, "|", ploidy_abs), levels = y_levels),
      model_name = factor(model_name, levels = c("baseline", "plus_integrated_state")),
      G0_factor = factor(G0, levels = sort(unique(G0)))
    ) %>%
    ggplot(aes(x = G0_factor, y = line_ploidy, fill = mean_rmse)) +
    geom_tile(color = "white", linewidth = 0.25) +
    facet_wrap(vars(model_name), ncol = 2, scales = "free_y") +
    scale_fill_viridis_c(option = "C", end = 0.95) +
    labs(
      title = title_stub,
      x = "G0",
      y = "cellLine | ploidy",
      fill = "Mean RMSE"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      strip.text = element_text(face = "bold"),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(face = "bold")
    )
}

plot_rmse_raster_by_response("growth_per_cell", "Condition-level mean RMSE raster | growth_per_cell") /
  plot_rmse_raster_by_response("uptake_per_cell", "Condition-level mean RMSE raster | uptake_per_cell")
```

```{r worst-residual-selection}
worst_groups <- rmse_raster_df %>%
  group_by(response, cellLine, ploidy_abs) %>%
  summarise(mean_rmse_across_G0 = mean(mean_rmse, na.rm = TRUE), .groups = "drop") %>%
  group_by(response) %>%
  arrange(desc(mean_rmse_across_G0), cellLine, ploidy_abs, .by_group = TRUE) %>%
  slice_head(n = 3L) %>%
  ungroup()

worst_groups
```

```{r worst-residual-plots, fig.width=14, fig.height=18}
extract_augmented_model_data <- function(cell_line, ploidy_value, response_name, model_name) {
  df_group <- rate_df %>%
    filter(cellLine == cell_line, ploidy_abs == ploidy_value)

  fit_rate_model(df_group, response_name, model_name)$augmented %>%
    mutate(
      model_name = factor(model_name, levels = c("baseline", "plus_integrated_state"))
    )
}

plot_group_residuals <- function(cell_line, ploidy_value, response_name) {
  aug_df <- compact(list(
    extract_augmented_model_data(cell_line, ploidy_value, response_name, "baseline"),
    extract_augmented_model_data(cell_line, ploidy_value, response_name, "plus_integrated_state")
  )) %>%
    bind_rows() %>%
    mutate(G0_factor = factor(G0, levels = sort(unique(G0))))

  title_stub <- paste(cell_line, "| ploidy =", ploidy_value, "|", response_name)

  p_hours <- ggplot(aug_df, aes(x = hours, y = resid, color = G0_factor)) +
    geom_point(size = 1.2) +
    facet_wrap(vars(model_name), scales = "fixed") +
    scale_color_viridis_d(option = "D", end = 0.9) +
    theme_minimal(base_size = 11) +
    labs(title = paste(title_stub, "| residual vs time"), x = "Time (hours)", y = "Residual", color = "G0")

  p_logn <- ggplot(aug_df, aes(x = log_N, y = resid, color = G0_factor)) +
    geom_point(size = 1.2) +
    facet_wrap(vars(model_name), scales = "fixed") +
    scale_color_viridis_d(option = "D", end = 0.9) +
    theme_minimal(base_size = 11) +
    labs(title = paste(title_stub, "| residual vs log10(N)"), x = "log10(N)", y = "Residual", color = "G0")

  p_hours / p_logn
}

for (response_name in c("growth_per_cell", "uptake_per_cell")) {
  response_hits <- worst_groups %>%
    filter(response == response_name)

  for (i in seq_len(nrow(response_hits))) {
    print(plot_group_residuals(
      cell_line = response_hits$cellLine[[i]],
      ploidy_value = response_hits$ploidy_abs[[i]],
      response_name = response_name
    ))
  }
}
```

## Panel-Level Diagnostics

This is a quick summary of well-level fit sizes and whether the spline could be fit.

```{r panel-diagnostics}
fit_qc %>%
  left_join(line_counts, by = c("cellLine", "variable")) %>%
  arrange(cellLine, variable, ploidy_abs, G0, well_idx)
```

```{r raw-count-point-diagnostics}
count_point_counts %>%
  arrange(cellLine, ploidy_abs, G0, well_idx, hours)
```

```{r rate-diagnostics}
rate_df %>%
  group_by(cellLine, ploidy_abs) %>%
  summarise(
    n_rate_points = n(),
    n_wells = n_distinct(well_idx),
    min_hours = min(hours),
    max_hours = max(hours),
    min_R1 = min(R1_fit),
    max_R1 = max(R1_fit),
    .groups = "drop"
  ) %>%
  arrange(cellLine, ploidy_abs)
```

## Current Read

The rate-relationship plots suggest that both uptake and growth often do have a meaningful dependence on glucose, but the residual screen indicates that `R1` alone is usually not the full state description.

Framed against the main modeling path, I do not read this as evidence that `gpath` is misguided. I read it as evidence that hidden state is justified. The latent resource / waste states already present in `gpath` are plausible carriers of the slow structure that appears here as time dependence. So the main conclusion is not "the data want time as a covariate"; it is "the data want more state than current glucose alone."

What worked:

1. adding `log10(N)` and time to the baseline `f(log10(R1))` model produced a large median improvement across groups,
2. for `growth_per_cell`, the `logN + time` model already cuts median RMSE sharply and strongly reduces residual autocorrelation,
3. adding a well random effect improves fit further, especially for uptake, which points to persistent well-level latent structure beyond `R1`, density, and time.

What partly worked:

1. glucose drawdown occasionally won for uptake in a few groups, so history of depletion seems relevant in some contexts,
2. lagged and integrated history terms improved on the baseline and often approached the plain time model, but in my screen they did not consistently replace `time`,
3. initial glucose alone helped less than current glucose plus density/time, so `G0` is not the dominant missing state once the trajectory is conditioned on current `R1`.

What did not work well:

1. `R1` alone leaves large residual autocorrelation and structured dependence on time / density,
2. time-only or density-only models help, but they are consistently weaker than using both together,
3. the lag / cumulative proxies I tried do not fully absorb the time effect, so they are not yet a sufficient mechanistic replacement for the hidden state,
4. the well random effect improves in-sample tightness but does not by itself identify whether the persistent latent state is intrinsic biology, local microenvironment, or some experimental artifact.

My current interpretation is that the tightest next model class should include:

1. a glucose-response term,
2. a density / confluence term,
3. a slow history or adaptation term, possibly better approximated by a richer mechanistic history variable than the simple lag/integral proxies screened here,
4. and probably a persistent well-level latent effect if the aim is descriptive fit rather than transfer.

Relative to the main workflow, that means:

1. the notebook supports the use of hidden-state models such as `gpath`,
2. the residual diagnostics can be used as a model-criticism layer to see where the existing latent-state parameterization is sufficient versus where it may be too rigid,
3. and the alternative latent-state screen here is best viewed as a complementary reduced-model probe, not as a replacement for the mechanistic path.
